### 2023-08-01 README.qmd
Add README

---
format: gfm
# in future: pdf!
editor: 
  markdown: 
    wrap: sentence
---

# 10th Transport Research Arena, TRA 2024, Dublin, Ireland

**CRUSE to Safe Cycling in Ireland: An Open Source Methodology to Support Active Travel**

*Dr. Robin Lovelace (a), Dr. Joey Talbot (a), Dan Brennan (b), Dr. Suzanne Meade (b), Elaine Brick (c), and Peter Wright (d)*

a University of Leeds, Institute for Transport Studies, Leeds, Ls2 9Jt, United Kingdom, Leeds LS2 9J, UK.

b Transport Infrastructure Ireland, Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland.

c AECOM, Unit 6, Galway Technology Park, Parkmore, Galway, Ireland.

d AECOM, Winslade House, Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY, UK.

## Abstract

Under the EU Road infrastructure safety management (RISM) directive, the National Road Safety Strategy (RSS), and the Climate Action Plan Transport Infrastructure Ireland (TII) has a remit for road safety and decarbonizing a predominantly road-based network in Ireland.
To address data needs for both safety and project evaluation on the National Road Network (NRN), the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was developed.
While cycling in Ireland represents only 3% of total modal share, with higher intensities in urban areas, the levels of cycling collisions are disproportionately high at 20% of all serious injuries and 7% of all fatalities.
If Ireland is to meet its climate and safety targets, data to establish baseline cycling levels and future cycling levels is needed.

Due to an absence of reliable data, particularly rural cycling levels, TII commissioned the Institute for Transport Studies (ITS) at the University of Leeds and AECOM to develop a new tool for this purpose.
ITS Leeds led the development of the PCT for England and Wales, which has "revolutionized the practice of strategic cycle planning nationally".
The tool is an open-source approach using recognized open-source methodology to enable planners, engineers, and other stakeholders to make evidence-based decisions for the NRN.
CRUSE is available at <https://cruse.bike/> and builds on the Propensity to Cycle Tool (PCT) for England and Wales.
CRUSE goes beyond the PCT in several important ways, higher resolution data, more trip types, including estimates for education, inter-urban, and recreational trips.
In addition to understanding cycling intensity, for asset planning and management purposes, the tool provides essential road safety information to enable reporting of disaggregate collision rates.

CRUSE is structured in a similar way to the traditional four-stage transport model, but its use of Open Street Map (OSM) data, used by [Cyclestreets.net](https://www.cyclestreets.net/) for routing, enables network quality to be assessed without costly surveys to record new infrastructure.
OSM tags generate "cycle friendliness" estimates of all links on the network, based on existing recorded infrastructure.
A range of networks is provided, highlighting routes for directness (Fastest) and "cycle friendliness" (Quietest).
It uses origin and destination data from the 2016 Census in combination with modeled demand data to estimate cycling levels and potential at the area, route, and network levels for each county in Ireland and offers estimates of the baseline level of cycling and several future scenario-based levels of cycling.

As countries, like Ireland, invest in cycling, the number of killed and seriously injured cyclists must reduce too.
The CRUSE Tool provides estimates of cycling potential and routing for each county in Ireland, and works in both urban and rural settings, to enable monitoring of cycling safety.
With growth in the E-bike market, the tool will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
The CRUSE Tool methodology and findings are directly relevant to addressing the challenges and opportunities faced by other NRAs.
The datasets resulting from the project are open access and can be used by both non-experts and professionals.

**Keywords:** Cycling, Cycling Exposure, Open-source, Road Safety, Active Travel.

### 2023-08-01 README.qmd
Add rendered result

@@
 ---
-format: gfm
+format:
+  gfm: default
+  pdf: default
+title: "CRUSE to Safe Cycling in Ireland: An Open Source Methodology to Support Active Travel"
 # in future: pdf!
@@
 
-# 10th Transport Research Arena, TRA 2024, Dublin, Ireland
-
-**CRUSE to Safe Cycling in Ireland: An Open Source Methodology to Support Active Travel**
+## 10th Transport Research Arena, TRA 2024, Dublin, Ireland
 
@@
 
-a University of Leeds, Institute for Transport Studies, Leeds, Ls2 9Jt, United Kingdom, Leeds LS2 9J, UK.
+a University of Leeds, Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom, Leeds LS2 9J, UK.
 

### 2023-08-01 cruse-paper.qmd
Multi format outputs

--- 2023-08-01 README.qmd
@@
 ---
+title: "CRUSE to Safe Cycling in Ireland: An Open Source Methodology to Support Active Travel"
 format: 
-  gfm: default
-  pdf: default
-title: "CRUSE to Safe Cycling in Ireland: An Open Source Methodology to Support Active Travel"
-# in future: pdf!
+  html:
+    output-file: index.html
+  pdf:
+    output-file: cruse-paper.pdf 
+  gfm:
+    output-file: README.md
+number-sections: true
 editor: 
+  # render-on-save: true
   markdown: 
@@
 
-## 10th Transport Research Arena, TRA 2024, Dublin, Ireland
+## 10th Transport Research Arena, TRA 2024, Dublin, Ireland {-}
 
@@
 
-## Abstract
+## Abstract {-}
 

### 2023-09-11 cruse-paper.qmd
Put in new format

--- 2023-08-01 README.qmd
@@
 ---
-format: gfm
-# in future: pdf!
-editor: 
-  markdown: 
-    wrap: sentence
----
+title: "CRUSE to Safe Cycling in Ireland"
+subtitle: An Open Source Methodology to Support Active Travel
+author:
+  - name: Dr. Robin Lovelace
+    affiliations:
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Dr. Joey Talbot
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Dan Brennan
+    affiliations:
+      - name: Transport Infrastructure Ireland
+        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+  - name: Dr. Suzanne Meade
+    affiliations:
+      - name: Transport Infrastructure Ireland
+        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+  - name: Elaine Brick
+    affiliations:
+      - name: AECOM
+        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+  - name: Peter Wright
+    affiliations:
+      - name: AECOM
+        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY, UK
+abstract: |
+  Under the EU Road infrastructure safety management (RISM) directive, the National Road Safety Strategy (RSS), and the Climate Action Plan Transport Infrastructure Ireland (TII) has a remit for road safety and decarbonizing a predominantly road-based network in Ireland.
 
-# 10th Transport Research Arena, TRA 2024, Dublin, Ireland
-
-**CRUSE to Safe Cycling in Ireland: An Open Source Methodology to Support Active Travel**
-
-*Dr. Robin Lovelace (a), Dr. Joey Talbot (a), Dan Brennan (b), Dr. Suzanne Meade (b), Elaine Brick (c), and Peter Wright (d)*
-
-a University of Leeds, Institute for Transport Studies, Leeds, Ls2 9Jt, United Kingdom, Leeds LS2 9J, UK.
-
-b Transport Infrastructure Ireland, Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland.
-
-c AECOM, Unit 6, Galway Technology Park, Parkmore, Galway, Ireland.
-
-d AECOM, Winslade House, Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY, UK.
-
-## Abstract
-
-Under the EU Road infrastructure safety management (RISM) directive, the National Road Safety Strategy (RSS), and the Climate Action Plan Transport Infrastructure Ireland (TII) has a remit for road safety and decarbonizing a predominantly road-based network in Ireland.
   To address data needs for both safety and project evaluation on the National Road Network (NRN), the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was developed.
@@
   The datasets resulting from the project are open access and can be used by both non-experts and professionals.
+keywords: 
+  - Cycling
+  - Open-source
+  - Road Safety
+  - Active Travel
+date: last-modified
+bibliography: bibliography.bib
+format:
+  elsevier-pdf:
+    keep-tex: true
+    journal:
+      name: Journal Name
+      formatting: preprint
+      model: 3p
+      cite-style: super
+# title: "CRUSE to Safe Cycling in Ireland: An Open Source Methodology to Support Active Travel"
+# format: 
+#   format:
+#   elsevier-pdf:
+#     keep-tex: true
+#     journal:
+#       name: Journal Name
+#       formatting: preprint
+#       model: 3p
+#       cite-style: super
+  # html:
+  #   output-file: index.html
+  # pdf:
+  #   output-file: cruse-paper.pdf 
+  # gfm:
+  #   output-file: README.md
+# affiliations:
+#   - id: cmu
+#     name: Carnegie Mellon University
+#     city: Pittsburgh
+#     state: PA
+#   - id: chicago
+#     name: University of Chicago
+#     city: Chicago
+#     state: IL
+# number-sections: true
+# editor: 
+#   # render-on-save: true
+#   markdown: 
+#     wrap: sentence
+# # Bibliography:
+# bibliography: references.bib
+# # APA style:
+csl: apa.csl
+---
 
-**Keywords:** Cycling, Cycling Exposure, Open-source, Road Safety, Active Travel.
+## 10th Transport Research Arena, TRA 2024, Dublin, Ireland {.unnumbered}
+
+{{< pagebreak >}}
+
+**Keywords:** .
+
+# Introduction
+
+[@lovelace2017]

### 2023-09-11 cruse-paper.qmd
Add paper stuff

@@
   - name: Dr. Robin Lovelace
+    email: r.lovelace@leeds.ac.uk
+    corresponding: true
     affiliations:
+      - name: University of Leeds
         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
   - name: Dr. Joey Talbot
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Eugeni Vidal Tortosa
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Hussein Mahfouz
     affiliations:
@@
       cite-style: super
-# title: "CRUSE to Safe Cycling in Ireland: An Open Source Methodology to Support Active Travel"
-# format: 
-#   format:
-#   elsevier-pdf:
-#     keep-tex: true
-#     journal:
-#       name: Journal Name
-#       formatting: preprint
-#       model: 3p
-#       cite-style: super
-  # html:
-  #   output-file: index.html
-  # pdf:
-  #   output-file: cruse-paper.pdf 
-  # gfm:
-  #   output-file: README.md
-# affiliations:
-#   - id: cmu
-#     name: Carnegie Mellon University
-#     city: Pittsburgh
-#     state: PA
-#   - id: chicago
-#     name: University of Chicago
-#     city: Chicago
-#     state: IL
-# number-sections: true
-# editor: 
-#   # render-on-save: true
-#   markdown: 
-#     wrap: sentence
-# # Bibliography:
-# bibliography: references.bib
-# # APA style:
 csl: apa.csl
@@
 
-**Keywords:** .
-
 # Introduction
@@
 [@lovelace2017]
+
+• Title page/main information (title, full names and institutional addresses for all
+authors, indication of the corresponding author)
+• Abstract
+• Introduction
+• Results
+• Discussion (can be combined in ‘Results and Discussion’ section)
+• Conclusions
+• Methods/Experimental (can also be placed after Introduction)
+• List of abbreviations
+• Declarations*
+o Availability of data and material
+o Funding
+o Acknowledgements
+o Competing interests
+o Authors' contributions
+o Authors' information (optional)
+o *If any of the sections are not relev

### 2023-09-11 cruse-paper.tex
Add paper stuff

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  super,
  preprint,
  3p]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={CRUSE to Safe Cycling in Ireland},
  pdfauthor={Dr.~Robin Lovelace; Dr.~Joey Talbot; Eugeni Vidal Tortosa; Hussein Mahfouz; Dan Brennan; Dr.~Suzanne Meade; Elaine Brick; Peter Wright},
  pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{CRUSE to Safe Cycling in Ireland \\\large{An Open Source
Methodology to Support Active Travel} }
\author[1]{Dr.~Robin Lovelace%
\corref{cor1}%
}
 \ead{r.lovelace@leeds.ac.uk} 
\author[1]{Dr.~Joey Talbot%
%
}

\author[1]{Eugeni Vidal Tortosa%
%
}

\author[1]{Hussein Mahfouz%
%
}

\author[2]{Dan Brennan%
%
}

\author[2]{Dr.~Suzanne Meade%
%
}

\author[3]{Elaine Brick%
%
}

\author[4]{Peter Wright%
%
}

\affiliation[1]{organization={University of
Leeds},addressline={Institute for Transport Studies, Leeds, LS2 9JT,
United Kingdom},postcodesep={}}
\affiliation[2]{organization={Transport Infrastructure
Ireland},addressline={Parkgate Business Centre, Parkgate Street, Dublin
8, D08 DK10, Ireland},postcodesep={}}
\affiliation[3]{organization={AECOM},addressline={Unit 6, Galway
Technology Park, Parkmore, Galway, Ireland},postcodesep={}}
\affiliation[4]{organization={AECOM},addressline={Winslade House,
Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY,
UK},postcodesep={}}

\cortext[cor1]{Corresponding author}

\begin{abstract}
Under the EU Road infrastructure safety management (RISM) directive, the
National Road Safety Strategy (RSS), and the Climate Action Plan
Transport Infrastructure Ireland (TII) has a remit for road safety and
decarbonizing a predominantly road-based network in Ireland.

To address data needs for both safety and project evaluation on the
National Road Network (NRN), the Cycle Route Uptake and Scenario
Estimation (CRUSE) Tool was developed. While cycling in Ireland
represents only 3\% of total modal share, with higher intensities in
urban areas, the levels of cycling collisions are disproportionately
high at 20\% of all serious injuries and 7\% of all fatalities. If
Ireland is to meet its climate and safety targets, data to establish
baseline cycling levels and future cycling levels is needed.

Due to an absence of reliable data, particularly rural cycling levels,
TII commissioned the Institute for Transport Studies (ITS) at the
University of Leeds and AECOM to develop a new tool for this purpose.
ITS Leeds led the development of the PCT for England and Wales, which
has ``revolutionized the practice of strategic cycle planning
nationally''. The tool is an open-source approach using recognized
open-source methodology to enable planners, engineers, and other
stakeholders to make evidence-based decisions for the NRN. CRUSE is
available at \url{https://cruse.bike/} and builds on the Propensity to
Cycle Tool (PCT) for England and Wales. CRUSE goes beyond the PCT in
several important ways, higher resolution data, more trip types,
including estimates for education, inter-urban, and recreational trips.
In addition to understanding cycling intensity, for asset planning and
management purposes, the tool provides essential road safety information
to enable reporting of disaggregate collision rates.

CRUSE is structured in a similar way to the traditional four-stage
transport model, but its use of Open Street Map (OSM) data, used by
\href{https://www.cyclestreets.net/}{Cyclestreets.net} for routing,
enables network quality to be assessed without costly surveys to record
new infrastructure. OSM tags generate ``cycle friendliness'' estimates
of all links on the network, based on existing recorded infrastructure.
A range of networks is provided, highlighting routes for directness
(Fastest) and ``cycle friendliness'' (Quietest). It uses origin and
destination data from the 2016 Census in combination with modeled demand
data to estimate cycling levels and potential at the area, route, and
network levels for each county in Ireland and offers estimates of the
baseline level of cycling and several future scenario-based levels of
cycling.

As countries, like Ireland, invest in cycling, the number of killed and
seriously injured cyclists must reduce too. The CRUSE Tool provides
estimates of cycling potential and routing for each county in Ireland,
and works in both urban and rural settings, to enable monitoring of
cycling safety. With growth in the E-bike market, the tool will help
inform inter-urban and rural networks to support the transfer of trips
to sustainable modes for longer journeys. The CRUSE Tool methodology and
findings are directly relevant to addressing the challenges and
opportunities faced by other NRAs. The datasets resulting from the
project are open access and can be used by both non-experts and
professionals.
\end{abstract}

\begin{keyword}
    Cycling \sep Open-source \sep Road Safety \sep 
    Active Travel
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi

\hypertarget{th-transport-research-arena-tra-2024-dublin-ireland}{%
\subsection*{10th Transport Research Arena, TRA 2024, Dublin,
Ireland}\label{th-transport-research-arena-tra-2024-dublin-ireland}}
\addcontentsline{toc}{subsection}{10th Transport Research Arena, TRA
2024, Dublin, Ireland}

\newpage{}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\citep{lovelace2017}

• Title page/main information (title, full names and institutional
addresses for all authors, indication of the corresponding author) •
Abstract • Introduction • Results • Discussion (can be combined in
`Results and Discussion' section) • Conclusions • Methods/Experimental
(can also be placed after Introduction) • List of abbreviations •
Declarations\emph{ o Availability of data and material o Funding o
Acknowledgements o Competing interests o Authors' contributions o
Authors' information (optional) o }If any of the sections are not relev

  \bibliography{bibliography.bib}

\end{document}

### 2023-09-11 paper.tex
Add paper stuff

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  super,
  preprint,
  3p]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Short Paper},
  pdfauthor={Alice Anonymous; Bob Security; Cat Memes; Derek Zoolander},
  pdfkeywords={keyword1, keyword2},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{Short Paper \\\large{A Short Subtitle} }
\author[1]{Alice Anonymous%
\corref{cor1}%
\fnref{fn1}}
 \ead{alice@example.com} 
\author[2]{Bob Security%
%
\fnref{fn2}}
 \ead{bob@example.com} 
\author[2]{Cat Memes%
%
\fnref{fn3}}
 \ead{cat@example.com} 
\author[]{Derek Zoolander%
%
}
 \ead{derek@example.com}

\affiliation[1]{organization={Some Institute of Technology, Department
Name},addressline={Street Address},city={City},postcode={Postal
Code},postcodesep={}}
\affiliation[2]{organization={Another University, Department
Name},addressline={Street Address},city={City},postcode={Postal
Code},postcodesep={}}

\cortext[cor1]{Corresponding author}
\fntext[fn1]{This is the first author footnote.}
\fntext[fn2]{Another author footnote, this is a very long footnote and
it should be a really long footnote. But this footnote is not yet
sufficiently long enough to make two lines of footnote text.}
\fntext[fn3]{Yet another author footnote.}

\begin{abstract}
This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing
elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget
velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris
sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta
egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet
auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo
molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae.
Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at
tempus elit.
\end{abstract}

\begin{keyword}
    keyword1 \sep 
    keyword2
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, boxrule=0pt, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

Please make sure that your manuscript follows the guidelines in the
Guide for Authors of the relevant journal. It is not necessary to
typeset your manuscript in exactly the same way as an article, unless
you are submitting to a camera-ready copy (CRC) journal.

For detailed instructions regarding the elsevier article class, see
\url{https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions}

\hypertarget{bibliography-styles}{%
\section{Bibliography styles}\label{bibliography-styles}}

Here are two sample references: \citet{Feynman1963118}
\citet{Dirac1953888}.

By default, natbib will be used with the \texttt{authoryear} style, set
in \texttt{classoption} variable in YAML. You can sets extra options
with \texttt{natbiboptions} variable in YAML header. Example

\begin{verbatim}
natbiboptions: longnamesfirst,angle,semicolon
\end{verbatim}

There are various more specific bibliography styles available at
\url{https://support.stmdocs.in/wiki/index.php?title=Model-wise_bibliographic_style_files}.
To use one of these, add it in the header using, for example,
\texttt{biblio-style:\ model1-num-names}.

\hypertarget{using-csl}{%
\subsection{Using CSL}\label{using-csl}}

If \texttt{cite-method} is set to \texttt{citeproc} in
\texttt{elsevier\_article()}, then pandoc is used for citations instead
of \texttt{natbib}. In this case, the \texttt{csl} option is used to
format the references. By default, this template will provide an
appropriate style, but alternative \texttt{csl} files are available from
\url{https://www.zotero.org/styles?q=elsevier}. These can be downloaded
and stored locally, or the url can be used as in the example header.

\hypertarget{equations}{%
\section{Equations}\label{equations}}

Here is an equation: \[ 
  f_{X}(x) = \left(\frac{\alpha}{\beta}\right)
  \left(\frac{x}{\beta}\right)^{\alpha-1}
  e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; 
  \alpha,\beta,x > 0 .
\]

Inline equations work as well: \(\sum_{i = 2}^\infty\{\alpha_i^\beta\}\)

\hypertarget{figures-and-tables}{%
\section{Figures and tables}\label{figures-and-tables}}

Figure~\ref{fig-meaningless} is generated using an R chunk.

\begin{figure}

{\centering \includegraphics[width=0.5\textwidth,height=\textheight]{paper_files/figure-pdf/fig-meaningless-1.pdf}

}

\caption{\label{fig-meaningless}A meaningless scatterplot}

\end{figure}

\hypertarget{tables-coming-from-r}{%
\section{Tables coming from R}\label{tables-coming-from-r}}

Tables can also be generated using R chunks, as shown in
Table~\ref{tbl-simple} example.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(}\FunctionTok{head}\NormalTok{(mtcars)[,}\DecValTok{1}\SpecialCharTok{:}\DecValTok{4}\NormalTok{])}
\end{Highlighting}
\end{Shaded}

\hypertarget{tbl-simple}{}
\begin{longtable}[]{@{}lrrrr@{}}
\caption{\label{tbl-simple}Caption centered above table}\tabularnewline
\toprule\noalign{}
& mpg & cyl & disp & hp \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& mpg & cyl & disp & hp \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mazda RX4 & 21.0 & 6 & 160 & 110 \\
Mazda RX4 Wag & 21.0 & 6 & 160 & 110 \\
Datsun 710 & 22.8 & 4 & 108 & 93 \\
Hornet 4 Drive & 21.4 & 6 & 258 & 110 \\
Hornet Sportabout & 18.7 & 8 & 360 & 175 \\
Valiant & 18.1 & 6 & 225 & 105 \\
\end{longtable}

\renewcommand\refname{References}
  \bibliography{bibliography.bib}

\end{document}

### 2023-09-11 cruse-paper.qmd
Update readme

--- 2023-08-01 cruse-paper.qmd
@@
 ---
-title: "CRUSE to Safe Cycling in Ireland: An Open Source Methodology to Support Active Travel"
+title: "CRUSE to Safe Cycling in Ireland"
+subtitle: An Open Source Methodology to Support Active Travel
+bibliography: bibliography.bib
+csl: apa.csl
+# Uncomment for draft versions:
 format: 
@@
 number-sections: true
-editor: 
-  # render-on-save: true
-  markdown: 
-    wrap: sentence
+
+# Uncomment the following line to produce the cover page
+# author:
+#   - name: Dr. Robin Lovelace
+#     email: r.lovelace@leeds.ac.uk
+#     corresponding: true
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Dr. Joey Talbot
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Eugeni Vidal Tortosa
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Hussein Mahfouz
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Dan Brennan
+#     affiliations:
+#       - name: Transport Infrastructure Ireland
+#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+#   - name: Dr. Suzanne Meade
+#     affiliations:
+#       - name: Transport Infrastructure Ireland
+#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+#   - name: Elaine Brick
+#     affiliations:
+#       - name: AECOM
+#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+#   - name: Peter Wright
+#     affiliations:
+#       - name: AECOM
+#         address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY, UK
+# abstract: |
+#   Under the EU Road infrastructure safety management (RISM) directive, the National Road Safety Strategy (RSS), and the Climate Action Plan Transport Infrastructure Ireland (TII) has a remit for road safety and decarbonizing a predominantly road-based network in Ireland.
+
+#   To address data needs for both safety and project evaluation on the National Road Network (NRN), the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was developed.
+#   While cycling in Ireland represents only 3% of total modal share, with higher intensities in urban areas, the levels of cycling collisions are disproportionately high at 20% of all serious injuries and 7% of all fatalities.
+#   If Ireland is to meet its climate and safety targets, data to establish baseline cycling levels and future cycling levels is needed.
+
+#   Due to an absence of reliable data, particularly rural cycling levels, TII commissioned the Institute for Transport Studies (ITS) at the University of Leeds and AECOM to develop a new tool for this purpose.
+#   ITS Leeds led the development of the PCT for England and Wales, which has "revolutionized the practice of strategic cycle planning nationally".
+#   The tool is an open-source approach using recognized open-source methodology to enable planners, engineers, and other stakeholders to make evidence-based decisions for the NRN.
+#   CRUSE is available at <https://cruse.bike/> and builds on the Propensity to Cycle Tool (PCT) for England and Wales.
+#   CRUSE goes beyond the PCT in several important ways, higher resolution data, more trip types, including estimates for education, inter-urban, and recreational trips.
+#   In addition to understanding cycling intensity, for asset planning and management purposes, the tool provides essential road safety information to enable reporting of disaggregate collision rates.
+
+#   CRUSE is structured in a similar way to the traditional four-stage transport model, but its use of Open Street Map (OSM) data, used by [Cyclestreets.net](https://www.cyclestreets.net/) for routing, enables network quality to be assessed without costly surveys to record new infrastructure.
+#   OSM tags generate "cycle friendliness" estimates of all links on the network, based on existing recorded infrastructure.
+#   A range of networks is provided, highlighting routes for directness (Fastest) and "cycle friendliness" (Quietest).
+#   It uses origin and destination data from the 2016 Census in combination with modeled demand data to estimate cycling levels and potential at the area, route, and network levels for each county in Ireland and offers estimates of the baseline level of cycling and several future scenario-based levels of cycling.
+
+#   As countries, like Ireland, invest in cycling, the number of killed and seriously injured cyclists must reduce too.
+#   The CRUSE Tool provides estimates of cycling potential and routing for each county in Ireland, and works in both urban and rural settings, to enable monitoring of cycling safety.
+#   With growth in the E-bike market, the tool will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
+#   The CRUSE Tool methodology and findings are directly relevant to addressing the challenges and opportunities faced by other NRAs.
+#   The datasets resulting from the project are open access and can be used by both non-experts and professionals.
+# keywords: 
+#   - Cycling
+#   - Open-source
+#   - Road Safety
+#   - Active Travel
+# date: last-modified
+# format:
+#   elsevier-pdf:
+#     keep-tex: true
+#     journal:
+#       name: Journal Name
+#       formatting: preprint
+#       model: 3p
+#       cite-style: super
 ---
 
-## 10th Transport Research Arena, TRA 2024, Dublin, Ireland {-}
+{{< pagebreak >}}
 
-*Dr. Robin Lovelace (a), Dr. Joey Talbot (a), Dan Brennan (b), Dr. Suzanne Meade (b), Elaine Brick (c), and Peter Wright (d)*
+# Introduction
 
-a University of Leeds, Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom, Leeds LS2 9J, UK.
+[@lovelace2017]
 
-b Transport Infrastructure Ireland, Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland.
+• Title page/main information (title, full names and institutional addresses for all
+authors, indication of the corresponding author)
+• Abstract
+• Introduction
 
-c AECOM, Unit 6, Galway Technology Park, Parkmore, Galway, Ireland.
+# Methods
 
-d AECOM, Winslade House, Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY, UK.
+# Results
 
-## Abstract {-}
+# Discussion
 
-Under the EU Road infrastructure safety management (RISM) directive, the National Road Safety Strategy (RSS), and the Climate Action Plan Transport Infrastructure Ireland (TII) has a remit for road safety and decarbonizing a predominantly road-based network in Ireland.
-To address data needs for both safety and project evaluation on the National Road Network (NRN), the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was developed.
-While cycling in Ireland represents only 3% of total modal share, with higher intensities in urban areas, the levels of cycling collisions are disproportionately high at 20% of all serious injuries and 7% of all fatalities.
-If Ireland is to meet its climate and safety targets, data to establish baseline cycling levels and future cycling levels is needed.
+# Conclusions
 
-Due to an absence of reliable data, particularly rural cycling levels, TII commissioned the Institute for Transport Studies (ITS) at the University of Leeds and AECOM to develop a new tool for this purpose.
-ITS Leeds led the development of the PCT for England and Wales, which has "revolutionized the practice of strategic cycle planning nationally".
-The tool is an open-source approach using recognized open-source methodology to enable planners, engineers, and other stakeholders to make evidence-based decisions for the NRN.
-CRUSE is available at <https://cruse.bike/> and builds on the Propensity to Cycle Tool (PCT) for England and Wales.
-CRUSE goes beyond the PCT in several important ways, higher resolution data, more trip types, including estimates for education, inter-urban, and recreational trips.
-In addition to understanding cycling intensity, for asset planning and management purposes, the tool provides essential road safety information to enable reporting of disaggregate collision rates.
+# List of abbreviations
 
-CRUSE is structured in a similar way to the traditional four-stage transport model, but its use of Open Street Map (OSM) data, used by [Cyclestreets.net](https://www.cyclestreets.net/) for routing, enables network quality to be assessed without costly surveys to record new infrastructure.
-OSM tags generate "cycle friendliness" estimates of all links on the network, based on existing recorded infrastructure.
-A range of networks is provided, highlighting routes for directness (Fastest) and "cycle friendliness" (Quietest).
-It uses origin and destination data from the 2016 Census in combination with modeled demand data to estimate cycling levels and potential at the area, route, and network levels for each county in Ireland and offers estimates of the baseline level of cycling and several future scenario-based levels of cycling.
+# Declarations
 
-As countries, like Ireland, invest in cycling, the number of killed and seriously injured cyclists must reduce too.
-The CRUSE Tool provides estimates of cycling potential and routing for each county in Ireland, and works in both urban and rural settings, to enable monitoring of cycling safety.
-With growth in the E-bike market, the tool will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
-The CRUSE Tool methodology and findings are directly relevant to addressing the challenges and opportunities faced by other NRAs.
-The datasets resulting from the project are open access and can be used by both non-experts and professionals.
+## Availability of data and material {.unnumbered}
+## Funding {.unnumbered}
+## Acknowledgements {.unnumbered}
+## Competing interests {.unnumbered}
+## Authors' contributions {.unnumbered}
 
-**Keywords:** Cycling, Cycling Exposure, Open-source, Road Safety, Active Travel.
+# References {.unnumbered}

### 2023-09-11 cruse-paper.qmd
PDF version

--- 2023-09-11 cruse-paper.qmd
@@
 subtitle: An Open Source Methodology to Support Active Travel
+bibliography: bibliography.bib
+csl: apa.csl
+
+# # Uncomment for draft versions:
+# format: 
+#   html:
+#     output-file: index.html
+#   pdf:
+#     output-file: cruse-paper.pdf 
+#   gfm:
+#     output-file: README.md
+# number-sections: true
+
+# Uncomment the following line to produce the cover page
 author:
@@
 date: last-modified
-bibliography: bibliography.bib
 format:
@@
       cite-style: super
-csl: apa.csl
 ---
-
-## 10th Transport Research Arena, TRA 2024, Dublin, Ireland {.unnumbered}
 
@@
 • Introduction
-• Results
-• Discussion (can be combined in ‘Results and Discussion’ section)
-• Conclusions
-• Methods/Experimental (can also be placed after Introduction)
-• List of abbreviations
-• Declarations*
-o Availability of data and material
-o Funding
-o Acknowledgements
-o Competing interests
-o Authors' contributions
-o Authors' information (optional)
-o *If any of the sections are not relev
+
+# Methods
+
+# Results
+
+# Discussion
+
+# Conclusions
+
+# List of abbreviations
+
+# Declarations
+
+## Availability of data and material {.unnumbered}
+## Funding {.unnumbered}
+## Acknowledgements {.unnumbered}
+## Competing interests {.unnumbered}
+## Authors' contributions {.unnumbered}
+
+# References {.unnumbered}

### 2023-09-12 cruse-paper.qmd
Update bibliography and intro

@@
 
-[@lovelace2017]
-
-• Title page/main information (title, full names and institutional addresses for all
-authors, indication of the corresponding author)
-• Abstract
-• Introduction
+Transport is a major contributor to climate change, premature death due to road traffic collisions and air pollution, yet has the potential to be a major contributor to decarbonization and improved public health.
+Transport is responsible for 23% of global emissions, 70% of which is from road transport, nearly half of which is from passenger cars [@jaramillo2022].
+The transport system encourages, enables and in some cases enforces unsustainable lifestyles, including over-consumption of goods due to excessive mobile storage space and dependency on services that are only accessible by car due to land use plans that have built up around roads [@gray2001; @shergold2012; @motte-baumvol2010].
 
@@
 ## Availability of data and material {.unnumbered}
+
 ## Funding {.unnumbered}
+
 ## Acknowledgements {.unnumbered}
+
 ## Competing interests {.unnumbered}
+
 ## Authors' contributions {.unnumbered}

### 2023-09-12 cruse-paper.qmd
Add placeholders

--- 2023-09-11 cruse-paper.qmd
@@
 csl: apa.csl
+
 # Uncomment for draft versions:
 format: 
+  docx: default
   html:
@@
 
-# Uncomment the following line to produce the cover page
+# # Uncomment the following line to produce the cover page
 # author:
@@
 
-[@lovelace2017]
+Transport is a major contributor to climate change, premature death due to road traffic collisions and air pollution, yet has the potential to be a major contributor to decarbonization and improved public health.
+Transport is responsible for 23% of global emissions, 70% of which is from road transport, nearly half of which is from passenger cars [@jaramillo2022].
+The transport system encourages, enables and in some cases enforces unsustainable lifestyles, including over-consumption of goods due to excessive mobile storage space and dependency on services that are only accessible by car due to land use plans that have built up around roads [@gray2001; @shergold2012; @motte-baumvol2010].
 
-• Title page/main information (title, full names and institutional addresses for all
-authors, indication of the corresponding author)
-• Abstract
-• Introduction
+Recognizing the impacts of poorly designed and performing transport systems on their citizens, governments in many countries have set targets and taken actions and set targets.
+In the context of climate, road safety and physicial inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
+The framework highlights the importance of demand reduction (*avoid*ing unnecessary trips), in addition to mode *shift* and *improvement* of existing energy converters, in that order.
+
+Uptake of cycling, the main topic of this paper, should be seen in this broader context of transport decarbonisation.
+Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips \[citation needed, any ideas Hussein?\]
+.
+The rapid uptake of highly efficient e-bikes can also be seen as an improvement on most electric vehicles, which are too heavy and expensive to be a sustainable alternative and could in fact delay the transition away from car dependency \[citation needed\].
+
+At the European Union level ...
+
+In Ireland...
+
+\[Gary, Suzanne, others input here please.\]
 
 # Methods
+
+Commute and school travel
+
+Other trip purposes
+
+\[Joey please add your bit here.\]
+
+Bit on Jittering [@lovelace2022]
 

### 2023-09-13 cruse-paper.qmd
Add website stuff, close #1

@@
 
-# Uncomment for draft versions:
-format: 
-  docx: default
-  html:
-    output-file: index.html
-  pdf:
-    output-file: cruse-paper.pdf 
-  gfm:
-    output-file: README.md
-number-sections: true
+# # Uncomment for draft versions:
+# format: 
+#   docx: default
+#   html:
+#     output-file: index.html
+#   pdf:
+#     output-file: cruse-paper.pdf 
+#   gfm:
+#     output-file: README.md
+# number-sections: true
 

### 2023-09-13 cruse-paper.tex
Add website stuff, close #1

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
-
-\hypertarget{th-transport-research-arena-tra-2024-dublin-ireland}{%
-\subsection*{10th Transport Research Arena, TRA 2024, Dublin,
-Ireland}\label{th-transport-research-arena-tra-2024-dublin-ireland}}
-\addcontentsline{toc}{subsection}{10th Transport Research Arena, TRA
-2024, Dublin, Ireland}
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-\citep{lovelace2017}
+Transport is a major contributor to climate change, premature death due
+to road traffic collisions and air pollution, yet has the potential to
+be a major contributor to decarbonization and improved public health.
+Transport is responsible for 23\% of global emissions, 70\% of which is
+from road transport, nearly half of which is from passenger cars
+\citep{jaramillo2022}. The transport system encourages, enables and in
+some cases enforces unsustainable lifestyles, including over-consumption
+of goods due to excessive mobile storage space and dependency on
+services that are only accessible by car due to land use plans that have
+built up around roads \citep{gray2001, shergold2012, motte-baumvol2010}.
 
-• Title page/main information (title, full names and institutional
-addresses for all authors, indication of the corresponding author) •
-Abstract • Introduction • Results • Discussion (can be combined in
-`Results and Discussion' section) • Conclusions • Methods/Experimental
-(can also be placed after Introduction) • List of abbreviations •
-Declarations\emph{ o Availability of data and material o Funding o
-Acknowledgements o Competing interests o Authors' contributions o
-Authors' information (optional) o }If any of the sections are not relev
+Recognizing the impacts of poorly designed and performing transport
+systems on their citizens, governments in many countries have set
+targets and taken actions and set targets. In the context of climate,
+road safety and physicial inactivity crises, policies to improve
+transport systems can be classified according to the
+\textquotesingle Avoid-Shift-Improve\textquotesingle{} (ASI) framework
+\citep{jaramillo2022}. The framework highlights the importance of demand
+reduction (\emph{avoid}ing unnecessary trips), in addition to mode
+\emph{shift} and \emph{improvement} of existing energy converters, in
+that order.
 
+Uptake of cycling, the main topic of this paper, should be seen in this
+broader context of transport decarbonisation. Although cycling uptake
+appears on the surface to only relate to the `shift' part of the ASI
+framework, closer consideration of the knock-on impacts of cycling
+uptake shows that it can also help avoid unnecessary trips {[}citation
+needed, any ideas Hussein?{]} . The rapid uptake of highly efficient
+e-bikes can also be seen as an improvement on most electric vehicles,
+which are too heavy and expensive to be a sustainable alternative and
+could in fact delay the transition away from car dependency {[}citation
+needed{]}.
+
+At the European Union level \ldots{}
+
+In Ireland\ldots{}
+
+{[}Gary, Suzanne, others input here please.{]}
+
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
+
+Commute and school travel
+
+Other trip purposes
+
+{[}Joey please add your bit here.{]}
+
+Bit on Jittering \citep{lovelace2022}
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
+
+\hypertarget{list-of-abbreviations}{%
+\section{List of abbreviations}\label{list-of-abbreviations}}
+
+\hypertarget{declarations}{%
+\section{Declarations}\label{declarations}}
+
+\hypertarget{availability-of-data-and-material}{%
+\subsection*{Availability of data and
+material}\label{availability-of-data-and-material}}
+\addcontentsline{toc}{subsection}{Availability of data and material}
+
+\hypertarget{funding}{%
+\subsection*{Funding}\label{funding}}
+\addcontentsline{toc}{subsection}{Funding}
+
+\hypertarget{acknowledgements}{%
+\subsection*{Acknowledgements}\label{acknowledgements}}
+\addcontentsline{toc}{subsection}{Acknowledgements}
+
+\hypertarget{competing-interests}{%
+\subsection*{Competing interests}\label{competing-interests}}
+\addcontentsline{toc}{subsection}{Competing interests}
+
+\hypertarget{authors-contributions}{%
+\subsection*{Authors' contributions}\label{authors-contributions}}
+\addcontentsline{toc}{subsection}{Authors' contributions}
+
+\renewcommand\refname{References}
   \bibliography{bibliography.bib}

### 2023-09-13 README.qmd
Add website stuff, close #1

--- 2023-08-01 README.qmd
@@
 ---
+title: CRUSE to Safe Cycling in Ireland
 format: gfm
-# in future: pdf!
-editor: 
-  markdown: 
-    wrap: sentence
 ---
 
-# 10th Transport Research Arena, TRA 2024, Dublin, Ireland
+<!-- # Introduction
 
-**CRUSE to Safe Cycling in Ireland: An Open Source Methodology to Support Active Travel**
+(Lovelace et al., 2017)
 
-*Dr. Robin Lovelace (a), Dr. Joey Talbot (a), Dan Brennan (b), Dr. Suzanne Meade (b), Elaine Brick (c), and Peter Wright (d)*
+• Title page/main information (title, full names and institutional
+addresses for all authors, indication of the corresponding author) •
+Abstract • Introduction
 
-a University of Leeds, Institute for Transport Studies, Leeds, Ls2 9Jt, United Kingdom, Leeds LS2 9J, UK.
+# Methods
 
-b Transport Infrastructure Ireland, Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland.
+# Results
 
-c AECOM, Unit 6, Galway Technology Park, Parkmore, Galway, Ireland.
+# Discussion
 
-d AECOM, Winslade House, Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY, UK.
+# Conclusions
 
-## Abstract
+# List of abbreviations
 
-Under the EU Road infrastructure safety management (RISM) directive, the National Road Safety Strategy (RSS), and the Climate Action Plan Transport Infrastructure Ireland (TII) has a remit for road safety and decarbonizing a predominantly road-based network in Ireland.
-To address data needs for both safety and project evaluation on the National Road Network (NRN), the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was developed.
-While cycling in Ireland represents only 3% of total modal share, with higher intensities in urban areas, the levels of cycling collisions are disproportionately high at 20% of all serious injuries and 7% of all fatalities.
-If Ireland is to meet its climate and safety targets, data to establish baseline cycling levels and future cycling levels is needed.
+# Declarations
 
-Due to an absence of reliable data, particularly rural cycling levels, TII commissioned the Institute for Transport Studies (ITS) at the University of Leeds and AECOM to develop a new tool for this purpose.
-ITS Leeds led the development of the PCT for England and Wales, which has "revolutionized the practice of strategic cycle planning nationally".
-The tool is an open-source approach using recognized open-source methodology to enable planners, engineers, and other stakeholders to make evidence-based decisions for the NRN.
-CRUSE is available at <https://cruse.bike/> and builds on the Propensity to Cycle Tool (PCT) for England and Wales.
-CRUSE goes beyond the PCT in several important ways, higher resolution data, more trip types, including estimates for education, inter-urban, and recreational trips.
-In addition to understanding cycling intensity, for asset planning and management purposes, the tool provides essential road safety information to enable reporting of disaggregate collision rates.
+## Availability of data and material
 
-CRUSE is structured in a similar way to the traditional four-stage transport model, but its use of Open Street Map (OSM) data, used by [Cyclestreets.net](https://www.cyclestreets.net/) for routing, enables network quality to be assessed without costly surveys to record new infrastructure.
-OSM tags generate "cycle friendliness" estimates of all links on the network, based on existing recorded infrastructure.
-A range of networks is provided, highlighting routes for directness (Fastest) and "cycle friendliness" (Quietest).
-It uses origin and destination data from the 2016 Census in combination with modeled demand data to estimate cycling levels and potential at the area, route, and network levels for each county in Ireland and offers estimates of the baseline level of cycling and several future scenario-based levels of cycling.
+## Funding
 
-As countries, like Ireland, invest in cycling, the number of killed and seriously injured cyclists must reduce too.
-The CRUSE Tool provides estimates of cycling potential and routing for each county in Ireland, and works in both urban and rural settings, to enable monitoring of cycling safety.
-With growth in the E-bike market, the tool will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
-The CRUSE Tool methodology and findings are directly relevant to addressing the challenges and opportunities faced by other NRAs.
-The datasets resulting from the project are open access and can be used by both non-experts and professionals.
+## Acknowledgements
 
-**Keywords:** Cycling, Cycling Exposure, Open-source, Road Safety, Active Travel.
+## Competing interests
+
+## Authors’ contributions
+
+# References
+
+<div id="refs" class="references csl-bib-body hanging-indent"
+line-spacing="2">
+
+<div id="ref-lovelace2017" class="csl-entry">
+
+Lovelace, R., Goodman, A., Aldred, R., Berkoff, N., Abbas, A., &
+Woodcock, J. (2017). The Propensity to Cycle Tool: An open source online
+system for sustainable transport planning. *Journal of Transport and
+Land Use*, *10*(1). <https://doi.org/10.5198/jtlu.2016.862>
+
+</div>
+
+</div> -->

### 2023-09-13 paper.tex
Add website stuff, close #1

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, boxrule=0pt, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, boxrule=0pt, breakable, sharp corners]}{\end{tcolorbox}}\fi
 

### 2023-09-13 paper.tex
Update paper

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, boxrule=0pt, breakable, sharp corners]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, sharp corners, frame hidden, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
 

### 2023-09-13 paper.qmd
Update paper
Simplify website
Add links to REF ICSs

--- 2023-09-13 cruse-paper.qmd
@@
 
-Commute and school travel
+Commute and school travel.
 
-Other trip purposes
+The methods used to generate the evidence presented in the CRUSE tool for Ireland build on previous work, particularly the Propensity to Cycle Tool (PCT), which was originally developed for the UK's Department for Transport and which is publicly available at www.pct.bike [@lovelace2017].
+The original PCT has had major policy impacts.
+It has "revolutionised strategic cycle planning in England and Wales"^[
+See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
+]
+by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential^[
+See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
+]
 
@@
 ## Authors' contributions {.unnumbered}
-
-# References {.unnumbered}

### 2023-09-13 paper.qmd
Tweak footnotes

@@
 The methods used to generate the evidence presented in the CRUSE tool for Ireland build on previous work, particularly the Propensity to Cycle Tool (PCT), which was originally developed for the UK's Department for Transport and which is publicly available at www.pct.bike [@lovelace2017].
-The original PCT has had major policy impacts.
-It has "revolutionised strategic cycle planning in England and Wales"^[
+The original PCT has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies submitted by the University of Leeds and the University of Westminster, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"^[
 See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
@@
 ]
+.
 

### 2023-09-13 paper.qmd
Add more pct refs

--- 2023-09-13 cruse-paper.qmd
@@
 
-Commute and school travel
+The methods used to generate the evidence presented in the CRUSE tool for Ireland build on previous work, particularly the Propensity to Cycle Tool (PCT), which was originally developed for the UK's Department for Transport and which is publicly available at www.pct.bike [@lovelace2017].
+The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies submitted by the University of Leeds and the University of Westminster, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^1] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^2]
+.
 
-Other trip purposes
+[^1]:  See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
 
-\[Joey please add your bit here.\]
+[^2]:  See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
+
+The first version of the PCT was based solely on current and future potential uptake of cycling for single stage travel to work at desire line, zone, route, and route network levels [@lovelace2016].
+It was launched in April 2017 as the government-endorsed tool for strategic cycle network planning, as part of the Cycling and Walking Investment Strategy [@cycling2017].
+Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improvement modelling of impacts on health, environmental and distributional outcomes [@woodcock2021].
 

### 2023-09-13 paper.qmd
Add comment about where to put content on the PCT
add biclaR case

@@
 
+\[Should this bit go in a subsection in the Introduction rather than in the methods?
+(RL)\]
+
 The methods used to generate the evidence presented in the CRUSE tool for Ireland build on previous work, particularly the Propensity to Cycle Tool (PCT), which was originally developed for the UK's Department for Transport and which is publicly available at www.pct.bike [@lovelace2017].
@@
 Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improvement modelling of impacts on health, environmental and distributional outcomes [@woodcock2021].
+In Portugal, a cycling uptake tool to suport decision making was developed based on PCT methods, for the Lisbon metro region (see [biclaR](https://biclar.tmlmobilidade.pt), and again, socioenvirnment impacts were added as a relevant information to support decisions, using the WHO HEAT for Cycling tool[^3] [@felix2023].
+biclaR added on the pevious methods by setting a intermodality scenario of combinining bicycle with the available transit modes (using GTFS), for a maximum time travel and distance on bicycle.
+
+[^3]:  See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
 

### 2023-09-13 paper.qmd
Post merge fixes
FYI made a few typo fixes @temospena, thanks for the quick and concise contribution : )

--- 2023-09-13 paper.qmd
@@
 Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improvement modelling of impacts on health, environmental and distributional outcomes [@woodcock2021].
+In Portugal, a cycling uptake tool to support decision making was developed based on PCT methods, for the Lisbon metro region (see [biclaR](https://biclar.tmlmobilidade.pt), and again, social and environment impacts were added as a relevant information to support decisions, using the WHO HEAT for Cycling tool[^3] [@felix2023].
+biclaR added on the previous methods by setting an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
+
+[^3]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
 

### 2023-09-13 paper.qmd
Sentence breaks

@@
 csl: apa.csl
+editor:
+  markdown:
+    wrap: sentence
 
@@
 
+Despite these advances, the PCT approach outlined in previous research paper had limitations:
+
+-   Limited coverage of modes beyond travel to work and school
+
+-   ...
+
 # Results

### 2023-09-13 paper.qmd
Add brand citation

@@
     wrap: sentence
+number-sections: true
 
@@
 # format: 
+#   html: default
 #   docx: default
-#   html:
-#     output-file: index.html
-#   pdf:
-#     output-file: cruse-paper.pdf 
-#   gfm:
-#     output-file: README.md
-# number-sections: true
 
-# # Uncomment the following line to produce the cover page
-# author:
-#   - name: Dr. Robin Lovelace
-#     email: r.lovelace@leeds.ac.uk
-#     corresponding: true
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Dr. Joey Talbot
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Eugeni Vidal Tortosa
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Hussein Mahfouz
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Dan Brennan
-#     affiliations:
-#       - name: Transport Infrastructure Ireland
-#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
-#   - name: Dr. Suzanne Meade
-#     affiliations:
-#       - name: Transport Infrastructure Ireland
-#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
-#   - name: Elaine Brick
-#     affiliations:
-#       - name: AECOM
-#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-#   - name: Peter Wright
-#     affiliations:
-#       - name: AECOM
-#         address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY, UK
-# abstract: |
-#   Under the EU Road infrastructure safety management (RISM) directive, the National Road Safety Strategy (RSS), and the Climate Action Plan Transport Infrastructure Ireland (TII) has a remit for road safety and decarbonizing a predominantly road-based network in Ireland.
+# Uncomment the following line to produce the cover page
+author:
+  - name: Dr. Robin Lovelace
+    email: r.lovelace@leeds.ac.uk
+    corresponding: true
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Dr. Joey Talbot
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Eugeni Vidal Tortosa
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Hussein Mahfouz
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Dan Brennan
+    affiliations:
+      - name: Transport Infrastructure Ireland
+        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+  - name: Dr. Suzanne Meade
+    affiliations:
+      - name: Transport Infrastructure Ireland
+        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+  - name: Elaine Brick
+    affiliations:
+      - name: AECOM
+        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+  - name: Peter Wright
+    affiliations:
+      - name: AECOM
+        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY, UK
+abstract: |
+  Under the EU Road infrastructure safety management (RISM) directive, the National Road Safety Strategy (RSS), and the Climate Action Plan Transport Infrastructure Ireland (TII) has a remit for road safety and decarbonizing a predominantly road-based network in Ireland.
 
-#   To address data needs for both safety and project evaluation on the National Road Network (NRN), the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was developed.
-#   While cycling in Ireland represents only 3% of total modal share, with higher intensities in urban areas, the levels of cycling collisions are disproportionately high at 20% of all serious injuries and 7% of all fatalities.
-#   If Ireland is to meet its climate and safety targets, data to establish baseline cycling levels and future cycling levels is needed.
+  To address data needs for both safety and project evaluation on the National Road Network (NRN), the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was developed.
+  While cycling in Ireland represents only 3% of total modal share, with higher intensities in urban areas, the levels of cycling collisions are disproportionately high at 20% of all serious injuries and 7% of all fatalities.
+  If Ireland is to meet its climate and safety targets, data to establish baseline cycling levels and future cycling levels is needed.
 
-#   Due to an absence of reliable data, particularly rural cycling levels, TII commissioned the Institute for Transport Studies (ITS) at the University of Leeds and AECOM to develop a new tool for this purpose.
-#   ITS Leeds led the development of the PCT for England and Wales, which has "revolutionized the practice of strategic cycle planning nationally".
-#   The tool is an open-source approach using recognized open-source methodology to enable planners, engineers, and other stakeholders to make evidence-based decisions for the NRN.
-#   CRUSE is available at <https://cruse.bike/> and builds on the Propensity to Cycle Tool (PCT) for England and Wales.
-#   CRUSE goes beyond the PCT in several important ways, higher resolution data, more trip types, including estimates for education, inter-urban, and recreational trips.
-#   In addition to understanding cycling intensity, for asset planning and management purposes, the tool provides essential road safety information to enable reporting of disaggregate collision rates.
+  Due to an absence of reliable data, particularly rural cycling levels, TII commissioned the Institute for Transport Studies (ITS) at the University of Leeds and AECOM to develop a new tool for this purpose.
+  ITS Leeds led the development of the PCT for England and Wales, which has "revolutionized the practice of strategic cycle planning nationally".
+  The tool is an open-source approach using recognized open-source methodology to enable planners, engineers, and other stakeholders to make evidence-based decisions for the NRN.
+  CRUSE is available at <https://cruse.bike/> and builds on the Propensity to Cycle Tool (PCT) for England and Wales.
+  CRUSE goes beyond the PCT in several important ways, higher resolution data, more trip types, including estimates for education, inter-urban, and recreational trips.
+  In addition to understanding cycling intensity, for asset planning and management purposes, the tool provides essential road safety information to enable reporting of disaggregate collision rates.
 
-#   CRUSE is structured in a similar way to the traditional four-stage transport model, but its use of Open Street Map (OSM) data, used by [Cyclestreets.net](https://www.cyclestreets.net/) for routing, enables network quality to be assessed without costly surveys to record new infrastructure.
-#   OSM tags generate "cycle friendliness" estimates of all links on the network, based on existing recorded infrastructure.
-#   A range of networks is provided, highlighting routes for directness (Fastest) and "cycle friendliness" (Quietest).
-#   It uses origin and destination data from the 2016 Census in combination with modeled demand data to estimate cycling levels and potential at the area, route, and network levels for each county in Ireland and offers estimates of the baseline level of cycling and several future scenario-based levels of cycling.
+  CRUSE is structured in a similar way to the traditional four-stage transport model, but its use of Open Street Map (OSM) data, used by [Cyclestreets.net](https://www.cyclestreets.net/) for routing, enables network quality to be assessed without costly surveys to record new infrastructure.
+  OSM tags generate "cycle friendliness" estimates of all links on the network, based on existing recorded infrastructure.
+  A range of networks is provided, highlighting routes for directness (Fastest) and "cycle friendliness" (Quietest).
+  It uses origin and destination data from the 2016 Census in combination with modeled demand data to estimate cycling levels and potential at the area, route, and network levels for each county in Ireland and offers estimates of the baseline level of cycling and several future scenario-based levels of cycling.
 
-#   As countries, like Ireland, invest in cycling, the number of killed and seriously injured cyclists must reduce too.
-#   The CRUSE Tool provides estimates of cycling potential and routing for each county in Ireland, and works in both urban and rural settings, to enable monitoring of cycling safety.
-#   With growth in the E-bike market, the tool will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
-#   The CRUSE Tool methodology and findings are directly relevant to addressing the challenges and opportunities faced by other NRAs.
-#   The datasets resulting from the project are open access and can be used by both non-experts and professionals.
-# keywords: 
-#   - Cycling
-#   - Open-source
-#   - Road Safety
-#   - Active Travel
-# date: last-modified
-# format:
-#   elsevier-pdf:
-#     keep-tex: true
-#     journal:
-#       name: Journal Name
-#       formatting: preprint
-#       model: 3p
-#       cite-style: super
+  As countries, like Ireland, invest in cycling, the number of killed and seriously injured cyclists must reduce too.
+  The CRUSE Tool provides estimates of cycling potential and routing for each county in Ireland, and works in both urban and rural settings, to enable monitoring of cycling safety.
+  With growth in the E-bike market, the tool will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
+  The CRUSE Tool methodology and findings are directly relevant to addressing the challenges and opportunities faced by other NRAs.
+  The datasets resulting from the project are open access and can be used by both non-experts and professionals.
+keywords: 
+  - Cycling
+  - Open-source
+  - Road Safety
+  - Active Travel
+date: last-modified
+
+format:
+  elsevier-pdf:
+    keep-tex: true
+    journal:
+      name: Journal Name
+      formatting: preprint
+      model: 3p
+      cite-style: super
 ---
@@
 
-Uptake of cycling, the main topic of this paper, should be seen in this broader context of transport decarbonisation.
+Uptake of cycling, the main topic of this paper, should be seen in this broader context of transport decarbonisation [@brand2020].
 Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips \[citation needed, any ideas Hussein?\]

### 2023-09-13 paper.tex
Add brand citation

@@
 \fi
-
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{241,243,245}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
-\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
 
@@
 \hypersetup{
-  pdftitle={Short Paper},
-  pdfauthor={Alice Anonymous; Bob Security; Cat Memes; Derek Zoolander},
-  pdfkeywords={keyword1, keyword2},
+  pdftitle={CRUSE to Safe Cycling in Ireland},
+  pdfauthor={Dr.~Robin Lovelace; Dr.~Joey Talbot; Eugeni Vidal Tortosa; Hussein Mahfouz; Dan Brennan; Dr.~Suzanne Meade; Elaine Brick; Peter Wright},
+  pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
   colorlinks=true,
@@
 \begin{frontmatter}
-\title{Short Paper \\\large{A Short Subtitle} }
-\author[1]{Alice Anonymous%
+\title{CRUSE to Safe Cycling in Ireland \\\large{An Open Source
+Methodology to Support Active Travel} }
+\author[1]{Dr.~Robin Lovelace%
 \corref{cor1}%
-\fnref{fn1}}
- \ead{alice@example.com} 
-\author[2]{Bob Security%
-%
-\fnref{fn2}}
- \ead{bob@example.com} 
-\author[2]{Cat Memes%
-%
-\fnref{fn3}}
- \ead{cat@example.com} 
-\author[]{Derek Zoolander%
+}
+ \ead{r.lovelace@leeds.ac.uk} 
+\author[1]{Dr.~Joey Talbot%
 %
 }
- \ead{derek@example.com}
 
-\affiliation[1]{organization={Some Institute of Technology, Department
-Name},addressline={Street Address},city={City},postcode={Postal
-Code},postcodesep={}}
-\affiliation[2]{organization={Another University, Department
-Name},addressline={Street Address},city={City},postcode={Postal
-Code},postcodesep={}}
+\author[1]{Eugeni Vidal Tortosa%
+%
+}
+
+\author[1]{Hussein Mahfouz%
+%
+}
+
+\author[2]{Dan Brennan%
+%
+}
+
+\author[2]{Dr.~Suzanne Meade%
+%
+}
+
+\author[3]{Elaine Brick%
+%
+}
+
+\author[4]{Peter Wright%
+%
+}
+
+\affiliation[1]{organization={University of
+Leeds},addressline={Institute for Transport Studies, Leeds, LS2 9JT,
+United Kingdom},postcodesep={}}
+\affiliation[2]{organization={Transport Infrastructure
+Ireland},addressline={Parkgate Business Centre, Parkgate Street, Dublin
+8, D08 DK10, Ireland},postcodesep={}}
+\affiliation[3]{organization={AECOM},addressline={Unit 6, Galway
+Technology Park, Parkmore, Galway, Ireland},postcodesep={}}
+\affiliation[4]{organization={AECOM},addressline={Winslade House,
+Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY,
+UK},postcodesep={}}
 
 \cortext[cor1]{Corresponding author}
-\fntext[fn1]{This is the first author footnote.}
-\fntext[fn2]{Another author footnote, this is a very long footnote and
-it should be a really long footnote. But this footnote is not yet
-sufficiently long enough to make two lines of footnote text.}
-\fntext[fn3]{Yet another author footnote.}
 
 \begin{abstract}
-This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing
-elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget
-velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris
-sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta
-egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet
-auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo
-molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae.
-Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at
-tempus elit.
+Under the EU Road infrastructure safety management (RISM) directive, the
+National Road Safety Strategy (RSS), and the Climate Action Plan
+Transport Infrastructure Ireland (TII) has a remit for road safety and
+decarbonizing a predominantly road-based network in Ireland.
+
+To address data needs for both safety and project evaluation on the
+National Road Network (NRN), the Cycle Route Uptake and Scenario
+Estimation (CRUSE) Tool was developed. While cycling in Ireland
+represents only 3\% of total modal share, with higher intensities in
+urban areas, the levels of cycling collisions are disproportionately
+high at 20\% of all serious injuries and 7\% of all fatalities. If
+Ireland is to meet its climate and safety targets, data to establish
+baseline cycling levels and future cycling levels is needed.
+
+Due to an absence of reliable data, particularly rural cycling levels,
+TII commissioned the Institute for Transport Studies (ITS) at the
+University of Leeds and AECOM to develop a new tool for this purpose.
+ITS Leeds led the development of the PCT for England and Wales, which
+has ``revolutionized the practice of strategic cycle planning
+nationally''. The tool is an open-source approach using recognized
+open-source methodology to enable planners, engineers, and other
+stakeholders to make evidence-based decisions for the NRN. CRUSE is
+available at \url{https://cruse.bike/} and builds on the Propensity to
+Cycle Tool (PCT) for England and Wales. CRUSE goes beyond the PCT in
+several important ways, higher resolution data, more trip types,
+including estimates for education, inter-urban, and recreational trips.
+In addition to understanding cycling intensity, for asset planning and
+management purposes, the tool provides essential road safety information
+to enable reporting of disaggregate collision rates.
+
+CRUSE is structured in a similar way to the traditional four-stage
+transport model, but its use of Open Street Map (OSM) data, used by
+\href{https://www.cyclestreets.net/}{Cyclestreets.net} for routing,
+enables network quality to be assessed without costly surveys to record
+new infrastructure. OSM tags generate ``cycle friendliness'' estimates
+of all links on the network, based on existing recorded infrastructure.
+A range of networks is provided, highlighting routes for directness
+(Fastest) and ``cycle friendliness'' (Quietest). It uses origin and
+destination data from the 2016 Census in combination with modeled demand
+data to estimate cycling levels and potential at the area, route, and
+network levels for each county in Ireland and offers estimates of the
+baseline level of cycling and several future scenario-based levels of
+cycling.
+
+As countries, like Ireland, invest in cycling, the number of killed and
+seriously injured cyclists must reduce too. The CRUSE Tool provides
+estimates of cycling potential and routing for each county in Ireland,
+and works in both urban and rural settings, to enable monitoring of
+cycling safety. With growth in the E-bike market, the tool will help
+inform inter-urban and rural networks to support the transfer of trips
+to sustainable modes for longer journeys. The CRUSE Tool methodology and
+findings are directly relevant to addressing the challenges and
+opportunities faced by other NRAs. The datasets resulting from the
+project are open access and can be used by both non-experts and
+professionals.
 \end{abstract}
@@
 \begin{keyword}
-    keyword1 \sep 
-    keyword2
+    Cycling \sep Open-source \sep Road Safety \sep 
+    Active Travel
 \end{keyword}
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, sharp corners, frame hidden, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, sharp corners, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
 
-Please make sure that your manuscript follows the guidelines in the
-Guide for Authors of the relevant journal. It is not necessary to
-typeset your manuscript in exactly the same way as an article, unless
-you are submitting to a camera-ready copy (CRC) journal.
+\newpage{}
 
-For detailed instructions regarding the elsevier article class, see
-\url{https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-\hypertarget{bibliography-styles}{%
-\section{Bibliography styles}\label{bibliography-styles}}
+Transport is a major contributor to climate change, premature death due
+to road traffic collisions and air pollution, yet has the potential to
+be a major contributor to decarbonization and improved public health.
+Transport is responsible for 23\% of global emissions, 70\% of which is
+from road transport, nearly half of which is from passenger cars
+\citep{jaramillo2022}. The transport system encourages, enables and in
+some cases enforces unsustainable lifestyles, including over-consumption
+of goods due to excessive mobile storage space and dependency on
+services that are only accessible by car due to land use plans that have
+built up around roads \citep{gray2001, shergold2012, motte-baumvol2010}.
 
-Here are two sample references: \citet{Feynman1963118}
-\citet{Dirac1953888}.
+Recognizing the impacts of poorly designed and performing transport
+systems on their citizens, governments in many countries have set
+targets and taken actions and set targets. In the context of climate,
+road safety and physicial inactivity crises, policies to improve
+transport systems can be classified according to the
+`Avoid-Shift-Improve' (ASI) framework \citep{jaramillo2022}. The
+framework highlights the importance of demand reduction (\emph{avoid}ing
+unnecessary trips), in addition to mode \emph{shift} and
+\emph{improvement} of existing energy converters, in that order.
 
-By default, natbib will be used with the \texttt{authoryear} style, set
-in \texttt{classoption} variable in YAML. You can sets extra options
-with \texttt{natbiboptions} variable in YAML header. Example
+Uptake of cycling, the main topic of this paper, should be seen in this
+broader context of transport decarbonisation. Although cycling uptake
+appears on the surface to only relate to the `shift' part of the ASI
+framework, closer consideration of the knock-on impacts of cycling
+uptake shows that it can also help avoid unnecessary trips {[}citation
+needed, any ideas Hussein?{]} . The rapid uptake of highly efficient
+e-bikes can also be seen as an improvement on most electric vehicles,
+which are too heavy and expensive to be a sustainable alternative and
+could in fact delay the transition away from car dependency {[}citation
+needed{]}.
 
-\begin{verbatim}
-natbiboptions: longnamesfirst,angle,semicolon
-\end{verbatim}
+At the European Union level \ldots{}
 
-There are various more specific bibliography styles available at
-\url{https://support.stmdocs.in/wiki/index.php?title=Model-wise_bibliographic_style_files}.
-To use one of these, add it in the header using, for example,
-\texttt{biblio-style:\ model1-num-names}.
+In Ireland\ldots{}
 
-\hypertarget{using-csl}{%
-\subsection{Using CSL}\label{using-csl}}
+{[}Gary, Suzanne, others input here please.{]}
 
-If \texttt{cite-method} is set to \texttt{citeproc} in
-\texttt{elsevier\_article()}, then pandoc is used for citations instead
-of \texttt{natbib}. In this case, the \texttt{csl} option is used to
-format the references. By default, this template will provide an
-appropriate style, but alternative \texttt{csl} files are available from
-\url{https://www.zotero.org/styles?q=elsevier}. These can be downloaded
-and stored locally, or the url can be used as in the example header.
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
-\hypertarget{equations}{%
-\section{Equations}\label{equations}}
+{[}Should this bit go in a subsection in the Introduction rather than in
+the methods? (RL){]}
 
-Here is an equation: \[ 
-  f_{X}(x) = \left(\frac{\alpha}{\beta}\right)
-  \left(\frac{x}{\beta}\right)^{\alpha-1}
-  e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; 
-  \alpha,\beta,x > 0 .
-\]
+The methods used to generate the evidence presented in the CRUSE tool
+for Ireland build on previous work, particularly the Propensity to Cycle
+Tool (PCT), which was originally developed for the UK's Department for
+Transport and which is publicly available at www.pct.bike
+\citep{lovelace2017}. The PCT approach has had major policy impacts, as
+outlined in Research Excellence Framework (REF) impact case studies
+submitted by the University of Leeds and the University of Westminster,
+which demonstrate that the tool ``revolutionised strategic cycle
+planning in England and Wales''\footnote{See REF Impact Case Study
+  ``Cycle network policy, planning and investment transformed by the
+  Propensity to Cycle Tool'' at
+  \href{https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f}{results2021.ref.ac.uk}
+  submitted by the University of Leeds.} by overcoming the barriers to
+cycling investment imposed by lack of evidence on cycling
+potential\footnote{See the REF Impact Case Study ``Creating Step Changes
+  in Cycling Policy and Infrastructure Planning across the UK'' at
+  \href{https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B}{results2021.ref.ac.uk}
+  submitted by the University of Westminster.} .
 
-Inline equations work as well: \(\sum_{i = 2}^\infty\{\alpha_i^\beta\}\)
+The first version of the PCT was based solely on current and future
+potential uptake of cycling for single stage travel to work at desire
+line, zone, route, and route network levels \citep{lovelace2016}. It was
+launched in April 2017 as the government-endorsed tool for strategic
+cycle network planning, as part of the Cycling and Walking Investment
+Strategy \citep{cycling2017}. Extensions of the PCT approach have
+included estimation of benefits at the individual level
+\citep{woodcock2018}, addition of travel to school network
+\citep{goodman2019}, and improvement modelling of impacts on health,
+environmental and distributional outcomes \citep{woodcock2021}. In
+Portugal, a cycling uptake tool to support decision making was developed
+based on PCT methods, for the Lisbon metro region (see
+\href{https://biclar.tmlmobilidade.pt}{biclaR}, and again, social and
+environment impacts were added as a relevant information to support
+decisions, using the WHO HEAT for Cycling tool\footnote{See the Health
+  economic assessment tool for walking and cycling at
+  \href{https://www.heatwalkingcycling.org/\#homepage}{www.heatwalkingcycling.org}.}
+\citep{felix2023}. biclaR added on the previous methods by setting an
+`intermodality' scenario that combines cycling with currently available
+public transit options based on General Transit Feed Specification
+(GTFS) data.
 
-\hypertarget{figures-and-tables}{%
-\section{Figures and tables}\label{figures-and-tables}}
+Bit on Jittering \citep{lovelace2022}
 
-Figure~\ref{fig-meaningless} is generated using an R chunk.
+Despite these advances, the PCT approach outlined in previous research
+paper had limitations:
 
-\begin{figure}
+\begin{itemize}
+\item
+  Limited coverage of modes beyond travel to work and school
+\item
+  \ldots{}
+\end{itemize}
 
-{\centering \includegraphics[width=0.5\textwidth,height=\textheight]{paper_files/figure-pdf/fig-meaningless-1.pdf}
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
-}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
-\caption{\label{fig-meaningless}A meaningless scatterplot}
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
 
-\end{figure}
+\hypertarget{list-of-abbreviations}{%
+\section{List of abbreviations}\label{list-of-abbreviations}}
 
-\hypertarget{tables-coming-from-r}{%
-\section{Tables coming from R}\label{tables-coming-from-r}}
+\hypertarget{declarations}{%
+\section{Declarations}\label{declarations}}
 
-Tables can also be generated using R chunks, as shown in
-Table~\ref{tbl-simple} example.
+\hypertarget{availability-of-data-and-material}{%
+\subsection*{Availability of data and
+material}\label{availability-of-data-and-material}}
+\addcontentsline{toc}{subsection}{Availability of data and material}
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(}\FunctionTok{head}\NormalTok{(mtcars)[,}\DecValTok{1}\SpecialCharTok{:}\DecValTok{4}\NormalTok{])}
-\end{Highlighting}
-\end{Shaded}
+\hypertarget{funding}{%
+\subsection*{Funding}\label{funding}}
+\addcontentsline{toc}{subsection}{Funding}
 
-\hypertarget{tbl-simple}{}
-\begin{longtable}[]{@{}lrrrr@{}}
-\caption{\label{tbl-simple}Caption centered above table}\tabularnewline
-\toprule\noalign{}
-& mpg & cyl & disp & hp \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-& mpg & cyl & disp & hp \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-Mazda RX4 & 21.0 & 6 & 160 & 110 \\
-Mazda RX4 Wag & 21.0 & 6 & 160 & 110 \\
-Datsun 710 & 22.8 & 4 & 108 & 93 \\
-Hornet 4 Drive & 21.4 & 6 & 258 & 110 \\
-Hornet Sportabout & 18.7 & 8 & 360 & 175 \\
-Valiant & 18.1 & 6 & 225 & 105 \\
-\end{longtable}
+\hypertarget{acknowledgements}{%
+\subsection*{Acknowledgements}\label{acknowledgements}}
+\addcontentsline{toc}{subsection}{Acknowledgements}
 
-\renewcommand\refname{References}
+\hypertarget{competing-interests}{%
+\subsection*{Competing interests}\label{competing-interests}}
+\addcontentsline{toc}{subsection}{Competing interests}
+
+\hypertarget{authors-contributions}{%
+\subsection*{Authors' contributions}\label{authors-contributions}}
+\addcontentsline{toc}{subsection}{Authors' contributions}
+
   \bibliography{bibliography.bib}

### 2023-09-14 paper.qmd
Fix website

@@
 #   docx: default
+
+# # Uncomment for formatted version
+# format:
+#   elsevier-pdf:
+#     keep-tex: true
+#     journal:
+#       name: Journal Name
+#       formatting: preprint
+#       model: 3p
+#       cite-style: super
 
@@
         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Elaine Brick
+    affiliations:
+      - name: AECOM
+        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+  - name: Peter Wright
+    affiliations:
+      - name: AECOM
+        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY, UK
+  - name: "Gary O'Tool"
+    affiliations:
+      - name: AECOM
+        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY, UK
   - name: Dan Brennan
@@
         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
-  - name: Elaine Brick
-    affiliations:
-      - name: AECOM
-        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-  - name: Peter Wright
-    affiliations:
-      - name: AECOM
-        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY, UK
 abstract: |
@@
 date: last-modified
-
-format:
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: Journal Name
-      formatting: preprint
-      model: 3p
-      cite-style: super
 ---
@@
 
+Mobility hub.
+
+Business cases need evidence.
+
+Como UK.
+
 \[Gary, Suzanne, others input here please.\]
+
+In this context, TII commissioned the CRUSE project which is...
+
+The rest of paper is structured as follows...
 
@@
 
+National planning
+
+Regional
+
+Local
+
 # Discussion

### 2023-09-14 paper.tex
Fix website

@@
   pdftitle={CRUSE to Safe Cycling in Ireland},
-  pdfauthor={Dr.~Robin Lovelace; Dr.~Joey Talbot; Eugeni Vidal Tortosa; Hussein Mahfouz; Dan Brennan; Dr.~Suzanne Meade; Elaine Brick; Peter Wright},
+  pdfauthor={Dr.~Robin Lovelace; Dr.~Joey Talbot; Eugeni Vidal Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Tool; Dan Brennan; Dr.~Suzanne Meade},
   pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
@@
 
-\author[2]{Dan Brennan%
+\author[2]{Elaine Brick%
 %
@@
 
-\author[2]{Dr.~Suzanne Meade%
+\author[3]{Peter Wright%
 %
@@
 
-\author[3]{Elaine Brick%
+\author[3]{Gary O'Tool%
 %
@@
 
-\author[4]{Peter Wright%
+\author[4]{Dan Brennan%
+%
+}
+
+\author[4]{Dr.~Suzanne Meade%
 %
@@
 United Kingdom},postcodesep={}}
-\affiliation[2]{organization={Transport Infrastructure
+\affiliation[2]{organization={AECOM},addressline={Unit 6, Galway
+Technology Park, Parkmore, Galway, Ireland},postcodesep={}}
+\affiliation[3]{organization={AECOM},addressline={Winslade House,
+Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY,
+UK},postcodesep={}}
+\affiliation[4]{organization={Transport Infrastructure
 Ireland},addressline={Parkgate Business Centre, Parkgate Street, Dublin
 8, D08 DK10, Ireland},postcodesep={}}
-\affiliation[3]{organization={AECOM},addressline={Unit 6, Galway
-Technology Park, Parkmore, Galway, Ireland},postcodesep={}}
-\affiliation[4]{organization={AECOM},addressline={Winslade House,
-Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY,
-UK},postcodesep={}}
 
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, sharp corners, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 Uptake of cycling, the main topic of this paper, should be seen in this
-broader context of transport decarbonisation. Although cycling uptake
-appears on the surface to only relate to the `shift' part of the ASI
-framework, closer consideration of the knock-on impacts of cycling
-uptake shows that it can also help avoid unnecessary trips {[}citation
-needed, any ideas Hussein?{]} . The rapid uptake of highly efficient
-e-bikes can also be seen as an improvement on most electric vehicles,
-which are too heavy and expensive to be a sustainable alternative and
-could in fact delay the transition away from car dependency {[}citation
-needed{]}.
+broader context of transport decarbonisation \citep{brand2020}. Although
+cycling uptake appears on the surface to only relate to the `shift' part
+of the ASI framework, closer consideration of the knock-on impacts of
+cycling uptake shows that it can also help avoid unnecessary trips
+{[}citation needed, any ideas Hussein?{]} . The rapid uptake of highly
+efficient e-bikes can also be seen as an improvement on most electric
+vehicles, which are too heavy and expensive to be a sustainable
+alternative and could in fact delay the transition away from car
+dependency {[}citation needed{]}.
 
@@
 
+Mobility hub.
+
+Business cases need evidence.
+
+Como UK.
+
 {[}Gary, Suzanne, others input here please.{]}
+
+In this context, TII commissioned the CRUSE project which is\ldots{}
+
+The rest of paper is structured as follows\ldots{}
 
@@
 
+National planning
+
+Regional
+
+Local
+
 \hypertarget{discussion}{%

### 2023-09-14 paper.qmd
Rm date

@@
 
-# Uncomment the following line to produce the cover page
+# # Uncomment the following line to produce the cover page
 author:
@@
   - Active Travel
-date: last-modified
+# date: last-modified
 ---

### 2023-09-14 paper.qmd
Tweak biclar bit

--- 2023-09-14 paper.qmd
@@
 The methods used to generate the evidence presented in the CRUSE tool for Ireland build on previous work, particularly the Propensity to Cycle Tool (PCT), which was originally developed for the UK's Department for Transport and which is publicly available at www.pct.bike [@lovelace2017].
-The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies submitted by the University of Leeds and the University of Westminster, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^1] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^2]
-.
+The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies submitted by the University of Leeds and the University of Westminster, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^1] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^2].
 
@@
 
-The first version of the PCT was based solely on current and future potential uptake of cycling for single stage travel to work at desire line, zone, route, and route network levels [@lovelace2016].
+The first version of the PCT was based on current and future potential uptake of cycling for single stage *travel to work* at desire line, zone, route, and route network levels [@lovelace2016].
 It was launched in April 2017 as the government-endorsed tool for strategic cycle network planning, as part of the Cycling and Walking Investment Strategy [@cycling2017].
-Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improvement modelling of impacts on health, environmental and distributional outcomes [@woodcock2021].
-In Portugal, a cycling uptake tool to support decision making was developed based on PCT methods, for the Lisbon metro region (see [biclaR](https://biclar.tmlmobilidade.pt), and again, social and environment impacts were added as a relevant information to support decisions, using the WHO HEAT for Cycling tool[^3] [@felix2023].
-biclaR added on the previous methods by setting an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
+Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modelling of impacts on health, environmental and distributional outcomes [@woodcock2021].
+Initially developed just for England, the PCT was extended to cover all of Wales (for commuter data only) in 2018.
+
+The PCT approach has been applied in other countries, including Ireland (the topic of this paper), Scotland, and Portugal.
+In Portugal, a strategic cycle network planning tool was developed based on PCT methods, for the Lisbon metro region.
+The resulting open and publicly available available evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
+The 'biclaR' project included estimation of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^3].
+biclaR extended the approach with an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
 

### 2023-09-14 paper.qmd
Add limitations of PCT approach

@@
 
-Bit on Jittering [@lovelace2022]
-
 Despite these advances, the PCT approach outlined in previous research paper had limitations:
 
--   Limited coverage of modes beyond travel to work and school
+- Limited coverage of trip purposes beyond travel to work and school
+- Low resolution of data, with routes starting and ending in administrative zone centroids
+- A web interface that was not user-friendly or intuitive
 
--   ...
+The CRUSE project addressed these limitations as outlined in this paper.
+
+Bit on Jittering [@lovelace2022]
 

### 2023-09-14 README.qmd
Add key link to README
Update readme

@@
 
-<!-- # Introduction
+The main purpose of this site is to provide broadly-defined documentation in support of the web application, research project, and planning support system hosted at [cruse.bike](https://cruse.bike/).
 
-(Lovelace et al., 2017)
+```{r}
+#| echo: false
+#| eval: false
+# Get the first page with pdftools
+remotes::install_cran("qpdf")
+qpdf::pdf_subset("_site/paper.pdf", pages = 1:2)
+```
 
-• Title page/main information (title, full names and institutional
-addresses for all authors, indication of the corresponding author) •
-Abstract • Introduction
-
-# Methods
-
-# Results
-
-# Discussion
-
-# Conclusions
-
-# List of abbreviations
-
-# Declarations
-
-## Availability of data and material
-
-## Funding
-
-## Acknowledgements
-
-## Competing interests
-
-## Authors’ contributions
-
-# References
-
-<div id="refs" class="references csl-bib-body hanging-indent"
-line-spacing="2">
-
-<div id="ref-lovelace2017" class="csl-entry">
-
-Lovelace, R., Goodman, A., Aldred, R., Berkoff, N., Abbas, A., &
-Woodcock, J. (2017). The Propensity to Cycle Tool: An open source online
-system for sustainable transport planning. *Journal of Transport and
-Land Use*, *10*(1). <https://doi.org/10.5198/jtlu.2016.862>
-
-</div>
-
-</div> -->
+See the [paper.qmd](paper.qmd) file for the source code of the paper.

### 2023-09-14 paper.qmd
Add author contributions

@@
 
+Data was obtained from the Central Statistics Office (CSO) and Transport Infrastructure Ireland (TII) under license and cannot be shared publicly.
+The code used to generate the results presented in this paper is available at https://github.com/cruse-bike.
+
 ## Funding {.unnumbered}
+
+The work was funded by Transport Infrastructure Ireland (TII).
 
@@
 
+Thanks to Paul MacDonald and Donal Hodgins (Kildare County Council),  for testing early versions of the tool and for their input into the project.
+Thanks to Ciaran Maguire, Catherine Swift, and others at AECOM for their input into the project.
+
 ## Competing interests {.unnumbered}
 
+The authors declare that they have no competing interests.
+
 ## Authors' contributions {.unnumbered}
+
+RL led the development of the CRUSE tool, with input from JT, EV, HM, EB, PW, GOT, DB, and SM.
+SM instigated the project and coordinated its policy impact.
+PW, EB and GOT provided project management and transport engineering expertise.

### 2023-09-14 paper.qmd
Add section x-refs

@@
 
-At the European Union level ...
+At the European level, the European Union has set a target of reducing greenhouse gas emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.^[
+https://climateactiontracker.org/countries/eu/targets/
+]
+Regarding road traffic casualties, another deadly consequence of inefficient transport systems, the Road Infrastructure Safety Management (RISM) directive (2008/96/EC) requires member states to implement a road safety management system (RSMS) for all public roads.
+Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out".^[
+https://eur-lex.europa.eu/legal-content/EN/TXT/?uri=celex%3A32008L0096
+]
+Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance travelled, the directive requires estimation of distance travelled by mode, down to the road link level.
+For cycling, on which there is a paucity of data, this is a major challenge.
 
-In Ireland...
+In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030.^[
+https://www.drugsandalcohol.ie/35347/
+]
+Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
 
-Mobility hub.
+<!-- \[Gary, Suzanne, others input here please.\] -->
 
-Business cases need evidence.
+It was in this context that TII commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) project.
+Based on the Propensity to Cycle Tool (PCT) for England and Wales [@lovelace2017], the CRUSE tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
 
-Como UK.
+The rest of paper is structured as follows.
+In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE tool for Ireland.
+In @sec-results, we present the results of the CRUSE tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
+In @sec-discussion, we discuss the implications of the results for policy and practice.
+In @sec-conclusions, we conclude the paper.
 
-\[Gary, Suzanne, others input here please.\]
-
-In this context, TII commissioned the CRUSE project which is...
-
-The rest of paper is structured as follows...
-
-# Methods
+# Methods {#sec-methods}
 
@@
 
-# Results
+# Results {#sec-results}
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 
-# Conclusions
+# Conclusions {#sec-conclusions}
 

### 2023-09-18 paper.qmd

--- 2023-09-14 paper.qmd
@@
 Uptake of cycling, the main topic of this paper, should be seen in this broader context of transport decarbonisation [@brand2020].
-Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips \[citation needed, any ideas Hussein?\]
-.
+Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
 The rapid uptake of highly efficient e-bikes can also be seen as an improvement on most electric vehicles, which are too heavy and expensive to be a sustainable alternative and could in fact delay the transition away from car dependency \[citation needed\].
 
-At the European Union level ...
+At the European level, the European Union has set a target of reducing greenhouse gas emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.[^1]
+Regarding road traffic casualties, another deadly consequence of inefficient transport systems, the Road Infrastructure Safety Management (RISM) directive (2008/96/EC) requires member states to implement a road safety management system (RSMS) for all public roads.
+Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out".
+[^2] Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance travelled, the directive requires estimation of distance travelled by mode, down to the road link level
+. For cycling, on which there is a paucity of data, this is a major challenge
+.
 
-In Ireland...
+[^1]: https://climateactiontracker.org/countries/eu/targets/
 
-Mobility hub.
+[^2]: https://eur-lex.europa.eu/legal-content/EN/TXT/?u
+    ri=celex%3A32008L0096
 
-Business cases need evidence.
+In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030.[^3]
+Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
 
-Como UK.
+[^3]: https://www.drugsandalcohol.ie/35347/
 
-\[Gary, Suzanne, others input here please.\]
+<!-- \[Gary, Suzanne, others input here please.\] -->
 
-In this context, TII commissioned the CRUSE project which is...
+It was in this context that TII commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) project.
+Based on the Propensity to Cycle Tool (PCT) for England and Wales [@lovelace2017], the CRUSE tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
 
-The rest of paper is structured as follows...
+The rest of paper is structured as follows.
+In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE tool for Ireland.
+In @sec-results, we present the results of the CRUSE tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
+In @sec-discussion, we discuss the implications of the results for policy and practice.
+In @sec-conclusions, we conclude the paper.
 
-# Methods
+# Methods {#sec-methods}
 
@@
 The methods used to generate the evidence presented in the CRUSE tool for Ireland build on previous work, particularly the Propensity to Cycle Tool (PCT), which was originally developed for the UK's Department for Transport and which is publicly available at www.pct.bike [@lovelace2017].
-The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies submitted by the University of Leeds and the University of Westminster, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^1] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^2].
+The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies submitted by the University of Leeds and the University of Westminster, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^4] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^5]
+.
 
-[^1]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
+[^4]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
 
-[^2]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
+[^5]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
 
@@
 The resulting open and publicly available available evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
-The 'biclaR' project included estimation of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^3].
+The 'biclaR' project included estimation of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^6].
 biclaR extended the approach with an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
 
-[^3]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
+[^6]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
 

### 2023-09-19 paper.qmd
Intro tweaks

--- 2023-09-14 paper.qmd
@@
 
-Transport is a major contributor to climate change, premature death due to road traffic collisions and air pollution, yet has the potential to be a major contributor to decarbonization and improved public health.
-Transport is responsible for 23% of global emissions, 70% of which is from road transport, nearly half of which is from passenger cars [@jaramillo2022].
+Transport kills.
+Transport systems are a major contributor to climate change, a leading cause of premature death and injury due to road traffic collisions, and a substantial cause of air pollution inhaled by citizens in many urban areas.
+Transport policies therefore have great potential to decarbonize economies, improve public health, and save lives.
+
+Transport is responsible for 23% of global emissions, 70% of which is from road transport, nearly half of which (around 10% of global emissions) is from passenger cars [@jaramillo2022].
 The transport system encourages, enables and in some cases enforces unsustainable lifestyles, including over-consumption of goods due to excessive mobile storage space and dependency on services that are only accessible by car due to land use plans that have built up around roads [@gray2001; @shergold2012; @motte-baumvol2010].
 
-Recognizing the impacts of poorly designed and performing transport systems on their citizens, governments in many countries have set targets and taken actions and set targets.
+Recognizing the growing evidence of such impacts of poorly designed and performing transport systems, governments in many countries have set targets and taken actions.
 In the context of climate, road safety and physicial inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
@@
 
-Uptake of cycling, the main topic of this paper, should be seen in this broader context of transport decarbonisation [@brand2020].
+Uptake of cycling, the main topic of this paper, should be seen in this broader context of transport decarbonisation [@brand2020] and sustainable mobility .
 Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips \[citation needed, any ideas Hussein?\]

### 2023-09-19 paper.qmd
Add citation

--- 2023-09-18 paper.qmd
@@
 
-Uptake of cycling, the main topic of this paper, should be seen in this broader context of transport decarbonisation [@brand2020].
-Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
+Uptake of cycling, the main topic of this paper, should be seen in this broader context of transport decarbonisation [@brand2020] and sustainable mobility [@burns2013].
+Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips \[citation needed, any ideas Hussein?\]
+.
 The rapid uptake of highly efficient e-bikes can also be seen as an improvement on most electric vehicles, which are too heavy and expensive to be a sustainable alternative and could in fact delay the transition away from car dependency \[citation needed\].

### 2023-09-19 paper.qmd

@@
 Uptake of cycling, the main topic of this paper, should be seen in this broader context of transport decarbonisation [@brand2020] and sustainable mobility [@burns2013].
-Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips \[citation needed, any ideas Hussein?\]
-.
+Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
 The rapid uptake of highly efficient e-bikes can also be seen as an improvement on most electric vehicles, which are too heavy and expensive to be a sustainable alternative and could in fact delay the transition away from car dependency \[citation needed\].
@@
 Regarding road traffic casualties, another deadly consequence of inefficient transport systems, the Road Infrastructure Safety Management (RISM) directive (2008/96/EC) requires member states to implement a road safety management system (RSMS) for all public roads.
-Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out".
-[^2] Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance travelled, the directive requires estimation of distance travelled by mode, down to the road link level
-. For cycling, on which there is a paucity of data, this is a major challenge
-.
-
-[^1]:  https://climateactiontracker.org/countries/eu/targets/
-
-[^2]:  https://eur-lex.europa.eu/legal-content/EN/TXT/?u
-    ri=celex%3A32008L0096
+Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out".^[
+https://eur-lex.europa.eu/legal-content/EN/TXT/?uri=celex%3A32008L0096
+]
+Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance travelled, the directive requires estimation of distance travelled by mode, down to the road link level.
+For cycling, on which there is a paucity of data, this is a major challenge.
 
@@
 Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
-
-[^3]:  https://www.drugsandalcohol.ie/35347/
 

### 2023-09-19 paper.qmd
Update doc

@@
 Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
-The rapid uptake of highly efficient e-bikes can also be seen as an improvement on most electric vehicles, which are too heavy and expensive to be a sustainable alternative and could in fact delay the transition away from car dependency \[citation needed\].
+The rapid uptake of highly efficient e-bikes can also be seen as an improvement on most electric vehicles, which are too heavy and expensive to be a sustainable alternative and could in fact delay the transition away from car dependency and inadvertently enable "high travel lock-in" [@shifting].
 
-At the European level, the European Union has set a target of reducing greenhouse gas emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.[^1]
-Regarding road traffic casualties, another deadly consequence of inefficient transport systems, the Road Infrastructure Safety Management (RISM) directive (2008/96/EC) requires member states to implement a road safety management system (RSMS) for all public roads.
-Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out".^[
-https://eur-lex.europa.eu/legal-content/EN/TXT/?uri=celex%3A32008L0096
-]
+At the European level, the European Union has set a target of reducing greenhouse gas emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.\[\^1\] Regarding road traffic casualties, another deadly consequence of inefficient transport systems, the Road Infrastructure Safety Management (RISM) directive (2008/96/EC) requires member states to implement a road safety management system (RSMS) for all public roads.
+Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out".[^1]
 Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance travelled, the directive requires estimation of distance travelled by mode, down to the road link level.
@@
 
-In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030.[^3]
-Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
+[^1]:  https://eur-lex.europa.eu/legal-content/EN/TXT/?u
+    ri=celex%3A32008L0096
+
+In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030.\[\^3\] Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
 
@@
 The methods used to generate the evidence presented in the CRUSE tool for Ireland build on previous work, particularly the Propensity to Cycle Tool (PCT), which was originally developed for the UK's Department for Transport and which is publicly available at www.pct.bike [@lovelace2017].
-The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies submitted by the University of Leeds and the University of Westminster, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^4] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^5]
+The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies submitted by the University of Leeds and the University of Westminster, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^2] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^3]
 .
 
-[^4]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
+[^2]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
 
-[^5]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
+[^3]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
 
@@
 The resulting open and publicly available available evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
-The 'biclaR' project included estimation of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^6].
+The 'biclaR' project included estimation of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^4].
 biclaR extended the approach with an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
 
-[^6]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
+[^4]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
 

### 2023-09-19 paper.qmd
Add anable reference

@@
 Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
-The rapid uptake of highly efficient e-bikes can also be seen as an improvement on most electric vehicles, which are too heavy and expensive to be a sustainable alternative and could in fact delay the transition away from car dependency and inadvertently enable "high travel lock-in" [@shifting].
+The rapid uptake of highly efficient e-bikes can also be seen as an improvement on most electric vehicles, which are too heavy and expensive to be a sustainable alternative and could in fact delay the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
 

### 2023-09-20 paper.qmd
More intro updates

@@
 
-Transport kills.
-Transport systems are a major contributor to climate change, a leading cause of premature death and injury due to road traffic collisions, and a substantial cause of air pollution inhaled by citizens in many urban areas.
-Transport policies therefore have great potential to decarbonize economies, improve public health, and save lives.
+Mobility kills.
+High energy transport systems are a major contributor to climate change, a leading cause of premature death and injury due to road traffic collisions, and a cause of disease due to airborne pollutants.
+Conversely, evidence-based and effective transport *policies* have great potential to decarbonize economies, improve public health, and save lives.
 
@@
 The rapid uptake of highly efficient e-bikes can also be seen as an improvement on most electric vehicles, which are too heavy and expensive to be a sustainable alternative and could in fact delay the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
+At the European level, the European Union has a target of reducing greenhouse gas emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.\[\^1\]
 
-At the European level, the European Union has set a target of reducing greenhouse gas emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.\[\^1\] Regarding road traffic casualties, another deadly consequence of inefficient transport systems, the Road Infrastructure Safety Management (RISM) directive (2008/96/EC) requires member states to implement a road safety management system (RSMS) for all public roads.
+Regarding road traffic casualties, another deadly consequence of inefficient transport systems, the Road Infrastructure Safety Management (RISM) directive (2008/96/EC) requires member states to implement a road safety management system (RSMS) for all public roads.
 Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out".[^1]
 Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance travelled, the directive requires estimation of distance travelled by mode, down to the road link level.
-For cycling, on which there is a paucity of data, this is a major challenge.
+For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
+Better data to inform road safety policies and interventions is a motivation for the CRUSE tool outlined in this paper, and modeling active travel more broadly.
 
@@
 
-It was in this context that TII commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) project.
+It was in this context that Transport Infrastructure Ireland (TII) commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) project.
 Based on the Propensity to Cycle Tool (PCT) for England and Wales [@lovelace2017], the CRUSE tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
@@
 # Methods {#sec-methods}
-
-\[Should this bit go in a subsection in the Introduction rather than in the methods?
-(RL)\]
 

### 2023-09-20 paper.qmd
New subsections

@@
 
-The methods used to generate the evidence presented in the CRUSE tool for Ireland build on previous work, particularly the Propensity to Cycle Tool (PCT), which was originally developed for the UK's Department for Transport and which is publicly available at www.pct.bike [@lovelace2017].
-The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies submitted by the University of Leeds and the University of Westminster, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^2] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^3]
+The methods used to generate the evidence presented in the CRUSE tool for Ireland build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
+An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
+The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^2] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^3]
 .
@@
 The PCT approach has been applied in other countries, including Ireland (the topic of this paper), Scotland, and Portugal.
-In Portugal, a strategic cycle network planning tool was developed based on PCT methods, for the Lisbon metro region.
-The resulting open and publicly available available evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
-The 'biclaR' project included estimation of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^4].
-biclaR extended the approach with an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
+In Portugal the 'biclaR' project, based on methods underlying the PCT, has been developed and deployed for the Lisbon metro region.
+The resulting evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
+biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^4] and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
 
@@
 
-Despite these advances, the PCT approach outlined in previous research paper had limitations:
+The CRUSE tool seeks to overcome the following methodological limitations of the original PCT:
 
+-   Low resolution of data, with routes starting and ending in administrative zone centroids
 -   Limited coverage of trip purposes beyond travel to work and school
--   Low resolution of data, with routes starting and ending in administrative zone centroids
 -   A web interface that was not user-friendly or intuitive
 
-The CRUSE project addressed these limitations as outlined in this paper.
+Methods were developed to overcome each of these, as outlined in the following subsections.
+
+## Disaggregation of origin-destination data
 
 Bit on Jittering [@lovelace2022]
+
+## Additional trip purposes
+
+bla bla
+
+## User interface
 

### 2023-09-20 paper.qmd
Number sections

@@
 
-# Methods {#sec-methods}
+# Methods and data {#sec-methods}
 
@@
 
-Methods were developed to overcome each of these, as outlined in the following subsections.
+Methods were developed to overcome each of these, as outlined in @sec-trip-purposes to @sec-ui.
 
-## Disaggregation of origin-destination data
+## Disaggregation of origin-destination data {#sec-disaggregation}
+
+A feature of active travel interventions is that they require dense networks of routes to be effective [TODO: cite parkin].
+This means that high levels of geographic resolution are needed in the data used to estimate cycling potential.
+However, datasets on travel patterns are often only available at the level of administrative zones.
 
@@
 
-## Additional trip purposes
+## Additional trip purposes {#sec-trip-purposes}
 
@@
 
-## User interface
+## User interface {#sec-ui}
 

### 2023-09-20 paper.qmd
Add abbreviations

@@
 
+<!-- Todo: add section on input datasets -->
+
 ## Disaggregation of origin-destination data {#sec-disaggregation}
@@
 However, datasets on travel patterns are often only available at the level of administrative zones.
+The Central Statistics Office (CSO) in Ireland provides Place of Work, School or College Census of Anonymised Records (POWSCAR) data on the number of people travelling to work and school at the Electoral Division (ED) level, for example.
 
@@
 # List of abbreviations
+
+CSO: Central Statistics Office
+
+CRUSE: Cycle Route Uptake and Scenario Estimation
+
+ED: Electoral Division
+
+GTFS: General Transit Feed Specification
+
+KSI: Killed and Seriously Injured
+
+NRA: National Road Authority
+
+NRN: National Road Network
+
+OSM: Open Street Map
+
+PCT: Propensity to Cycle Tool
+
+POWSCAR: Place of Work, School or College Census of Anonymised Records
+
+TII: Transport Infrastructure Ireland
 

### 2023-09-20 paper.qmd
Add image

@@
 
-A feature of active travel interventions is that they require dense networks of routes to be effective [TODO: cite parkin].
+A feature of active travel interventions is that they require dense networks of routes to be effective \[TODO: cite parkin\].
 This means that high levels of geographic resolution are needed in the data used to estimate cycling potential.
 However, datasets on travel patterns are often only available at the level of administrative zones.
-The Central Statistics Office (CSO) in Ireland provides Place of Work, School or College Census of Anonymised Records (POWSCAR) data on the number of people travelling to work and school at the Electoral Division (ED) level, for example.
+The Central Statistics Office (CSO) in Ireland provides Place of Work, School or College Census of Anonymised Records ([POWSCAR](https://www.cso.ie/en/census/census2016reports/powscar/)) data on the number of people travelling to work and school at the Electoral Division (ED) level, for example.
 
-Bit on Jittering [@lovelace2022]
+The method used to convert OD data to route networks used in the PCT was to calculate a single route between the population weighted centroids associated with each OD pair.
+This method works fine when the OD data represents movement between small areas, but was not appropriate for generating route networks from the POWSCAR data because zone centroids are so far apart that the resulting route networks would be sparse and unrealistic.
+To tackle this issue we developed a new method for OD data disaggregation called 'jittering' [@lovelace2022].
+The method works by first disaggregating the OD data based on a 'disaggregation threshold' and then assigning each disaggregated 'sub-OD' pair to 'subpoints' within each zone.
+As shown in @fig-dublin, the resulting route networks are dense, even in rural areas.
+
+![](images/paste-1.png)
 
@@
 
+OD: Origin-Destination, typically referring to origin-destination data which contains information on the number of people travelling between each pair of zones
+
 OSM: Open Street Map

### 2023-09-20 paper.qmd
Add parkin2018 citation
Updates for .tex version

@@
 
-# # Uncomment for formatted version
-# format:
-#   elsevier-pdf:
-#     keep-tex: true
-#     journal:
-#       name: Journal Name
-#       formatting: preprint
-#       model: 3p
-#       cite-style: super
+# Uncomment for formatted version
+format:
+  elsevier-pdf:
+    keep-tex: true
+    journal:
+      name: Journal Name
+      formatting: preprint
+      model: 3p
+      cite-style: super
 
@@
 
-A feature of active travel interventions is that they require dense networks of routes to be effective \[TODO: cite parkin\].
+A feature of active travel interventions is that they require dense networks of routes to be effective [@parkin2018].
 This means that high levels of geographic resolution are needed in the data used to estimate cycling potential.
@@
 
+::: {#fig-dublin}
 ![](images/paste-1.png)
+
+![](images/paste-3.png)
+
+Illustration of the density of route networks generated by the CRUSE tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland.
+Source: CRUSE tool, available at <https://cruse.bike/>.
+:::
 

### 2023-09-20 paper.tex
Updates for .tex version

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, frame hidden, sharp corners, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-Transport is a major contributor to climate change, premature death due
-to road traffic collisions and air pollution, yet has the potential to
-be a major contributor to decarbonization and improved public health.
+Mobility kills. High energy transport systems are a major contributor to
+climate change, a leading cause of premature death and injury due to
+road traffic collisions, and a cause of disease due to airborne
+pollutants. Conversely, evidence-based and effective transport
+\emph{policies} have great potential to decarbonize economies, improve
+public health, and save lives.
+
 Transport is responsible for 23\% of global emissions, 70\% of which is
-from road transport, nearly half of which is from passenger cars
-\citep{jaramillo2022}. The transport system encourages, enables and in
-some cases enforces unsustainable lifestyles, including over-consumption
-of goods due to excessive mobile storage space and dependency on
-services that are only accessible by car due to land use plans that have
-built up around roads \citep{gray2001, shergold2012, motte-baumvol2010}.
+from road transport, nearly half of which (around 10\% of global
+emissions) is from passenger cars \citep{jaramillo2022}. The transport
+system encourages, enables and in some cases enforces unsustainable
+lifestyles, including over-consumption of goods due to excessive mobile
+storage space and dependency on services that are only accessible by car
+due to land use plans that have built up around roads
+\citep{gray2001, shergold2012, motte-baumvol2010}.
 
-Recognizing the impacts of poorly designed and performing transport
-systems on their citizens, governments in many countries have set
-targets and taken actions and set targets. In the context of climate,
-road safety and physicial inactivity crises, policies to improve
-transport systems can be classified according to the
-`Avoid-Shift-Improve' (ASI) framework \citep{jaramillo2022}. The
-framework highlights the importance of demand reduction (\emph{avoid}ing
-unnecessary trips), in addition to mode \emph{shift} and
-\emph{improvement} of existing energy converters, in that order.
+Recognizing the growing evidence of such impacts of poorly designed and
+performing transport systems, governments in many countries have set
+targets and taken actions. In the context of climate, road safety and
+physicial inactivity crises, policies to improve transport systems can
+be classified according to the `Avoid-Shift-Improve' (ASI) framework
+\citep{jaramillo2022}. The framework highlights the importance of demand
+reduction (\emph{avoid}ing unnecessary trips), in addition to mode
+\emph{shift} and \emph{improvement} of existing energy converters, in
+that order.
 
 Uptake of cycling, the main topic of this paper, should be seen in this
-broader context of transport decarbonisation \citep{brand2020}. Although
-cycling uptake appears on the surface to only relate to the `shift' part
-of the ASI framework, closer consideration of the knock-on impacts of
-cycling uptake shows that it can also help avoid unnecessary trips
-{[}citation needed, any ideas Hussein?{]} . The rapid uptake of highly
-efficient e-bikes can also be seen as an improvement on most electric
-vehicles, which are too heavy and expensive to be a sustainable
-alternative and could in fact delay the transition away from car
-dependency {[}citation needed{]}.
+broader context of transport decarbonisation \citep{brand2020} and
+sustainable mobility \citep{burns2013}. Although cycling uptake appears
+on the surface to only relate to the `shift' part of the ASI framework,
+closer consideration of the knock-on impacts of cycling uptake shows
+that it can also help avoid unnecessary trips \citep{nello-deakin2020}.
+The rapid uptake of highly efficient e-bikes can also be seen as an
+improvement on most electric vehicles, which are too heavy and expensive
+to be a sustainable alternative and could in fact delay the transition
+away from car dependency and inadvertently enable ``high travel
+lock-in'' \citep{anable2019}. At the European level, the European Union
+has a target of reducing greenhouse gas emissions by 55\% by 2030,
+compared to 1990 levels, and to achieve `net-zero' by 2050.{[}\^{}1{]}
 
-At the European Union level \ldots{}
+Regarding road traffic casualties, another deadly consequence of
+inefficient transport systems, the Road Infrastructure Safety Management
+(RISM) directive (2008/96/EC) requires member states to implement a road
+safety management system (RSMS) for all public roads. Specifically,
+``Member States shall ensure that the ranking of high accident
+concentration sections and the network safety ranking are carried
+out''.\footnote{https://eur-lex.europa.eu/legal-content/EN/TXT/?u
+  ri=celex\%3A32008L0096} Given that `safety' in this context is
+usefully quantified as the number of people killed and seriously injured
+(KSI) per distance travelled, the directive requires estimation of
+distance travelled by mode, down to the road link level. For active
+modes, about which there is a paucity of data compared with motorized
+modes, this is a major challenge. Better data to inform road safety
+policies and interventions is a motivation for the CRUSE tool outlined
+in this paper, and modeling active travel more broadly.
 
-In Ireland\ldots{}
+In Ireland, the Road Safety Authority (RSA) has set the target of
+halving the number of road traffic deaths and serious injuries by
+2030.{[}\^{}3{]} Doing so while simultaneously enabling rapid uptake of
+active modes will require key active travel routes to be identified and
+improved.
 
-Mobility hub.
+It was in this context that Transport Infrastructure Ireland (TII)
+commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE)
+project. Based on the Propensity to Cycle Tool (PCT) for England and
+Wales \citep{lovelace2017}, the CRUSE tool was developed to provide
+evidence on current cycling levels and future cycling potential
+nationwide across Ireland.
 
-Business cases need evidence.
+The rest of paper is structured as follows. In
+Section~\ref{sec-methods}, we outline the methods used to generate the
+evidence presented in the CRUSE tool for Ireland. In
+Section~\ref{sec-results}, we present the results of the CRUSE tool,
+including estimates of current cycling levels and future cycling
+potential at the national, regional and local levels. In
+Section~\ref{sec-discussion}, we discuss the implications of the results
+for policy and practice. In Section~\ref{sec-conclusions}, we conclude
+the paper.
 
-Como UK.
-
-{[}Gary, Suzanne, others input here please.{]}
-
-In this context, TII commissioned the CRUSE project which is\ldots{}
-
-The rest of paper is structured as follows\ldots{}
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-{[}Should this bit go in a subsection in the Introduction rather than in
-the methods? (RL){]}
+\hypertarget{sec-methods}{%
+\section{Methods and data}\label{sec-methods}}
 
 The methods used to generate the evidence presented in the CRUSE tool
-for Ireland build on previous work, particularly the Propensity to Cycle
-Tool (PCT), which was originally developed for the UK's Department for
-Transport and which is publicly available at www.pct.bike
-\citep{lovelace2017}. The PCT approach has had major policy impacts, as
-outlined in Research Excellence Framework (REF) impact case studies
-submitted by the University of Leeds and the University of Westminster,
-which demonstrate that the tool ``revolutionised strategic cycle
-planning in England and Wales''\footnote{See REF Impact Case Study
-  ``Cycle network policy, planning and investment transformed by the
-  Propensity to Cycle Tool'' at
+for Ireland build on the Propensity to Cycle Tool (PCT), which was
+originally funded by the UK's Department for Transport and developed by
+a multi-university team. An important feature of the PCT is that it is
+open source and publicly available (at
+\href{https://www.pct.bike/}{www.pct.bike}), allowing its use by all
+stakeholders in the transport planning process \citep{lovelace2017}. The
+PCT approach has had major policy impacts, as outlined in Research
+Excellence Framework (REF) impact case studies, which demonstrate that
+the tool ``revolutionised strategic cycle planning in England and
+Wales''\footnote{See REF Impact Case Study ``Cycle network policy,
+  planning and investment transformed by the Propensity to Cycle Tool''
+  at
   \href{https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f}{results2021.ref.ac.uk}
@@
 
-The first version of the PCT was based solely on current and future
-potential uptake of cycling for single stage travel to work at desire
-line, zone, route, and route network levels \citep{lovelace2016}. It was
+The first version of the PCT was based on current and future potential
+uptake of cycling for single stage \emph{travel to work} at desire line,
+zone, route, and route network levels \citep{lovelace2016}. It was
 launched in April 2017 as the government-endorsed tool for strategic
@@
 \citep{woodcock2018}, addition of travel to school network
-\citep{goodman2019}, and improvement modelling of impacts on health,
-environmental and distributional outcomes \citep{woodcock2021}. In
-Portugal, a cycling uptake tool to support decision making was developed
-based on PCT methods, for the Lisbon metro region (see
-\href{https://biclar.tmlmobilidade.pt}{biclaR}, and again, social and
-environment impacts were added as a relevant information to support
-decisions, using the WHO HEAT for Cycling tool\footnote{See the Health
-  economic assessment tool for walking and cycling at
+\citep{goodman2019}, and improved modelling of impacts on health,
+environmental and distributional outcomes \citep{woodcock2021}.
+Initially developed just for England, the PCT was extended to cover all
+of Wales (for commuter data only) in 2018.
+
+The PCT approach has been applied in other countries, including Ireland
+(the topic of this paper), Scotland, and Portugal. In Portugal the
+`biclaR' project, based on methods underlying the PCT, has been
+developed and deployed for the Lisbon metro region. The resulting
+evidence is available in an interactive web application hosted at
+\href{https://biclar.tmlmobilidade.pt}{biclar.tmlmobilidade.pt}
+\citep{felix2023}. biclaR includes estimates of impacts, using the World
+Health Organisation (WHO) `HEAT for Cycling' tool\footnote{See the
+  Health economic assessment tool for walking and cycling at
   \href{https://www.heatwalkingcycling.org/\#homepage}{www.heatwalkingcycling.org}.}
-\citep{felix2023}. biclaR added on the previous methods by setting an
-`intermodality' scenario that combines cycling with currently available
-public transit options based on General Transit Feed Specification
-(GTFS) data.
+and an `intermodality' scenario that combines cycling with currently
+available public transit options based on General Transit Feed
+Specification (GTFS) data.
 
-Bit on Jittering \citep{lovelace2022}
-
-Despite these advances, the PCT approach outlined in previous research
-paper had limitations:
+The CRUSE tool seeks to overcome the following methodological
+limitations of the original PCT:
 
 \begin{itemize}
+\tightlist
 \item
-  Limited coverage of modes beyond travel to work and school
+  Low resolution of data, with routes starting and ending in
+  administrative zone centroids
 \item
-  \ldots{}
+  Limited coverage of trip purposes beyond travel to work and school
+\item
+  A web interface that was not user-friendly or intuitive
 \end{itemize}
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
+Methods were developed to overcome each of these, as outlined in
+Section~\ref{sec-trip-purposes} to Section~\ref{sec-ui}.
+
+\hypertarget{sec-disaggregation}{%
+\subsection{Disaggregation of origin-destination
+data}\label{sec-disaggregation}}
+
+A feature of active travel interventions is that they require dense
+networks of routes to be effective \citep{parkin2018}. This means that
+high levels of geographic resolution are needed in the data used to
+estimate cycling potential. However, datasets on travel patterns are
+often only available at the level of administrative zones. The Central
+Statistics Office (CSO) in Ireland provides Place of Work, School or
+College Census of Anonymised Records
+(\href{https://www.cso.ie/en/census/census2016reports/powscar/}{POWSCAR})
+data on the number of people travelling to work and school at the
+Electoral Division (ED) level, for example.
+
+The method used to convert OD data to route networks used in the PCT was
+to calculate a single route between the population weighted centroids
+associated with each OD pair. This method works fine when the OD data
+represents movement between small areas, but was not appropriate for
+generating route networks from the POWSCAR data because zone centroids
+are so far apart that the resulting route networks would be sparse and
+unrealistic. To tackle this issue we developed a new method for OD data
+disaggregation called `jittering' \citep{lovelace2022}. The method works
+by first disaggregating the OD data based on a `disaggregation
+threshold' and then assigning each disaggregated `sub-OD' pair to
+`subpoints' within each zone. As shown in Figure~\ref{fig-dublin}, the
+resulting route networks are dense, even in rural areas.
+
+\begin{figure}
+
+{\centering
+
+\includegraphics{images/paste-1.png}
+
+\includegraphics{images/paste-3.png}
+
+}
+
+\caption{\label{fig-dublin}Illustration of the density of route networks
+generated by the CRUSE tool for Dublin City and surroundings (top) and
+rural County Mayo on the West coast of Ireland. Source: CRUSE tool,
+available at \url{https://cruse.bike/}.}
+
+\end{figure}
+
+\hypertarget{sec-trip-purposes}{%
+\subsection{Additional trip purposes}\label{sec-trip-purposes}}
+
+bla bla
+
+\hypertarget{sec-ui}{%
+\subsection{User interface}\label{sec-ui}}
+
+\hypertarget{sec-results}{%
+\section{Results}\label{sec-results}}
 
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\hypertarget{sec-discussion}{%
+\section{Discussion}\label{sec-discussion}}
 
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
+\hypertarget{sec-conclusions}{%
+\section{Conclusions}\label{sec-conclusions}}
 
@@
 \section{List of abbreviations}\label{list-of-abbreviations}}
+
+CSO: Central Statistics Office
+
+CRUSE: Cycle Route Uptake and Scenario Estimation
+
+ED: Electoral Division
+
+GTFS: General Transit Feed Specification
+
+KSI: Killed and Seriously Injured
+
+NRA: National Road Authority
+
+NRN: National Road Network
+
+OD: Origin-Destination, typically referring to origin-destination data
+which contains information on the number of people travelling between
+each pair of zones
+
+OSM: Open Street Map
+
+PCT: Propensity to Cycle Tool
+
+POWSCAR: Place of Work, School or College Census of Anonymised Records
+
+TII: Transport Infrastructure Ireland
 
@@
 
+Data was obtained from the Central Statistics Office (CSO) and Transport
+Infrastructure Ireland (TII) under license and cannot be shared
+publicly. The code used to generate the results presented in this paper
+is available at https://github.com/cruse-bike.
+
 \hypertarget{funding}{%
@@
 \addcontentsline{toc}{subsection}{Funding}
+
+The work was funded by Transport Infrastructure Ireland (TII).
 
@@
 
+Thanks to Paul MacDonald and Donal Hodgins (Kildare County Council), for
+testing early versions of the tool and for their input into the project.
+Thanks to Ciaran Maguire, Catherine Swift, and others at AECOM for their
+input into the project.
+
 \hypertarget{competing-interests}{%
@@
 \addcontentsline{toc}{subsection}{Competing interests}
+
+The authors declare that they have no competing interests.
 
@@
 
+RL led the development of the CRUSE tool, with input from JT, EV, HM,
+EB, PW, GOT, DB, and SM. SM instigated the project and coordinated its
+policy impact. PW, EB and GOT provided project management and transport
+engineering expertise.
+
   \bibliography{bibliography.bib}

### 2023-09-20 paper.qmd
Back to html version

@@
 
-# Uncomment for formatted version
-format:
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: Journal Name
-      formatting: preprint
-      model: 3p
-      cite-style: super
+# # Uncomment for formatted version
+# format:
+#   elsevier-pdf:
+#     keep-tex: true
+#     journal:
+#       name: Journal Name
+#       formatting: preprint
+#       model: 3p
+#       cite-style: super
 

### 2023-09-20 paper.qmd
Update names and affiliations

@@
 author:
-  - name: Dr. Robin Lovelace
+  - name: Robin Lovelace
     email: r.lovelace@leeds.ac.uk
@@
         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Dr. Joey Talbot
+  - name: Joey Talbot
     affiliations:
@@
       - name: AECOM
-        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY, UK
+        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
   - name: Dan Brennan
@@
         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
-  - name: Dr. Suzanne Meade
+  - name: Suzanne Meade
     affiliations:

### 2023-09-20 paper.qmd
Add trip purposes figure and text
@joeytalbot can you proof read and contribute to this bit please?

@@
 
-bla bla
+A limitation of the original PCT was that it only included travel to work data.
+This was partially addressed by the inclusion of travel to school based on data from the Department for Education in England [@goodman2019].
+An advantage of the POWSCAR OD data over OD datasets derived from census surveys in many countries is that it includes travel to school data.
+We commissioned a version of POWSCAR that included a breakdown of the total flows between OD pairs by purpose and mode, enabling a more realistic estimation of the 'Baseline' cycling network.
 
+However, travel to school and work only constitute a small proportion of all trips.
+To overcome this issue, we developed a spatial interaction modelling methodology to estimate the number of trips between each OD pair for additional trip purposes.
+
+The classification of trip purposes used in the CRUSE tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
+An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes.
+
+::: {#fig-trip-purposes}
+![](images/paste-4.png)
+
+Trip purposes used in the CRUSE tool.
+:::
+
+```{=html}
+<!-- 
+The POWSCAR-based categories are:
+
+- work (commute)
+- primary education
+- secondary education
+- tertiary education
+
+The NHTS-based categories are:
+
+- social
+- shopping
+- personal/other
+
+And the recreational trips were classified as follows:
+
+- Domestic with cycle trips starting at home
+- Domestic trips involving travel by other mode to the start of the cycle trip
+- Cycle tourism, where start and end trips are different
+- Sport cycling 
+-->
+```
 ## User interface {#sec-ui}

### 2023-09-21 paper.qmd
Add content from GOT

@@
 
-# # Uncomment for draft versions:
-# format: 
-#   html: default
-#   docx: default
+# Uncomment for draft versions:
+format: 
+  html: default
+  docx: default
 
@@
 
-<!-- \[Gary, Suzanne, others input here please.\] -->
+A proactive approach to the development of cycling policy and
+infrastructure is taking place in Ireland.
+The main framework
+underpinning these efforts is the Climate Action Plan[^2] and The National
+Development Plan[^3]
+. Strategic alignment is also outlined by the
+Phoenix Park Transport and Mobility Options Study, \"83% of respondents
+supported enhanced walking and cycling facilities\"
+. The recently
+published Greater Dublin Area Transport Strategy builds on this,
+suggesting that \"nearly a quarter of adults cycle at least once a week
+in the Dublin Metropolitan area\" and that \"cycling in the Dublin area
+takes up to 60,000 cars off the road today\"
+. (Q for RL: This could be
+extrapolated out across other counties in method section if needed?!)
+.
+Clear evidence of the importance of cycling is provided by the National
+Strategic Objective (NSO) from the National Development Plan as \"under
+the NSO, sustainable mobility, the Government has committed to investing
+€8.6bn in public transport between 2018 and 2027
+. This investment will
+deliver a wide range of improved infrastructure across both public
+transport and active travel, including
+
+[^2]: <https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/>
+
+[^3]: <https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/>
+
+BusConnects (including
+improved cycling infrastructure)\".
+This highlights that cycling
+infrastructure improvements are to take place concurrently with the
+BusConnects project, an entire redesign of the bus network in Dublin and
+Cork.
+To support this modal shift, TII commissioned AECOM to develop
+the CRUSE tool to use as an instrument to understand different cycling
+scenarios to de-risk investment in mobility and active travel
+infrastructure and promote the benefits of a modal shift towards
+cycling.
+A key consideration for TII was to ensure both urban and rural
+Ireland were targeted for cycling infrastructure in the future.
+To this
+end, TII extensively reviewed activity on Irish Roads and how the road
+network was being used by commuters, cyclists, heavy goods vehicles and
+other road users across the Irish road network.
+These efforts to analyse
+the wider picture of the Irish road network and to identify the
+greatest benefits from investment in cycling infrastructure within the
+national, regional and local frameworks all provided further
+justification to build, design and maintain the CRUSE tool.
+
+The
+ability to validate cycling trips, prioritise locations for development
+of cycling infrastructure and estimate demand of various cycling
+scenarios in Ireland was prioritized by TII given the commitment to
+invest in cycling infrastructure by the Irish government.
+Furthermore,
+TII and Local Authorities would use data generated from the CRUSE tool
+to monitor cycling friendliness, comfort and safety in Ireland to ensure
+strategic alignment with national, regional and local policies.
+
+When
+developing the CRUSE tool, TII, the Institute of Transport Studies
+(ITS) Leeds and AECOM to develop the CRUSE Tool to focus on Ireland
+specifically(https://cruse.bike/).
+ITS Leeds were part of the team that
+developed the PCT for England and Wales, with the aim of providing \"an
+evidence base to inform cycling investment\".
+This experience ensured
+that ITS Leeds had extensive experience in developing cutting edge
+cycling tools.
+The project recognized the importance of an online tool,
+using a recognised open-source methodology that will be discussed in
+more detail in this paper.
+The decision to ensure the tool is opensource
+that is publicly available ensures transport planners and members of
+the public can utilize the tool for scenario planning to meet their
+requirements.
+Other use cases in the Irish context includes monitoring
+the need for existing cycling infrastructure upgrades, undertaking
+appraisal, as well as safety and exposure information in line with the
+European Cycling Federation, Dutch Cycling Embassy recommendations and
+the European RISM Directive, for reporting collision rates of vulnerable
+road users by 2024.
 
@@
 An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
-The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^2] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^3]
+The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^4] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^5]
 .
 
-[^2]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
+[^4]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
 
-[^3]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
+[^5]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
 
@@
 The resulting evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
-biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^4] and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
+biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^6] and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
 
-[^4]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
+[^6]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
 

### 2023-09-21 paper.qmd
Merge GOT input

--- 2023-09-20 paper.qmd
@@
 
-<!-- \[Gary, Suzanne, others input here please.\] -->
+A proactive approach to the development of cycling policy and
+infrastructure is taking place in Ireland.
+The main framework
+underpinning these efforts is the Climate Action Plan[^2] and The National
+Development Plan[^3]
+. Strategic alignment is also outlined by the
+Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: \"83% of respondents supported enhanced walking and cycling facilities\"
+.
+
+[^2]: <https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/>
+
+[^3]: <https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/>
+
+At the regional level, the recently published Greater Dublin Area Transport Strategy[^4] reinforces these findings: \"nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area\" with cycling in the Dublin area taking up to 60,000 cars off the road today.
+Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, this suggests that around 150,000 cars could be removed nationwide just by achieving Dubline levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
+
+[^4]: <https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/>
+
+Further evidence of the importance of cycling in Ireland *already* is provided by the National Strategic Objective (NSO) from the National Development Plan, with €8.6 billion allocated to sustainable transport infrastructure including public transport and active travel interventions.
+Cycle infrastructure will be developed in synchrony with the BusConnects project, an entire redesign of the bus network in Dublin and Cork.
 
 It was in this context that Transport Infrastructure Ireland (TII) commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) project.
-Based on the Propensity to Cycle Tool (PCT) for England and Wales [@lovelace2017], the CRUSE tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
+Based on the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
+A key consideration for TII is to ensure both urban and rural Ireland were targeted for cycling infrastructure.
+To this end, TII extensively reviewed activity on Irish Roads and how the road network was being used by commuters, cyclists, heavy goods vehicles and other road users across the Irish transport network.
+The aim was for the tool to provide strong, national, systematic but locally-specific evidence to monitor cycling friendliness and safety in Ireland, to ensure strategic alignment with national, regional and local policies.
+The open source and publicly available nature of the tool ensures more inclusive and evidence-based conversations around cycle network planning between all stakeholders in the transport planning process.
+Other use cases in the Irish context includes monitoring the need for existing cycling infrastructure upgrades, undertaking appraisal, as well as safety and exposure information in line with the European Cycling Federation, Dutch Cycling Embassy recommendations and the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
+These factors mean that the tool can be seen as an open access 'leverage point' in the planning system [@lovelace2020].
 
@@
 An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
-The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^2] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^3]
+The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^5] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^6]
 .
 
-[^2]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
+[^5]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
 
-[^3]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
+[^6]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
 
@@
 The resulting evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
-biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^4] and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
+biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^7] and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
 
-[^4]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
+[^7]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
 

### 2023-09-21 paper.qmd
Typo fix

@@
 At the regional level, the recently published Greater Dublin Area Transport Strategy[^4] reinforces these findings: \"nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area\" with cycling in the Dublin area taking up to 60,000 cars off the road today.
-Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, this suggests that around 150,000 cars could be removed nationwide just by achieving Dubline levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
+Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, this suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
 

### 2023-09-22 paper.qmd
Add shiny citation

@@
 
-# Uncomment for draft versions:
+# # Uncomment for draft versions:
+# format: 
+#   html: default
+#   docx: default
+
+# Uncomment for formatted version
 format:
-  html: default
-  docx: default
-
-# # Uncomment for formatted version
-# format:
-#   elsevier-pdf:
-#     keep-tex: true
-#     journal:
-#       name: Journal Name
-#       formatting: preprint
-#       model: 3p
-#       cite-style: super
+  elsevier-pdf:
+    keep-tex: true
+    journal:
+      name: Journal Name
+      formatting: preprint
+      model: 3p
+      cite-style: super
 
@@
 
-A proactive approach to the development of cycling policy and
-infrastructure is taking place in Ireland.
-The main framework
-underpinning these efforts is the Climate Action Plan[^2] and The National
-Development Plan[^3]
-. Strategic alignment is also outlined by the
-Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: \"83% of respondents supported enhanced walking and cycling facilities\"
+A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
+The main framework underpinning these efforts is the Climate Action Plan[^2] and The National Development Plan[^3]
+.
+Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: "83% of respondents supported enhanced walking and cycling facilities"
 .
@@
 
-At the regional level, the recently published Greater Dublin Area Transport Strategy[^4] reinforces these findings: \"nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area\" with cycling in the Dublin area taking up to 60,000 cars off the road today.
+At the regional level, the recently published Greater Dublin Area Transport Strategy[^4] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
 Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, this suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
@@
 Based on the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
-A key consideration for TII is to ensure both urban and rural Ireland were targeted for cycling infrastructure.
+A key consideration for TII is to ensure both urban and rural Ireland were targeted for cycling infrastructure, as shown in @fig-dublin.
 To this end, TII extensively reviewed activity on Irish Roads and how the road network was being used by commuters, cyclists, heavy goods vehicles and other road users across the Irish transport network.
@@
 These factors mean that the tool can be seen as an open access 'leverage point' in the planning system [@lovelace2020].
+
+::: {#fig-dublin}
+![](images/paste-1.png)
+
+![](images/paste-3.png)
+
+Illustration of the density of route networks generated by the CRUSE tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland.
+Source: CRUSE tool, available at <https://cruse.bike/>.
+:::
 
@@
 
-::: {#fig-dublin}
-![](images/paste-1.png)
-
-![](images/paste-3.png)
-
-Illustration of the density of route networks generated by the CRUSE tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland.
-Source: CRUSE tool, available at <https://cruse.bike/>.
-:::
-
 ## Additional trip purposes {#sec-trip-purposes}
@@
 To overcome this issue, we developed a spatial interaction modelling methodology to estimate the number of trips between each OD pair for additional trip purposes.
-
 The classification of trip purposes used in the CRUSE tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
@@
 ## User interface {#sec-ui}
+
+The CRUSE web application is statically hosted, meaning that it can be accessed without the need for a server running software such as the R package 'shiny' or Python packages such as 'streamlit' or 'flask' in the background [@wickham2021].
 

### 2023-09-22 paper.tex
Add shiny citation

@@
   pdftitle={CRUSE to Safe Cycling in Ireland},
-  pdfauthor={Dr.~Robin Lovelace; Dr.~Joey Talbot; Eugeni Vidal Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Tool; Dan Brennan; Dr.~Suzanne Meade},
+  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Tool; Dan Brennan; Suzanne Meade},
   pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
@@
 Methodology to Support Active Travel} }
-\author[1]{Dr.~Robin Lovelace%
+\author[1]{Robin Lovelace%
 \corref{cor1}%
@@
  \ead{r.lovelace@leeds.ac.uk} 
-\author[1]{Dr.~Joey Talbot%
+\author[1]{Joey Talbot%
 %
@@
 
-\author[3]{Gary O'Tool%
+\author[2]{Gary O'Tool%
 %
@@
 
-\author[4]{Dr.~Suzanne Meade%
+\author[4]{Suzanne Meade%
 %
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, frame hidden, sharp corners, enhanced]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 
+A proactive approach to the development of cycling policy and
+infrastructure is taking place in Ireland. The main framework
+underpinning these efforts is the Climate Action Plan\footnote{\url{https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/}}
+and The National Development Plan\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}}
+. Strategic alignment is also outlined by the Phoenix Park Transport and
+Mobility Options Study, which highlights public support for active
+travel interventions: ``83\% of respondents supported enhanced walking
+and cycling facilities'' .
+
+At the regional level, the recently published Greater Dublin Area
+Transport Strategy\footnote{\url{https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}}
+reinforces these findings: ``nearly a quarter of adults cycle at least
+once a week in the Dublin Metropolitan Area'' with cycling in the Dublin
+area taking up to 60,000 cars off the road today. Extrapolating this on
+a per population basis across Ireland, with around 40\% of the
+population living in Dublin, this suggests that around 150,000 cars
+could be removed nationwide just by achieving Dublin levels of cycling
+in all counties (notwithstanding existing cycling trips and differences
+in trip distances).
+
+Further evidence of the importance of cycling in Ireland \emph{already}
+is provided by the National Strategic Objective (NSO) from the National
+Development Plan, with €8.6 billion allocated to sustainable transport
+infrastructure including public transport and active travel
+interventions. Cycle infrastructure will be developed in synchrony with
+the BusConnects project, an entire redesign of the bus network in Dublin
+and Cork.
+
 It was in this context that Transport Infrastructure Ireland (TII)
@@
 project. Based on the Propensity to Cycle Tool (PCT) for England and
-Wales \citep{lovelace2017}, the CRUSE tool was developed to provide
-evidence on current cycling levels and future cycling potential
-nationwide across Ireland.
+Wales, the CRUSE tool was developed to provide evidence on current
+cycling levels and future cycling potential nationwide across Ireland. A
+key consideration for TII is to ensure both urban and rural Ireland were
+targeted for cycling infrastructure, as shown in
+Figure~\ref{fig-dublin}. To this end, TII extensively reviewed activity
+on Irish Roads and how the road network was being used by commuters,
+cyclists, heavy goods vehicles and other road users across the Irish
+transport network. The aim was for the tool to provide strong, national,
+systematic but locally-specific evidence to monitor cycling friendliness
+and safety in Ireland, to ensure strategic alignment with national,
+regional and local policies. The open source and publicly available
+nature of the tool ensures more inclusive and evidence-based
+conversations around cycle network planning between all stakeholders in
+the transport planning process. Other use cases in the Irish context
+includes monitoring the need for existing cycling infrastructure
+upgrades, undertaking appraisal, as well as safety and exposure
+information in line with the European Cycling Federation, Dutch Cycling
+Embassy recommendations and the European RISM Directive, for reporting
+collision rates of vulnerable road users by 2024. These factors mean
+that the tool can be seen as an open access `leverage point' in the
+planning system \citep{lovelace2020}.
+
+\begin{figure}
+
+{\centering
+
+\includegraphics{images/paste-1.png}
+
+\includegraphics{images/paste-3.png}
+
+}
+
+\caption{\label{fig-dublin}Illustration of the density of route networks
+generated by the CRUSE tool for Dublin City and surroundings (top) and
+rural County Mayo on the West coast of Ireland. Source: CRUSE tool,
+available at \url{https://cruse.bike/}.}
+
+\end{figure}
 
@@
   \href{https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B}{results2021.ref.ac.uk}
-  submitted by the University of Westminster.} .
+  submitted by the University of Westminster.}.
 
@@
 
+\hypertarget{sec-trip-purposes}{%
+\subsection{Additional trip purposes}\label{sec-trip-purposes}}
+
+A limitation of the original PCT was that it only included travel to
+work data. This was partially addressed by the inclusion of travel to
+school based on data from the Department for Education in England
+\citep{goodman2019}. An advantage of the POWSCAR OD data over OD
+datasets derived from census surveys in many countries is that it
+includes travel to school data. We commissioned a version of POWSCAR
+that included a breakdown of the total flows between OD pairs by purpose
+and mode, enabling a more realistic estimation of the `Baseline' cycling
+network.
+
+However, travel to school and work only constitute a small proportion of
+all trips. To overcome this issue, we developed a spatial interaction
+modelling methodology to estimate the number of trips between each OD
+pair for additional trip purposes.
+
+The classification of trip purposes used in the CRUSE tool was guided
+mainly by the trip purpose classification found within the National
+Household Travel Survey (NHTS), but with the addition of categories
+based on the comprehensive POWSCAR data, and the need to include
+recreational trips and multi-stage trips (not yet implemented). An
+overview of the trip purposes used in CRUSE is presented in
+Figure~\ref{fig-trip-purposes}.
+
 \begin{figure}
@@
 
-\includegraphics{images/paste-1.png}
-
-\includegraphics{images/paste-3.png}
+\includegraphics{images/paste-4.png}
 
@@
 
-\caption{\label{fig-dublin}Illustration of the density of route networks
-generated by the CRUSE tool for Dublin City and surroundings (top) and
-rural County Mayo on the West coast of Ireland. Source: CRUSE tool,
-available at \url{https://cruse.bike/}.}
+\caption{\label{fig-trip-purposes}Trip purposes used in the CRUSE tool.}
 
 \end{figure}
-
-\hypertarget{sec-trip-purposes}{%
-\subsection{Additional trip purposes}\label{sec-trip-purposes}}
-
-bla bla
 

### 2023-09-22 paper.qmd
Finish UI section

@@
 The CRUSE web application is statically hosted, meaning that it can be accessed without the need for a server running software such as the R package 'shiny' or Python packages such as 'streamlit' or 'flask' in the background [@wickham2021].
+Typical intended user stories are illustrated in @fig-user-stories, which shows that the tool is designed to be used by both professional and non-professional users.
+For the main target audience, professional transport planners working at the county level, the tool provides a range of outputs, including estimates of cycling potential at the county and network level.
+The provision of balanced (the default), quietest and fastest route networks enables planners to consider the trade-offs between directness and 'cycle friendliness' when planning new infrastructure.
+The tool also provides data downloads, enabling estimates of cycling potential on the network to be visualised and analysed with other tools such as QGIS, Python or R.
+
+For non-professional users, the tool provides a simple interface to explore the cycling potential of the network.
+By providing a single landing page that is suitable for both professional and non-professional users, the tool aims to facilitate communication between these groups.
+
+```{mermaid}
+%%| fig-width: 100%
+%%| fig-cap: User stories for the CRUSE tool
+%%| label: fig-user-stories
+graph LR
+    A(Professional user) --> B[CRUSE landing page]
+    B --> C[CRUSE national map]
+    C --> D[CRUSE county map]
+    D --> F[Full county map]
+    D --> CS[County statistics]
+    D --> Q[Quiet and fast route networks]
+    D --> DD[Data downloads]
+    DD --> EX{External tools}
+    X(Non-professional user) --> Y[CRUSE landing page]
+    Y --> Z[CRUSE national map]
+    Z --> R[CRUSE county level results]
+    Z --> E(Exploration)
+    R --> E
+    E --> COMS{Communication}
+```
 

### 2023-09-22 paper.tex
Finish UI section

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
   \href{https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B}{results2021.ref.ac.uk}
-  submitted by the University of Westminster.}.
+  submitted by the University of Westminster.} .
 
@@
 modelling methodology to estimate the number of trips between each OD
-pair for additional trip purposes.
-
-The classification of trip purposes used in the CRUSE tool was guided
-mainly by the trip purpose classification found within the National
-Household Travel Survey (NHTS), but with the addition of categories
-based on the comprehensive POWSCAR data, and the need to include
-recreational trips and multi-stage trips (not yet implemented). An
-overview of the trip purposes used in CRUSE is presented in
-Figure~\ref{fig-trip-purposes}.
+pair for additional trip purposes. The classification of trip purposes
+used in the CRUSE tool was guided mainly by the trip purpose
+classification found within the National Household Travel Survey (NHTS),
+but with the addition of categories based on the comprehensive POWSCAR
+data, and the need to include recreational trips and multi-stage trips
+(not yet implemented). An overview of the trip purposes used in CRUSE is
+presented in Figure~\ref{fig-trip-purposes}.
 
@@
 \subsection{User interface}\label{sec-ui}}
+
+The CRUSE web application is statically hosted, meaning that it can be
+accessed without the need for a server running software such as the R
+package `shiny' or Python packages such as `streamlit' or `flask' in the
+background \citep{wickham2021}. Typical intended user stories are
+illustrated in Figure~\ref{fig-user-stories}, which shows that the tool
+is designed to be used by both professional and non-professional users.
+For the main target audience, professional transport planners working at
+the county level, the tool provides a range of outputs, including
+estimates of cycling potential at the county and network level. The
+provision of balanced (the default), quietest and fastest route networks
+enables planners to consider the trade-offs between directness and
+`cycle friendliness' when planning new infrastructure. The tool also
+provides data downloads, enabling estimates of cycling potential on the
+network to be visualised and analysed with other tools such as QGIS,
+Python or R.
+
+For non-professional users, the tool provides a simple interface to
+explore the cycling potential of the network. By providing a single
+landing page that is suitable for both professional and non-professional
+users, the tool aims to facilitate communication between these groups.
+
+\begin{figure}
+
+{\centering
+
+\begin{figure}[H]
+
+{\centering \includegraphics{paper_files/figure-latex/mermaid-figure-1.png}
+
+}
+
+\end{figure}
+
+}
+
+\caption{\label{fig-user-stories}User stories for the CRUSE tool}
+
+\end{figure}
 

### 2023-09-22 paper.qmd
Update formats

@@
 
-# # Uncomment for draft versions:
+# Uncomment for draft versions:
+format: 
+  html: default
+  docx: default
+  pdf: default
+
+# # Uncomment for formatted version
 # format:
-#   html: default
-#   docx: default
-
-# Uncomment for formatted version
-format:
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: Journal Name
-      formatting: preprint
-      model: 3p
-      cite-style: super
+#   elsevier-pdf:
+#     keep-tex: true
+#     journal:
+#       name: Journal Name
+#       formatting: preprint
+#       model: 3p
+#       cite-style: super
 

### 2023-09-25 paper.qmd
Update formats, not latex
Update description of user stories figure

@@
   html: default
-  docx: default
-  pdf: default
+  # docx: default
+  # pdf: default
 
@@
 For non-professional users, the tool provides a simple interface to explore the cycling potential of the network.
-By providing a single landing page that is suitable for both professional and non-professional users, the tool aims to facilitate communication between these groups.
+By providing a single landing page that is suitable for both professional and non-professional users (such as an advocate or parent interested in safe routes to school), the tool aims to facilitate communication between these groups.
 
@@
 %%| fig-width: 100%
-%%| fig-cap: User stories for the CRUSE tool
+%%| fig-cap: Hypothetical user stories for the CRUSE tool illustrating that it can be used in different ways by different user groups.
 %%| label: fig-user-stories
@@
 
+The main result of the CRUSE project is a new and open access web application and dataset on current and potential future cycling levels in Ireland, with evidence provided at county and route segment levels.
+Like the PCT on which CRUSE builds, a key feature of the results is that they are systematic and available for all counties, and most routes on which cycling is permitted, in Ireland.
+
+## National results {#sec-national-results}
+
+The CRUSE tool provides estimates of cycling potential at the national level, as shown in @fig-national-results.
+
 National planning
@@
 Local
+
+## Local results
+
+To illustrate the results in urban areas, we took the top 4 cities in Ireland by population: Dublin, Cork, Limerick and Galway.
+The results for these cities are presented in @fig-city-results.
+
+```{r}
+# Results for one city
+
+```
 

### 2023-09-25 paper.qmd
Update with new results figure

@@
   html: default
-  # docx: default
-  # pdf: default
+#   # docx: default
+#   # pdf: default
 
@@
 {{< pagebreak >}}
+
+```{r}
+#| include: false
+library(tidyverse)
+library(tmap)
+```
 
@@
 ```{r}
-# Results for one city
+#| eval: false
+#| echo: false
+# Results for one city are shown below
+city_name = "Dublin"
+city_buffer = zonebuilder::zb_zone(city_name)
+city_buffer_5km = city_buffer |>
+  filter(circle_id <= 2)
+# qtm(city_buffer_5km)
+city_buffer_union = sf::st_union(city_buffer_5km)
+rnet_fastest = sf::read_sf("./../testout/dublin/rnet_combined_fastest.gpkg")
+rnet_fastest_city = sf::st_intersection(rnet_fastest, city_buffer_union)#
+# nrow(rnet_fastest_city)
+# table(rnet_fastest_city$`Bicycle (Go Dutch)`)
+rnet_fastest_city_top = rnet_fastest_city |>
+  filter(`Bicycle (Go Dutch)` != "1-10") |>
+  filter(`Bicycle (Go Dutch)` != "0") |>
+  mutate(across(`Bicycle (Baseline)`:`Quietness`, as.numeric)) |>
+  slice_max(`Bicycle (Go Dutch)`, n = 10000)
+m = tm_shape(rnet_fastest_city_top) +
+  tm_lines(col = "Quietness", palette = "RdYlGn", lwd = 5)
+tmap_save(m, "./images/rnet_fastest_city_top.png", width = 10, height = 10)
+```
 
-```
+::: {#fig-city-results}
+
+![](images/rnet_fastest_city_top.png)
+
+Top 10,000 routes in Dublin by 'cycle friendliness' (quietness) based on the 'fastest' route network.
+
+:::
 

### 2023-09-25 paper.tex
Update with new results figure

@@
 \fi
+
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{241,243,245}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
+\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
 
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, enhanced, sharp corners, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 landing page that is suitable for both professional and non-professional
-users, the tool aims to facilitate communication between these groups.
+users (such as an advocate or parent interested in safe routes to
+school), the tool aims to facilitate communication between these groups.
 
@@
 
-\caption{\label{fig-user-stories}User stories for the CRUSE tool}
+\caption{\label{fig-user-stories}Hypothetical user stories for the CRUSE
+tool illustrating that it can be used in different ways by different
+user groups.}
 
@@
 
+The main result of the CRUSE project is a new and open access web
+application and dataset on current and potential future cycling levels
+in Ireland, with evidence provided at county and route segment levels.
+Like the PCT on which CRUSE builds, a key feature of the results is that
+they are systematic and available for all counties, and most routes on
+which cycling is permitted, in Ireland.
+
+\hypertarget{sec-national-results}{%
+\subsection{National results}\label{sec-national-results}}
+
+The CRUSE tool provides estimates of cycling potential at the national
+level, as shown in \textbf{?@fig-national-results}.
+
 National planning
@@
 Local
+
+\hypertarget{local-results}{%
+\subsection{Local results}\label{local-results}}
+
+To illustrate the results in urban areas, we took the top 4 cities in
+Ireland by population: Dublin, Cork, Limerick and Galway. The results
+for these cities are presented in Figure~\ref{fig-city-results}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# Results for one city are shown below}
+\NormalTok{city\_name }\OtherTok{=} \StringTok{"Dublin"}
+\NormalTok{city\_buffer }\OtherTok{=}\NormalTok{ zonebuilder}\SpecialCharTok{::}\FunctionTok{zb\_zone}\NormalTok{(city\_name)}
+\NormalTok{city\_buffer\_5km }\OtherTok{=}\NormalTok{ city\_buffer }\SpecialCharTok{|\textgreater{}}
+  \FunctionTok{filter}\NormalTok{(circle\_id }\SpecialCharTok{\textless{}=} \DecValTok{2}\NormalTok{)}
+\CommentTok{\# qtm(city\_buffer\_5km)}
+\NormalTok{city\_buffer\_union }\OtherTok{=}\NormalTok{ sf}\SpecialCharTok{::}\FunctionTok{st\_union}\NormalTok{(city\_buffer\_5km)}
+\NormalTok{rnet\_fastest }\OtherTok{=}\NormalTok{ sf}\SpecialCharTok{::}\FunctionTok{read\_sf}\NormalTok{(}\StringTok{"./../testout/dublin/rnet\_combined\_fastest.gpkg"}\NormalTok{)}
+\NormalTok{rnet\_fastest\_city }\OtherTok{=}\NormalTok{ sf}\SpecialCharTok{::}\FunctionTok{st\_intersection}\NormalTok{(rnet\_fastest, city\_buffer\_union)}\CommentTok{\#}
+\CommentTok{\# nrow(rnet\_fastest\_city)}
+\CommentTok{\# table(rnet\_fastest\_city$\textasciigrave{}Bicycle (Go Dutch)\textasciigrave{})}
+\NormalTok{rnet\_fastest\_city\_top }\OtherTok{=}\NormalTok{ rnet\_fastest\_city }\SpecialCharTok{|\textgreater{}}
+  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Bicycle (Go Dutch)}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \StringTok{"1{-}10"}\NormalTok{) }\SpecialCharTok{|\textgreater{}}
+  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Bicycle (Go Dutch)}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \StringTok{"0"}\NormalTok{) }\SpecialCharTok{|\textgreater{}}
+  \FunctionTok{mutate}\NormalTok{(}\FunctionTok{across}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Bicycle (Baseline)}\StringTok{\textasciigrave{}}\SpecialCharTok{:}\StringTok{\textasciigrave{}}\AttributeTok{Quietness}\StringTok{\textasciigrave{}}\NormalTok{, as.numeric)) }\SpecialCharTok{|\textgreater{}}
+  \FunctionTok{slice\_max}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Bicycle (Go Dutch)}\StringTok{\textasciigrave{}}\NormalTok{, }\AttributeTok{n =} \DecValTok{10000}\NormalTok{)}
+\NormalTok{m }\OtherTok{=} \FunctionTok{tm\_shape}\NormalTok{(rnet\_fastest\_city\_top) }\SpecialCharTok{+}
+  \FunctionTok{tm\_lines}\NormalTok{(}\AttributeTok{col =} \StringTok{"Quietness"}\NormalTok{, }\AttributeTok{palette =} \StringTok{"RdYlGn"}\NormalTok{, }\AttributeTok{lwd =} \DecValTok{5}\NormalTok{)}
+\FunctionTok{tmap\_save}\NormalTok{(m, }\StringTok{"./images/rnet\_fastest\_city\_top.png"}\NormalTok{, }\AttributeTok{width =} \DecValTok{10}\NormalTok{, }\AttributeTok{height =} \DecValTok{10}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+
+{\centering
+
+\includegraphics{images/rnet_fastest_city_top.png}
+
+}
+
+\caption{\label{fig-city-results}Top 10,000 routes in Dublin by `cycle
+friendliness' (quietness) based on the `fastest' route network.}
+
+\end{figure}
 

### 2023-09-25 paper.qmd
Add findings and figures

--- 2023-09-25 paper.qmd
@@
 {{< pagebreak >}}
+
+```{r}
+#| include: false
+library(tidyverse)
+remotes::install_dev("tmap")
+library(tmap)
+```
 
@@
 
-The main result of the CRUSE project is a new and open access web application and dataset on current and potential future cycling levels in Ireland, with evidence provided at county and route segment levels.
+The main result of the CRUSE project is an open access web application and datasets on current and potential future cycling levels in Ireland, with evidence provided at county and route segment levels.
 Like the PCT on which CRUSE builds, a key feature of the results is that they are systematic and available for all counties, and most routes on which cycling is permitted, in Ireland.
+The full set of results are therefore too extensive to present in their entirety in this paper.
+Through the interactive web application, users can explore the data to generate the results that are most relevant to their needs, whether that is finding the cycling potential on a particular road or finding 'weak links' in the cycle network associated with a particular school, work place or other destination.
 
-## National results {#sec-national-results}
+Instead of trying to show such use cases, of which there are many hundreds, we present a selection of results to illustrate the main features of the CRUSE tool.
+
+<!-- ## National results {#sec-national-results}
 
@@
 
-Local
+Local -->
 
-## Local results
+<!-- ## Local results -->
 
-To illustrate the results in urban areas, we took the top 4 cities in Ireland by population: Dublin, Cork, Limerick and Galway.
-The results for these cities are presented in @fig-city-results.
+To illustrate the results in urban areas, we took the top 4 cities in Ireland by population: Dublin City, Cork, Limerick and Galway (with city populations of 0.6, 0.2, 0.1 and 0.1 million respectively).
+The results for central areas demarcated with circular areas with a radius of 3km for these cities are presented in @fig-city-results.
 
 ```{r}
-# Results for one city
+#| eval: false
+#| echo: false
+# Results for one city are shown below
+cities = c("dublin", "cork", "limerick", "galway")
+city_name = "dublin"
+city_buffer = zonebuilder::zb_zone(city_name)
+city_buffer_5km = city_buffer |>
+  filter(circle_id <= 2)
+# qtm(city_buffer_5km)
+city_buffer_union = sf::st_union(city_buffer_5km)
+rnet_fastest = sf::read_sf(glue::glue("./../testout/{city_name}/rnet_combined_fastest.gpkg"))
+rnet_fastest_city = sf::st_intersection(rnet_fastest, city_buffer_union)#
+# nrow(rnet_fastest_city)
+# table(rnet_fastest_city$`Bicycle (Go Dutch)`)
+rnet_fastest_city_top = rnet_fastest_city |>
+  filter(`Bicycle (Go Dutch)` != "1-10") |>
+  filter(`Bicycle (Go Dutch)` != "0") |>
+  mutate(across(`Bicycle (Baseline)`:`Quietness`, as.numeric)) |>
+  slice_max(`Bicycle (Go Dutch)`, n = 10000)
+m = tm_shape(rnet_fastest_city_top) +
+  tm_lines(col = "Quietness", palette = "RdYlGn", lwd = 5)
+tmap_save(m, "./images/rnet_fastest_city_top.png", width = 10, height = 10)
+# Generalisation to balanced and quiestest route types:
+route_types = c("fastest", "balanced", "quietest")
+dublin_rnets = purrr::map(route_types, function(route_type) {
+  rnet = sf::read_sf(glue::glue("./../testout/{city_name}/rnet_combined_{route_type}.gpkg"))
+  rnet_city = sf::st_intersection(rnet, city_buffer_union)
+  rnet_city_top = rnet_city |>
+    filter(`Bicycle (Go Dutch)` != "1-10") |>
+    filter(`Bicycle (Go Dutch)` != "0") |>
+    mutate(across(`Bicycle (Baseline)`:`Quietness`, as.numeric)) |>
+    slice_max(`Bicycle (Go Dutch)`, n = 1000) |>
+    mutate(`Route type` = route_type) |>
+    rename(geometry = geom)
+  rnet_city_top
+})
+dublin_rnets = do.call(rbind, dublin_rnets)
+dublin_rnets$`Route type` = factor(dublin_rnets$`Route type`, levels = route_types)
+m_types_dublin = tm_shape(dublin_rnets) +
+  tm_lines(col = "col", lwd = "Bicycle (Go Dutch)", scale = 9) +
+  tm_legend_hide() +
+  tm_facets(by = "Route type", nrow = 1)
 
+cities_rnet_types = purrr::map(cities, function(city_name) {
+  city_buffer = zonebuilder::zb_zone(city_name)
+  city_buffer_5km = city_buffer |>
+    filter(circle_id <= 2)
+  city_buffer_union = sf::st_union(city_buffer_5km)
+  route_types = c("fastest", "balanced", "quietest")
+  rnet_types = purrr::map(route_types, function(route_type) {
+    rnet = sf::read_sf(glue::glue("./../testout/{city_name}/rnet_combined_{route_type}.gpkg"))
+    rnet_city = sf::st_intersection(rnet, city_buffer_union)
+    rnet_city_top = rnet_city |>
+      filter(`Bicycle (Go Dutch)` != "1-10") |>
+      filter(`Bicycle (Go Dutch)` != "0") |>
+      mutate(across(`Bicycle (Baseline)`:`Quietness`, as.numeric)) |>
+      # slice_max(`Bicycle (Go Dutch)`, n = 1000) |>
+      # Title Case city name:
+      mutate(`Route type` = str_to_title(route_type), `City` = str_to_title(city_name))
+    rnet_city_top
+  })
+  rnet_types = do.call(rbind, rnet_types)
+  rnet_types$`Route type` = factor(rnet_types$`Route type`, levels = str_to_title(route_types))
+  rnet_types$`City` = factor(rnet_types$`City`, levels = str_to_title(cities))
+  rnet_types
+})
+cities_rnet_types = do.call(rbind, cities_rnet_types)
+summary(cities_rnet_types$`Bicycle (Go Dutch)`)
+# Set max to 1000:
+cities_rnet_types$`Bicycle (Go Dutch)` = pmin(cities_rnet_types$`Bicycle (Go Dutch)`, 1000)
+m_types = tm_shape(cities_rnet_types) +
+  tm_lines(col = "col", lwd = "Bicycle (Go Dutch)", scale = 20) +
+  tm_legend_hide() +
+  tm_facets_grid(rows = "City", columns = "Route type")
+m_types
+tmap_save(m_types, "./images/rnet_types.png", width = 12, height = 10)
+browseURL("./images/rnet_types.png")
 ```
 
+::: {#fig-city-results}
+
+![](images/rnet_types.png)
+
+Fastest, balanced and quietest cycle networks within 3km circular areas containing the four largest cities in Ireland.
+The networks are coloured by cycle friendliness and with widths varying in proportion to the potential level of cycling under the 'Go Dutch' scenario.
+
+:::
+
+Aside from the widely varying sizes and route networks of each city, with Dublin having by far the densest network, it is clear from the visualisation of cycling friendliness that there is a lot of room for improvement in all cities.
+Dublin has had received most investment in cycling infrastructure and this is apparent from the high proportion of the route network that is green, even in the 'Fastest' route network representing routes that prioritise speed and directness over comfort and safety.
+Dublin also has the highest mode share of cycling of the four cities, adding to the evidence base showing that investment in cycling infrastructure is effective in increasing cycling uptake.
+
+Although Dublin City has made progress, there are still many parts of the Fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
+According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated
+from traffic and pedestrians".^[
+https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22_DublinMetropolitanArea_v35_DIGITAL_v2.pdf
+]
+The results for the Dublin area could help prioritise investment in cycling infrastructure in the city.
+
+The results for the other
+
 # Discussion {#sec-discussion}
+
+The CRUSE tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
+This provision of multiple scenarios of behaviour changes *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE tool.
+
+As highlighted in a paper on cycling infrastructure preferences based on a case study of Dublin, both directness and quietness are important :
+
+> Direct routes with short journey times were found to be the most important positive variable for existing cyclists and non-cyclists in determining route choice.
+This is followed by infrastructure type, the number of junctions along the route, traffic speed and cyclist volumes.
+In terms if infrastructure, regardless of the level of cycling confidence, routes which have ‘no facilities’ or ‘bus/cycle lanes’ are the least favoured cycle route types.
+
+The CRUSE tool can help both in terms of describing the current situation, and also prioritise investment in those routes lacking infrastructure highlighted by  as being least favoured by cyclists.
 

### 2023-09-25 paper.qmd
City Caulfield et al 2012

@@
 
+```{=html}
 <!-- ## National results {#sec-national-results}
@@
 Local -->
-
+```
 <!-- ## Local results -->
@@
 ::: {#fig-city-results}
-
 ![](images/rnet_types.png)
@@
 The networks are coloured by cycle friendliness and with widths varying in proportion to the potential level of cycling under the 'Go Dutch' scenario.
-
 :::
@@
 Although Dublin City has made progress, there are still many parts of the Fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
-According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated
-from traffic and pedestrians".^[
-https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22_DublinMetropolitanArea_v35_DIGITAL_v2.pdf
-]
+According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians".[^8]
 The results for the Dublin area could help prioritise investment in cycling infrastructure in the city.
+
+[^8]:  https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22_DublinMetropolitanArea_v35_DIGITAL_v2.pdf
 
@@
 
-As highlighted in a paper on cycling infrastructure preferences based on a case study of Dublin, both directness and quietness are important :
+As highlighted in a paper on cycling infrastructure preferences based on a case study of Dublin, both directness and quietness are important [@caulfield2012]:
 
 > Direct routes with short journey times were found to be the most important positive variable for existing cyclists and non-cyclists in determining route choice.
-This is followed by infrastructure type, the number of junctions along the route, traffic speed and cyclist volumes.
-In terms if infrastructure, regardless of the level of cycling confidence, routes which have ‘no facilities’ or ‘bus/cycle lanes’ are the least favoured cycle route types.
+> This is followed by infrastructure type, the number of junctions along the route, traffic speed and cyclist volumes.
+> In terms if infrastructure, regardless of the level of cycling confidence, routes which have 'no facilities' or 'bus/cycle lanes' are the least favoured cycle route types.
 
-The CRUSE tool can help both in terms of describing the current situation, and also prioritise investment in those routes lacking infrastructure highlighted by  as being least favoured by cyclists.
+The CRUSE tool can help both in terms of describing the current situation, and also prioritise investment in those routes lacking infrastructure highlighted by @caulfield2012 as being least favoured by cyclists.
 

### 2023-09-25 paper.qmd
Minor rewording

@@
 
-The results for the other
+<!-- The results for the other -->
 
@@
 
-The CRUSE tool can help both in terms of describing the current situation, and also prioritise investment in those routes lacking infrastructure highlighted by @caulfield2012 as being least favoured by cyclists.
+The CRUSE tool can help both in terms of describing the current situation, and also prioritise investment in those direct routes with high cycling potential that lack adequate cycling infrastructure, highlighted by @caulfield2012.
 

### 2023-09-25 paper.qmd
Fix up footnotes

@@
 The rapid uptake of highly efficient e-bikes can also be seen as an improvement on most electric vehicles, which are too heavy and expensive to be a sustainable alternative and could in fact delay the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
-At the European level, the European Union has a target of reducing greenhouse gas emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.\[\^1\]
+At the European level, the European Union has a target of reducing greenhouse gas emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.[^1]
 
@@
 
-In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030.\[\^3\] Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
+In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030.[^3] Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
 

### 2023-09-26 paper.qmd
eval=FALSE
Add docx

@@
   html: default
-#   # docx: default
+  docx: default
 #   # pdf: default
@@
 #| include: false
+#| eval: false
 library(tidyverse)

### 2023-09-29 paper.qmd
Minor edit from Peter

@@
 
-  To address data needs for both safety and project evaluation on the National Road Network (NRN), the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was developed.
+  To address data needs for both safety, business case development and project evaluation on the National Road Network (NRN), the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was developed.
   While cycling in Ireland represents only 3% of total modal share, with higher intensities in urban areas, the levels of cycling collisions are disproportionately high at 20% of all serious injuries and 7% of all fatalities.

### 2023-09-29 paper.qmd
Tweak abstract

@@
   To address data needs for both safety, business case development and project evaluation on the National Road Network (NRN), the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was developed.
-  While cycling in Ireland represents only 3% of total modal share, with higher intensities in urban areas, the levels of cycling collisions are disproportionately high at 20% of all serious injuries and 7% of all fatalities.
-  If Ireland is to meet its climate and safety targets, data to establish baseline cycling levels and future cycling levels is needed.
+  Cycling in Ireland represents only 3% of total modal share as of the latest Census (2016), but accounts for 20% of serious injuries and 7% of all fatalities.
+  High resolution datasets on baseline and potential cycling levels are needed to meet Ireland's climate and safety targets.
 

### 2023-09-29 paper.qmd
Add link supporting 3% figure

@@
 A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
-The main framework underpinning these efforts is the Climate Action Plan[^2] and The National Development Plan[^3]
-.
-Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: "83% of respondents supported enhanced walking and cycling facilities"
-.
+The main framework underpinning these efforts is the Climate Action Plan[^2] and The National Development Plan[^3].
+Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[
+https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf
+] "83% of respondents supported enhanced walking and cycling facilities".
 

### 2023-09-29 paper.qmd
Update abstract

@@
   CRUSE is available at <https://cruse.bike/> and builds on the Propensity to Cycle Tool (PCT) for England and Wales.
-  CRUSE goes beyond the PCT in several important ways, higher resolution data, more trip types, including estimates for education, inter-urban, and recreational trips.
-  In addition to understanding cycling intensity, for asset planning and management purposes, the tool provides essential road safety information to enable reporting of disaggregate collision rates.
+  CRUSE goes beyond the PCT in several important ways, modeling denser networks at higher levels of geographic resolution and additional purposes, including simulated recreational, social, shopping and personal trips, in addition trips to work and primary, secondary and tertiary education locations.
+  The tool provides key 'denominator' information to enable reporting of collision rates per billion km cycled.
 

### 2023-09-29 paper.qmd
Shorten abstract

@@
   OSM tags generate "cycle friendliness" estimates of all links on the network, based on existing recorded infrastructure.
-  A range of networks is provided, highlighting routes for directness (Fastest) and "cycle friendliness" (Quietest).
-  It uses origin and destination data from the 2016 Census in combination with modeled demand data to estimate cycling levels and potential at the area, route, and network levels for each county in Ireland and offers estimates of the baseline level of cycling and several future scenario-based levels of cycling.
+  Many networks are provided, highlighting routes for directness (Fastest) and "cycle friendliness" (Quietest).
+  It uses origin and destination data from the 2016 Census in combination with modeled demand data to estimate current and potential future cycling levels at area, route, and network levels for each county in Ireland.
 

### 2023-09-29 paper.qmd
Update abstract ending

@@
 
-  As countries, like Ireland, invest in cycling, the number of killed and seriously injured cyclists must reduce too.
-  The CRUSE Tool provides estimates of cycling potential and routing for each county in Ireland, and works in both urban and rural settings, to enable monitoring of cycling safety.
+  For investment in cycling to deliver on the huge potential benefits cycling must become safer.
   With growth in the E-bike market, the tool will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
-  The CRUSE Tool methodology and findings are directly relevant to addressing the challenges and opportunities faced by other NRAs.
-  The datasets resulting from the project are open access and can be used by both non-experts and professionals.
+  The findings and resulting tool are directly relevant to addressing the challenges and opportunities faced by National Road Authorities (NRAs) across Ireland.
+  The provision of open access evidence, that can be used non-experts and professionals, is relevant to countries with sustainable transport policies worldwide.
 keywords: 

### 2023-09-29 paper.qmd
Complete abstract tweaks
Shorten bit about car dependency

@@
 
-  For investment in cycling to deliver on the huge potential benefits cycling must become safer.
+  For investment in cycling to deliver on the huge potential benefits, cycling must become safer.
   With growth in the E-bike market, the tool will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
@@
 
-Transport is responsible for 23% of global emissions, 70% of which is from road transport, nearly half of which (around 10% of global emissions) is from passenger cars [@jaramillo2022].
-The transport system encourages, enables and in some cases enforces unsustainable lifestyles, including over-consumption of goods due to excessive mobile storage space and dependency on services that are only accessible by car due to land use plans that have built up around roads [@gray2001; @shergold2012; @motte-baumvol2010].
+Transport is responsible for 23% of global greenhouse gas emissions, 70% of which are from road transport, nearly half of which (around 10% of global emissions) is from passenger cars [@jaramillo2022].
+The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
+Services that are only accessible by car lock-in car dependency [@gray2001; @shergold2012; @motte-baumvol2010].
 

### 2023-09-29 paper.qmd
Improve discussion of evs

@@
 In the context of climate, road safety and physicial inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
-The framework highlights the importance of demand reduction (*avoid*ing unnecessary trips), in addition to mode *shift* and *improvement* of existing energy converters, in that order.
+The framework highlights the importance of demand reduction (*avoid*ing unnecessary trips), in addition to mode *shift* to sustainable modes and *improvement* of existing energy converters, in that order.
 
@@
 Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
-The rapid uptake of highly efficient e-bikes can also be seen as an improvement on most electric vehicles, which are too heavy and expensive to be a sustainable alternative and could in fact delay the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
+Furthermore, highly efficient e-bikes --- which are seeing rapid uptake in Ireland and other countries --- outperform electric electric cars, which are too heavy and expensive.
+Heavy electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
 At the European level, the European Union has a target of reducing greenhouse gas emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.[^1]

### 2023-09-29 paper.qmd
CRUSE Tool not tool

@@
 Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
-Furthermore, highly efficient e-bikes --- which are seeing rapid uptake in Ireland and other countries --- outperform electric electric cars, which are too heavy and expensive.
-Heavy electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
+Furthermore, highly efficient e-bikes --- which are seeing rapid uptake in Ireland and other countries --- outperform electric cars, which are too heavy and expensive.
+Over-reliance on electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
 At the European level, the European Union has a target of reducing greenhouse gas emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.[^1]
@@
 For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
-Better data to inform road safety policies and interventions is a motivation for the CRUSE tool outlined in this paper, and modeling active travel more broadly.
+Better data to inform road safety policies and interventions is a motivation for the CRUSE Tool outlined in this paper, and modeling active travel more broadly.
 
@@
 It was in this context that Transport Infrastructure Ireland (TII) commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) project.
-Based on the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
+Based on the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE Tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
 A key consideration for TII is to ensure both urban and rural Ireland were targeted for cycling infrastructure, as shown in @fig-dublin.
@@
 
-Illustration of the density of route networks generated by the CRUSE tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland.
-Source: CRUSE tool, available at <https://cruse.bike/>.
+Illustration of the density of route networks generated by the CRUSE Tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland.
+Source: CRUSE Tool, available at <https://cruse.bike/>.
 :::
@@
 The rest of paper is structured as follows.
-In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE tool for Ireland.
-In @sec-results, we present the results of the CRUSE tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
+In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool for Ireland.
+In @sec-results, we present the results of the CRUSE Tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
 In @sec-discussion, we discuss the implications of the results for policy and practice.
@@
 
-The methods used to generate the evidence presented in the CRUSE tool for Ireland build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
+The methods used to generate the evidence presented in the CRUSE Tool for Ireland build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
 An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
@@
 
-The CRUSE tool seeks to overcome the following methodological limitations of the original PCT:
+The CRUSE Tool seeks to overcome the following methodological limitations of the original PCT:
 
@@
 To overcome this issue, we developed a spatial interaction modelling methodology to estimate the number of trips between each OD pair for additional trip purposes.
-The classification of trip purposes used in the CRUSE tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
+The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
 An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes.
@@
 
-Trip purposes used in the CRUSE tool.
+Trip purposes used in the CRUSE Tool.
 :::
@@
 %%| fig-width: 100%
-%%| fig-cap: Hypothetical user stories for the CRUSE tool illustrating that it can be used in different ways by different user groups.
+%%| fig-cap: Hypothetical user stories for the CRUSE Tool illustrating that it can be used in different ways by different user groups.
 %%| label: fig-user-stories
@@
 
-Instead of trying to show such use cases, of which there are many hundreds, we present a selection of results to illustrate the main features of the CRUSE tool.
+Instead of trying to show such use cases, of which there are many hundreds, we present a selection of results to illustrate the main features of the CRUSE Tool.
 
@@
 
-The CRUSE tool provides estimates of cycling potential at the national level, as shown in @fig-national-results.
+The CRUSE Tool provides estimates of cycling potential at the national level, as shown in @fig-national-results.
 
@@
 
-The CRUSE tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
-This provision of multiple scenarios of behaviour changes *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE tool.
+The CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
+This provision of multiple scenarios of behaviour changes *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.
 
@@
 
-The CRUSE tool can help both in terms of describing the current situation, and also prioritise investment in those direct routes with high cycling potential that lack adequate cycling infrastructure, highlighted by @caulfield2012.
+The CRUSE Tool can help both in terms of describing the current situation, and also prioritise investment in those direct routes with high cycling potential that lack adequate cycling infrastructure, highlighted by @caulfield2012.
 
@@
 
-RL led the development of the CRUSE tool, with input from JT, EV, HM, EB, PW, GOT, DB, and SM.
+RL led the development of the CRUSE Tool, with input from JT, EV, HM, EB, PW, GOT, DB, and SM.
 SM instigated the project and coordinated its policy impact.

### 2023-09-29 paper.qmd
Plurality fix
Comment-out random study bit

@@
 A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
-The main framework underpinning these efforts is the Climate Action Plan[^2] and The National Development Plan[^3].
-Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[
+The main frameworks underpinning these efforts are the Climate Action Plan[^2], the National Development Plan[^3], and the National Roads 2040 strategy^[
+https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf
+].
+<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ -->
+These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[
 https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf

### 2023-09-29 paper.qmd
7 out of 10 trips by car stat

@@
 Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, this suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
+7 in 10 trips in Ireland are by car^[
+https://www.nationaltransport.ie/wp-content/uploads/2019/01/National_Household_Travel_Survey_2017_Report_-_December_2018.pdf
+].
+Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by
+this mode have the potential to replace trips by private car, most
+specifically for short to medium distance trips, but increasingly
+for longer trips as e-bikes extend the range of this mode"^[
+https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/
+].
 

### 2023-09-29 paper.qmd
Tweak wording

@@
 Based on the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE Tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
+
 A key consideration for TII is to ensure both urban and rural Ireland were targeted for cycling infrastructure, as shown in @fig-dublin.
@@
 The open source and publicly available nature of the tool ensures more inclusive and evidence-based conversations around cycle network planning between all stakeholders in the transport planning process.
-Other use cases in the Irish context includes monitoring the need for existing cycling infrastructure upgrades, undertaking appraisal, as well as safety and exposure information in line with the European Cycling Federation, Dutch Cycling Embassy recommendations and the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
+Other use cases in the Irish context includes monitoring the need for existing cycling infrastructure upgrades, undertaking appraisal for business case development, as well as safety and exposure information.
+The latter is in line with the European Cycling Federation, Dutch Cycling Embassy recommendations and the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
 These factors mean that the tool can be seen as an open access 'leverage point' in the planning system [@lovelace2020].

### 2023-09-29 paper.qmd
Add bit to discussion, finish additions

@@
 
-However, travel to school and work only constitute a small proportion of all trips.
+However, travel to school and work only constitute around only 30% of all trips in Ireland.
 To overcome this issue, we developed a spatial interaction modelling methodology to estimate the number of trips between each OD pair for additional trip purposes.
@@
 The full set of results are therefore too extensive to present in their entirety in this paper.
-Through the interactive web application, users can explore the data to generate the results that are most relevant to their needs, whether that is finding the cycling potential on a particular road or finding 'weak links' in the cycle network associated with a particular school, work place or other destination.
+Through the interactive web application, users can explore the data to generate the results that are most relevant to their needs, whether that is finding the cycling potential on a particular road or finding 'weak links' or barriers in the cycle network associated with a particular school, work place or other destination.
 
@@
 The CRUSE Tool can help both in terms of describing the current situation, and also prioritise investment in those direct routes with high cycling potential that lack adequate cycling infrastructure, highlighted by @caulfield2012.
+It will also support reporting for the RISM Directive on road casualties as part of the road safety management system.
+
+While the CRUSE Tool is not the first national and publicly available tool for cycling planning, it has some unique features that make it relevant for other countries, and other national road authorities tasked with making their transport systems safe.
+
+TODO: outline these features.
+
+The CRUSE Tool is designed to be used by a wide range of stakeholders, including transport planners, engineers, and other professionals, as well as non-professionals.
 

### 2023-10-03 paper.qmd
Update abstract

@@
 title: "CRUSE to Safe Cycling in Ireland"
-subtitle: An Open Source Methodology to Support Active Travel
+subtitle: An Open Source Tool to Support Active Travel
 bibliography: bibliography.bib
@@
 
-# Uncomment for draft versions:
+# # Uncomment for draft versions:
 format: 
@@
   docx: default
-#   # pdf: default
-
-# # Uncomment for formatted version
-# format:
-#   elsevier-pdf:
-#     keep-tex: true
-#     journal:
-#       name: Journal Name
-#       formatting: preprint
-#       model: 3p
-#       cite-style: super
+# Uncomment for formatted version
+  elsevier-pdf:
+    keep-tex: true
+    journal:
+      name: Journal Name
+      formatting: preprint
+      model: 3p
+      cite-style: super
+# #   # pdf: default
 
@@
 
-  To address data needs for both safety, business case development and project evaluation on the National Road Network (NRN), the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was developed.
+  The Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was commissioned by TII and developed by the Leeds Institute for Transport Studies (ITS) and AECOM To address data needs for safety, business case development and project evaluation on the National Road Network (NRN).
   Cycling in Ireland represents only 3% of total modal share as of the latest Census (2016), but accounts for 20% of serious injuries and 7% of all fatalities.
@@
 
-  Due to an absence of reliable data, particularly rural cycling levels, TII commissioned the Institute for Transport Studies (ITS) at the University of Leeds and AECOM to develop a new tool for this purpose.
-  ITS Leeds led the development of the PCT for England and Wales, which has "revolutionized the practice of strategic cycle planning nationally".
-  The tool is an open-source approach using recognized open-source methodology to enable planners, engineers, and other stakeholders to make evidence-based decisions for the NRN.
-  CRUSE is available at <https://cruse.bike/> and builds on the Propensity to Cycle Tool (PCT) for England and Wales.
-  CRUSE goes beyond the PCT in several important ways, modeling denser networks at higher levels of geographic resolution and additional purposes, including simulated recreational, social, shopping and personal trips, in addition trips to work and primary, secondary and tertiary education locations.
+  The CRUSE project builds on the methods underlying the Propensity to Cycle Tool for England and Wales (PCT), which had a major impact on strategic cycle network planning in the UK.
+  Like the PCT, CRUSE is based on open source software and reproducible data science.
+The resulting web application is available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions for the NRN.
+  CRUSE goes beyond the PCT in several important ways, modeling denser networks at higher levels of geographic resolution and additional purposes, including simulated recreational, social, shopping and personal trips, in addition to trips to work and primary, secondary and tertiary education locations.
   The tool provides key 'denominator' information to enable reporting of collision rates per billion km cycled.
 
-  CRUSE is structured in a similar way to the traditional four-stage transport model, but its use of Open Street Map (OSM) data, used by [Cyclestreets.net](https://www.cyclestreets.net/) for routing, enables network quality to be assessed without costly surveys to record new infrastructure.
-  OSM tags generate "cycle friendliness" estimates of all links on the network, based on existing recorded infrastructure.
-  Many networks are provided, highlighting routes for directness (Fastest) and "cycle friendliness" (Quietest).
-  It uses origin and destination data from the 2016 Census in combination with modeled demand data to estimate current and potential future cycling levels at area, route, and network levels for each county in Ireland.
+  Another key feature of CRUSE is its use of Open Street Map (OSM) data, which is continuously updated, and used by [Cyclestreets.net](https://www.cyclestreets.net/) for route calculation and "cycle friendliness" estimates of all links on the network.
+  CRUSE provides three network types: the 'Fastest' network highlights routes for directness; 'Quietest' prioritises cycle friendliness, and 'Balanced' represents a balance between speed and comfort.
+  Trips are generated based on origin and destination data from the 2016 Census in combination with modeled demand data to estimate current and potential future cycling levels at area, route, and network levels for each county in Ireland.
 
   For investment in cycling to deliver on the huge potential benefits, cycling must become safer.
-  With growth in the E-bike market, the tool will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
-  The findings and resulting tool are directly relevant to addressing the challenges and opportunities faced by National Road Authorities (NRAs) across Ireland.
-  The provision of open access evidence, that can be used non-experts and professionals, is relevant to countries with sustainable transport policies worldwide.
+  With growth in the E-bike market, CRUSE will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
+  Provision of open access data in a publicly available web tool will address the challenges faced by National Road Authorities (NRAs) across Ireland in meeting their sustainability commitments.
+  The experience of developing a deploying CRUSE are relevant to countries with sustainable transport policies worldwide.
 keywords: 

### 2023-10-03 paper.tex
Update abstract

--- 2023-09-22 paper.tex
@@
 
-To address data needs for both safety and project evaluation on the
-National Road Network (NRN), the Cycle Route Uptake and Scenario
-Estimation (CRUSE) Tool was developed. While cycling in Ireland
-represents only 3\% of total modal share, with higher intensities in
-urban areas, the levels of cycling collisions are disproportionately
-high at 20\% of all serious injuries and 7\% of all fatalities. If
-Ireland is to meet its climate and safety targets, data to establish
-baseline cycling levels and future cycling levels is needed.
+To address data needs for both safety, business case development and
+project evaluation on the National Road Network (NRN), the Cycle Route
+Uptake and Scenario Estimation (CRUSE) Tool was developed. Cycling in
+Ireland represents only 3\% of total modal share as of the latest Census
+(2016), but accounts for 20\% of serious injuries and 7\% of all
+fatalities. High resolution datasets on baseline and potential cycling
+levels are needed to meet Ireland's climate and safety targets.
 
@@
 Cycle Tool (PCT) for England and Wales. CRUSE goes beyond the PCT in
-several important ways, higher resolution data, more trip types,
-including estimates for education, inter-urban, and recreational trips.
-In addition to understanding cycling intensity, for asset planning and
-management purposes, the tool provides essential road safety information
-to enable reporting of disaggregate collision rates.
+several important ways, modeling denser networks at higher levels of
+geographic resolution and additional purposes, including simulated
+recreational, social, shopping and personal trips, in addition trips to
+work and primary, secondary and tertiary education locations. The tool
+provides key `denominator' information to enable reporting of collision
+rates per billion km cycled.
 
@@
 of all links on the network, based on existing recorded infrastructure.
-A range of networks is provided, highlighting routes for directness
-(Fastest) and ``cycle friendliness'' (Quietest). It uses origin and
-destination data from the 2016 Census in combination with modeled demand
-data to estimate cycling levels and potential at the area, route, and
-network levels for each county in Ireland and offers estimates of the
-baseline level of cycling and several future scenario-based levels of
-cycling.
+Many networks are provided, highlighting routes for directness (Fastest)
+and ``cycle friendliness'' (Quietest). It uses origin and destination
+data from the 2016 Census in combination with modeled demand data to
+estimate current and potential future cycling levels at area, route, and
+network levels for each county in Ireland.
 
-As countries, like Ireland, invest in cycling, the number of killed and
-seriously injured cyclists must reduce too. The CRUSE Tool provides
-estimates of cycling potential and routing for each county in Ireland,
-and works in both urban and rural settings, to enable monitoring of
-cycling safety. With growth in the E-bike market, the tool will help
-inform inter-urban and rural networks to support the transfer of trips
-to sustainable modes for longer journeys. The CRUSE Tool methodology and
-findings are directly relevant to addressing the challenges and
-opportunities faced by other NRAs. The datasets resulting from the
-project are open access and can be used by both non-experts and
-professionals.
+For investment in cycling to deliver on the huge potential benefits,
+cycling must become safer. With growth in the E-bike market, the tool
+will help inform inter-urban and rural networks to support the transfer
+of trips to sustainable modes for longer journeys. The findings and
+resulting tool are directly relevant to addressing the challenges and
+opportunities faced by National Road Authorities (NRAs) across Ireland.
+The provision of open access evidence, that can be used non-experts and
+professionals, is relevant to countries with sustainable transport
+policies worldwide.
 \end{abstract}
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, boxrule=0pt, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-Transport is responsible for 23\% of global emissions, 70\% of which is
-from road transport, nearly half of which (around 10\% of global
-emissions) is from passenger cars \citep{jaramillo2022}. The transport
-system encourages, enables and in some cases enforces unsustainable
-lifestyles, including over-consumption of goods due to excessive mobile
-storage space and dependency on services that are only accessible by car
-due to land use plans that have built up around roads
+Transport is responsible for 23\% of global greenhouse gas emissions,
+70\% of which are from road transport, nearly half of which (around 10\%
+of global emissions) is from passenger cars \citep{jaramillo2022}. The
+transport system encourages, enables and in some cases enforces
+unsustainable lifestyles. Services that are only accessible by car
+lock-in car dependency
 \citep{gray2001, shergold2012, motte-baumvol2010}.
@@
 reduction (\emph{avoid}ing unnecessary trips), in addition to mode
-\emph{shift} and \emph{improvement} of existing energy converters, in
-that order.
+\emph{shift} to sustainable modes and \emph{improvement} of existing
+energy converters, in that order.
 
@@
 that it can also help avoid unnecessary trips \citep{nello-deakin2020}.
-The rapid uptake of highly efficient e-bikes can also be seen as an
-improvement on most electric vehicles, which are too heavy and expensive
-to be a sustainable alternative and could in fact delay the transition
-away from car dependency and inadvertently enable ``high travel
-lock-in'' \citep{anable2019}. At the European level, the European Union
-has a target of reducing greenhouse gas emissions by 55\% by 2030,
-compared to 1990 levels, and to achieve `net-zero' by 2050.{[}\^{}1{]}
+Furthermore, highly efficient e-bikes --- which are seeing rapid uptake
+in Ireland and other countries --- outperform electric cars, which are
+too heavy and expensive. Over-reliance on electric cars could slow the
+transition away from car dependency and inadvertently enable ``high
+travel lock-in'' \citep{anable2019}. At the European level, the European
+Union has a target of reducing greenhouse gas emissions by 55\% by 2030,
+compared to 1990 levels, and to achieve `net-zero' by 2050.\footnote{https://eur-lex.europa.eu/legal-content/EN/TXT/?u
+  ri=celex\%3A32008L0096}
 
@@
 modes, this is a major challenge. Better data to inform road safety
-policies and interventions is a motivation for the CRUSE tool outlined
+policies and interventions is a motivation for the CRUSE Tool outlined
 in this paper, and modeling active travel more broadly.
@@
 halving the number of road traffic deaths and serious injuries by
-2030.{[}\^{}3{]} Doing so while simultaneously enabling rapid uptake of
-active modes will require key active travel routes to be identified and
-improved.
+2030.\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}}
+Doing so while simultaneously enabling rapid uptake of active modes will
+require key active travel routes to be identified and improved.
 
 A proactive approach to the development of cycling policy and
-infrastructure is taking place in Ireland. The main framework
-underpinning these efforts is the Climate Action Plan\footnote{\url{https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/}}
-and The National Development Plan\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}}
-. Strategic alignment is also outlined by the Phoenix Park Transport and
-Mobility Options Study, which highlights public support for active
-travel interventions: ``83\% of respondents supported enhanced walking
-and cycling facilities'' .
+infrastructure is taking place in Ireland. The main frameworks
+underpinning these efforts are the Climate Action Plan\footnote{\url{https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/}},
+the National Development Plan\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}},
+and the National Roads 2040 strategy\footnote{
+  https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf}.
+These policy frameworks are backed-up by public support for active
+travel interventions: while only 3\% of trips in Ireland are made by
+cycling according to the latest Census data from 2016,\footnote{
+  https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf}
+``83\% of respondents supported enhanced walking and cycling
+facilities''.
 
@@
 in all counties (notwithstanding existing cycling trips and differences
-in trip distances).
+in trip distances). 7 in 10 trips in Ireland are by car\footnote{
+  https://www.nationaltransport.ie/wp-content/uploads/2019/01/National\_Household\_Travel\_Survey\_2017\_Report\_-\_December\_2018.pdf}.
+Cycling has the potential to replace a large proportion of these trips:
+``A high priority must also be given to cyclists, because trips by this
+mode have the potential to replace trips by private car, most
+specifically for short to medium distance trips, but increasingly for
+longer trips as e-bikes extend the range of this mode''\footnote{
+  https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}.
 
@@
 project. Based on the Propensity to Cycle Tool (PCT) for England and
-Wales, the CRUSE tool was developed to provide evidence on current
-cycling levels and future cycling potential nationwide across Ireland. A
-key consideration for TII is to ensure both urban and rural Ireland were
-targeted for cycling infrastructure, as shown in
+Wales, the CRUSE Tool was developed to provide evidence on current
+cycling levels and future cycling potential nationwide across Ireland.
+
+A key consideration for TII is to ensure both urban and rural Ireland
+were targeted for cycling infrastructure, as shown in
 Figure~\ref{fig-dublin}. To this end, TII extensively reviewed activity
@@
 includes monitoring the need for existing cycling infrastructure
-upgrades, undertaking appraisal, as well as safety and exposure
-information in line with the European Cycling Federation, Dutch Cycling
-Embassy recommendations and the European RISM Directive, for reporting
-collision rates of vulnerable road users by 2024. These factors mean
-that the tool can be seen as an open access `leverage point' in the
-planning system \citep{lovelace2020}.
+upgrades, undertaking appraisal for business case development, as well
+as safety and exposure information. The latter is in line with the
+European Cycling Federation, Dutch Cycling Embassy recommendations and
+the European RISM Directive, for reporting collision rates of vulnerable
+road users by 2024. These factors mean that the tool can be seen as an
+open access `leverage point' in the planning system
+\citep{lovelace2020}.
 
@@
 \caption{\label{fig-dublin}Illustration of the density of route networks
-generated by the CRUSE tool for Dublin City and surroundings (top) and
-rural County Mayo on the West coast of Ireland. Source: CRUSE tool,
+generated by the CRUSE Tool for Dublin City and surroundings (top) and
+rural County Mayo on the West coast of Ireland. Source: CRUSE Tool,
 available at \url{https://cruse.bike/}.}
@@
 Section~\ref{sec-methods}, we outline the methods used to generate the
-evidence presented in the CRUSE tool for Ireland. In
-Section~\ref{sec-results}, we present the results of the CRUSE tool,
+evidence presented in the CRUSE Tool for Ireland. In
+Section~\ref{sec-results}, we present the results of the CRUSE Tool,
 including estimates of current cycling levels and future cycling
@@
 
-The methods used to generate the evidence presented in the CRUSE tool
+The methods used to generate the evidence presented in the CRUSE Tool
 for Ireland build on the Propensity to Cycle Tool (PCT), which was
@@
 
-The CRUSE tool seeks to overcome the following methodological
+The CRUSE Tool seeks to overcome the following methodological
 limitations of the original PCT:
@@
 
-However, travel to school and work only constitute a small proportion of
-all trips. To overcome this issue, we developed a spatial interaction
-modelling methodology to estimate the number of trips between each OD
-pair for additional trip purposes. The classification of trip purposes
-used in the CRUSE tool was guided mainly by the trip purpose
-classification found within the National Household Travel Survey (NHTS),
-but with the addition of categories based on the comprehensive POWSCAR
-data, and the need to include recreational trips and multi-stage trips
-(not yet implemented). An overview of the trip purposes used in CRUSE is
-presented in Figure~\ref{fig-trip-purposes}.
+However, travel to school and work only constitute around only 30\% of
+all trips in Ireland. To overcome this issue, we developed a spatial
+interaction modelling methodology to estimate the number of trips
+between each OD pair for additional trip purposes. The classification of
+trip purposes used in the CRUSE Tool was guided mainly by the trip
+purpose classification found within the National Household Travel Survey
+(NHTS), but with the addition of categories based on the comprehensive
+POWSCAR data, and the need to include recreational trips and multi-stage
+trips (not yet implemented). An overview of the trip purposes used in
+CRUSE is presented in Figure~\ref{fig-trip-purposes}.
 
@@
 
-\caption{\label{fig-trip-purposes}Trip purposes used in the CRUSE tool.}
+\caption{\label{fig-trip-purposes}Trip purposes used in the CRUSE Tool.}
 
@@
 
-\caption{\label{fig-user-stories}User stories for the CRUSE tool}
+\caption{\label{fig-user-stories}Hypothetical user stories for the CRUSE
+Tool illustrating that it can be used in different ways by different
+user groups.}
 
@@
 
-National planning
+The main result of the CRUSE project is an open access web application
+and datasets on current and potential future cycling levels in Ireland,
+with evidence provided at county and route segment levels. Like the PCT
+on which CRUSE builds, a key feature of the results is that they are
+systematic and available for all counties, and most routes on which
+cycling is permitted, in Ireland. The full set of results are therefore
+too extensive to present in their entirety in this paper. Through the
+interactive web application, users can explore the data to generate the
+results that are most relevant to their needs, whether that is finding
+the cycling potential on a particular road or finding `weak links' or
+barriers in the cycle network associated with a particular school, work
+place or other destination.
 
-Regional
+Instead of trying to show such use cases, of which there are many
+hundreds, we present a selection of results to illustrate the main
+features of the CRUSE Tool.
 
-Local
+To illustrate the results in urban areas, we took the top 4 cities in
+Ireland by population: Dublin City, Cork, Limerick and Galway (with city
+populations of 0.6, 0.2, 0.1 and 0.1 million respectively). The results
+for central areas demarcated with circular areas with a radius of 3km
+for these cities are presented in Figure~\ref{fig-city-results}.
+
+\begin{figure}
+
+{\centering
+
+\includegraphics{images/rnet_types.png}
+
+}
+
+\caption{\label{fig-city-results}Fastest, balanced and quietest cycle
+networks within 3km circular areas containing the four largest cities in
+Ireland. The networks are coloured by cycle friendliness and with widths
+varying in proportion to the potential level of cycling under the `Go
+Dutch' scenario.}
+
+\end{figure}
+
+Aside from the widely varying sizes and route networks of each city,
+with Dublin having by far the densest network, it is clear from the
+visualisation of cycling friendliness that there is a lot of room for
+improvement in all cities. Dublin has had received most investment in
+cycling infrastructure and this is apparent from the high proportion of
+the route network that is green, even in the `Fastest' route network
+representing routes that prioritise speed and directness over comfort
+and safety. Dublin also has the highest mode share of cycling of the
+four cities, adding to the evidence base showing that investment in
+cycling infrastructure is effective in increasing cycling uptake.
+
+Although Dublin City has made progress, there are still many parts of
+the Fastest route network, and even some parts of the Balanced and
+Quietest route networks, that are not cycle friendly and which have high
+cycling potential. According to a recent report, 71\% of residents in
+the Dublin Metropolitan Area support ``more cycle tracks along roads,
+physically separated from traffic and pedestrians''.\footnote{https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22\_DublinMetropolitanArea\_v35\_DIGITAL\_v2.pdf}
+The results for the Dublin area could help prioritise investment in
+cycling infrastructure in the city.
 
@@
 \section{Discussion}\label{sec-discussion}}
+
+The CRUSE Tool provides evidence on current and potential future cycling
+levels across Ireland down to the street level, with potential assigned
+to `fastest', `balanced' and `quietest' route networks. This provision
+of multiple scenarios of behaviour changes \emph{and} multiple scenarios
+of investment, for example in cycling infrastructure next to major roads
+vs quiet residential streets, is a key feature of the CRUSE Tool.
+
+As highlighted in a paper on cycling infrastructure preferences based on
+a case study of Dublin, both directness and quietness are important
+\citep{caulfield2012}:
+
+\begin{quote}
+Direct routes with short journey times were found to be the most
+important positive variable for existing cyclists and non-cyclists in
+determining route choice. This is followed by infrastructure type, the
+number of junctions along the route, traffic speed and cyclist volumes.
+In terms if infrastructure, regardless of the level of cycling
+confidence, routes which have `no facilities' or `bus/cycle lanes' are
+the least favoured cycle route types.
+\end{quote}
+
+The CRUSE Tool can help both in terms of describing the current
+situation, and also prioritise investment in those direct routes with
+high cycling potential that lack adequate cycling infrastructure,
+highlighted by \citet{caulfield2012}. It will also support reporting for
+the RISM Directive on road casualties as part of the road safety
+management system.
+
+While the CRUSE Tool is not the first national and publicly available
+tool for cycling planning, it has some unique features that make it
+relevant for other countries, and other national road authorities tasked
+with making their transport systems safe.
+
+TODO: outline these features.
+
+The CRUSE Tool is designed to be used by a wide range of stakeholders,
+including transport planners, engineers, and other professionals, as
+well as non-professionals.
 
@@
 
-RL led the development of the CRUSE tool, with input from JT, EV, HM,
+RL led the development of the CRUSE Tool, with input from JT, EV, HM,
 EB, PW, GOT, DB, and SM. SM instigated the project and coordinated its

### 2023-10-03 paper.qmd
Add aim

@@
 
-Mobility kills.
-High energy transport systems are a major contributor to climate change, a leading cause of premature death and injury due to road traffic collisions, and a cause of disease due to airborne pollutants.
+<!-- Mobility kills. -->
+High energy transport systems are a major contributor to climate change, a leading cause of premature death and injury due to road traffic collisions, and a cause of disease due to airborne and noise pollutants.
 Conversely, evidence-based and effective transport *policies* have great potential to decarbonize economies, improve public health, and save lives.
 
-Transport is responsible for 23% of global greenhouse gas emissions, 70% of which are from road transport, nearly half of which (around 10% of global emissions) is from passenger cars [@jaramillo2022].
+Transport is responsible for 23% of global greenhouse gas emissions, 70% of which are from road transport.
+Nearly nearly half of these emissions (around 10% of global emissions) are from passenger cars [@jaramillo2022].
 The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
@@
 In the context of climate, road safety and physicial inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
-The framework highlights the importance of demand reduction (*avoid*ing unnecessary trips), in addition to mode *shift* to sustainable modes and *improvement* of existing energy converters, in that order.
+The ASI framework highlights the importance of demand reduction (*avoid*ing unnecessary trips), in addition to mode *shift* to sustainable modes and *improvement* of existing energy converters, in that order.
 
@@
 The open source and publicly available nature of the tool ensures more inclusive and evidence-based conversations around cycle network planning between all stakeholders in the transport planning process.
-Other use cases in the Irish context includes monitoring the need for existing cycling infrastructure upgrades, undertaking appraisal for business case development, as well as safety and exposure information.
-The latter is in line with the European Cycling Federation, Dutch Cycling Embassy recommendations and the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
-These factors mean that the tool can be seen as an open access 'leverage point' in the planning system [@lovelace2020].
+
+TII also requires evidence to support monitoring of growth in cycling, upgrades of existing cycling infrastructure, business case development and appraisal, and exposure information to estimate crash rates.
+Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
+These factors mean that the tool can be seen as an open access 'leverage point', providing key information for and supporting ambitious plans in many aspects the planning system, from network design to post-build monitoring [@lovelace2020].
 
@@
 
-The rest of paper is structured as follows.
+The main aim of the paper is to present the CRUSE Tool and describe the methods and datasets on which the estimates of current and future potential cycling levels are based.
+We will also highlight key results, discuss the implications for policy and practice, and outline how similar tools could be developed in other countries.
+It is structured as follows.
 In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool for Ireland.

### 2023-10-03 paper.qmd
Add section on routing

@@
 The first version of the PCT was based on current and future potential uptake of cycling for single stage *travel to work* at desire line, zone, route, and route network levels [@lovelace2016].
-It was launched in April 2017 as the government-endorsed tool for strategic cycle network planning, as part of the Cycling and Walking Investment Strategy [@cycling2017].
+It was launched in April 2017 as the government-endorsed tool for strategic cycle network planning and part of the Cycling and Walking Investment Strategy [@cycling2017].
 Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modelling of impacts on health, environmental and distributional outcomes [@woodcock2021].
@@
 ```
+
+## Routing
+
+Routes in CRUSE are generated by CycleStreets, a not-for profit transport consultancy and web development company who provide application programming interfaces (APIs) supplying a range of datasets for cycle planning and advocacy.
+Although based in Cambridge, UK, CycleStreets provide routing services 'by cyclists, for cyclists' internationally, including in Ireland.
+The CycleStreets routing engine is based on OpenStreetMap (OSM) data, which is continuously updated by a global community of volunteers.
+
+CycleStreets' services can be used by anyone for free, but we commissioned a bespoke routing service for CRUSE for the following reasons:
+
+- We needed to calculate hundreds-of-thousands of routes, which is against the terms of service of the free API.
+- We needed to make changes to the routing profiles, including allowing routing on trunk roads, which are not permitted in the default routing profiles.
+- We needed to be able to control the version of OSM data being used for the routing, enabling regular updates to the route networks as OSM data is updated.
+
+## Scenarios of cycling potential {#sec-scenarios}
+
+The datasets outlined above were used to generate estimates of cycling currently (the 'Baseline' scenario) and under four scenarios of cycling potential: Near Market, Climate Action Plan, Go Dutch, and Ebike.
+Each of these scenarios is described below.
+<!-- TODO: add table. -->
+
+The Baseline scenario approximates current cycling levels.
+As outlined
+
+Near market
+Climate Action Plan
+Go Dutch
+Ebike
+
 ## User interface {#sec-ui}

### 2023-10-03 paper.qmd
Add route type section

@@
 
-CycleStreets' services can be used by anyone for free, but we commissioned a bespoke routing service for CRUSE for the following reasons:
+CycleStreets' services can be used by anyone for free, but we commissioned a bespoke routing service for CRUSE for the following reasons, to enable:
 
-- We needed to calculate hundreds-of-thousands of routes, which is against the terms of service of the free API.
-- We needed to make changes to the routing profiles, including allowing routing on trunk roads, which are not permitted in the default routing profiles.
-- We needed to be able to control the version of OSM data being used for the routing, enabling regular updates to the route networks as OSM data is updated.
+- Calculation of hundreds-of-thousands of routes, which is against the terms of service of the free API.
+- Making changes to the routing profiles, including allowing routing on trunk roads, which are not permitted in the default routing profiles.
+- Control over the version of OSM data being used for the routing, allowing regular updates to the route networks as OSM data is updated.
+
+We calculated three network types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest'.
+As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the 'Fastest' routes minimise journey time, accounting for traffic lights and surface type.
+The 'Quietest' route type minimises busy sections of road, allowing high 'diversion factors' from the fastest route on quieter but less direct ways (often avoiding roads and interactions with motor vehicles altogether where possible).
+The 'Balanced' route type is a compromise between the two, minimising journey time while avoiding the busiest roads.
+Allowing users to switch between these three route types enables them to consider the trade-offs between directness and 'cycle friendliness' when planning new infrastructure, as shown in @fig-route-types.
 

### 2023-10-03 paper.qmd
Add figures to illustrate fastest and quietest network types

@@
 <!-- Mobility kills. -->
+
 High energy transport systems are a major contributor to climate change, a leading cause of premature death and injury due to road traffic collisions, and a cause of disease due to airborne and noise pollutants.
@@
 
+[^1]: https://eur-lex.europa.eu/legal-content/EN/TXT/?u
+    ri=celex%3A32008L0096
+
 Regarding road traffic casualties, another deadly consequence of inefficient transport systems, the Road Infrastructure Safety Management (RISM) directive (2008/96/EC) requires member states to implement a road safety management system (RSMS) for all public roads.
-Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out".[^1]
+Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out".[^2]
 Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance travelled, the directive requires estimation of distance travelled by mode, down to the road link level.
@@
 
-[^1]: https://eur-lex.europa.eu/legal-content/EN/TXT/?u
+[^2]: https://eur-lex.europa.eu/legal-content/EN/TXT/?u
     ri=celex%3A32008L0096
 
-In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030.[^3] Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
-
-A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
-The main frameworks underpinning these efforts are the Climate Action Plan[^2], the National Development Plan[^3], and the National Roads 2040 strategy^[
-https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf
-].
-<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ -->
-These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[
-https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf
-] "83% of respondents supported enhanced walking and cycling facilities".
-
-[^2]: <https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/>
+In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030.[^3]
+Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
 
@@
 
-At the regional level, the recently published Greater Dublin Area Transport Strategy[^4] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
+A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
+The main frameworks underpinning these efforts are the Climate Action Plan[^4], the National Development Plan[^5], and the National Roads 2040 strategy
+[^6].
+<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ --> These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,[^7] "83% of respondents supported enhanced walking and cycling facilitie
+s".
+
+[^4]: <https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/>
+
+[^5]: <https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/>
+
+[^6]:  https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf
+
+[^7]:  https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf
+
+At the regional level, the recently published Greater Dublin Area Transport Strategy[^8] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
 Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, this suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
-7 in 10 trips in Ireland are by car^[
-https://www.nationaltransport.ie/wp-content/uploads/2019/01/National_Household_Travel_Survey_2017_Report_-_December_2018.pdf
-].
-Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by
-this mode have the potential to replace trips by private car, most
-specifically for short to medium distance trips, but increasingly
-for longer trips as e-bikes extend the range of this mode"^[
-https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/
-].
+7 in 10 trips in Ireland are by car[^9]
+.
+Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as e-bikes extend the range of this mode"
+[^10].
 
-[^4]: <https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/>
+[^8]: <https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/>
+
+[^9]:  https://www.nationaltransport.ie/wp-content/uploads/2019/01/National_Household_Travel_Survey_2017_Report\_-\_December_2018.pdf
+
+[^10]:  https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/
 
@@
 An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
-The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^5] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^6]
+The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^11] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^12]
 .
 
-[^5]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
+[^11]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
 
-[^6]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
+[^12]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
 
@@
 The resulting evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
-biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^7] and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
+biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^13] and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
 
-[^7]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
+[^13]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
 
@@
 ```
-
 ## Routing
@@
 
+![](images/paste-5.png)
+
+![](images/paste-6.png)
+
 ## Scenarios of cycling potential {#sec-scenarios}
@@
 
-Near market
-Climate Action Plan
-Go Dutch
-Ebike
+Near market Climate Action Plan Go Dutch Ebike
 
@@
 Although Dublin City has made progress, there are still many parts of the Fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
-According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians".[^8]
+According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians".[^14]
 The results for the Dublin area could help prioritise investment in cycling infrastructure in the city.
 
-[^8]:  https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22_DublinMetropolitanArea_v35_DIGITAL_v2.pdf
+[^14]: https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22_DublinMetropolitanArea_v35_DIGITAL_v2.pdf
 

### 2023-10-03 paper.qmd
Figure caption

@@
 
-![](images/paste-5.png)
+<!-- Add figure with a single row and subcaptions for Fastest and Quietest route types: -->
 
-![](images/paste-6.png)
+::: {#fig-route-types}
+
+!['Fastest' route network for Limerick](images/paste-5.png)
+
+!['Quietest' route network for Limerick](images/paste-6.png)
+
+Illustration of the 'Fastest' and 'Quietest' route networks for central and East Limmerick.
+
+:::
 

### 2023-10-03 paper.tex
Figure caption

@@
 \begin{frontmatter}
-\title{CRUSE to Safe Cycling in Ireland \\\large{An Open Source
-Methodology to Support Active Travel} }
+\title{CRUSE to Safe Cycling in Ireland \\\large{An Open Source Tool to
+Support Active Travel} }
 \author[1]{Robin Lovelace%
@@
 
-To address data needs for both safety, business case development and
-project evaluation on the National Road Network (NRN), the Cycle Route
-Uptake and Scenario Estimation (CRUSE) Tool was developed. Cycling in
-Ireland represents only 3\% of total modal share as of the latest Census
-(2016), but accounts for 20\% of serious injuries and 7\% of all
-fatalities. High resolution datasets on baseline and potential cycling
-levels are needed to meet Ireland's climate and safety targets.
+The Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was
+commissioned by TII and developed by the Leeds Institute for Transport
+Studies (ITS) and AECOM To address data needs for safety, business case
+development and project evaluation on the National Road Network (NRN).
+Cycling in Ireland represents only 3\% of total modal share as of the
+latest Census (2016), but accounts for 20\% of serious injuries and 7\%
+of all fatalities. High resolution datasets on baseline and potential
+cycling levels are needed to meet Ireland's climate and safety targets.
 
-Due to an absence of reliable data, particularly rural cycling levels,
-TII commissioned the Institute for Transport Studies (ITS) at the
-University of Leeds and AECOM to develop a new tool for this purpose.
-ITS Leeds led the development of the PCT for England and Wales, which
-has ``revolutionized the practice of strategic cycle planning
-nationally''. The tool is an open-source approach using recognized
-open-source methodology to enable planners, engineers, and other
-stakeholders to make evidence-based decisions for the NRN. CRUSE is
-available at \url{https://cruse.bike/} and builds on the Propensity to
-Cycle Tool (PCT) for England and Wales. CRUSE goes beyond the PCT in
-several important ways, modeling denser networks at higher levels of
-geographic resolution and additional purposes, including simulated
-recreational, social, shopping and personal trips, in addition trips to
-work and primary, secondary and tertiary education locations. The tool
-provides key `denominator' information to enable reporting of collision
-rates per billion km cycled.
+The CRUSE project builds on the methods underlying the Propensity to
+Cycle Tool for England and Wales (PCT), which had a major impact on
+strategic cycle network planning in the UK. Like the PCT, CRUSE is based
+on open source software and reproducible data science. The resulting web
+application is available at \url{https://cruse.bike/}, enabling
+planners, engineers, and other stakeholders to make more evidence-based
+decisions for the NRN. CRUSE goes beyond the PCT in several important
+ways, modeling denser networks at higher levels of geographic resolution
+and additional purposes, including simulated recreational, social,
+shopping and personal trips, in addition to trips to work and primary,
+secondary and tertiary education locations. The tool provides key
+`denominator' information to enable reporting of collision rates per
+billion km cycled.
 
-CRUSE is structured in a similar way to the traditional four-stage
-transport model, but its use of Open Street Map (OSM) data, used by
-\href{https://www.cyclestreets.net/}{Cyclestreets.net} for routing,
-enables network quality to be assessed without costly surveys to record
-new infrastructure. OSM tags generate ``cycle friendliness'' estimates
-of all links on the network, based on existing recorded infrastructure.
-Many networks are provided, highlighting routes for directness (Fastest)
-and ``cycle friendliness'' (Quietest). It uses origin and destination
-data from the 2016 Census in combination with modeled demand data to
-estimate current and potential future cycling levels at area, route, and
-network levels for each county in Ireland.
+Another key feature of CRUSE is its use of Open Street Map (OSM) data,
+which is continuously updated, and used by
+\href{https://www.cyclestreets.net/}{Cyclestreets.net} for route
+calculation and ``cycle friendliness'' estimates of all links on the
+network. CRUSE provides three network types: the `Fastest' network
+highlights routes for directness; `Quietest' prioritises cycle
+friendliness, and `Balanced' represents a balance between speed and
+comfort. Trips are generated based on origin and destination data from
+the 2016 Census in combination with modeled demand data to estimate
+current and potential future cycling levels at area, route, and network
+levels for each county in Ireland.
 
 For investment in cycling to deliver on the huge potential benefits,
-cycling must become safer. With growth in the E-bike market, the tool
-will help inform inter-urban and rural networks to support the transfer
-of trips to sustainable modes for longer journeys. The findings and
-resulting tool are directly relevant to addressing the challenges and
-opportunities faced by National Road Authorities (NRAs) across Ireland.
-The provision of open access evidence, that can be used non-experts and
-professionals, is relevant to countries with sustainable transport
-policies worldwide.
+cycling must become safer. With growth in the E-bike market, CRUSE will
+help inform inter-urban and rural networks to support the transfer of
+trips to sustainable modes for longer journeys. Provision of open access
+data in a publicly available web tool will address the challenges faced
+by National Road Authorities (NRAs) across Ireland in meeting their
+sustainability commitments. The experience of developing a deploying
+CRUSE are relevant to countries with sustainable transport policies
+worldwide.
 \end{abstract}
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, boxrule=0pt, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, sharp corners, boxrule=0pt, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 
-Mobility kills. High energy transport systems are a major contributor to
-climate change, a leading cause of premature death and injury due to
-road traffic collisions, and a cause of disease due to airborne
-pollutants. Conversely, evidence-based and effective transport
-\emph{policies} have great potential to decarbonize economies, improve
-public health, and save lives.
+High energy transport systems are a major contributor to climate change,
+a leading cause of premature death and injury due to road traffic
+collisions, and a cause of disease due to airborne and noise pollutants.
+Conversely, evidence-based and effective transport \emph{policies} have
+great potential to decarbonize economies, improve public health, and
+save lives.
 
 Transport is responsible for 23\% of global greenhouse gas emissions,
-70\% of which are from road transport, nearly half of which (around 10\%
-of global emissions) is from passenger cars \citep{jaramillo2022}. The
-transport system encourages, enables and in some cases enforces
-unsustainable lifestyles. Services that are only accessible by car
-lock-in car dependency
+70\% of which are from road transport. Nearly nearly half of these
+emissions (around 10\% of global emissions) are from passenger cars
+\citep{jaramillo2022}. The transport system encourages, enables and in
+some cases enforces unsustainable lifestyles. Services that are only
+accessible by car lock-in car dependency
 \citep{gray2001, shergold2012, motte-baumvol2010}.
@@
 be classified according to the `Avoid-Shift-Improve' (ASI) framework
-\citep{jaramillo2022}. The framework highlights the importance of demand
-reduction (\emph{avoid}ing unnecessary trips), in addition to mode
-\emph{shift} to sustainable modes and \emph{improvement} of existing
-energy converters, in that order.
+\citep{jaramillo2022}. The ASI framework highlights the importance of
+demand reduction (\emph{avoid}ing unnecessary trips), in addition to
+mode \emph{shift} to sustainable modes and \emph{improvement} of
+existing energy converters, in that order.
 
@@
 the National Development Plan\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}},
-and the National Roads 2040 strategy\footnote{
-  https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf}.
+and the National Roads 2040 strategy \footnote{https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf}.
 These policy frameworks are backed-up by public support for active
 travel interventions: while only 3\% of trips in Ireland are made by
-cycling according to the latest Census data from 2016,\footnote{
-  https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf}
-``83\% of respondents supported enhanced walking and cycling
-facilities''.
+cycling according to the latest Census data from 2016,\footnote{https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf}
+``83\% of respondents supported enhanced walking and cycling facilitie
+s''.
 
@@
 in all counties (notwithstanding existing cycling trips and differences
-in trip distances). 7 in 10 trips in Ireland are by car\footnote{
-  https://www.nationaltransport.ie/wp-content/uploads/2019/01/National\_Household\_Travel\_Survey\_2017\_Report\_-\_December\_2018.pdf}.
-Cycling has the potential to replace a large proportion of these trips:
-``A high priority must also be given to cyclists, because trips by this
-mode have the potential to replace trips by private car, most
+in trip distances). 7 in 10 trips in Ireland are by car\footnote{https://www.nationaltransport.ie/wp-content/uploads/2019/01/National\_Household\_Travel\_Survey\_2017\_Report\_-\_December\_2018.pdf}
+. Cycling has the potential to replace a large proportion of these
+trips: ``A high priority must also be given to cyclists, because trips
+by this mode have the potential to replace trips by private car, most
 specifically for short to medium distance trips, but increasingly for
-longer trips as e-bikes extend the range of this mode''\footnote{
-  https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}.
+longer trips as e-bikes extend the range of this mode'' \footnote{https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}.
 
@@
 conversations around cycle network planning between all stakeholders in
-the transport planning process. Other use cases in the Irish context
-includes monitoring the need for existing cycling infrastructure
-upgrades, undertaking appraisal for business case development, as well
-as safety and exposure information. The latter is in line with the
-European Cycling Federation, Dutch Cycling Embassy recommendations and
-the European RISM Directive, for reporting collision rates of vulnerable
-road users by 2024. These factors mean that the tool can be seen as an
-open access `leverage point' in the planning system
-\citep{lovelace2020}.
+the transport planning process.
+
+TII also requires evidence to support monitoring of growth in cycling,
+upgrades of existing cycling infrastructure, business case development
+and appraisal, and exposure information to estimate crash rates.
+Exposure information is a key part of the the European RISM Directive,
+for reporting collision rates of vulnerable road users by 2024. These
+factors mean that the tool can be seen as an open access `leverage
+point', providing key information for and supporting ambitious plans in
+many aspects the planning system, from network design to post-build
+monitoring \citep{lovelace2020}.
 
@@
 
-The rest of paper is structured as follows. In
-Section~\ref{sec-methods}, we outline the methods used to generate the
-evidence presented in the CRUSE Tool for Ireland. In
+The main aim of the paper is to present the CRUSE Tool and describe the
+methods and datasets on which the estimates of current and future
+potential cycling levels are based. We will also highlight key results,
+discuss the implications for policy and practice, and outline how
+similar tools could be developed in other countries. It is structured as
+follows. In Section~\ref{sec-methods}, we outline the methods used to
+generate the evidence presented in the CRUSE Tool for Ireland. In
 Section~\ref{sec-results}, we present the results of the CRUSE Tool,
@@
 launched in April 2017 as the government-endorsed tool for strategic
-cycle network planning, as part of the Cycling and Walking Investment
+cycle network planning and part of the Cycling and Walking Investment
 Strategy \citep{cycling2017}. Extensions of the PCT approach have
@@
 \end{figure}
+
+\hypertarget{routing}{%
+\subsection{Routing}\label{routing}}
+
+Routes in CRUSE are generated by CycleStreets, a not-for profit
+transport consultancy and web development company who provide
+application programming interfaces (APIs) supplying a range of datasets
+for cycle planning and advocacy. Although based in Cambridge, UK,
+CycleStreets provide routing services `by cyclists, for cyclists'
+internationally, including in Ireland. The CycleStreets routing engine
+is based on OpenStreetMap (OSM) data, which is continuously updated by a
+global community of volunteers.
+
+CycleStreets' services can be used by anyone for free, but we
+commissioned a bespoke routing service for CRUSE for the following
+reasons, to enable:
+
+\begin{itemize}
+\tightlist
+\item
+  Calculation of hundreds-of-thousands of routes, which is against the
+  terms of service of the free API.
+\item
+  Making changes to the routing profiles, including allowing routing on
+  trunk roads, which are not permitted in the default routing profiles.
+\item
+  Control over the version of OSM data being used for the routing,
+  allowing regular updates to the route networks as OSM data is updated.
+\end{itemize}
+
+We calculated three network types for each disaggregated (`jittered') OD
+pair: `Fastest', `Balanced' and `Quietest'. As outlined on the
+CycleStreets'
+\href{https://www.cyclestreets.net/help/journey/howitworks/}{website},
+the `Fastest' routes minimise journey time, accounting for traffic
+lights and surface type. The `Quietest' route type minimises busy
+sections of road, allowing high `diversion factors' from the fastest
+route on quieter but less direct ways (often avoiding roads and
+interactions with motor vehicles altogether where possible). The
+`Balanced' route type is a compromise between the two, minimising
+journey time while avoiding the busiest roads. Allowing users to switch
+between these three route types enables them to consider the trade-offs
+between directness and `cycle friendliness' when planning new
+infrastructure, as shown in Figure~\ref{fig-route-types}.
+
+\begin{figure}
+
+{\centering
+
+\begin{figure}
+
+{\centering \includegraphics{images/paste-5.png}
+
+}
+
+\caption{`Fastest' route network for Limerick}
+
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics{images/paste-6.png}
+
+}
+
+\caption{`Quietest' route network for Limerick}
+
+\end{figure}
+
+}
+
+\caption{\label{fig-route-types}Illustration of the `Fastest' and
+`Quietest' route networks for central and East Limmerick.}
+
+\end{figure}
+
+\hypertarget{sec-scenarios}{%
+\subsection{Scenarios of cycling potential}\label{sec-scenarios}}
+
+The datasets outlined above were used to generate estimates of cycling
+currently (the `Baseline' scenario) and under four scenarios of cycling
+potential: Near Market, Climate Action Plan, Go Dutch, and Ebike. Each
+of these scenarios is described below.
+
+The Baseline scenario approximates current cycling levels. As outlined
+
+Near market Climate Action Plan Go Dutch Ebike
 

### 2023-10-03 paper.tex
Update with route network type figures

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, sharp corners, boxrule=0pt, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden, enhanced, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
+\begin{minipage}[t]{\linewidth}
+
 {\centering
 
-\begin{figure}
+\raisebox{-\height}{
 
-{\centering \includegraphics{images/paste-5.png}
+\includegraphics{images/paste-5.png}
 
@@
 
-\caption{`Fastest' route network for Limerick}
+}
 
-\end{figure}
+\end{minipage}%
+\newline
+\begin{minipage}[t]{\linewidth}
 
-\begin{figure}
+{\centering
 
-{\centering \includegraphics{images/paste-6.png}
+\raisebox{-\height}{
+
+\includegraphics{images/paste-6.png}
 
@@
 
-\caption{`Quietest' route network for Limerick}
-
-\end{figure}
-
 }
 
-\caption{\label{fig-route-types}Illustration of the `Fastest' and
-`Quietest' route networks for central and East Limmerick.}
+\end{minipage}%
+
+\caption{\label{fig-route-types}Illustration of the Fastest and Quietest
+route networks for central and East Limmerick.}
 

### 2023-10-03 paper.qmd
Update with route network type figures
Add stats

--- 2023-10-03 paper.qmd
@@
 
+These map are available on the 'Route types' page for each county (see [cruse.bike/limerick/route-types](https://cruse.bike/limerick/route-types)).
+In addition to showing the estimated routes under each scenario, the page also presents summary statistics on the network:
+
+- On the fastest network, 25% of the distance cycled occurs in non-hostile segments and 6% in cycle-friendly segments.
+
+- Under the baseline scenario, 55% of the distance cycled on the quietest network occurs in non-hostile segments and 13% in cycle-friendly segments.
+
+These statistics can be revealing: in Limerick it suggests that around half of all cycling activity occurs on network segments that are hostile (with a high inferred level of traffic stress), even when efforts are taken to avoid busy segments.
+For the Fastest network, which may be more realistic for utility cycling, the proportion of cycling on hostile segments is even higher at 75%.
+
+TODO: Add description of route types in Limerick based on local knowledge and @fig-route-types.
+
+<!-- Add figure with a single row and subcaptions for Fastest and Quietest route types: -->
+
+::: {#fig-route-types layout-nrow="2"}
+
 ![](images/paste-5.png)
@@
 ![](images/paste-6.png)
+
+Illustration of the Fastest and Quietest route networks for central and East Limmerick.
+
+:::
 

### 2023-10-03 paper.qmd
Describe Near Market

@@
 ```
+
 ## Routing
@@
 The Baseline scenario approximates current cycling levels.
-As outlined
+As outlined in @sec-trip-purposes, the Baseline scenario includes travel to work and school, and additional trip purposes.
+Cycling levels were taken from the POWSCAR data, which includes the number of people travelling by each mode between each OD pair.
 
-Near market Climate Action Plan Go Dutch Ebike
+The Near Market scenario approximates the level of cycling that would be achieved if levels of cycling uptake observed in areas of Ireland with high levels of cycling according to the 2016 Census were achieved everywhere, accounting for differences in trip distances and hilliness levels.
+The scenario is implemented as follows:
+
+-   Calculate distance decay curves for Dublin for the base year (2016, using POWSCAR data) by fitting a model to the relevant OD data after it has been converted to a route network datasest
+-   Apply the Near Market model to the hilliness and distance values for each county during the build process
+-   Add the current level of cycling to the Near Market model
+
+An illustration of the level of cycling under various scenarios in the county of Limerick is shown in Figure \@ref(fig:distance-frequency).
+The results will vary from county to county.
+
+ Climate Action Plan Go Dutch Ebike
 

### 2023-10-03 paper.qmd
Add scenarios, close #12

@@
   docx: default
-# Uncomment for formatted version
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: Journal Name
-      formatting: preprint
-      model: 3p
-      cite-style: super
-# #   # pdf: default
+# # Uncomment for formatted version
+#   elsevier-pdf:
+#     keep-tex: true
+#     journal:
+#       name: Journal Name
+#       formatting: preprint
+#       model: 3p
+#       cite-style: super
+# # #   # pdf: default
 
@@
 ::: {#fig-route-types layout-nrow="2"}
-
 ![](images/paste-5.png)
@@
 Illustration of the Fastest and Quietest route networks for central and East Limmerick.
-
 :::
@@
 The datasets outlined above were used to generate estimates of cycling currently (the 'Baseline' scenario) and under four scenarios of cycling potential: Near Market, Climate Action Plan, Go Dutch, and Ebike.
-Each of these scenarios is described below.
+Each of these scenarios is described below, as illustrated in Figure @fig-scenarios.
 <!-- TODO: add table. -->
+
+::: {#fig-scenarios}
+
+![](images/paste-7.png)
+
+Distance-mode share graph for Limerick used to communicate the different scenarios. Source: county level view for Limerick in the open access CRUSE Tool at <https://cruse.bike/limerick>.
+
+:::
+
+### Baseline
 
@@
 Cycling levels were taken from the POWSCAR data, which includes the number of people travelling by each mode between each OD pair.
+
+### Near Market
 
@@
 
-An illustration of the level of cycling under various scenarios in the county of Limerick is shown in Figure \@ref(fig:distance-frequency).
-The results will vary from county to county.
+### Climate Action Plan
 
- Climate Action Plan Go Dutch Ebike
+The Climate Action Plan scenario is loosely based on the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which contains policies for action to achieve a 51% reduction in overall greenhouse gas emissions by 2030, on a path towards net-zero by 2050.
+For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.
+In terms of car travel, the target is to "Increas[e] the proportion of kilometres driven by passenger electric cars to between 40 and 45% by 2030, in addition to a reduction of 10% in kilometres driven by the remaining internal combustion engine cars."
+This equates to a 5.5 to 6% reduction in total car km driven.
+
+To model this decrease in car driver km, cycling uptake increases in line with the Go Dutch scenario.
+However, we only model shift from driving to cycling.
+There is no shift from other modes of transport to cycling.
+
+### Go Dutch
+
+Under the Go Dutch scenario cycling reaches levels equivalent to those found in the Netherlands, taking account of the effects of route hilliness (measured as mean gradient) and route distance.
+This scenario uses the same model that was used PCT, allowing trips to shift from any other mode to cycling [@lovelace2017].
+
+## Ebike
+
+Also based on the PCT scenario with the same name, the Eike scenario in CRUSE takes Go Dutch cycling uptake, and adds onto this the impact of increased ebike usage, which allows for longer cycle trips.
+However, travel to primary and secondary schools still uses Go Dutch uptake, since no ebike scenario has been developed for school journeys, and children may be less likely to own ebikes than adults.
 

### 2023-10-03 paper.tex
12 changes from eugeni (#13)
* Update abstract

* Add aim

* Add section on routing

* Add route type section

* Add figures to illustrate fastest and quietest network types

* Figure caption

* Update with route network type figures

* Add stats

* Describe Near Market

* Add scenarios, close #12

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden, enhanced, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, interior hidden, breakable]}{\end{tcolorbox}}\fi
 
@@
 
+These map are available on the `Route types' page for each county (see
+\href{https://cruse.bike/limerick/route-types}{cruse.bike/limerick/route-types}).
+In addition to showing the estimated routes under each scenario, the
+page also presents summary statistics on the network:
+
+\begin{itemize}
+\item
+  On the fastest network, 25\% of the distance cycled occurs in
+  non-hostile segments and 6\% in cycle-friendly segments.
+\item
+  Under the baseline scenario, 55\% of the distance cycled on the
+  quietest network occurs in non-hostile segments and 13\% in
+  cycle-friendly segments.
+\end{itemize}
+
+These statistics can be revealing: in Limerick it suggests that around
+half of all cycling activity occurs on network segments that are hostile
+(with a high inferred level of traffic stress), even when efforts are
+taken to avoid busy segments. For the Fastest network, which may be more
+realistic for utility cycling, the proportion of cycling on hostile
+segments is even higher at 75\%.
+
+TODO: Add description of route types in Limerick based on local
+knowledge and Figure~\ref{fig-route-types}.
+
 \begin{figure}
@@
 potential: Near Market, Climate Action Plan, Go Dutch, and Ebike. Each
-of these scenarios is described below.
+of these scenarios is described below, as illustrated in Figure
+Figure~\ref{fig-scenarios}.
+
+\begin{figure}
+
+{\centering
+
+\includegraphics{images/paste-7.png}
+
+}
+
+\caption{\label{fig-scenarios}Distance-mode share graph for Limerick
+used to communicate the different scenarios. Source: county level view
+for Limerick in the open access CRUSE Tool at
+\url{https://cruse.bike/limerick}.}
+
+\end{figure}
+
+\hypertarget{baseline}{%
+\subsubsection{Baseline}\label{baseline}}
 
 The Baseline scenario approximates current cycling levels. As outlined
+in Section~\ref{sec-trip-purposes}, the Baseline scenario includes
+travel to work and school, and additional trip purposes. Cycling levels
+were taken from the POWSCAR data, which includes the number of people
+travelling by each mode between each OD pair.
 
-Near market Climate Action Plan Go Dutch Ebike
+\hypertarget{near-market}{%
+\subsubsection{Near Market}\label{near-market}}
+
+The Near Market scenario approximates the level of cycling that would be
+achieved if levels of cycling uptake observed in areas of Ireland with
+high levels of cycling according to the 2016 Census were achieved
+everywhere, accounting for differences in trip distances and hilliness
+levels. The scenario is implemented as follows:
+
+\begin{itemize}
+\tightlist
+\item
+  Calculate distance decay curves for Dublin for the base year (2016,
+  using POWSCAR data) by fitting a model to the relevant OD data after
+  it has been converted to a route network datasest
+\item
+  Apply the Near Market model to the hilliness and distance values for
+  each county during the build process
+\item
+  Add the current level of cycling to the Near Market model
+\end{itemize}
+
+\hypertarget{climate-action-plan}{%
+\subsubsection{Climate Action Plan}\label{climate-action-plan}}
+
+The Climate Action Plan scenario is loosely based on the Irish
+Government's
+\href{https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/}{Climate
+Action Plan 2021}, which contains policies for action to achieve a 51\%
+reduction in overall greenhouse gas emissions by 2030, on a path towards
+net-zero by 2050. For transport, this includes 500,000 extra walking,
+cycling and public transport trips per day by 2030. In terms of car
+travel, the target is to ``Increas{[}e{]} the proportion of kilometres
+driven by passenger electric cars to between 40 and 45\% by 2030, in
+addition to a reduction of 10\% in kilometres driven by the remaining
+internal combustion engine cars.'' This equates to a 5.5 to 6\%
+reduction in total car km driven.
+
+To model this decrease in car driver km, cycling uptake increases in
+line with the Go Dutch scenario. However, we only model shift from
+driving to cycling. There is no shift from other modes of transport to
+cycling.
+
+\hypertarget{go-dutch}{%
+\subsubsection{Go Dutch}\label{go-dutch}}
+
+Under the Go Dutch scenario cycling reaches levels equivalent to those
+found in the Netherlands, taking account of the effects of route
+hilliness (measured as mean gradient) and route distance. This scenario
+uses the same model that was used PCT, allowing trips to shift from any
+other mode to cycling \citep{lovelace2017}.
+
+\hypertarget{ebike}{%
+\subsection{Ebike}\label{ebike}}
+
+Also based on the PCT scenario with the same name, the Eike scenario in
+CRUSE takes Go Dutch cycling uptake, and adds onto this the impact of
+increased ebike usage, which allows for longer cycle trips. However,
+travel to primary and secondary schools still uses Go Dutch uptake,
+since no ebike scenario has been developed for school journeys, and
+children may be less likely to own ebikes than adults.
 

### 2023-10-06 paper.qmd
Fix typo in scenario name heading
Add SIM x-reference

@@
 ::: {#fig-scenarios}
-
 ![](images/paste-7.png)
 
-Distance-mode share graph for Limerick used to communicate the different scenarios. Source: county level view for Limerick in the open access CRUSE Tool at <https://cruse.bike/limerick>.
-
+Distance-mode share graph for Limerick used to communicate the different scenarios.
+Source: county level view for Limerick in the open access CRUSE Tool at <https://cruse.bike/limerick>.
 :::
@@
 For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.
-In terms of car travel, the target is to "Increas[e] the proportion of kilometres driven by passenger electric cars to between 40 and 45% by 2030, in addition to a reduction of 10% in kilometres driven by the remaining internal combustion engine cars."
-This equates to a 5.5 to 6% reduction in total car km driven.
+In terms of car travel, the target is to "Increas\[e\] the proportion of kilometres driven by passenger electric cars to between 40 and 45% by 2030, in addition to a reduction of 10% in kilometres driven by the remaining internal combustion engine cars." This equates to a 5.5 to 6% reduction in total car km driven.
 
@@
 
-## Ebike
+### Ebike
 
@@
 
+The CRUSE Tool is an open access web application for strategic cycle network planning and prioritisation of road safety interventions across Ireland.
+Building on previous work, it provides a nationally consistent evidence base, providing valuable insights to planners and other stakeholders in the transport planning process, at national and local levels.
+Because the results are available at the route segment level, the tool can be used to identify 'weak links' in the cycle network, and to prioritise investment in cycling infrastructure.
+Furthermore, the provision of the evidence in a free and publicly available website, hosted at [cruse.bike](https://cruse.bike), means that it can be used by anyone, encouraging wider participation and more evidence-based debate about transport planning.
+
+A key feature of the project methodologically is its calculation of current and future potential not only for travel to work and travel to school, but also for other trip purposes, including recreational trips.
+This required the development of spatial interaction models and estimation of the relative attractiveness of different destinations for different trip purposes, an area of active research where new developments could be incorporated into the tool in future [@hasova2022].
+
 # List of abbreviations

### 2023-10-06 paper.qmd
Complete draft of conclusion

@@
 
+The CRUSE Tool and the underlying research and methods are not without limitations, suggesting future areas of research, data collection needs, and policy application.
+The route network level results have not been validated to the extent we had planned at the outset of the project.
+We compared route network level results with 'ground truth' data from cycle counter datasets across Dublin to test different network types and to ensure that additional trip types added to work and educational trips increased the quality of fit under the baseline scenario.
+However, the size of the counter network was insufficient to provide an opportunity for robust evaluation of model performance, suggesting a combined program of new count data collection and analysis should be a priority for future work, with findings feeding directly into work to improve the outputs of the the CRUSE Tool.
+
+More broadly, the tool's outputs are limited to just one more of active travel (cycling), ignoring walking and wheeling, including wheelchair use and a range of wheeled devices such as scooters, bike trailers and e-cargo bikes that can be used to escort children to school.
+Sustainable transport policies should plan for walking, wheeling and cycling, and there are many co-benefits of broadly defined active travel interventions that benefit all active modes, such as measures to reduce heavy motor traffic speeds and volumes in areas and along corridors with high active travel potential.
+This raises the question of whether other active modes should be incorporated into the results using the OD-based approach outlined in this paper, or whether different modelling approaches are needed to properly capture the shorter distance trips typically made by walking [@cooper2018].
+Furthermore, the estimates of cycling presented in the CRUSE Tool omit multi-modal trips including public transport, and omit trip chaining, due to the need to capture a large portion of cycling potential within the resource constraints of the project.
+
+The CRUSE Tool is already used in practice to support more ambitious and data-driven planning for safe cycling routes in multiple counties across Ireland.
+We hope that the underlying approach, and the publicly available evidence provided in the CRUSE Tool itself, provide a basis for future reproducible research, open source software development into strategic cycle network planning tools in Ireland and other countries.
+Active travel represents a win-win-win for physical activity, environment, and local economic opportunities.
+In combination with broader sustainable mobility measures and policies to reduce motor traffic speeds and volumes, tools such as CRUSE can support the fast and fair decarbonisation of transport systems worldwide.
+
 # List of abbreviations

### 2023-10-06 paper.qmd
Update discussion, first cohesive draft

@@
 
-While the CRUSE Tool is not the first national and publicly available tool for cycling planning, it has some unique features that make it relevant for other countries, and other national road authorities tasked with making their transport systems safe.
+While the CRUSE Tool is not the first national and publicly available tool for cycling planning, it has some key features that make it relevant for other countries, and other national road authorities tasked with making their transport systems safe:
 
-TODO: outline these features.
-
-The CRUSE Tool is designed to be used by a wide range of stakeholders, including transport planners, engineers, and other professionals, as well as non-professionals.
+-   The results are *open access* meaning that any stakeholder in the planning system can access the evidence.
+This will help to democratise the transport planning process and make wider conversations about transport planning more evidence-based and less polarised [@lovelace2020], something that is particularly relevant given the potentially polarising nature of pro-cycling interventions [@wild2017].
+-   The results are fully reproducible (code to be released pending sign-off by TII's IT team), preventing 'cloud lock in' to a potentially monopolistic consultancy, and encouraging input from the wider open source community [@lovelace2021; @dhir2017].
+-   By covering a wide range of trip purposes --- not just travel to work [@lovelace2017; @heinen2010] and travel school [@goodman2019] as covered in previously published research on open access national cycle netework planning tools --- the results capture a higher proportion of cycling potential including key rural trips which are often under-represented in models of active travel.
 

### 2023-10-06 paper.qmd
replace limerick with naas (kildare) - routing section

@@
 
-These map are available on the 'Route types' page for each county (see [cruse.bike/limerick/route-types](https://cruse.bike/limerick/route-types)).
+These map are available on the 'Route types' page for each county (see [cruse.bike/kildare/route-types](https://cruse.bike/kildare/route-types)).
 In addition to showing the estimated routes under each scenario, the page also presents summary statistics on the network:
 
--   On the fastest network, 25% of the distance cycled occurs in non-hostile segments and 6% in cycle-friendly segments.
+-   On the fastest network, 25% of the distance cycled occurs in non-hostile segments and 7% in cycle-friendly segments.
 
--   Under the baseline scenario, 55% of the distance cycled on the quietest network occurs in non-hostile segments and 13% in cycle-friendly segments.
+-   Under the baseline scenario, 52% of the distance cycled on the quietest network occurs in non-hostile segments and 15% in cycle-friendly segments.
 
-These statistics can be revealing: in Limerick it suggests that around half of all cycling activity occurs on network segments that are hostile (with a high inferred level of traffic stress), even when efforts are taken to avoid busy segments.
+These statistics can be revealing: in Kildare it suggests that around half of all cycling activity occurs on network segments that are hostile (with a high inferred level of traffic stress), even when efforts are taken to avoid busy segments.
 For the Fastest network, which may be more realistic for utility cycling, the proportion of cycling on hostile segments is even higher at 75%.
 
-TODO: Add description of route types in Limerick based on local knowledge and @fig-route-types.
+TODO: Add description of route types in Naas based on local knowledge and @fig-route-types.
 
@@
 
+<!-- ::: {#fig-route-types layout-nrow="2"} -->
+<!-- ![](images/paste-5.png) -->
+
+<!-- ![](images/paste-6.png) -->
+
+<!-- Illustration of the Fastest and Quietest route networks for central and East Limmerick. -->
+<!-- ::: -->
+
 ::: {#fig-route-types layout-nrow="2"}
-![](images/paste-5.png)
+![](images/naas_quietest.png){width="250"}
 
-![](images/paste-6.png)
+![](images/naas_fastest.png){width="250"}
 
-Illustration of the Fastest and Quietest route networks for central and East Limmerick.
+<!-- ![](images/newbridge_quietest.png){width="250"} -->
+
+<!-- ![](images/newbridge_fastest.png){width="250"} -->
+
+Illustration of the Quietest (top) and Fastest (bottom) route networks for Naas.
 :::
@@
 ::: {#fig-scenarios}
-![](images/paste-7.png)
+<!-- ![](images/paste-7.png) -->
+![](images/scenarios_kildare.png)
 
-Distance-mode share graph for Limerick used to communicate the different scenarios.
-Source: county level view for Limerick in the open access CRUSE Tool at <https://cruse.bike/limerick>.
+Distance-mode share graph for Kildare used to communicate the different scenarios.
+Source: county level view for Kildare in the open access CRUSE Tool at <https://cruse.bike/kildare>.
 :::

### 2023-10-06 paper.tex
Review by CycleStreets.tex (#16)
Have reviewed the paragraphs where we are mentioned.

@@
 \item
-  Calculation of hundreds-of-thousands of routes, which is against the
+  Calculation of hundreds-of-thousands of routes, which is beyond the
   terms of service of the free API.
@@
   Making changes to the routing profiles, including allowing routing on
-  trunk roads, which are not permitted in the default routing profiles.
+  trunk roads, which are sometimes omitted in the default routing profiles.
 \item

### 2023-10-06 paper.qmd
Port changes to .qmd, close #17 (#18)

--- 2023-10-06 paper.qmd
@@
 
--   Calculation of hundreds-of-thousands of routes, which is against the terms of service of the free API.
--   Making changes to the routing profiles, including allowing routing on trunk roads, which are not permitted in the default routing profiles.
+-   Calculation of hundreds-of-thousands of routes, which is beyond the terms of service of the free API.
+-   Making changes to the routing profiles, including allowing routing on trunk roads, which are sometimes avoided in the default routing profiles.
 -   Control over the version of OSM data being used for the routing, allowing regular updates to the route networks as OSM data is updated.

### 2023-10-06 paper.qmd
A couple of tweaks

@@
         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Eugeni Vidal Tortosa
+  - name: Eugeni Vidal-Tortosa
     affiliations:
@@
 abstract: |
-  Under the EU Road infrastructure safety management (RISM) directive, the National Road Safety Strategy (RSS), and the Climate Action Plan Transport Infrastructure Ireland (TII) has a remit for road safety and decarbonizing a predominantly road-based network in Ireland.
+  Under the EU Road infrastructure safety management (RISM) directive, the National Road Safety Strategy (RSS) and the Climate Action Plan, Transport Infrastructure Ireland (TII) has a remit for road safety and decarbonizing a predominantly road-based network in Ireland.
 

### 2023-10-06 paper.qmd
Uptake of cycling

@@
 
-Uptake of cycling, the main topic of this paper, should be seen in this broader context of transport decarbonisation [@brand2020] and sustainable mobility [@burns2013].
+Uptake of cycling should be seen in this broader context of transport decarbonisation [@brand2020] and sustainable mobility [@burns2013].
 Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].

### 2023-10-06 paper.qmd
A couple of tweaks (#20)
* A couple of tweaks

* Uptake of cycling

* Be concise

@@
 
-The main aim of the paper is to present the CRUSE Tool and describe the methods and datasets on which the estimates of current and future potential cycling levels are based.
-We will also highlight key results, discuss the implications for policy and practice, and outline how similar tools could be developed in other countries.
-It is structured as follows.
+This paper aims to describe the design, features and potential use of the CRUSE Tool.
 In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool for Ireland.

### 2023-10-07 paper.qmd
naas and newbridge godutch

--- 2023-10-06 paper.qmd
@@
         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Eugeni Vidal-Tortosa
+  - name: Eugeni Vidal Tortosa
     affiliations:
@@
 abstract: |
-  Under the EU Road infrastructure safety management (RISM) directive, the National Road Safety Strategy (RSS) and the Climate Action Plan, Transport Infrastructure Ireland (TII) has a remit for road safety and decarbonizing a predominantly road-based network in Ireland.
+  Under the EU Road infrastructure safety management (RISM) directive, the National Road Safety Strategy (RSS), and the Climate Action Plan Transport Infrastructure Ireland (TII) has a remit for road safety and decarbonizing a predominantly road-based network in Ireland.
 
@@
 
--   Calculation of hundreds-of-thousands of routes, which is beyond the terms of service of the free API.
--   Making changes to the routing profiles, including allowing routing on trunk roads, which are sometimes avoided in the default routing profiles.
+-   Calculation of hundreds-of-thousands of routes, which is against the terms of service of the free API.
+-   Making changes to the routing profiles, including allowing routing on trunk roads, which are not permitted in the default routing profiles.
 -   Control over the version of OSM data being used for the routing, allowing regular updates to the route networks as OSM data is updated.
@@
 
+<!-- ::: {#fig-route-types layout-nrow="2"} -->
+
+<!-- ![](images/paste-5.png) -->
+
+<!-- ![](images/paste-6.png) -->
+
+<!-- Illustration of the Fastest and Quietest route networks for central and East Limmerick. -->
+
+<!-- ::: -->
+
 ::: {#fig-route-types layout-nrow="2"}
-![](images/paste-5.png)
+![](images/naas_quietest_godutch.png){width="250"}
 
-![](images/paste-6.png)
+![](images/naas_fastest_godutch.png){width="250"}
 
-Illustration of the Fastest and Quietest route networks for central and East Limmerick.
+![](images/newbridge_quietest_godutch.png){width="250"}
+
+![](images/newbridge_fastest_godutch.png){width="250"}
+
+Illustration of the Quietest (left) and Fastest (right) route networks for Naas (top) and Newbridge (bottom).
 :::
@@
 ::: {#fig-scenarios}
-![](images/paste-7.png)
+<!-- ![](images/paste-7.png) -->
 
-Distance-mode share graph for Limerick used to communicate the different scenarios.
-Source: county level view for Limerick in the open access CRUSE Tool at <https://cruse.bike/limerick>.
+![](images/scenarios_kildare.png)
+
+Distance-mode share graph for Kildare used to communicate the different scenarios.
+Source: county level view for Kildare in the open access CRUSE Tool at <https://cruse.bike/kildare>.
 :::

### 2023-10-07 paper.qmd
More acknowledgements

@@
 
-Transport is responsible for 23% of global greenhouse gas emissions, 70% of which are from road transport.
+Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport.
 Nearly nearly half of these emissions (around 10% of global emissions) are from passenger cars [@jaramillo2022].
@@
 Over-reliance on electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
-At the European level, the European Union has a target of reducing greenhouse gas emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.[^1]
+At the European level, the European Union has a target of reducing GHG emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.[^1]
 
@@
 
-The Climate Action Plan scenario is loosely based on the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which contains policies for action to achieve a 51% reduction in overall greenhouse gas emissions by 2030, on a path towards net-zero by 2050.
+The Climate Action Plan scenario is loosely based on the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which contains policies for action to achieve a 51% reduction in overall GHG emissions by 2030, on a path towards net-zero by 2050.
 For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.
@@
 
+GHG: Greehouse Gas
+
 GTFS: General Transit Feed Specification
@@
 Thanks to Ciaran Maguire, Catherine Swift, and others at AECOM for their input into the project.
+Thanks to Martin Lucas-Smith and Simon Nutall from CycleStreets for their input into the project.
 

### 2023-10-08 cover-letter.qmd
Add cover letter for #21

---
format: pdf
title: "Cover letter"
bibliography: references.bib
---

```{=html}
<!-- Cover letter file, answering the following questions:

    1–2 sentences about the importance of the manuscript.
    1–2 sentences about the experience the author group has on the topic (for example, previous publications).
    1–2 sentences about the (expected) clinical impact of the manuscript. -->
```
Dear Editor,

We are submitting our manuscript entitled "CRUSE to Safe Cycling in Ireland: An Open Source Tool to Support Active Travel" for consideration for publication in the European Transport Research Review.

This manuscript describes the development of an open source tool to support active travel in Ireland.
The paper describes the design, features and potential uses of the CRUSE Tool.
We believe that the paper is important because it marks a substantial step forward in the development of open source tools to support active travel.
The work presented goes substantially beyond previously published research such as the Propensity to Cycle Tool for England and Wales.

The Tool itself is important, providing the first comprehensive evidence base on cycling across Ireland.
The evidence is actionable, provided for free to all stakeholders in the transport planning process, with information on cycling potential and cycle friendliness down to the segment level nationwide.
We believe that our experiences developing the tool will be of use to others tasked with investing in cost effective cycle infrastructure.
Furthermore, and unusually for this kind of tool, a major policy driver was road safety: a key output of the tool is estimates of baseline levels of cycling, which can be used to generate estimates of road safety levels in units of killed or seriously injured casualties per billion km (KSI/bkm).

As a group we have wide experience in this space...

We expect the manuscript will have the following impacts:

### 2023-10-09 cover-letter.qmd
21 cover letter (#23)
* Add cover letter for #21

* Update cover letter

@@
 We are submitting our manuscript entitled "CRUSE to Safe Cycling in Ireland: An Open Source Tool to Support Active Travel" for consideration for publication in the European Transport Research Review.
+It has a word count of 4,721, excluding references.
 
@@
 
-As a group we have wide experience in this space...
+As a group we have wide experience in this space, as highlighted by the below selection of our previous publications:
+
+-   A paper on cycling safety based on route network analysis [@meade2018]
+
+-   Methods enabling inequalities to be accounted for in cycle network analysis and design [@mahfouz]
+
+-   Methods enabling accurate modeling and effective visualization of origin-destination data and route networks [@morgan2020; @lovelace2022]
+
+-   Analysis of infrastructure preferences for cycling in Ireland [@caulfield2012]
+
+-   Papers describing previous approaches to developing strategic cycle network planning tools [@lovelace2017]
 
 We expect the manuscript will have the following impacts:
+
+-   Highlight the open availability of the CRUSE tool to stakeholders in Ireland, increasing its use, leading to more evidence-based and likely more effective network level designs, accounting for trade-offs between investment in 'fast' vs 'quiet' networks
+
+-   Increase awareness of open evidence to support effective and data-driven investment in and planning for cycling internationally
+
+-   Provide information and inspiration to researchers and other stakeholders to imagine, explore, and in some cases invest time in developing new strategic cycle network planning tools
+
+All the best,
+
+Robin Lovelace on behalf of all authors, Joey Talbot, Eugeni Vidal-Tortosa, and Hussein Mahfouz (University of Leeds), Elaine Brick, Peter Wright, Gary O'Tool (AECOM), Dan Brennan and Suzanne Meade (Transport Infrastructure Ireland)
+
+# References

### 2023-10-09 paper.qmd
Comment out TODO
Shorten abstract

@@
 format: 
-  html: default
-  docx: default
-# # Uncomment for formatted version
-#   elsevier-pdf:
-#     keep-tex: true
-#     journal:
-#       name: Journal Name
-#       formatting: preprint
-#       model: 3p
-#       cite-style: super
+  # html: default
+  # docx: default
+# Uncomment for formatted version
+  elsevier-pdf:
+    linestretch: 2
+    keep-tex: true
+    journal:
+      name: European Transport Research Review
+      formatting: preprint
+      model: 3p
+      cite-style: super
 # # #   # pdf: default
@@
 abstract: |
-  Under the EU Road infrastructure safety management (RISM) directive, the National Road Safety Strategy (RSS) and the Climate Action Plan, Transport Infrastructure Ireland (TII) has a remit for road safety and decarbonizing a predominantly road-based network in Ireland.
-
-  The Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was commissioned by TII and developed by the Leeds Institute for Transport Studies (ITS) and AECOM To address data needs for safety, business case development and project evaluation on the National Road Network (NRN).
+  Legislation mandates improved road safety and decarbonisation of predominantly road-based transport network in Ireland and other countries.
+The Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was commissioned to support cycle network planning, safety interventions, business case development and project evaluation.
   Cycling in Ireland represents only 3% of total modal share as of the latest Census (2016), but accounts for 20% of serious injuries and 7% of all fatalities.
@@
 
-  The CRUSE project builds on the methods underlying the Propensity to Cycle Tool for England and Wales (PCT), which had a major impact on strategic cycle network planning in the UK.
-  Like the PCT, CRUSE is based on open source software and reproducible data science.
-The resulting web application is available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions for the NRN.
-  CRUSE goes beyond the PCT in several important ways, modeling denser networks at higher levels of geographic resolution and additional purposes, including simulated recreational, social, shopping and personal trips, in addition to trips to work and primary, secondary and tertiary education locations.
+  The CRUSE Tool is based on open source software and reproducible data science.
+The resulting web application is available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions.
+  CRUSE goes beyond previous work in several important ways.
+  It models networks at high levels of geographic resolution, simulates a wide range of trips including recreational, social, shopping and personal trips, in addition to trips to work and primary, secondary and tertiary education locations.
   The tool provides key 'denominator' information to enable reporting of collision rates per billion km cycled.
 
-  Another key feature of CRUSE is its use of Open Street Map (OSM) data, which is continuously updated, and used by [Cyclestreets.net](https://www.cyclestreets.net/) for route calculation and "cycle friendliness" estimates of all links on the network.
+  Another key feature of CRUSE is its use of frequently-updated OpenStreetMap data for route calculation and "cycle friendliness" estimates of all links on the network.
   CRUSE provides three network types: the 'Fastest' network highlights routes for directness; 'Quietest' prioritises cycle friendliness, and 'Balanced' represents a balance between speed and comfort.
@@
   With growth in the E-bike market, CRUSE will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
-  Provision of open access data in a publicly available web tool will address the challenges faced by National Road Authorities (NRAs) across Ireland in meeting their sustainability commitments.
+  Provision of open access data in a publicly available web tool will address the challenges faced by counties across Ireland in meeting their sustainability commitments.
   The experience of developing a deploying CRUSE are relevant to countries with sustainable transport policies worldwide.
@@
 
-TODO: Add description of route types in Naas based on local knowledge and @fig-route-types.
+<!-- TODO: Add description of route types in Naas based on local knowledge and @fig-route-types. -->
 

### 2023-10-09 paper.tex
Shorten abstract

@@
 \usepackage{amsmath,amssymb}
+\usepackage{setspace}
 \usepackage{iftex}
@@
 \makeatother
-\journal{Journal Name}
+\journal{European Transport Research Review}
 \ifLuaTeX
@@
   pdftitle={CRUSE to Safe Cycling in Ireland},
-  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Tool; Dan Brennan; Suzanne Meade},
+  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Tool; Dan Brennan; Suzanne Meade},
   pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
@@
 
-\author[1]{Eugeni Vidal Tortosa%
+\author[1]{Eugeni Vidal-Tortosa%
 %
@@
 \begin{abstract}
-Under the EU Road infrastructure safety management (RISM) directive, the
-National Road Safety Strategy (RSS), and the Climate Action Plan
-Transport Infrastructure Ireland (TII) has a remit for road safety and
-decarbonizing a predominantly road-based network in Ireland.
-
-The Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was
-commissioned by TII and developed by the Leeds Institute for Transport
-Studies (ITS) and AECOM To address data needs for safety, business case
-development and project evaluation on the National Road Network (NRN).
-Cycling in Ireland represents only 3\% of total modal share as of the
-latest Census (2016), but accounts for 20\% of serious injuries and 7\%
-of all fatalities. High resolution datasets on baseline and potential
-cycling levels are needed to meet Ireland's climate and safety targets.
-
-The CRUSE project builds on the methods underlying the Propensity to
-Cycle Tool for England and Wales (PCT), which had a major impact on
-strategic cycle network planning in the UK. Like the PCT, CRUSE is based
-on open source software and reproducible data science. The resulting web
-application is available at \url{https://cruse.bike/}, enabling
-planners, engineers, and other stakeholders to make more evidence-based
-decisions for the NRN. CRUSE goes beyond the PCT in several important
-ways, modeling denser networks at higher levels of geographic resolution
-and additional purposes, including simulated recreational, social,
-shopping and personal trips, in addition to trips to work and primary,
-secondary and tertiary education locations. The tool provides key
-`denominator' information to enable reporting of collision rates per
-billion km cycled.
-
-Another key feature of CRUSE is its use of Open Street Map (OSM) data,
-which is continuously updated, and used by
-\href{https://www.cyclestreets.net/}{Cyclestreets.net} for route
-calculation and ``cycle friendliness'' estimates of all links on the
-network. CRUSE provides three network types: the `Fastest' network
-highlights routes for directness; `Quietest' prioritises cycle
-friendliness, and `Balanced' represents a balance between speed and
-comfort. Trips are generated based on origin and destination data from
-the 2016 Census in combination with modeled demand data to estimate
-current and potential future cycling levels at area, route, and network
-levels for each county in Ireland.
+Legislation mandates improved road safety and decarbonisation of
+predominantly road-based transport network in Ireland and other
+countries. The Cycle Route Uptake and Scenario Estimation (CRUSE) Tool
+was commissioned to support cycle network planning, safety
+interventions, business case development and project evaluation. Cycling
+in Ireland represents only 3\% of total modal share as of the latest
+Census (2016), but accounts for 20\% of serious injuries and 7\% of all
+fatalities. High resolution datasets on baseline and potential cycling
+levels are needed to meet Ireland's climate and safety targets.
+
+The CRUSE Tool is based on open source software and reproducible data
+science. The resulting web application is available at
+\url{https://cruse.bike/}, enabling planners, engineers, and other
+stakeholders to make more evidence-based decisions. CRUSE goes beyond
+previous work in several important ways. It models networks at high
+levels of geographic resolution, simulates a wide range of trips
+including recreational, social, shopping and personal trips, in addition
+to trips to work and primary, secondary and tertiary education
+locations. The tool provides key `denominator' information to enable
+reporting of collision rates per billion km cycled.
+
+Another key feature of CRUSE is its use of frequently-updated
+OpenStreetMap data for route calculation and ``cycle friendliness''
+estimates of all links on the network. CRUSE provides three network
+types: the `Fastest' network highlights routes for directness;
+`Quietest' prioritises cycle friendliness, and `Balanced' represents a
+balance between speed and comfort. Trips are generated based on origin
+and destination data from the 2016 Census in combination with modeled
+demand data to estimate current and potential future cycling levels at
+area, route, and network levels for each county in Ireland.
 
@@
 data in a publicly available web tool will address the challenges faced
-by National Road Authorities (NRAs) across Ireland in meeting their
-sustainability commitments. The experience of developing a deploying
-CRUSE are relevant to countries with sustainable transport policies
-worldwide.
+by counties across Ireland in meeting their sustainability commitments.
+The experience of developing a deploying CRUSE are relevant to countries
+with sustainable transport policies worldwide.
 \end{abstract}
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, interior hidden, breakable]}{\end{tcolorbox}}\fi
-
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, interior hidden]}{\end{tcolorbox}}\fi
+
+\setstretch{2}
 \newpage{}
@@
 
-Transport is responsible for 23\% of global greenhouse gas emissions,
-70\% of which are from road transport. Nearly nearly half of these
-emissions (around 10\% of global emissions) are from passenger cars
-\citep{jaramillo2022}. The transport system encourages, enables and in
-some cases enforces unsustainable lifestyles. Services that are only
+Transport is responsible for 23\% of global greenhouse gas (GHG)
+emissions, 70\% of which are from road transport. Nearly nearly half of
+these emissions (around 10\% of global emissions) are from passenger
+cars \citep{jaramillo2022}. The transport system encourages, enables and
+in some cases enforces unsustainable lifestyles. Services that are only
 accessible by car lock-in car dependency
@@
 
-Uptake of cycling, the main topic of this paper, should be seen in this
-broader context of transport decarbonisation \citep{brand2020} and
-sustainable mobility \citep{burns2013}. Although cycling uptake appears
-on the surface to only relate to the `shift' part of the ASI framework,
-closer consideration of the knock-on impacts of cycling uptake shows
-that it can also help avoid unnecessary trips \citep{nello-deakin2020}.
+Uptake of cycling should be seen in this broader context of transport
+decarbonisation \citep{brand2020} and sustainable mobility
+\citep{burns2013}. Although cycling uptake appears on the surface to
+only relate to the `shift' part of the ASI framework, closer
+consideration of the knock-on impacts of cycling uptake shows that it
+can also help avoid unnecessary trips \citep{nello-deakin2020}.
 Furthermore, highly efficient e-bikes --- which are seeing rapid uptake
@@
 travel lock-in'' \citep{anable2019}. At the European level, the European
-Union has a target of reducing greenhouse gas emissions by 55\% by 2030,
-compared to 1990 levels, and to achieve `net-zero' by 2050.\footnote{https://eur-lex.europa.eu/legal-content/EN/TXT/?u
+Union has a target of reducing GHG emissions by 55\% by 2030, compared
+to 1990 levels, and to achieve `net-zero' by 2050.\footnote{https://eur-lex.europa.eu/legal-content/EN/TXT/?u
   ri=celex\%3A32008L0096}
@@
 
-The main aim of the paper is to present the CRUSE Tool and describe the
-methods and datasets on which the estimates of current and future
-potential cycling levels are based. We will also highlight key results,
-discuss the implications for policy and practice, and outline how
-similar tools could be developed in other countries. It is structured as
-follows. In Section~\ref{sec-methods}, we outline the methods used to
-generate the evidence presented in the CRUSE Tool for Ireland. In
-Section~\ref{sec-results}, we present the results of the CRUSE Tool,
+This paper aims to describe the design, features and potential use of
+the CRUSE Tool. In Section~\ref{sec-methods}, we outline the methods
+used to generate the evidence presented in the CRUSE Tool for Ireland.
+In Section~\ref{sec-results}, we present the results of the CRUSE Tool,
 including estimates of current cycling levels and future cycling
@@
   Making changes to the routing profiles, including allowing routing on
-  trunk roads, which are sometimes omitted in the default routing profiles.
+  trunk roads, which are sometimes avoided in the default routing
+  profiles.
 \item
@@
 These map are available on the `Route types' page for each county (see
-\href{https://cruse.bike/limerick/route-types}{cruse.bike/limerick/route-types}).
+\href{https://cruse.bike/kildare/route-types}{cruse.bike/kildare/route-types}).
 In addition to showing the estimated routes under each scenario, the
@@
   On the fastest network, 25\% of the distance cycled occurs in
-  non-hostile segments and 6\% in cycle-friendly segments.
-\item
-  Under the baseline scenario, 55\% of the distance cycled on the
-  quietest network occurs in non-hostile segments and 13\% in
+  non-hostile segments and 7\% in cycle-friendly segments.
+\item
+  Under the baseline scenario, 52\% of the distance cycled on the
+  quietest network occurs in non-hostile segments and 15\% in
   cycle-friendly segments.
@@
 
-These statistics can be revealing: in Limerick it suggests that around
+These statistics can be revealing: in Kildare it suggests that around
 half of all cycling activity occurs on network segments that are hostile
@@
 
-TODO: Add description of route types in Limerick based on local
-knowledge and Figure~\ref{fig-route-types}.
-
 \begin{figure}
 
-\begin{minipage}[t]{\linewidth}
+\begin{minipage}[t]{0.50\linewidth}
 
@@
 
-\includegraphics{images/paste-5.png}
+\includegraphics[width=2.60417in,height=\textheight]{images/naas_quietest_godutch.png}
+
+}
+
+}
+
+\end{minipage}%
+%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics[width=2.60417in,height=\textheight]{images/naas_fastest_godutch.png}
 
@@
 \newline
-\begin{minipage}[t]{\linewidth}
+\begin{minipage}[t]{0.50\linewidth}
 
@@
 
-\includegraphics{images/paste-6.png}
+\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_quietest_godutch.png}
 
@@
 \end{minipage}%
-
-\caption{\label{fig-route-types}Illustration of the Fastest and Quietest
-route networks for central and East Limmerick.}
+%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_fastest_godutch.png}
+
+}
+
+}
+
+\end{minipage}%
+
+\caption{\label{fig-route-types}Illustration of the Quietest (left) and
+Fastest (right) route networks for Naas (top) and Newbridge (bottom).}
 
@@
 
-\includegraphics{images/paste-7.png}
-
-}
-
-\caption{\label{fig-scenarios}Distance-mode share graph for Limerick
-used to communicate the different scenarios. Source: county level view
-for Limerick in the open access CRUSE Tool at
-\url{https://cruse.bike/limerick}.}
+\includegraphics{images/scenarios_kildare.png}
+
+}
+
+\caption{\label{fig-scenarios}Distance-mode share graph for Kildare used
+to communicate the different scenarios. Source: county level view for
+Kildare in the open access CRUSE Tool at
+\url{https://cruse.bike/kildare}.}
 
@@
 Action Plan 2021}, which contains policies for action to achieve a 51\%
-reduction in overall greenhouse gas emissions by 2030, on a path towards
-net-zero by 2050. For transport, this includes 500,000 extra walking,
-cycling and public transport trips per day by 2030. In terms of car
-travel, the target is to ``Increas{[}e{]} the proportion of kilometres
-driven by passenger electric cars to between 40 and 45\% by 2030, in
-addition to a reduction of 10\% in kilometres driven by the remaining
-internal combustion engine cars.'' This equates to a 5.5 to 6\%
-reduction in total car km driven.
+reduction in overall GHG emissions by 2030, on a path towards net-zero
+by 2050. For transport, this includes 500,000 extra walking, cycling and
+public transport trips per day by 2030. In terms of car travel, the
+target is to ``Increas{[}e{]} the proportion of kilometres driven by
+passenger electric cars to between 40 and 45\% by 2030, in addition to a
+reduction of 10\% in kilometres driven by the remaining internal
+combustion engine cars.'' This equates to a 5.5 to 6\% reduction in
+total car km driven.
 
@@
 \hypertarget{ebike}{%
-\subsection{Ebike}\label{ebike}}
+\subsubsection{Ebike}\label{ebike}}
 
@@
 While the CRUSE Tool is not the first national and publicly available
-tool for cycling planning, it has some unique features that make it
+tool for cycling planning, it has some key features that make it
 relevant for other countries, and other national road authorities tasked
-with making their transport systems safe.
-
-TODO: outline these features.
-
-The CRUSE Tool is designed to be used by a wide range of stakeholders,
-including transport planners, engineers, and other professionals, as
-well as non-professionals.
+with making their transport systems safe:
+
+\begin{itemize}
+\tightlist
+\item
+  The results are \emph{open access} meaning that any stakeholder in the
+  planning system can access the evidence. This will help to democratise
+  the transport planning process and make wider conversations about
+  transport planning more evidence-based and less polarised
+  \citep{lovelace2020}, something that is particularly relevant given
+  the potentially polarising nature of pro-cycling interventions
+  \citep{wild2017}.
+\item
+  The results are fully reproducible (code to be released pending
+  sign-off by TII's IT team), preventing `cloud lock in' to a
+  potentially monopolistic consultancy, and encouraging input from the
+  wider open source community \citep{lovelace2021, dhir2017}.
+\item
+  By covering a wide range of trip purposes --- not just travel to work
+  \citep{lovelace2017, heinen2010} and travel school \citep{goodman2019}
+  as covered in previously published research on open access national
+  cycle netework planning tools --- the results capture a higher
+  proportion of cycling potential including key rural trips which are
+  often under-represented in models of active travel.
+\end{itemize}
 
@@
 
+The CRUSE Tool is an open access web application for strategic cycle
+network planning and prioritisation of road safety interventions across
+Ireland. Building on previous work, it provides a nationally consistent
+evidence base, providing valuable insights to planners and other
+stakeholders in the transport planning process, at national and local
+levels. Because the results are available at the route segment level,
+the tool can be used to identify `weak links' in the cycle network, and
+to prioritise investment in cycling infrastructure. Furthermore, the
+provision of the evidence in a free and publicly available website,
+hosted at \href{https://cruse.bike}{cruse.bike}, means that it can be
+used by anyone, encouraging wider participation and more evidence-based
+debate about transport planning.
+
+A key feature of the project methodologically is its calculation of
+current and future potential not only for travel to work and travel to
+school, but also for other trip purposes, including recreational trips.
+This required the development of spatial interaction models and
+estimation of the relative attractiveness of different destinations for
+different trip purposes, an area of active research where new
+developments could be incorporated into the tool in future
+\citep{hasova2022}.
+
+The CRUSE Tool and the underlying research and methods are not without
+limitations, suggesting future areas of research, data collection needs,
+and policy application. The route network level results have not been
+validated to the extent we had planned at the outset of the project. We
+compared route network level results with `ground truth' data from cycle
+counter datasets across Dublin to test different network types and to
+ensure that additional trip types added to work and educational trips
+increased the quality of fit under the baseline scenario. However, the
+size of the counter network was insufficient to provide an opportunity
+for robust evaluation of model performance, suggesting a combined
+program of new count data collection and analysis should be a priority
+for future work, with findings feeding directly into work to improve the
+outputs of the the CRUSE Tool.
+
+More broadly, the tool's outputs are limited to just one more of active
+travel (cycling), ignoring walking and wheeling, including wheelchair
+use and a range of wheeled devices such as scooters, bike trailers and
+e-cargo bikes that can be used to escort children to school. Sustainable
+transport policies should plan for walking, wheeling and cycling, and
+there are many co-benefits of broadly defined active travel
+interventions that benefit all active modes, such as measures to reduce
+heavy motor traffic speeds and volumes in areas and along corridors with
+high active travel potential. This raises the question of whether other
+active modes should be incorporated into the results using the OD-based
+approach outlined in this paper, or whether different modelling
+approaches are needed to properly capture the shorter distance trips
+typically made by walking \citep{cooper2018}. Furthermore, the estimates
+of cycling presented in the CRUSE Tool omit multi-modal trips including
+public transport, and omit trip chaining, due to the need to capture a
+large portion of cycling potential within the resource constraints of
+the project.
+
+The CRUSE Tool is already used in practice to support more ambitious and
+data-driven planning for safe cycling routes in multiple counties across
+Ireland. We hope that the underlying approach, and the publicly
+available evidence provided in the CRUSE Tool itself, provide a basis
+for future reproducible research, open source software development into
+strategic cycle network planning tools in Ireland and other countries.
+Active travel represents a win-win-win for physical activity,
+environment, and local economic opportunities. In combination with
+broader sustainable mobility measures and policies to reduce motor
+traffic speeds and volumes, tools such as CRUSE can support the fast and
+fair decarbonisation of transport systems worldwide.
+
 \hypertarget{list-of-abbreviations}{%
@@
 ED: Electoral Division
+
+GHG: Greehouse Gas
 
@@
 Thanks to Ciaran Maguire, Catherine Swift, and others at AECOM for their
-input into the project.
+input into the project. Thanks to Martin Lucas-Smith and Simon Nutall
+from CycleStreets for their input into the project.
 

### 2023-10-09 paper.qmd
Exeter name tweak, title page

@@
       - name: AECOM
-        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY, UK
+        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
   - name: "Gary O'Tool"
@@
 library(tmap)
+# Save title page (1st page of paper.pdf):
+system("pdftk _site/paper.pdf cat 1 output _site/title_page.pdf")
 ```

### 2023-10-09 paper.qmd
Extra citations

@@
 Over-reliance on electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
-At the European level, the European Union has a target of reducing GHG emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.[^1]
-
-[^1]: https://eur-lex.europa.eu/legal-content/EN/TXT/?u
-    ri=celex%3A32008L0096
+At the European level, the European Union has a target of reducing GHG emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050 [@rosenow2022].
 
 Regarding road traffic casualties, another deadly consequence of inefficient transport systems, the Road Infrastructure Safety Management (RISM) directive (2008/96/EC) requires member states to implement a road safety management system (RSMS) for all public roads.
-Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out".[^2]
+Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out" [@directiv2008].
 Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance travelled, the directive requires estimation of distance travelled by mode, down to the road link level.
@@
 
-[^2]: https://eur-lex.europa.eu/legal-content/EN/TXT/?u
-    ri=celex%3A32008L0096
+In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030.[^1]
+Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
 
-In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030.[^3]
-Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
+[^1]: <https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/>
+
+A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
+The main frameworks underpinning these efforts are the Climate Action Plan[^2], the National Development Plan[^3], and the National Roads 2040 strategy
+[^4].
+<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ --> These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,[^5] "83% of respondents supported enhanced walking and cycling facilitie
+s".
+
+[^2]: <https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/>
 
@@
 
-A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
-The main frameworks underpinning these efforts are the Climate Action Plan[^4], the National Development Plan[^5], and the National Roads 2040 strategy
-[^6].
-<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ --> These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,[^7] "83% of respondents supported enhanced walking and cycling facilitie
-s".
+[^4]: https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf
 
-[^4]: <https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/>
+[^5]: https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf
 
-[^5]: <https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/>
-
-[^6]: https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf
-
-[^7]: https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf
-
-At the regional level, the recently published Greater Dublin Area Transport Strategy[^8] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
+At the regional level, the recently published Greater Dublin Area Transport Strategy[^6] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
 Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, this suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
-7 in 10 trips in Ireland are by car[^9]
+7 in 10 trips in Ireland are by car[^7]
 .
 Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as e-bikes extend the range of this mode"
-[^10].
+[^8].
 
-[^8]: <https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/>
+[^6]: <https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/>
 
-[^9]: https://www.nationaltransport.ie/wp-content/uploads/2019/01/National_Household_Travel_Survey_2017_Report\_-\_December_2018.pdf
+[^7]: https://www.nationaltransport.ie/wp-content/uploads/2019/01/National_Household_Travel_Survey_2017_Report\_-\_December_2018.pdf
 
-[^10]: https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/
+[^8]: https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/
 
@@
 An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
-The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^11] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^12]
+The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^9] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^10]
 .
 
-[^11]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
+[^9]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
 
-[^12]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
+[^10]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
 
@@
 The resulting evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
-biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^13] and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
+biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^11] and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
 
-[^13]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
+[^11]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
 
@@
 Although Dublin City has made progress, there are still many parts of the Fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
-According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians".[^14]
+According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians".[^12]
 The results for the Dublin area could help prioritise investment in cycling infrastructure in the city.
 
-[^14]: https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22_DublinMetropolitanArea_v35_DIGITAL_v2.pdf
+[^12]: https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22_DublinMetropolitanArea_v35_DIGITAL_v2.pdf
 

### 2023-10-09 paper.tex
Extra citations

@@
 \affiliation[3]{organization={AECOM},addressline={Winslade House,
-Winslade Park, Manor Drive, Clyst St Mary, EXETER, EX5 1FY,
+Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY,
 UK},postcodesep={}}
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, interior hidden]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, boxrule=0pt, enhanced, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 Union has a target of reducing GHG emissions by 55\% by 2030, compared
-to 1990 levels, and to achieve `net-zero' by 2050.\footnote{https://eur-lex.europa.eu/legal-content/EN/TXT/?u
-  ri=celex\%3A32008L0096}
+to 1990 levels, and to achieve `net-zero' by 2050 \citep{directiv2008}.
 

### 2023-10-09 paper.qmd
Update paper, no footnotes

@@
 
-In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030.[^1]
+In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030 [@national2021].
 Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
 
-[^1]: <https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/>
+A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
+The main frameworks underpinning these efforts are the Climate Action Plan [@climate2022a], the National Development Plan, and the National Roads 2040 strategy [@national2023].
+<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ --> These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016 "83% of respondents supported enhanced walking and cycling facilities" [@sustaina2019].
 
-A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
-The main frameworks underpinning these efforts are the Climate Action Plan[^2], the National Development Plan[^3], and the National Roads 2040 strategy
-[^4].
-<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ --> These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,[^5] "83% of respondents supported enhanced walking and cycling facilitie
-s".
-
-[^2]: <https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/>
-
-[^3]: <https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/>
-
-[^4]: https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf
-
-[^5]: https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf
-
-At the regional level, the recently published Greater Dublin Area Transport Strategy[^6] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
+At the regional level, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
 Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, this suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
-7 in 10 trips in Ireland are by car[^7]
-.
-Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as e-bikes extend the range of this mode"
-[^8].
-
-[^6]: <https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/>
-
-[^7]: https://www.nationaltransport.ie/wp-content/uploads/2019/01/National_Household_Travel_Survey_2017_Report\_-\_December_2018.pdf
-
-[^8]: https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/
+7 in 10 trips in Ireland are by car [@national] .
+Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as e-bikes extend the range of this mode" [@greater].
 
@@
 An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
-The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^9] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^10]
-.
-
-[^9]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.
-
-[^10]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.
+The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales" by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[@lovelace2023].
 
@@
 The resulting evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
-biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^11] and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
-
-[^11]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).
+biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
 
@@
 Although Dublin City has made progress, there are still many parts of the Fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
-According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians".[^12]
+According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians".[^1]
 The results for the Dublin area could help prioritise investment in cycling infrastructure in the city.
 
-[^12]: https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22_DublinMetropolitanArea_v35_DIGITAL_v2.pdf
+[^1]: https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22_DublinMetropolitanArea_v35_DIGITAL_v2.pdf
 

### 2023-10-09 paper.tex
Update paper, no footnotes

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, boxrule=0pt, enhanced, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 Union has a target of reducing GHG emissions by 55\% by 2030, compared
-to 1990 levels, and to achieve `net-zero' by 2050 \citep{directiv2008}.
+to 1990 levels, and to achieve `net-zero' by 2050 \citep{rosenow2022}.
 
@@
 ``Member States shall ensure that the ranking of high accident
-concentration sections and the network safety ranking are carried
-out''.\footnote{https://eur-lex.europa.eu/legal-content/EN/TXT/?u
-  ri=celex\%3A32008L0096} Given that `safety' in this context is
-usefully quantified as the number of people killed and seriously injured
-(KSI) per distance travelled, the directive requires estimation of
-distance travelled by mode, down to the road link level. For active
-modes, about which there is a paucity of data compared with motorized
-modes, this is a major challenge. Better data to inform road safety
-policies and interventions is a motivation for the CRUSE Tool outlined
-in this paper, and modeling active travel more broadly.
+concentration sections and the network safety ranking are carried out''
+\citep{directiv2008}. Given that `safety' in this context is usefully
+quantified as the number of people killed and seriously injured (KSI)
+per distance travelled, the directive requires estimation of distance
+travelled by mode, down to the road link level. For active modes, about
+which there is a paucity of data compared with motorized modes, this is
+a major challenge. Better data to inform road safety policies and
+interventions is a motivation for the CRUSE Tool outlined in this paper,
+and modeling active travel more broadly.
 
 In Ireland, the Road Safety Authority (RSA) has set the target of
-halving the number of road traffic deaths and serious injuries by
-2030.\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}}
-Doing so while simultaneously enabling rapid uptake of active modes will
-require key active travel routes to be identified and improved.
+halving the number of road traffic deaths and serious injuries by 2030
+\citep{national2021}. Doing so while simultaneously enabling rapid
+uptake of active modes will require key active travel routes to be
+identified and improved.
 
@@
 infrastructure is taking place in Ireland. The main frameworks
-underpinning these efforts are the Climate Action Plan\footnote{\url{https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/}},
-the National Development Plan\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}},
-and the National Roads 2040 strategy \footnote{https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf}.
-These policy frameworks are backed-up by public support for active
-travel interventions: while only 3\% of trips in Ireland are made by
-cycling according to the latest Census data from 2016,\footnote{https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf}
-``83\% of respondents supported enhanced walking and cycling facilitie
-s''.
+underpinning these efforts are the Climate Action Plan
+\citep{climate2022a}, the National Development Plan, and the National
+Roads 2040 strategy \citep{national2023}. These policy frameworks are
+backed-up by public support for active travel interventions: while only
+3\% of trips in Ireland are made by cycling according to the latest
+Census data from 2016 ``83\% of respondents supported enhanced walking
+and cycling facilities'' \citep{sustaina2019}.
 
 At the regional level, the recently published Greater Dublin Area
-Transport Strategy\footnote{\url{https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}}
-reinforces these findings: ``nearly a quarter of adults cycle at least
-once a week in the Dublin Metropolitan Area'' with cycling in the Dublin
-area taking up to 60,000 cars off the road today. Extrapolating this on
-a per population basis across Ireland, with around 40\% of the
-population living in Dublin, this suggests that around 150,000 cars
-could be removed nationwide just by achieving Dublin levels of cycling
-in all counties (notwithstanding existing cycling trips and differences
-in trip distances). 7 in 10 trips in Ireland are by car\footnote{https://www.nationaltransport.ie/wp-content/uploads/2019/01/National\_Household\_Travel\_Survey\_2017\_Report\_-\_December\_2018.pdf}
-. Cycling has the potential to replace a large proportion of these
-trips: ``A high priority must also be given to cyclists, because trips
-by this mode have the potential to replace trips by private car, most
-specifically for short to medium distance trips, but increasingly for
-longer trips as e-bikes extend the range of this mode'' \footnote{https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}.
+Transport Strategy \citep{greater} reinforces these findings: ``nearly a
+quarter of adults cycle at least once a week in the Dublin Metropolitan
+Area'' with cycling in the Dublin area taking up to 60,000 cars off the
+road today. Extrapolating this on a per population basis across Ireland,
+with around 40\% of the population living in Dublin, this suggests that
+around 150,000 cars could be removed nationwide just by achieving Dublin
+levels of cycling in all counties (notwithstanding existing cycling
+trips and differences in trip distances). 7 in 10 trips in Ireland are
+by car \citep{national} . Cycling has the potential to replace a large
+proportion of these trips: ``A high priority must also be given to
+cyclists, because trips by this mode have the potential to replace trips
+by private car, most specifically for short to medium distance trips,
+but increasingly for longer trips as e-bikes extend the range of this
+mode'' \citep{greater}.
 
@@
 the tool ``revolutionised strategic cycle planning in England and
-Wales''\footnote{See REF Impact Case Study ``Cycle network policy,
-  planning and investment transformed by the Propensity to Cycle Tool''
-  at
-  \href{https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f}{results2021.ref.ac.uk}
-  submitted by the University of Leeds.} by overcoming the barriers to
-cycling investment imposed by lack of evidence on cycling
-potential\footnote{See the REF Impact Case Study ``Creating Step Changes
-  in Cycling Policy and Infrastructure Planning across the UK'' at
-  \href{https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B}{results2021.ref.ac.uk}
-  submitted by the University of Westminster.} .
+Wales'' by overcoming the barriers to cycling investment imposed by lack
+of evidence on cycling potential\citep{lovelace2023}.
 
@@
 \citep{felix2023}. biclaR includes estimates of impacts, using the World
-Health Organisation (WHO) `HEAT for Cycling' tool\footnote{See the
-  Health economic assessment tool for walking and cycling at
-  \href{https://www.heatwalkingcycling.org/\#homepage}{www.heatwalkingcycling.org}.}
-and an `intermodality' scenario that combines cycling with currently
-available public transit options based on General Transit Feed
-Specification (GTFS) data.
+Health Organisation (WHO) `HEAT for Cycling' tool and an `intermodality'
+scenario that combines cycling with currently available public transit
+options based on General Transit Feed Specification (GTFS) data.
 

### 2023-10-10 paper.qmd
Blinded review version

@@
 format: 
-  # html: default
+  html: default
   # docx: default
@@
 
-# # Uncomment the following line to produce the cover page
-author:
-  - name: Robin Lovelace
-    email: r.lovelace@leeds.ac.uk
-    corresponding: true
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Joey Talbot
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Eugeni Vidal-Tortosa
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Hussein Mahfouz
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Elaine Brick
-    affiliations:
-      - name: AECOM
-        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-  - name: Peter Wright
-    affiliations:
-      - name: AECOM
-        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
-  - name: "Gary O'Tool"
-    affiliations:
-      - name: AECOM
-        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-  - name: Dan Brennan
-    affiliations:
-      - name: Transport Infrastructure Ireland
-        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
-  - name: Suzanne Meade
-    affiliations:
-      - name: Transport Infrastructure Ireland
-        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+# # # Uncomment the following line to produce the cover page
+# author:
+#   - name: Robin Lovelace
+#     email: r.lovelace@leeds.ac.uk
+#     corresponding: true
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Joey Talbot
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Eugeni Vidal-Tortosa
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Hussein Mahfouz
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Elaine Brick
+#     affiliations:
+#       - name: AECOM
+#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+#   - name: Peter Wright
+#     affiliations:
+#       - name: AECOM
+#         address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
+#   - name: "Gary O'Tool"
+#     affiliations:
+#       - name: AECOM
+#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+#   - name: Dan Brennan
+#     affiliations:
+#       - name: Transport Infrastructure Ireland
+#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+#   - name: Suzanne Meade
+#     affiliations:
+#       - name: Transport Infrastructure Ireland
+#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
 abstract: |

### 2023-10-10 paper.tex
Blinded review version

@@
   pdftitle={CRUSE to Safe Cycling in Ireland},
-  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Tool; Dan Brennan; Suzanne Meade},
   pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
@@
 Support Active Travel} }
-\author[1]{Robin Lovelace%
-\corref{cor1}%
-}
- \ead{r.lovelace@leeds.ac.uk} 
-\author[1]{Joey Talbot%
-%
-}
-
-\author[1]{Eugeni Vidal-Tortosa%
-%
-}
-
-\author[1]{Hussein Mahfouz%
-%
-}
-
-\author[2]{Elaine Brick%
-%
-}
-
-\author[3]{Peter Wright%
-%
-}
-
-\author[2]{Gary O'Tool%
-%
-}
-
-\author[4]{Dan Brennan%
-%
-}
-
-\author[4]{Suzanne Meade%
-%
-}
-
-\affiliation[1]{organization={University of
-Leeds},addressline={Institute for Transport Studies, Leeds, LS2 9JT,
-United Kingdom},postcodesep={}}
-\affiliation[2]{organization={AECOM},addressline={Unit 6, Galway
-Technology Park, Parkmore, Galway, Ireland},postcodesep={}}
-\affiliation[3]{organization={AECOM},addressline={Winslade House,
-Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY,
-UK},postcodesep={}}
-\affiliation[4]{organization={Transport Infrastructure
-Ireland},addressline={Parkgate Business Centre, Parkgate Street, Dublin
-8, D08 DK10, Ireland},postcodesep={}}
 
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, frame hidden]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 

### 2023-10-10 paper.qmd
Add another reference

@@
 Although Dublin City has made progress, there are still many parts of the Fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
-According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians".[^1]
+According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians" [@walking2021].
 The results for the Dublin area could help prioritise investment in cycling infrastructure in the city.
-
-[^1]: https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22_DublinMetropolitanArea_v35_DIGITAL_v2.pdf
 

### 2023-10-10 paper.qmd
Anonymisation

@@
 Data was obtained from the Central Statistics Office (CSO) and Transport Infrastructure Ireland (TII) under license and cannot be shared publicly.
-The code used to generate the results presented in this paper is available at https://github.com/cruse-bike.
+The code used to generate the results presented in this paper is available 
+<!-- at https://github.com/cruse-bike. -->
+[To be added after paper is accepted.]
 
@@
 
-RL led the development of the CRUSE Tool, with input from JT, EV, HM, EB, PW, GOT, DB, and SM.
+[To be added after paper is accepted.]
+
+<!-- RL led the development of the CRUSE Tool, with input from JT, EV, HM, EB, PW, GOT, DB, and SM.
 SM instigated the project and coordinated its policy impact.
-PW, EB and GOT provided project management and transport engineering expertise.
+PW, EB and GOT provided project management and transport engineering expertise. -->

### 2023-10-10 paper.tex
Anonymisation

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, breakable, interior hidden, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 the Dublin Metropolitan Area support ``more cycle tracks along roads,
-physically separated from traffic and pedestrians''.\footnote{https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22\_DublinMetropolitanArea\_v35\_DIGITAL\_v2.pdf}
+physically separated from traffic and pedestrians'' \citep{walking2021}.
 The results for the Dublin area could help prioritise investment in

### 2023-10-10 paper.qmd
edits

@@
 High energy transport systems are a major contributor to climate change, a leading cause of premature death and injury due to road traffic collisions, and a cause of disease due to airborne and noise pollutants.
-Conversely, evidence-based and effective transport *policies* have great potential to decarbonize economies, improve public health, and save lives.
+Conversely, evidence-based and effective transport *policies* have great potential to decarbonise economies, improve public health, and save lives.
 
 Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport.
-Nearly nearly half of these emissions (around 10% of global emissions) are from passenger cars [@jaramillo2022].
+Nearly half of these emissions (around 10% of global emissions) are from passenger cars [@jaramillo2022].
 The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
@@
 Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance travelled, the directive requires estimation of distance travelled by mode, down to the road link level.
-For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
+For active modes, about which there is a paucity of data compared with motorised modes, this is a major challenge.
 Better data to inform road safety policies and interventions is a motivation for the CRUSE Tool outlined in this paper, and modeling active travel more broadly.
@@
 
-The method used to convert OD data to route networks used in the PCT was to calculate a single route between the population weighted centroids associated with each OD pair.
+In the PCT, the method used to convert OD data to route networks was to calculate a single route between the population weighted centroids of the zones associated with each OD pair.
 This method works fine when the OD data represents movement between small areas, but was not appropriate for generating route networks from the POWSCAR data because zone centroids are so far apart that the resulting route networks would be sparse and unrealistic.
@@
 
-Also based on the PCT scenario with the same name, the Eike scenario in CRUSE takes Go Dutch cycling uptake, and adds onto this the impact of increased ebike usage, which allows for longer cycle trips.
+Also based on the PCT scenario with the same name, the Ebike scenario in CRUSE takes Go Dutch cycling uptake, and adds onto this the impact of increased ebike usage, which allows for longer cycle trips.
 However, travel to primary and secondary schools still uses Go Dutch uptake, since no ebike scenario has been developed for school journeys, and children may be less likely to own ebikes than adults.
@@
 To illustrate the results in urban areas, we took the top 4 cities in Ireland by population: Dublin City, Cork, Limerick and Galway (with city populations of 0.6, 0.2, 0.1 and 0.1 million respectively).
-The results for central areas demarcated with circular areas with a radius of 3km for these cities are presented in @fig-city-results.
+In @fig-city-results we present the cycle route networks within a 3km radius circle centred on each of these cities.
 
@@
 The CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
-This provision of multiple scenarios of behaviour changes *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.
+This provision of multiple scenarios of behaviour change *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.
 
@@
 
-More broadly, the tool's outputs are limited to just one more of active travel (cycling), ignoring walking and wheeling, including wheelchair use and a range of wheeled devices such as scooters, bike trailers and e-cargo bikes that can be used to escort children to school.
+More broadly, the tool's outputs are limited to just one mode of active travel (cycling), ignoring walking and wheeling, including wheelchair use and a range of wheeled devices such as scooters, bike trailers and e-cargo bikes that can be used to escort children to school.
 Sustainable transport policies should plan for walking, wheeling and cycling, and there are many co-benefits of broadly defined active travel interventions that benefit all active modes, such as measures to reduce heavy motor traffic speeds and volumes in areas and along corridors with high active travel potential.

### 2023-10-10 paper.qmd
Minor tweaks

--- 2023-10-10 paper.qmd
@@
 format: 
-  html: default
+  # html: default
   # docx: default

### 2023-10-10 paper.tex
Minor tweaks

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, breakable, interior hidden, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, frame hidden, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 publicly. The code used to generate the results presented in this paper
-is available at https://github.com/cruse-bike.
+is available {[}To be added after paper is accepted.{]}
 
@@
 
-RL led the development of the CRUSE Tool, with input from JT, EV, HM,
-EB, PW, GOT, DB, and SM. SM instigated the project and coordinated its
-policy impact. PW, EB and GOT provided project management and transport
-engineering expertise.
+{[}To be added after paper is accepted.{]}
 

### 2023-10-10 paper.qmd
En-us, close #26 (#27)

--- 2023-10-10 paper.qmd
@@
 format: 
-  html: default
-  # docx: default
+  # html: default
+  docx: default
 # Uncomment for formatted version
@@
 High energy transport systems are a major contributor to climate change, a leading cause of premature death and injury due to road traffic collisions, and a cause of disease due to airborne and noise pollutants.
-Conversely, evidence-based and effective transport *policies* have great potential to decarbonise economies, improve public health, and save lives.
+Conversely, evidence-based and effective transport *policies* have great potential to decarbonize economies, improve public health, and save lives.
 
@@
 Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out" [@directiv2008].
-Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance travelled, the directive requires estimation of distance travelled by mode, down to the road link level.
-For active modes, about which there is a paucity of data compared with motorised modes, this is a major challenge.
+Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance traveled, the directive requires estimation of distance traveled by mode, down to the road link level.
+For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
 Better data to inform road safety policies and interventions is a motivation for the CRUSE Tool outlined in this paper, and modeling active travel more broadly.
@@
 An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
-The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales" by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[@lovelace2023].
+The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionized strategic cycle planning in England and Wales" by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[@lovelace2023].
 
@@
 It was launched in April 2017 as the government-endorsed tool for strategic cycle network planning and part of the Cycling and Walking Investment Strategy [@cycling2017].
-Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modelling of impacts on health, environmental and distributional outcomes [@woodcock2021].
+Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modeling of impacts on health, environmental and distributional outcomes [@woodcock2021].
 Initially developed just for England, the PCT was extended to cover all of Wales (for commuter data only) in 2018.
@@
 However, datasets on travel patterns are often only available at the level of administrative zones.
-The Central Statistics Office (CSO) in Ireland provides Place of Work, School or College Census of Anonymised Records ([POWSCAR](https://www.cso.ie/en/census/census2016reports/powscar/)) data on the number of people travelling to work and school at the Electoral Division (ED) level, for example.
+The Central Statistics Office (CSO) in Ireland provides Place of Work, School or College Census of Anonymized Records ([POWSCAR](https://www.cso.ie/en/census/census2016reports/powscar/)) data on the number of people traveling to work and school at the Electoral Division (ED) level, for example.
 
@@
 However, travel to school and work only constitute around only 30% of all trips in Ireland.
-To overcome this issue, we developed a spatial interaction modelling methodology to estimate the number of trips between each OD pair for additional trip purposes.
+To overcome this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
 The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
@@
 We calculated three network types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest'.
-As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the 'Fastest' routes minimise journey time, accounting for traffic lights and surface type.
-The 'Quietest' route type minimises busy sections of road, allowing high 'diversion factors' from the fastest route on quieter but less direct ways (often avoiding roads and interactions with motor vehicles altogether where possible).
-The 'Balanced' route type is a compromise between the two, minimising journey time while avoiding the busiest roads.
+As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the 'Fastest' routes minimize journey time, accounting for traffic lights and surface type.
+The 'Quietest' route type minimizes busy sections of road, allowing high 'diversion factors' from the fastest route on quieter but less direct ways (often avoiding roads and interactions with motor vehicles altogether where possible).
+The 'Balanced' route type is a compromize between the two, minimising journey time while avoiding the busiest roads.
 Allowing users to switch between these three route types enables them to consider the trade-offs between directness and 'cycle friendliness' when planning new infrastructure, as shown in @fig-route-types.
@@
 As outlined in @sec-trip-purposes, the Baseline scenario includes travel to work and school, and additional trip purposes.
-Cycling levels were taken from the POWSCAR data, which includes the number of people travelling by each mode between each OD pair.
+Cycling levels were taken from the POWSCAR data, which includes the number of people traveling by each mode between each OD pair.
 
@@
 
--   Calculate distance decay curves for Dublin for the base year (2016, using POWSCAR data) by fitting a model to the relevant OD data after it has been converted to a route network datasest
+-   Calculate distance decay curves for Dublin for the base year (2016, using POWSCAR data) by fitting a model to the relevant OD data after it has been converted to a route network dataset
 -   Apply the Near Market model to the hilliness and distance values for each county during the build process
@@
 For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.
-In terms of car travel, the target is to "Increas\[e\] the proportion of kilometres driven by passenger electric cars to between 40 and 45% by 2030, in addition to a reduction of 10% in kilometres driven by the remaining internal combustion engine cars." This equates to a 5.5 to 6% reduction in total car km driven.
+In terms of car travel, the target is to "Increas\[e\] the proportion of kilometers driven by passenger electric cars to between 40 and 45% by 2030, in addition to a reduction of 10% in kilometers driven by the remaining internal combustion engine cars." This equates to a 5.5 to 6% reduction in total car km driven.
 
@@
 Aside from the widely varying sizes and route networks of each city, with Dublin having by far the densest network, it is clear from the visualisation of cycling friendliness that there is a lot of room for improvement in all cities.
-Dublin has had received most investment in cycling infrastructure and this is apparent from the high proportion of the route network that is green, even in the 'Fastest' route network representing routes that prioritise speed and directness over comfort and safety.
+Dublin has had received most investment in cycling infrastructure and this is apparent from the high proportion of the route network that is green, even in the 'Fastest' route network representing routes that prioritize speed and directness over comfort and safety.
 Dublin also has the highest mode share of cycling of the four cities, adding to the evidence base showing that investment in cycling infrastructure is effective in increasing cycling uptake.
@@
 -   The results are *open access* meaning that any stakeholder in the planning system can access the evidence.
-This will help to democratise the transport planning process and make wider conversations about transport planning more evidence-based and less polarised [@lovelace2020], something that is particularly relevant given the potentially polarising nature of pro-cycling interventions [@wild2017].
+This will help to democratize the transport planning process and make wider conversations about transport planning more evidence-based and less polarized [@lovelace2020], something that is particularly relevant given the potentially polarizing nature of pro-cycling interventions [@wild2017].
 -   The results are fully reproducible (code to be released pending sign-off by TII's IT team), preventing 'cloud lock in' to a potentially monopolistic consultancy, and encouraging input from the wider open source community [@lovelace2021; @dhir2017].
--   By covering a wide range of trip purposes --- not just travel to work [@lovelace2017; @heinen2010] and travel school [@goodman2019] as covered in previously published research on open access national cycle netework planning tools --- the results capture a higher proportion of cycling potential including key rural trips which are often under-represented in models of active travel.
+-   By covering a wide range of trip purposes --- not just travel to work [@lovelace2017; @heinen2010] and travel school [@goodman2019] as covered in previously published research on open access national cycle network planning tools --- the results capture a higher proportion of cycling potential including key rural trips which are often under-represented in models of active travel.
 
@@
 Building on previous work, it provides a nationally consistent evidence base, providing valuable insights to planners and other stakeholders in the transport planning process, at national and local levels.
-Because the results are available at the route segment level, the tool can be used to identify 'weak links' in the cycle network, and to prioritise investment in cycling infrastructure.
+Because the results are available at the route segment level, the tool can be used to identify 'weak links' in the cycle network, and to prioritize investment in cycling infrastructure.
 Furthermore, the provision of the evidence in a free and publicly available website, hosted at [cruse.bike](https://cruse.bike), means that it can be used by anyone, encouraging wider participation and more evidence-based debate about transport planning.
@@
 Sustainable transport policies should plan for walking, wheeling and cycling, and there are many co-benefits of broadly defined active travel interventions that benefit all active modes, such as measures to reduce heavy motor traffic speeds and volumes in areas and along corridors with high active travel potential.
-This raises the question of whether other active modes should be incorporated into the results using the OD-based approach outlined in this paper, or whether different modelling approaches are needed to properly capture the shorter distance trips typically made by walking [@cooper2018].
+This raises the question of whether other active modes should be incorporated into the results using the OD-based approach outlined in this paper, or whether different modeling approaches are needed to properly capture the shorter distance trips typically made by walking [@cooper2018].
 Furthermore, the estimates of cycling presented in the CRUSE Tool omit multi-modal trips including public transport, and omit trip chaining, due to the need to capture a large portion of cycling potential within the resource constraints of the project.
@@
 
-OD: Origin-Destination, typically referring to origin-destination data which contains information on the number of people travelling between each pair of zones
+OD: Origin-Destination, typically referring to origin-destination data which contains information on the number of people traveling between each pair of zones
 
@@
 
-POWSCAR: Place of Work, School or College Census of Anonymised Records
+POWSCAR: Place of Work, School or College Census of Anonymized Records
 
@@
 Data was obtained from the Central Statistics Office (CSO) and Transport Infrastructure Ireland (TII) under license and cannot be shared publicly.
-The code used to generate the results presented in this paper is available 
-<!-- at https://github.com/cruse-bike. -->
-[To be added after paper is accepted.]
+The code used to generate the results presented in this paper is available <!-- at https://github.com/cruse-bike. --> \[To be added after paper is accepted.\]
 
@@
 
-[To be added after paper is accepted.]
+\[To be added after paper is accepted.\]
 
+```{=html}
 <!-- RL led the development of the CRUSE Tool, with input from JT, EV, HM, EB, PW, GOT, DB, and SM.
@@
 PW, EB and GOT provided project management and transport engineering expertise. -->
+```

### 2023-10-10 paper.qmd
Incorporate changes from @eugenividal

@@
   # html: default
-  docx: default
+  # docx: default
 # Uncomment for formatted version
@@
   Legislation mandates improved road safety and decarbonisation of predominantly road-based transport network in Ireland and other countries.
-The Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was commissioned to support cycle network planning, safety interventions, business case development and project evaluation.
+Increasing cycling transport, making it safer and more attractive, is essential to this end.
 Cycling in Ireland represents only 3% of total modal share as of the latest Census (2016), but accounts for 20% of serious injuries and 7% of all fatalities.
-  High resolution datasets on baseline and potential cycling levels are needed to meet Ireland's climate and safety targets.
+A planning tool based on high resolution datasets on baseline and potential cycling levels could greatly assist in achieving this goal.
 
-  The CRUSE Tool is based on open source software and reproducible data science.
+  This paper describes the design, features and potential use of the Cycle Route Acceptance and Scenario Estimation (CRUSE) Tool for Ireland was commissioned to support cycle network planning, safety interventions, business cases and project evaluation.
 The resulting web application is available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions.
@@
 
-Regarding road traffic casualties, another deadly consequence of inefficient transport systems, the Road Infrastructure Safety Management (RISM) directive (2008/96/EC) requires member states to implement a road safety management system (RSMS) for all public roads.
+Another deadly consequence of inefficient transport systems is tackled in the Road Infrastructure Safety Management (RISM) directive (2008/96/EC), which requires member states to implement a road safety management system (RSMS) for all public roads.
 Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out" [@directiv2008].
@@
 At the regional level, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
-Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, this suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
+Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
 7 in 10 trips in Ireland are by car [@national] .
@@
 
-It was in this context that Transport Infrastructure Ireland (TII) commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) project.
+It was in this context that Transport Infrastructure Ireland (TII) commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
 Based on the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE Tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
@@
 This paper aims to describe the design, features and potential use of the CRUSE Tool.
-In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool for Ireland.
+In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool.
 In @sec-results, we present the results of the CRUSE Tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
@@
 
-The methods used to generate the evidence presented in the CRUSE Tool for Ireland build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
+The methods used to generate the evidence presented in the CRUSE Tool build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
 An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
@@
 
-CycleStreets' services can be used by anyone for free, but we commissioned a bespoke routing service for CRUSE for the following reasons, to enable:
+CycleStreets' services can be used by anyone for free, but we commissioned a bespoke routing service for CRUSE, to enable:
 
@@
 The datasets outlined above were used to generate estimates of cycling currently (the 'Baseline' scenario) and under four scenarios of cycling potential: Near Market, Climate Action Plan, Go Dutch, and Ebike.
-Each of these scenarios is described below, as illustrated in Figure @fig-scenarios.
+Each of these scenarios is described below, as illustrated in @fig-scenarios.
 <!-- TODO: add table. -->
@@
 
-The main result of the CRUSE project is an open access web application and datasets on current and potential future cycling levels in Ireland, with evidence provided at county and route segment levels.
+The main result of the work presented in this paper is an open access web application and datasets on current and potential future cycling levels in Ireland, with evidence provided at county and route segment levels.
 Like the PCT on which CRUSE builds, a key feature of the results is that they are systematic and available for all counties, and most routes on which cycling is permitted, in Ireland.
@@
 Fastest, balanced and quietest cycle networks within 3km circular areas containing the four largest cities in Ireland.
-The networks are coloured by cycle friendliness and with widths varying in proportion to the potential level of cycling under the 'Go Dutch' scenario.
+The networks are coloured by cycle friendliness, with greener segments representing more cycle friendly segments.
+Widths vary in proportion to the potential level of cycling under the 'Go Dutch' scenario.
 :::
 
-Aside from the widely varying sizes and route networks of each city, with Dublin having by far the densest network, it is clear from the visualisation of cycling friendliness that there is a lot of room for improvement in all cities.
+Aside from the widely varying sizes and route networks of each city, with Dublin having by far the densest network, it is clear from the visualisation of cycling friendliness that there are many gaps in the networks.
 Dublin has had received most investment in cycling infrastructure and this is apparent from the high proportion of the route network that is green, even in the 'Fastest' route network representing routes that prioritize speed and directness over comfort and safety.
@@
 
-The CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
+The aim of the paper was to describe the design, features and potential use of the CRUSE Tool.
+As outlined in the previous section, the CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
 This provision of multiple scenarios of behaviour change *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.
@@
 
-The CRUSE Tool can help both in terms of describing the current situation, and also prioritise investment in those direct routes with high cycling potential that lack adequate cycling infrastructure, highlighted by @caulfield2012.
+The CRUSE Tool can help both in terms of describing the current situation, and also prioritise investment in those direct routes with high cycling potential that lack adequate cycling infrastructure [@caulfield2012].
 It will also support reporting for the RISM Directive on road casualties as part of the road safety management system.
@@
 
-The CRUSE Tool is an open access web application for strategic cycle network planning and prioritisation of road safety interventions across Ireland.
+The CRUSE Tool is an open access web application for strategic cycle network planning and prioritization of road safety interventions across Ireland.
 Building on previous work, it provides a nationally consistent evidence base, providing valuable insights to planners and other stakeholders in the transport planning process, at national and local levels.
@@
 # List of abbreviations
+
+APIs: Application Programming Interfaces
 

### 2023-10-10 paper.tex
Incorporate changes from @eugenividal

@@
 predominantly road-based transport network in Ireland and other
-countries. The Cycle Route Uptake and Scenario Estimation (CRUSE) Tool
-was commissioned to support cycle network planning, safety
-interventions, business case development and project evaluation. Cycling
-in Ireland represents only 3\% of total modal share as of the latest
-Census (2016), but accounts for 20\% of serious injuries and 7\% of all
-fatalities. High resolution datasets on baseline and potential cycling
-levels are needed to meet Ireland's climate and safety targets.
-
-The CRUSE Tool is based on open source software and reproducible data
-science. The resulting web application is available at
-\url{https://cruse.bike/}, enabling planners, engineers, and other
-stakeholders to make more evidence-based decisions. CRUSE goes beyond
-previous work in several important ways. It models networks at high
-levels of geographic resolution, simulates a wide range of trips
+countries. Increasing cycling transport, making it safer and more
+attractive, is essential to this end. Cycling in Ireland represents only
+3\% of total modal share as of the latest Census (2016), but accounts
+for 20\% of serious injuries and 7\% of all fatalities. A planning tool
+based on high resolution datasets on baseline and potential cycling
+levels could greatly assist in achieving this goal.
+
+This paper describes the design, features and potential use of the Cycle
+Route Acceptance and Scenario Estimation (CRUSE) Tool for Ireland was
+commissioned to support cycle network planning, safety interventions,
+business cases and project evaluation. The resulting web application is
+available at \url{https://cruse.bike/}, enabling planners, engineers,
+and other stakeholders to make more evidence-based decisions. CRUSE goes
+beyond previous work in several important ways. It models networks at
+high levels of geographic resolution, simulates a wide range of trips
 including recreational, social, shopping and personal trips, in addition
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, frame hidden, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, frame hidden, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 Transport is responsible for 23\% of global greenhouse gas (GHG)
-emissions, 70\% of which are from road transport. Nearly nearly half of
-these emissions (around 10\% of global emissions) are from passenger
-cars \citep{jaramillo2022}. The transport system encourages, enables and
-in some cases enforces unsustainable lifestyles. Services that are only
+emissions, 70\% of which are from road transport. Nearly half of these
+emissions (around 10\% of global emissions) are from passenger cars
+\citep{jaramillo2022}. The transport system encourages, enables and in
+some cases enforces unsustainable lifestyles. Services that are only
 accessible by car lock-in car dependency
@@
 
-Regarding road traffic casualties, another deadly consequence of
-inefficient transport systems, the Road Infrastructure Safety Management
-(RISM) directive (2008/96/EC) requires member states to implement a road
-safety management system (RSMS) for all public roads. Specifically,
-``Member States shall ensure that the ranking of high accident
-concentration sections and the network safety ranking are carried out''
+Another deadly consequence of inefficient transport systems is tackled
+in the Road Infrastructure Safety Management (RISM) directive
+(2008/96/EC), which requires member states to implement a road safety
+management system (RSMS) for all public roads. Specifically, ``Member
+States shall ensure that the ranking of high accident concentration
+sections and the network safety ranking are carried out''
 \citep{directiv2008}. Given that `safety' in this context is usefully
 quantified as the number of people killed and seriously injured (KSI)
-per distance travelled, the directive requires estimation of distance
-travelled by mode, down to the road link level. For active modes, about
+per distance traveled, the directive requires estimation of distance
+traveled by mode, down to the road link level. For active modes, about
 which there is a paucity of data compared with motorized modes, this is
@@
 road today. Extrapolating this on a per population basis across Ireland,
-with around 40\% of the population living in Dublin, this suggests that
+with around 40\% of the population living in Dublin, suggests that
 around 150,000 cars could be removed nationwide just by achieving Dublin
@@
 It was in this context that Transport Infrastructure Ireland (TII)
-commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE)
-project. Based on the Propensity to Cycle Tool (PCT) for England and
+commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool
+for Ireland. Based on the Propensity to Cycle Tool (PCT) for England and
 Wales, the CRUSE Tool was developed to provide evidence on current
@@
 the CRUSE Tool. In Section~\ref{sec-methods}, we outline the methods
-used to generate the evidence presented in the CRUSE Tool for Ireland.
-In Section~\ref{sec-results}, we present the results of the CRUSE Tool,
+used to generate the evidence presented in the CRUSE Tool. In
+Section~\ref{sec-results}, we present the results of the CRUSE Tool,
 including estimates of current cycling levels and future cycling
@@
 The methods used to generate the evidence presented in the CRUSE Tool
-for Ireland build on the Propensity to Cycle Tool (PCT), which was
-originally funded by the UK's Department for Transport and developed by
-a multi-university team. An important feature of the PCT is that it is
-open source and publicly available (at
-\href{https://www.pct.bike/}{www.pct.bike}), allowing its use by all
-stakeholders in the transport planning process \citep{lovelace2017}. The
-PCT approach has had major policy impacts, as outlined in Research
-Excellence Framework (REF) impact case studies, which demonstrate that
-the tool ``revolutionised strategic cycle planning in England and
-Wales'' by overcoming the barriers to cycling investment imposed by lack
-of evidence on cycling potential\citep{lovelace2023}.
+build on the Propensity to Cycle Tool (PCT), which was originally funded
+by the UK's Department for Transport and developed by a multi-university
+team. An important feature of the PCT is that it is open source and
+publicly available (at \href{https://www.pct.bike/}{www.pct.bike}),
+allowing its use by all stakeholders in the transport planning process
+\citep{lovelace2017}. The PCT approach has had major policy impacts, as
+outlined in Research Excellence Framework (REF) impact case studies,
+which demonstrate that the tool ``revolutionized strategic cycle
+planning in England and Wales'' by overcoming the barriers to cycling
+investment imposed by lack of evidence on cycling
+potential\citep{lovelace2023}.
 
@@
 \citep{woodcock2018}, addition of travel to school network
-\citep{goodman2019}, and improved modelling of impacts on health,
+\citep{goodman2019}, and improved modeling of impacts on health,
 environmental and distributional outcomes \citep{woodcock2021}.
@@
 Statistics Office (CSO) in Ireland provides Place of Work, School or
-College Census of Anonymised Records
+College Census of Anonymized Records
 (\href{https://www.cso.ie/en/census/census2016reports/powscar/}{POWSCAR})
-data on the number of people travelling to work and school at the
+data on the number of people traveling to work and school at the
 Electoral Division (ED) level, for example.
 
-The method used to convert OD data to route networks used in the PCT was
-to calculate a single route between the population weighted centroids
-associated with each OD pair. This method works fine when the OD data
-represents movement between small areas, but was not appropriate for
-generating route networks from the POWSCAR data because zone centroids
-are so far apart that the resulting route networks would be sparse and
-unrealistic. To tackle this issue we developed a new method for OD data
-disaggregation called `jittering' \citep{lovelace2022}. The method works
-by first disaggregating the OD data based on a `disaggregation
-threshold' and then assigning each disaggregated `sub-OD' pair to
-`subpoints' within each zone. As shown in Figure~\ref{fig-dublin}, the
-resulting route networks are dense, even in rural areas.
+In the PCT, the method used to convert OD data to route networks was to
+calculate a single route between the population weighted centroids of
+the zones associated with each OD pair. This method works fine when the
+OD data represents movement between small areas, but was not appropriate
+for generating route networks from the POWSCAR data because zone
+centroids are so far apart that the resulting route networks would be
+sparse and unrealistic. To tackle this issue we developed a new method
+for OD data disaggregation called `jittering' \citep{lovelace2022}. The
+method works by first disaggregating the OD data based on a
+`disaggregation threshold' and then assigning each disaggregated
+`sub-OD' pair to `subpoints' within each zone. As shown in
+Figure~\ref{fig-dublin}, the resulting route networks are dense, even in
+rural areas.
 
@@
 all trips in Ireland. To overcome this issue, we developed a spatial
-interaction modelling methodology to estimate the number of trips
-between each OD pair for additional trip purposes. The classification of
-trip purposes used in the CRUSE Tool was guided mainly by the trip
-purpose classification found within the National Household Travel Survey
-(NHTS), but with the addition of categories based on the comprehensive
-POWSCAR data, and the need to include recreational trips and multi-stage
-trips (not yet implemented). An overview of the trip purposes used in
-CRUSE is presented in Figure~\ref{fig-trip-purposes}.
+interaction modeling methodology to estimate the number of trips between
+each OD pair for additional trip purposes. The classification of trip
+purposes used in the CRUSE Tool was guided mainly by the trip purpose
+classification found within the National Household Travel Survey (NHTS),
+but with the addition of categories based on the comprehensive POWSCAR
+data, and the need to include recreational trips and multi-stage trips
+(not yet implemented). An overview of the trip purposes used in CRUSE is
+presented in Figure~\ref{fig-trip-purposes}.
 
@@
 CycleStreets' services can be used by anyone for free, but we
-commissioned a bespoke routing service for CRUSE for the following
-reasons, to enable:
+commissioned a bespoke routing service for CRUSE, to enable:
 
@@
 \href{https://www.cyclestreets.net/help/journey/howitworks/}{website},
-the `Fastest' routes minimise journey time, accounting for traffic
-lights and surface type. The `Quietest' route type minimises busy
+the `Fastest' routes minimize journey time, accounting for traffic
+lights and surface type. The `Quietest' route type minimizes busy
 sections of road, allowing high `diversion factors' from the fastest
@@
 interactions with motor vehicles altogether where possible). The
-`Balanced' route type is a compromise between the two, minimising
+`Balanced' route type is a compromize between the two, minimising
 journey time while avoiding the busiest roads. Allowing users to switch
@@
 potential: Near Market, Climate Action Plan, Go Dutch, and Ebike. Each
-of these scenarios is described below, as illustrated in Figure
+of these scenarios is described below, as illustrated in
 Figure~\ref{fig-scenarios}.
@@
 were taken from the POWSCAR data, which includes the number of people
-travelling by each mode between each OD pair.
+traveling by each mode between each OD pair.
 
@@
   using POWSCAR data) by fitting a model to the relevant OD data after
-  it has been converted to a route network datasest
+  it has been converted to a route network dataset
 \item
@@
 public transport trips per day by 2030. In terms of car travel, the
-target is to ``Increas{[}e{]} the proportion of kilometres driven by
+target is to ``Increas{[}e{]} the proportion of kilometers driven by
 passenger electric cars to between 40 and 45\% by 2030, in addition to a
-reduction of 10\% in kilometres driven by the remaining internal
+reduction of 10\% in kilometers driven by the remaining internal
 combustion engine cars.'' This equates to a 5.5 to 6\% reduction in
@@
 
-Also based on the PCT scenario with the same name, the Eike scenario in
+Also based on the PCT scenario with the same name, the Ebike scenario in
 CRUSE takes Go Dutch cycling uptake, and adds onto this the impact of
@@
 
-The main result of the CRUSE project is an open access web application
-and datasets on current and potential future cycling levels in Ireland,
-with evidence provided at county and route segment levels. Like the PCT
-on which CRUSE builds, a key feature of the results is that they are
-systematic and available for all counties, and most routes on which
-cycling is permitted, in Ireland. The full set of results are therefore
-too extensive to present in their entirety in this paper. Through the
-interactive web application, users can explore the data to generate the
-results that are most relevant to their needs, whether that is finding
-the cycling potential on a particular road or finding `weak links' or
-barriers in the cycle network associated with a particular school, work
-place or other destination.
+The main result of the work presented in this paper is an open access
+web application and datasets on current and potential future cycling
+levels in Ireland, with evidence provided at county and route segment
+levels. Like the PCT on which CRUSE builds, a key feature of the results
+is that they are systematic and available for all counties, and most
+routes on which cycling is permitted, in Ireland. The full set of
+results are therefore too extensive to present in their entirety in this
+paper. Through the interactive web application, users can explore the
+data to generate the results that are most relevant to their needs,
+whether that is finding the cycling potential on a particular road or
+finding `weak links' or barriers in the cycle network associated with a
+particular school, work place or other destination.
 
@@
 Ireland by population: Dublin City, Cork, Limerick and Galway (with city
-populations of 0.6, 0.2, 0.1 and 0.1 million respectively). The results
-for central areas demarcated with circular areas with a radius of 3km
-for these cities are presented in Figure~\ref{fig-city-results}.
+populations of 0.6, 0.2, 0.1 and 0.1 million respectively). In
+Figure~\ref{fig-city-results} we present the cycle route networks within
+a 3km radius circle centred on each of these cities.
 
@@
 networks within 3km circular areas containing the four largest cities in
-Ireland. The networks are coloured by cycle friendliness and with widths
-varying in proportion to the potential level of cycling under the `Go
-Dutch' scenario.}
+Ireland. The networks are coloured by cycle friendliness, with greener
+segments representing more cycle friendly segments. Widths vary in
+proportion to the potential level of cycling under the `Go Dutch'
+scenario.}
 
@@
 with Dublin having by far the densest network, it is clear from the
-visualisation of cycling friendliness that there is a lot of room for
-improvement in all cities. Dublin has had received most investment in
-cycling infrastructure and this is apparent from the high proportion of
-the route network that is green, even in the `Fastest' route network
-representing routes that prioritise speed and directness over comfort
+visualisation of cycling friendliness that there are many gaps in the
+networks. Dublin has had received most investment in cycling
+infrastructure and this is apparent from the high proportion of the
+route network that is green, even in the `Fastest' route network
+representing routes that prioritize speed and directness over comfort
 and safety. Dublin also has the highest mode share of cycling of the
@@
 
-The CRUSE Tool provides evidence on current and potential future cycling
-levels across Ireland down to the street level, with potential assigned
-to `fastest', `balanced' and `quietest' route networks. This provision
-of multiple scenarios of behaviour changes \emph{and} multiple scenarios
-of investment, for example in cycling infrastructure next to major roads
-vs quiet residential streets, is a key feature of the CRUSE Tool.
+The aim of the paper was to describe the design, features and potential
+use of the CRUSE Tool. As outlined in the previous section, the CRUSE
+Tool provides evidence on current and potential future cycling levels
+across Ireland down to the street level, with potential assigned to
+`fastest', `balanced' and `quietest' route networks. This provision of
+multiple scenarios of behaviour change \emph{and} multiple scenarios of
+investment, for example in cycling infrastructure next to major roads vs
+quiet residential streets, is a key feature of the CRUSE Tool.
 
@@
 situation, and also prioritise investment in those direct routes with
-high cycling potential that lack adequate cycling infrastructure,
-highlighted by \citet{caulfield2012}. It will also support reporting for
-the RISM Directive on road casualties as part of the road safety
-management system.
+high cycling potential that lack adequate cycling infrastructure
+\citep{caulfield2012}. It will also support reporting for the RISM
+Directive on road casualties as part of the road safety management
+system.
 
@@
   The results are \emph{open access} meaning that any stakeholder in the
-  planning system can access the evidence. This will help to democratise
+  planning system can access the evidence. This will help to democratize
   the transport planning process and make wider conversations about
-  transport planning more evidence-based and less polarised
+  transport planning more evidence-based and less polarized
   \citep{lovelace2020}, something that is particularly relevant given
-  the potentially polarising nature of pro-cycling interventions
+  the potentially polarizing nature of pro-cycling interventions
   \citep{wild2017}.
@@
   as covered in previously published research on open access national
-  cycle netework planning tools --- the results capture a higher
+  cycle network planning tools --- the results capture a higher
   proportion of cycling potential including key rural trips which are
@@
 The CRUSE Tool is an open access web application for strategic cycle
-network planning and prioritisation of road safety interventions across
+network planning and prioritization of road safety interventions across
 Ireland. Building on previous work, it provides a nationally consistent
@@
 the tool can be used to identify `weak links' in the cycle network, and
-to prioritise investment in cycling infrastructure. Furthermore, the
+to prioritize investment in cycling infrastructure. Furthermore, the
 provision of the evidence in a free and publicly available website,
@@
 
-More broadly, the tool's outputs are limited to just one more of active
+More broadly, the tool's outputs are limited to just one mode of active
 travel (cycling), ignoring walking and wheeling, including wheelchair
@@
 active modes should be incorporated into the results using the OD-based
-approach outlined in this paper, or whether different modelling
+approach outlined in this paper, or whether different modeling
 approaches are needed to properly capture the shorter distance trips
@@
 
+APIs: Application Programming Interfaces
+
 CSO: Central Statistics Office
@@
 OD: Origin-Destination, typically referring to origin-destination data
-which contains information on the number of people travelling between
+which contains information on the number of people traveling between
 each pair of zones

### 2023-10-10 paper.qmd
Add new regional cycle network citation

@@
   For investment in cycling to deliver on the huge potential benefits, cycling must become safer.
-  With growth in the E-bike market, CRUSE will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
+  With growth in the ebike market, CRUSE will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
   Provision of open access data in a publicly available web tool will address the challenges faced by counties across Ireland in meeting their sustainability commitments.
@@
 
-Uptake of cycling should be seen in this broader context of transport decarbonisation [@brand2020] and sustainable mobility [@burns2013].
+Building cycle networks represents a relatively 'quick win' within the context of decarbonization [@brand2020] and sustainable mobility [@burns2013].
 Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
-Furthermore, highly efficient e-bikes --- which are seeing rapid uptake in Ireland and other countries --- outperform electric cars, which are too heavy and expensive.
+Furthermore, highly efficient ebikes --- which are seeing rapid uptake in Ireland and other countries --- outperform electric cars, which are too heavy and expensive.
 Over-reliance on electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
 At the European level, the European Union has a target of reducing GHG emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050 [@rosenow2022].
+Climate change mitigation is a major motivation for cycle network plans [@scappini2022].
 
-Another deadly consequence of inefficient transport systems is tackled in the Road Infrastructure Safety Management (RISM) directive (2008/96/EC), which requires member states to implement a road safety management system (RSMS) for all public roads.
+Another motivation for cycle network planning is the Road Infrastructure Safety Management (RISM) directive (2008/96/EC), which requires member states to implement a road safety management system (RSMS) for all public roads.
 Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out" [@directiv2008].
@@
 7 in 10 trips in Ireland are by car [@national] .
-Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as e-bikes extend the range of this mode" [@greater].
+Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as ebikes extend the range of this mode" [@greater].
 

### 2023-10-10 paper.qmd
Final tweaks

@@
 Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
-
 A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
@@
 As outlined in the previous section, the CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
-This provision of multiple scenarios of behaviour change *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.
+This provision of multiple scenarios of behavior change *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.
 
@@
 
-While the CRUSE Tool is not the first national and publicly available tool for cycling planning, it has some key features that make it relevant for other countries, and other national road authorities tasked with making their transport systems safe:
+While the CRUSE Tool is not the first national and publicly available tool for cycling planning, it has some key features that make it relevant for other countries, regions and road authorities tasked with making their transport systems safe and sustainable:
 
@@
 
+As the benefits of strategic planning for cycling become more apparent [@scappini2022] we expect the demand for tools such as CRUSE to grow.
+
 # Conclusions {#sec-conclusions}
@@
 Building on previous work, it provides a nationally consistent evidence base, providing valuable insights to planners and other stakeholders in the transport planning process, at national and local levels.
-Because the results are available at the route segment level, the tool can be used to identify 'weak links' in the cycle network, and to prioritize investment in cycling infrastructure.
+Because the results are available at the route segment level, the tool can be used to identify 'weak links' in the cycle network [@vybornova2022], and to prioritize investment in cycling infrastructure to maximize health, equality, and other benefits [@mahfouz; @woodcock2021].
 Furthermore, the provision of the evidence in a free and publicly available website, hosted at [cruse.bike](https://cruse.bike), means that it can be used by anyone, encouraging wider participation and more evidence-based debate about transport planning.

### 2023-10-10 paper.tex
Final tweaks

@@
 For investment in cycling to deliver on the huge potential benefits,
-cycling must become safer. With growth in the E-bike market, CRUSE will
+cycling must become safer. With growth in the ebike market, CRUSE will
 help inform inter-urban and rural networks to support the transfer of
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, frame hidden, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden, sharp corners, breakable, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-Uptake of cycling should be seen in this broader context of transport
-decarbonisation \citep{brand2020} and sustainable mobility
+Building cycle networks represents a relatively `quick win' within the
+context of decarbonization \citep{brand2020} and sustainable mobility
 \citep{burns2013}. Although cycling uptake appears on the surface to
@@
 can also help avoid unnecessary trips \citep{nello-deakin2020}.
-Furthermore, highly efficient e-bikes --- which are seeing rapid uptake
+Furthermore, highly efficient ebikes --- which are seeing rapid uptake
 in Ireland and other countries --- outperform electric cars, which are
@@
 to 1990 levels, and to achieve `net-zero' by 2050 \citep{rosenow2022}.
-
-Another deadly consequence of inefficient transport systems is tackled
-in the Road Infrastructure Safety Management (RISM) directive
-(2008/96/EC), which requires member states to implement a road safety
-management system (RSMS) for all public roads. Specifically, ``Member
-States shall ensure that the ranking of high accident concentration
-sections and the network safety ranking are carried out''
-\citep{directiv2008}. Given that `safety' in this context is usefully
-quantified as the number of people killed and seriously injured (KSI)
-per distance traveled, the directive requires estimation of distance
-traveled by mode, down to the road link level. For active modes, about
-which there is a paucity of data compared with motorized modes, this is
-a major challenge. Better data to inform road safety policies and
-interventions is a motivation for the CRUSE Tool outlined in this paper,
-and modeling active travel more broadly.
+Climate change mitigation is a major motivation for cycle network plans
+\citep{scappini2022}.
+
+Another motivation for cycle network planning is the Road Infrastructure
+Safety Management (RISM) directive (2008/96/EC), which requires member
+states to implement a road safety management system (RSMS) for all
+public roads. Specifically, ``Member States shall ensure that the
+ranking of high accident concentration sections and the network safety
+ranking are carried out'' \citep{directiv2008}. Given that `safety' in
+this context is usefully quantified as the number of people killed and
+seriously injured (KSI) per distance traveled, the directive requires
+estimation of distance traveled by mode, down to the road link level.
+For active modes, about which there is a paucity of data compared with
+motorized modes, this is a major challenge. Better data to inform road
+safety policies and interventions is a motivation for the CRUSE Tool
+outlined in this paper, and modeling active travel more broadly.
 
@@
 uptake of active modes will require key active travel routes to be
-identified and improved.
-
-A proactive approach to the development of cycling policy and
-infrastructure is taking place in Ireland. The main frameworks
-underpinning these efforts are the Climate Action Plan
+identified and improved. A proactive approach to the development of
+cycling policy and infrastructure is taking place in Ireland. The main
+frameworks underpinning these efforts are the Climate Action Plan
 \citep{climate2022a}, the National Development Plan, and the National
@@
 by private car, most specifically for short to medium distance trips,
-but increasingly for longer trips as e-bikes extend the range of this
+but increasingly for longer trips as ebikes extend the range of this
 mode'' \citep{greater}.
@@
 `fastest', `balanced' and `quietest' route networks. This provision of
-multiple scenarios of behaviour change \emph{and} multiple scenarios of
+multiple scenarios of behavior change \emph{and} multiple scenarios of
 investment, for example in cycling infrastructure next to major roads vs
@@
 tool for cycling planning, it has some key features that make it
-relevant for other countries, and other national road authorities tasked
-with making their transport systems safe:
+relevant for other countries, regions and road authorities tasked with
+making their transport systems safe and sustainable:
 
@@
 
+As the benefits of strategic planning for cycling become more apparent
+\citep{scappini2022} we expect the demand for tools such as CRUSE to
+grow.
+
 \hypertarget{sec-conclusions}{%
@@
 levels. Because the results are available at the route segment level,
-the tool can be used to identify `weak links' in the cycle network, and
-to prioritize investment in cycling infrastructure. Furthermore, the
-provision of the evidence in a free and publicly available website,
-hosted at \href{https://cruse.bike}{cruse.bike}, means that it can be
-used by anyone, encouraging wider participation and more evidence-based
-debate about transport planning.
+the tool can be used to identify `weak links' in the cycle network
+\citep{vybornova2022}, and to prioritize investment in cycling
+infrastructure to maximize health, equality, and other benefits
+\citep{mahfouz, woodcock2021}. Furthermore, the provision of the
+evidence in a free and publicly available website, hosted at
+\href{https://cruse.bike}{cruse.bike}, means that it can be used by
+anyone, encouraging wider participation and more evidence-based debate
+about transport planning.
 

### 2023-10-11 cover-letter.qmd
Update paper

@@
 
-We are submitting our manuscript entitled "CRUSE to Safe Cycling in Ireland: An Open Source Tool to Support Active Travel" for consideration for publication in the European Transport Research Review.
+We are submitting our manuscript entitled "CRUSE to Safe Cycling in Ireland: An Open Source Tool to Support Active Travel" for consideration for publication in the European Transport Research Review, as part of a **Full Paper Submission for TRA 2024 (note: I this option was not available the dropdown menu in the Thematic Series question, please ensure it is in the correct category)**.
 It has a word count of 4,721, excluding references.

### 2023-10-11 paper.qmd
Update paper

@@
 format: 
-  # html: default
+  html: default
   # docx: default
-# Uncomment for formatted version
-  elsevier-pdf:
-    linestretch: 2
-    keep-tex: true
-    journal:
-      name: European Transport Research Review
-      formatting: preprint
-      model: 3p
-      cite-style: super
+# # Uncomment for formatted version
+#   elsevier-pdf:
+#     linestretch: 2
+#     keep-tex: true
+#     journal:
+#       name: European Transport Research Review
+#       formatting: preprint
+#       model: 3p
+#       cite-style: super
 # # #   # pdf: default
@@
 
-  This paper describes the design, features and potential use of the Cycle Route Acceptance and Scenario Estimation (CRUSE) Tool for Ireland was commissioned to support cycle network planning, safety interventions, business cases and project evaluation.
+  This paper describes the design, features and potential use of the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland was commissioned to support cycle network planning, safety interventions, business cases and project evaluation.
 The resulting web application is available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions.
@@
 system("pdftk _site/paper.pdf cat 1 output _site/title_page.pdf")
+# Combine cv.pdf with visions.pdf:
+system("pdftk _site/cv.pdf _site/visions.pdf cat output _site/cv_and_visions.pdf")
 ```

### 2023-10-11 paper.tex
Update paper

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden, sharp corners, breakable, enhanced]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
 

### 2023-10-11 paper.qmd
Only build html

@@
   # docx: default
-# # Uncomment for formatted version
+# # # Uncomment for formatted version
 #   elsevier-pdf:

### 2023-10-11 paper.tex
Only build html

@@
 This paper describes the design, features and potential use of the Cycle
-Route Acceptance and Scenario Estimation (CRUSE) Tool for Ireland was
+Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland was
 commissioned to support cycle network planning, safety interventions,
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
 

### 2023-10-17 cover-letter.qmd
Updates for ETRR

--- 2023-10-09 cover-letter.qmd
@@
 
-We are submitting our manuscript entitled "CRUSE to Safe Cycling in Ireland: An Open Source Tool to Support Active Travel" for consideration for publication in the European Transport Research Review.
+We are submitting our manuscript entitled "CRUSE to Safe Cycling in Ireland: An Open Source Tool to Support Active Travel" for consideration for publication in the European Transport Research Review, as part of a **Full Paper Submission for TRA 2024 (note: this option was not available the dropdown menu in the Thematic Series question, please ensure it is in the correct category)**.
 It has a word count of 4,721, excluding references.
 
-This manuscript describes the development of an open source tool to support active travel in Ireland.
+Please see the full list of authors and the abstract in the title page below.
+The manuscript describes the development of an open source tool to support active travel in Ireland.
 The paper describes the design, features and potential uses of the CRUSE Tool.
@@
 
-Robin Lovelace on behalf of all authors, Joey Talbot, Eugeni Vidal-Tortosa, and Hussein Mahfouz (University of Leeds), Elaine Brick, Peter Wright, Gary O'Tool (AECOM), Dan Brennan and Suzanne Meade (Transport Infrastructure Ireland)
+Robin Lovelace on behalf of all authors:
+
+Robin Lovelace (University of Leeds, r.lovelace@leeds.ac.uk)
+
+Joey Talbot (University of Leeds, J.D.Talbot@leeds.ac.uk)
+
+Eugeni Vidal-Tortosa (University of Leeds, E.VidalTortosa@leeds.ac.uk)
+
+Hussein Mahfouz (University of Leeds, tshma@leeds.ac.uk)
+
+Suzanne Meade (Transport Infrastructure Ireland, Suzanne.Meade@tii.ie)
+
+Dan Brennan (Transport Infrastructure Ireland, Dan.Brennan@tii.ie)
+
+Peter Wright (AECOM, peter.wright1@aecom.com)
+
+Gary O'Toole (AECOM, Gary.OToole@aecom.com)
+
+Elaine Brick (AECOM, elaine.brick@aecom.com)
+
+# Declarations
+
+## Availability of data and material {.unnumbered}
+
+Data was obtained from the Central Statistics Office (CSO) and Transport Infrastructure Ireland (TII) under license and cannot be shared publicly.
+The code used to generate the results presented in this paper is available <!-- at https://github.com/cruse-bike. --> \[To be added after paper is accepted.\]
+
+## Funding {.unnumbered}
+
+The work was funded by Transport Infrastructure Ireland (TII).
+
+## Acknowledgements {.unnumbered}
+
+Thanks to Paul MacDonald and Donal Hodgins (Kildare County Council), for testing early versions of the tool and for their input into the project.
+Thanks to Ciaran Maguire, Catherine Swift, and others at AECOM for their input into the project.
+Thanks to Martin Lucas-Smith and Simon Nutall from CycleStreets for their input into the project.
+
+## Competing interests {.unnumbered}
+
+The authors declare that they have no competing interests.
+
+## Authors' contributions {.unnumbered}
+
+RL led the development of the CRUSE Tool, with input from JT, EV, HM, EB, PW, GOT, DB, and SM.
+SM instigated the project and coordinated its policy impact.
+PW, EB and GOT provided project management and transport engineering expertise.
 

### 2023-10-17 paper.qmd
Updates for ETRR

--- 2023-10-11 paper.qmd
@@
 format: 
-  html: default
+  # html: default
   # docx: default
-# # Uncomment for formatted version
-#   elsevier-pdf:
-#     linestretch: 2
-#     keep-tex: true
-#     journal:
-#       name: European Transport Research Review
-#       formatting: preprint
-#       model: 3p
-#       cite-style: super
+# # # Uncomment for formatted version
+  elsevier-pdf:
+    linestretch: 2
+    keep-tex: true
+    journal:
+      name: European Transport Research Review
+      formatting: preprint
+      model: 3p
+      cite-style: super
 # # #   # pdf: default
 
-# # # Uncomment the following line to produce the cover page
-# author:
-#   - name: Robin Lovelace
-#     email: r.lovelace@leeds.ac.uk
-#     corresponding: true
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Joey Talbot
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Eugeni Vidal-Tortosa
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Hussein Mahfouz
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Elaine Brick
-#     affiliations:
-#       - name: AECOM
-#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-#   - name: Peter Wright
-#     affiliations:
-#       - name: AECOM
-#         address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
-#   - name: "Gary O'Tool"
-#     affiliations:
-#       - name: AECOM
-#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-#   - name: Dan Brennan
-#     affiliations:
-#       - name: Transport Infrastructure Ireland
-#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
-#   - name: Suzanne Meade
-#     affiliations:
-#       - name: Transport Infrastructure Ireland
-#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+# # Uncomment the following line to produce the cover page
+author:
+  - name: Robin Lovelace
+    email: r.lovelace@leeds.ac.uk
+    corresponding: true
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Joey Talbot
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Eugeni Vidal-Tortosa
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Hussein Mahfouz
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Elaine Brick
+    affiliations:
+      - name: AECOM
+        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+  - name: Peter Wright
+    affiliations:
+      - name: AECOM
+        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
+  - name: "Gary O'Tool"
+    affiliations:
+      - name: AECOM
+        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+  - name: Dan Brennan
+    affiliations:
+      - name: Transport Infrastructure Ireland
+        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+  - name: Suzanne Meade
+    affiliations:
+      - name: Transport Infrastructure Ireland
+        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
 abstract: |
@@
 Data was obtained from the Central Statistics Office (CSO) and Transport Infrastructure Ireland (TII) under license and cannot be shared publicly.
-The code used to generate the results presented in this paper is available <!-- at https://github.com/cruse-bike. --> \[To be added after paper is accepted.\]
+The code used to generate the results presented in this paper is fully reproducible and available at \[to be added post review.\]
 
@@
 
-The work was funded by Transport Infrastructure Ireland (TII).
+This work was funded by Transport Infrastructure Ireland (TII).
 
@@
 
-Thanks to Paul MacDonald and Donal Hodgins (Kildare County Council), for testing early versions of the tool and for their input into the project.
-Thanks to Ciaran Maguire, Catherine Swift, and others at AECOM for their input into the project.
-Thanks to Martin Lucas-Smith and Simon Nutall from CycleStreets for their input into the project.
-
-## Competing interests {.unnumbered}
-
-The authors declare that they have no competing interests.
-
-## Authors' contributions {.unnumbered}
-
-\[To be added after paper is accepted.\]
-
-```{=html}
-<!-- RL led the development of the CRUSE Tool, with input from JT, EV, HM, EB, PW, GOT, DB, and SM.
-SM instigated the project and coordinated its policy impact.
-PW, EB and GOT provided project management and transport engineering expertise. -->
-```
+\[To be added post review.\]

### 2023-10-17 paper.tex
Updates for ETRR

@@
   pdftitle={CRUSE to Safe Cycling in Ireland},
+  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Tool; Dan Brennan; Suzanne Meade},
   pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
@@
 Support Active Travel} }
+\author[1]{Robin Lovelace%
+\corref{cor1}%
+}
+ \ead{r.lovelace@leeds.ac.uk} 
+\author[1]{Joey Talbot%
+%
+}
+
+\author[1]{Eugeni Vidal-Tortosa%
+%
+}
+
+\author[1]{Hussein Mahfouz%
+%
+}
+
+\author[2]{Elaine Brick%
+%
+}
+
+\author[3]{Peter Wright%
+%
+}
+
+\author[2]{Gary O'Tool%
+%
+}
+
+\author[4]{Dan Brennan%
+%
+}
+
+\author[4]{Suzanne Meade%
+%
+}
+
+\affiliation[1]{organization={University of
+Leeds},addressline={Institute for Transport Studies, Leeds, LS2 9JT,
+United Kingdom},postcodesep={}}
+\affiliation[2]{organization={AECOM},addressline={Unit 6, Galway
+Technology Park, Parkmore, Galway, Ireland},postcodesep={}}
+\affiliation[3]{organization={AECOM},addressline={Winslade House,
+Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY,
+UK},postcodesep={}}
+\affiliation[4]{organization={Transport Infrastructure
+Ireland},addressline={Parkgate Business Centre, Parkgate Street, Dublin
+8, D08 DK10, Ireland},postcodesep={}}
 
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 publicly. The code used to generate the results presented in this paper
-is available {[}To be added after paper is accepted.{]}
+is fully reproducible and available at {[}to be added post review.{]}
 
@@
 
-Thanks to Paul MacDonald and Donal Hodgins (Kildare County Council), for
-testing early versions of the tool and for their input into the project.
-Thanks to Ciaran Maguire, Catherine Swift, and others at AECOM for their
-input into the project. Thanks to Martin Lucas-Smith and Simon Nutall
-from CycleStreets for their input into the project.
-
-\hypertarget{competing-interests}{%
-\subsection*{Competing interests}\label{competing-interests}}
-\addcontentsline{toc}{subsection}{Competing interests}
-
-The authors declare that they have no competing interests.
-
-\hypertarget{authors-contributions}{%
-\subsection*{Authors' contributions}\label{authors-contributions}}
-\addcontentsline{toc}{subsection}{Authors' contributions}
-
-{[}To be added after paper is accepted.{]}
+{[}To be added post review.{]}
 

### 2023-10-17 paper.qmd
Updates for submission

@@
 number-sections: true
-
-# # Uncomment for draft versions:
+latex-clean: false
+pdf-engine: pdflatex
+knitr:
+  opts_chunk:
+    fig.path: images/
 format: 
@@
 
-```{mermaid}
+```{r}
+#| eval: false
+#| echo: false
+# ```{mermaid}
+%%| eval: false
 %%| fig-width: 100%
@@
 ```
+
+::: {#fig-user-stories}
+
+![](images/mermaid-figure-1.png)
+
+Hypothetical user stories for the CRUSE Tool illustrating that it can be used in different ways by different user groups.
+
+:::
 

### 2023-10-17 paper.tex
Updates for submission

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, sharp corners]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, sharp corners, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{paper_files/figure-latex/mermaid-figure-1.png}
-
-}
-
-\end{figure}
+\includegraphics{images/mermaid-figure-1.png}
 

### 2023-10-17 novelty.qmd
Prep for submission

---
format: pdf
title: "Novelty statement"
---

<!-- Details of most impactful recent EC project
Research must be based on transport and predominantly funded by the EC. Research with a
mix of EC and other funding (e.g. national government) is acceptable but maybe given a
lower weighting. For PROJECT TITLE give the EC acronym followed by full title. The TIME
SPAN should include start and end dates while the PROJECT VALUES should include total
values (in Euro) and the value to your company/institution, best guess is acceptable for now.
Research impact statement
Your impact statement forms one of the key sets of evidence that will be used to select the
short list. Successful candidates will have to demonstrate excellence and proven impacts in
their field of research. The statement must describe the impact of your recent EU funded
research. It is important to describe and evidence 
• how this research is/was supported by EU funding
• your contribution to the research, this might be in a leadership role or as a
practitioner
• the impact of your research at national, EU and global levels, this can be a
combination of scientific impact; economic impact; policy development and public
outreach
• plans for further development and how your research will contribute towards
rethinking transport for clean and inclusive mobility.
The statement should be no more than 500 words, may contain up to 2 pictures/graphs and
should be uploaded as Word file. -->

The Cycle Route Uptake and Scenario Estimation (CRUSE) Tool presented in the paper is the first strategic cycle network planning tool of this level of sophistication to the best of our knowledge. The inclusion of multiple trip types, including travel to work and education from official sources, and simulated trips for other purposes, is a world first for a national tool.
The tool required substantial methodological development, including the development of a new R package (`simodels`, for modelling origin-destination data with spatial interaction models).
The tool is already having major impacts in Ireland and looks set to transform the practice of strategic cycle network planning in Ireland.

I led the project from it's inception, in conversation with Dr Suzanne Meade, Head of Safety at TII, through to the development of the tool and deployment on a public facing website.
The impact of the research is national for Ireland and, if the approach gains traction, global.

I plan to take the methodology and tool to other countries, and would like to develop an European wide tool, with the support of the European Commission, e.g. as part of a Horizon funded project.

### 2023-10-17 paper.qmd
Prep for submission

--- 2023-10-11 paper.qmd
@@
 number-sections: true
-
-# # Uncomment for draft versions:
+latex-clean: false
+pdf-engine: pdflatex
+knitr:
+  opts_chunk:
+    fig.path: images/
 format: 
-  html: default
+  # html: default
   # docx: default
 # # # Uncomment for formatted version
-#   elsevier-pdf:
-#     linestretch: 2
-#     keep-tex: true
-#     journal:
-#       name: European Transport Research Review
-#       formatting: preprint
-#       model: 3p
-#       cite-style: super
+  elsevier-pdf:
+    linestretch: 2
+    keep-tex: true
+    journal:
+      name: European Transport Research Review
+      formatting: preprint
+      model: 3p
+      cite-style: super
 # # #   # pdf: default
@@
 ::: {#fig-dublin}
-![](images/paste-1.png)
+![](paste-1.png)
 
-![](images/paste-3.png)
+![](paste-3.png)
 
@@
 ::: {#fig-trip-purposes}
-![](images/paste-4.png)
+![](paste-4.png)
 
@@
 ::: {#fig-route-types layout-nrow="2"}
-![](images/naas_quietest_godutch.png){width="250"}
+![](naas_quietest_godutch.png){width="250"}
 
-![](images/naas_fastest_godutch.png){width="250"}
+![](naas_fastest_godutch.png){width="250"}
 
-![](images/newbridge_quietest_godutch.png){width="250"}
+![](newbridge_quietest_godutch.png){width="250"}
 
-![](images/newbridge_fastest_godutch.png){width="250"}
+![](newbridge_fastest_godutch.png){width="250"}
 
@@
 
-![](images/scenarios_kildare.png)
+![](scenarios_kildare.png)
 
@@
 ```
+
+::: {#fig-user-stories}
+
+![](mermaid-figure-1.png)
+
+Hypothetical user stories for the CRUSE Tool illustrating that it can be used in different ways by different user groups.
+
+:::
 
@@
 ::: {#fig-city-results}
-![](images/rnet_types.png)
+![](rnet_types.png)
 

### 2023-10-17 paper.tex
Prep for submission

--- 2023-10-11 paper.tex
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, sharp corners, frame hidden, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 
-\includegraphics{images/paste-1.png}
-
-\includegraphics{images/paste-3.png}
+\includegraphics{paste-1.png}
+
+\includegraphics{paste-3.png}
 
@@
 
-\includegraphics{images/paste-4.png}
+\includegraphics{paste-4.png}
 
@@
 
-\includegraphics[width=2.60417in,height=\textheight]{images/naas_quietest_godutch.png}
+\includegraphics[width=2.60417in,height=\textheight]{naas_quietest_godutch.png}
 
@@
 
-\includegraphics[width=2.60417in,height=\textheight]{images/naas_fastest_godutch.png}
+\includegraphics[width=2.60417in,height=\textheight]{naas_fastest_godutch.png}
 
@@
 
-\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_quietest_godutch.png}
+\includegraphics[width=2.60417in,height=\textheight]{newbridge_quietest_godutch.png}
 
@@
 
-\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_fastest_godutch.png}
+\includegraphics[width=2.60417in,height=\textheight]{newbridge_fastest_godutch.png}
 
@@
 
-\includegraphics{images/scenarios_kildare.png}
+\includegraphics{scenarios_kildare.png}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{paper_files/figure-latex/mermaid-figure-1.png}
-
-}
-
-\end{figure}
+\includegraphics{mermaid-figure-1.png}
 
@@
 
-\includegraphics{images/rnet_types.png}
+\includegraphics{rnet_types.png}
 

### 2023-10-17 visions.qmd
Prep for submission

--- 2023-10-17 novelty.qmd
@@
 
-The Cycle Route Uptake and Scenario Estimation (CRUSE) Tool presented in the paper is the first strategic cycle network planning tool of this level of sophistication to the best of our knowledge.
+The Cycle Route Uptake and Scenario Estimation (CRUSE) Tool presented in the paper received funding from the EU via Transport Infrastructure Ireland (TII), with funding in the region of €200k.
+
+This is the first strategic cycle network planning tool of this level of sophistication to the best of our knowledge.
 The inclusion of multiple trip types, including travel to work and education from official sources, and simulated trips for other purposes, is a world first for a national tool.

### 2023-10-17 paper.qmd
Update title page

@@
 # Save title page (1st page of paper.pdf):
-system("pdftk _site/paper.pdf cat 1 output _site/title_page.pdf")
+system("pdftk paper.pdf cat 1 output _site/title_page.pdf")
+# Combine title page and cover-letter.pdf:
+system("pdftk _site/cover-letter.pdf _site/title_page.pdf cat output _site/cover_letter_full.pdf")
+browseURL("_site/cover_letter_full.pdf")
 # Combine cv.pdf with visions.pdf:

### 2023-10-19 paper.qmd
Update letter

--- 2023-10-17 paper.qmd
@@
 format: 
-  # html: default
+  html: default
   # docx: default
-# # # Uncomment for formatted version
-  elsevier-pdf:
-    linestretch: 2
-    keep-tex: true
-    journal:
-      name: European Transport Research Review
-      formatting: preprint
-      model: 3p
-      cite-style: super
-# # #   # pdf: default
+# # # # Uncomment for formatted version
+#   elsevier-pdf:
+#     linestretch: 2
+#     keep-tex: true
+#     journal:
+#       name: European Transport Research Review
+#       formatting: preprint
+#       model: 3p
+#       cite-style: super
+# # # #   # pdf: default
 

### 2024-02-07 resubmission/responses1.qmd
Add resubmission starter

---
title: "Reviewer responses"
format: gfm
bibliography: references.bib
---

```{=html}
<!-- ETRE-D-23-00321
CRUSE to Safe Cycling in Ireland An Open Source Tool to Support Active Travel
Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O’Tool; Dan Brennan; Suzanne Meade
European Transport Research Review

Dear Prof. Lovelace,

Your manuscript 'CRUSE to Safe Cycling in Ireland An Open Source Tool to Support Active Travel' (ETRE-D-23-00321) has been assessed by our reviewers. Although it is of interest, we are unable to consider it for publication in its current form. The reviewers have raised a number of points which we believe would improve the manuscript and may allow a revised version to be published in European Transport Research Review.

Their reports, together with any other comments, are below. Please also take a moment to check our website at https://www.editorialmanager.com/etre/ for any additional comments that were saved as attachments.

If you are able to fully address these points, we would encourage you to submit a revised manuscript to European Transport Research Review.
Once you have made the necessary corrections, please submit online at:

https://www.editorialmanager.com/etre/

If you have forgotten your username or password please use the 'Send Login Details' link to get your login information. For security reasons, your password will be reset.

Please include a point-by-point response within the 'Response to Reviewers' box in the submission system and highlight (with 'tracked changes'/coloured/underlines/highlighted text) all changes made when revising the manuscript. Please ensure you describe additional experiments that were carried out and include a detailed rebuttal of any criticisms or requested revisions that you disagreed with. Please also ensure that your revised manuscript conforms to the journal style, which can be found in the Submission Guidelines on the journal homepage.

The due date for submitting the revised version of your article is 02 Mar 2024.

Please note, if your manuscript is accepted you will not be able to make any changes to the authors, or order of authors, of your manuscript once the editor has accepted your manuscript for publication. If you wish to make any changes to authorship before you resubmit your revisions, please reply to this email and ask for a 'Request for change in authorship' form which should be completed by all authors (including those to be removed) and returned to this email address. Please ensure that any changes in authorship fulfil the criteria for authorship as outlined in SpringerOpen's editorial policies (http://www.springeropen.com/get-published/editorial-policies#authorship).

Once you have completed and returned the form, your request will be considered and you will be advised whether the requested changes will be allowed.
By resubmitting your manuscript you confirm that all author details on the revised version are correct, that all authors have agreed to authorship and order of authorship for this manuscript and that all authors have the appropriate permissions and rights to the reported data.

Please be aware that we may investigate, or ask your institute to investigate, any unauthorised attempts to change authorship or discrepancies in authorship between the submitted and revised versions of your manuscript.

We look forward to receiving your revised manuscript soon.

Best wishes,

European Transport Research Review
https://www.editorialmanager.com/etre 
-->
```
# Response to the editor

We thank the editor and reviewers for their constructive comments.
We have revised the manuscript in response to the reviewers' comments and believe that the revised manuscript is much improved.
We have made the following changes:

[@lovelace2017]

Reviewer reports:

Reviewer #1: REVIEW: CRUSE to Safe Cycling in Ireland An Open Source Tool to Support Active Travel

C1) General comments: Thank you for the opportunity to review this paper.
This paper aims to describe the design, features, and potential uses of the Cycle Route Uptake and Scenario Estimation tool (CRUSE), an upgraded version of the PCT tool.
This research paper is interesting, but I believe it fails to address one of the three main research goals, which refers to the potential use of the tool in practice.
I recommend against publication.
Although the authors have put considerable thought and effort into this research, portions of the research are not presented clearly, and it is difficult for me to determine how the cycling conditions are being analysed/calculated.
In addition, the authors overstate their case when they claim a contribution is an upgraded version of the PTC tool.
Similarly, while the contribution of variables that are analysed may be new, and the site may be new, the conceptual approach, as I understand it, is similar to other standard approaches that involve using GIS-based tools to measure environmental attributes and cycling potential based on OSM.

C2) The theoretical framework needs to be improved.
This paper misses a literature overview of cycling planning support systems (PSS) tools developed in the last decades.
After the introduction, which gives an interesting overview of the Irish scenario on cycling levels and road safety measures, the authors directly present the methodology encompassing the CRUSE tool.
I believe it would be necessary to provide a new section dedicated to the literature review on PSS tools and cycling planning, as well as the most applied metrics to measure cycling potential in practice, to provide a critical overview of missing links between this paper's added value and the current literature and empirical evidence.

C3) The authors describe the comprehensive nature of the measures they incorporate as indicators, and these all seem appropriate, but it is unclear whether the municipal officials responsible for the local cycling strategy would prioritise the same measures.
At the very least, the authors need to discuss strategies for engaging the tool users and those affected by the tool.
Based on my experience, while elegant and comprehensive, many technical tools have proven to be of limited value in practice because they do not anticipate the constraints of practice.
My point is that the real-world complexities of use need to be considered if the claim is a public process and meant to be used in real-world applications.
What difficulties might be expected in the use of CRUSE in planning practice?

C4) In line with the previous question, this paper primarily focuses on presenting the tool in methodological terms, lacking a critical theoretical discussion and a broad understanding of the potential usefulness for planning practice.
Likewise, it would be desirable to explore this tool from a qualitative perspective based on the perception of planning practitioners and potential users regarding the tool's user-friendliness.
In fact, this research has the potential to build a bridge between academia and practice, but it is limited to a superficial discussion of the tool in methodological terms only.

C5) Compared to the original PCT tool, the authors argue that the CRUSE tool overcomes the issues of low data resolution, limited coverage of trip purposes, and a web interface that is considered user-friendly or intuitive.
Yet, I could not fully grasp such attributes when accessing the provided CRUSE link.
First, the data is organised into counties, which does not precisely bring a high resolution represented by the scale of the cyclist.
Second, how could the authors claim that the tool is user-friendly without gathering feedback from practitioners or users who will have to use the tool in their professional tasks?

C6) How exactly were adjustments made for topography?
For e-bike users, topography has less impact on travel speed.
Were adjustments also made for delays at traffic signals?
In determining whether an opportunity is accessible, is any consideration given to whether it can be reached via the low-stress, traffic-free bicycle network?

C7) Limitations of Research.
I think the authors need to address the limitations of their research systematically.
While I understand the purpose of the research is to illustrate the technical, analytic approach, I believe the authors need to discuss the limitations of the CRUSE tool and how these limitations can be addressed in practice.

C8) Fix grammar and formatting inconsistencies throughout the whole manuscript.

Reviewer #2: European Transport Research Review Manuscript Number: ETRE-D-23-00321 Manuscript type: Original Paper Title: CRUSE to Safe Cycling in Ireland An Open Source Tool to Support Active Travel This paper describes the design, features, and potential use of the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland to support cycle network planning, safety interventions, business cases, and project evaluation.

Overall this is an important paper that contributes to the practical means for modeling networks at high levels of geographic resolution.
While its contribution to the existing literature is in doubt.

However, some issues require attention and revisions before I can recommend publication.

Comments:

The title: those who will read the title cannot understand the meaning of CRUSE.
So my recommendation is to take out this word and replace it with an application or something else.

The introduction is very general.
While the paper attempted to examine a new application for modeling networks at high levels of geographic resolution, the introduction did not include any literature about the existence of similar applications and the limitations of such applications.

The authors did not highlight the contribution of this paper to the existing literature.

Results:

Figure 6 is not clear and one cannot learn much based on it.
I think that authors should try to present the results more clearly.

Discussion: The discussion is insufficient it needs more work.
The discussion includes findings about the infrastructure features that encourage the use of the bicycle so why nothing is mentioned in the introduction, this is an important issue.

Here is some Python code to calculate the distance between Dublin and London:

```{python}
from geopy.distance import geodesic
coords_1 = (53.349805, -6.26031)
coords_2 = (51.5074, -0.1278)
print(geodesic(coords_1, coords_2).km)
```

### 2024-03-20 paper.qmd
docx

--- 2023-10-19 paper.qmd
@@
 format: 
-  html: default
-  # docx: default
+  # html: default
+  docx: default
 # # # # Uncomment for formatted version

### 2024-03-22 paper.qmd
update title

@@
 ---
-title: "CRUSE to Safe Cycling in Ireland"
-subtitle: An Open Source Tool to Support Active Travel
+title: "Cycle Route Uptake and Scenario Estimation (CRUSE): an open and nationally scalable approach for strategic cycle network planning tool development"
+subtitle: A case study of Ireland
 bibliography: bibliography.bib
@@
 format: 
-  # html: default
-  docx: default
+  html: default
+  # docx: default
 # # # # Uncomment for formatted version
@@
 abstract: |
-  Legislation mandates improved road safety and decarbonisation of predominantly road-based transport network in Ireland and other countries.
-Increasing cycling transport, making it safer and more attractive, is essential to this end.
-Cycling in Ireland represents only 3% of total modal share as of the latest Census (2016), but accounts for 20% of serious injuries and 7% of all fatalities.
-A planning tool based on high resolution datasets on baseline and potential cycling levels could greatly assist in achieving this goal.
-
-  This paper describes the design, features and potential use of the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland was commissioned to support cycle network planning, safety interventions, business cases and project evaluation.
+  This paper describes the design, features and potential use of the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
+The tool was commissioned to support cycle network planning, safety interventions, business cases and project evaluation.
 The resulting web application is available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions.
@@
 Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
+Cycling in Ireland represents only 3% of total modal share as of the latest Census (2016), but accounts for 20% of serious injuries and 7% of all fatalities.
 A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.

### 2024-03-22 paper.qmd
Update abstract

@@
 abstract: |
-  This paper describes the design, features and potential use of the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
-The tool was commissioned to support cycle network planning, safety interventions, business cases and project evaluation.
-The resulting web application is available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions.
-  CRUSE goes beyond previous work in several important ways.
-  It models networks at high levels of geographic resolution, simulates a wide range of trips including recreational, social, shopping and personal trips, in addition to trips to work and primary, secondary and tertiary education locations.
-  The tool provides key 'denominator' information to enable reporting of collision rates per billion km cycled.
-
-  Another key feature of CRUSE is its use of frequently-updated OpenStreetMap data for route calculation and "cycle friendliness" estimates of all links on the network.
-  CRUSE provides three network types: the 'Fastest' network highlights routes for directness; 'Quietest' prioritises cycle friendliness, and 'Balanced' represents a balance between speed and comfort.
-  Trips are generated based on origin and destination data from the 2016 Census in combination with modeled demand data to estimate current and potential future cycling levels at area, route, and network levels for each county in Ireland.
-
-  For investment in cycling to deliver on the huge potential benefits, cycling must become safer.
-  With growth in the ebike market, CRUSE will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
-  Provision of open access data in a publicly available web tool will address the challenges faced by counties across Ireland in meeting their sustainability commitments.
-  The experience of developing a deploying CRUSE are relevant to countries with sustainable transport policies worldwide.
+  This paper describes an approach to the development of strategic cycle network planning tools.
+Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the methods used to generate the evidence presented in the tool, including disaggregation of origin-destination data, additional trip purposes, routing, scenarios of cycling potential, and the user interface.
+The CRUSE Tool provides estimates of cycling currently and under four scenarios of cycling potential: Near Market, Climate Action Plan, Go Dutch, and Ebike.
+The tool was commissioned by Transport Infrastructure Ireland and is publicly available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions.
+CRUSE goes beyond previous work by: modelling networks at high levels of geographic resolution; simulating multiple trip purposes (social, shopping and personal utility trips plus recreational and cycle touring trips), in addition to using official data on trips to work and education locations; provision of estimates of 'quietness' (a proxy for cyclability) at the route segment level; and its provision of three network types: 'Fastest', 'Balanced' and 'Quietest' to help plan both arterial and residential cycle networks.
+Workshops and unstructure interview is stakeholders show that the tool is already being used in practice, to inform urban, inter-urban and rural network development.
+The approach and methods for strategic cycle network development tools outlined in this paper can be applied and built-on in other places to support the generation of evidence needed to support ambitious cycling policies and interventions internationally.
 keywords: 

### 2024-03-22 paper.qmd
New images for Figure 1

@@
 ::: {#fig-dublin}
-![](images/paste-1.png)
+<!-- ![](images/paste-1.png) -->
+![](images/paste-8.png)
+:::
 
-![](images/paste-3.png)
+<div>
 
-Illustration of the density of route networks generated by the CRUSE Tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland.
+<!-- ![](images/paste-3.png) -->
+![](images/paste-9.png)
+
+Illustration of the density of route networks generated by the CRUSE Tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland (botton).
 Source: CRUSE Tool, available at <https://cruse.bike/>.
-:::
+
+</div>
 
@@
 ::: {#fig-user-stories}
-
 ![](images/mermaid-figure-1.png)
@@
 Hypothetical user stories for the CRUSE Tool illustrating that it can be used in different ways by different user groups.
-
 :::

### 2024-03-22 paper.qmd
Update paper.tex

@@
 format: 
-  html: default
+  # html: default
   # docx: default
 # # # # Uncomment for formatted version
-#   elsevier-pdf:
-#     linestretch: 2
-#     keep-tex: true
-#     journal:
-#       name: European Transport Research Review
-#       formatting: preprint
-#       model: 3p
-#       cite-style: super
+  elsevier-pdf:
+    linestretch: 2
+    keep-tex: true
+    journal:
+      name: European Transport Research Review
+      formatting: preprint
+      model: 3p
+      cite-style: super
 # # # #   # pdf: default
@@
 ![](images/paste-8.png)
-:::
-
-<div>
-
 <!-- ![](images/paste-3.png) -->
@@
 Source: CRUSE Tool, available at <https://cruse.bike/>.
-
-</div>
+:::
 

### 2024-03-22 paper.tex
Update paper.tex

--- 2023-10-17 paper.tex
@@
 \makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
 \@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
 \makeatother
@@
 \bibliographystyle{elsarticle-num}
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \hypersetup{
-  pdftitle={CRUSE to Safe Cycling in Ireland},
+  pdftitle={Cycle Route Uptake and Scenario Estimation (CRUSE): an open and nationally scalable approach for strategic cycle network planning tool development},
   pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Tool; Dan Brennan; Suzanne Meade},
@@
 \begin{frontmatter}
-\title{CRUSE to Safe Cycling in Ireland \\\large{An Open Source Tool to
-Support Active Travel} }
+\title{Cycle Route Uptake and Scenario Estimation (CRUSE): an open and
+nationally scalable approach for strategic cycle network planning tool
+development \\\large{A case study of Ireland} }
 \author[1]{Robin Lovelace%
@@
 \begin{abstract}
-Legislation mandates improved road safety and decarbonisation of
-predominantly road-based transport network in Ireland and other
-countries. Increasing cycling transport, making it safer and more
-attractive, is essential to this end. Cycling in Ireland represents only
-3\% of total modal share as of the latest Census (2016), but accounts
-for 20\% of serious injuries and 7\% of all fatalities. A planning tool
-based on high resolution datasets on baseline and potential cycling
-levels could greatly assist in achieving this goal.
-
-This paper describes the design, features and potential use of the Cycle
-Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland was
-commissioned to support cycle network planning, safety interventions,
-business cases and project evaluation. The resulting web application is
-available at \url{https://cruse.bike/}, enabling planners, engineers,
-and other stakeholders to make more evidence-based decisions. CRUSE goes
-beyond previous work in several important ways. It models networks at
-high levels of geographic resolution, simulates a wide range of trips
-including recreational, social, shopping and personal trips, in addition
-to trips to work and primary, secondary and tertiary education
-locations. The tool provides key `denominator' information to enable
-reporting of collision rates per billion km cycled.
-
-Another key feature of CRUSE is its use of frequently-updated
-OpenStreetMap data for route calculation and ``cycle friendliness''
-estimates of all links on the network. CRUSE provides three network
-types: the `Fastest' network highlights routes for directness;
-`Quietest' prioritises cycle friendliness, and `Balanced' represents a
-balance between speed and comfort. Trips are generated based on origin
-and destination data from the 2016 Census in combination with modeled
-demand data to estimate current and potential future cycling levels at
-area, route, and network levels for each county in Ireland.
-
-For investment in cycling to deliver on the huge potential benefits,
-cycling must become safer. With growth in the ebike market, CRUSE will
-help inform inter-urban and rural networks to support the transfer of
-trips to sustainable modes for longer journeys. Provision of open access
-data in a publicly available web tool will address the challenges faced
-by counties across Ireland in meeting their sustainability commitments.
-The experience of developing a deploying CRUSE are relevant to countries
-with sustainable transport policies worldwide.
+This paper describes an approach to the development of strategic cycle
+network planning tools. Based on our experience developing and deploying
+the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland,
+we outline the methods used to generate the evidence presented in the
+tool, including disaggregation of origin-destination data, additional
+trip purposes, routing, scenarios of cycling potential, and the user
+interface. The CRUSE Tool provides estimates of cycling currently and
+under four scenarios of cycling potential: Near Market, Climate Action
+Plan, Go Dutch, and Ebike. The tool was commissioned by Transport
+Infrastructure Ireland and is publicly available at
+\url{https://cruse.bike/}, enabling planners, engineers, and other
+stakeholders to make more evidence-based decisions. CRUSE goes beyond
+previous work by: modelling networks at high levels of geographic
+resolution; simulating multiple trip purposes (social, shopping and
+personal utility trips plus recreational and cycle touring trips), in
+addition to using official data on trips to work and education
+locations; provision of estimates of `quietness' (a proxy for
+cyclability) at the route segment level; and its provision of three
+network types: `Fastest', `Balanced' and `Quietest' to help plan both
+arterial and residential cycle networks. Workshops and unstructure
+interview is stakeholders show that the tool is already being used in
+practice, to inform urban, inter-urban and rural network development.
+The approach and methods for strategic cycle network development tools
+outlined in this paper can be applied and built-on in other places to
+support the generation of evidence needed to support ambitious cycling
+policies and interventions internationally.
 \end{abstract}
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, sharp corners, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 uptake of active modes will require key active travel routes to be
-identified and improved. A proactive approach to the development of
-cycling policy and infrastructure is taking place in Ireland. The main
-frameworks underpinning these efforts are the Climate Action Plan
-\citep{climate2022a}, the National Development Plan, and the National
-Roads 2040 strategy \citep{national2023}. These policy frameworks are
-backed-up by public support for active travel interventions: while only
-3\% of trips in Ireland are made by cycling according to the latest
-Census data from 2016 ``83\% of respondents supported enhanced walking
-and cycling facilities'' \citep{sustaina2019}.
+identified and improved. Cycling in Ireland represents only 3\% of total
+modal share as of the latest Census (2016), but accounts for 20\% of
+serious injuries and 7\% of all fatalities. A proactive approach to the
+development of cycling policy and infrastructure is taking place in
+Ireland. The main frameworks underpinning these efforts are the Climate
+Action Plan \citep{climate2022a}, the National Development Plan, and the
+National Roads 2040 strategy \citep{national2023}. These policy
+frameworks are backed-up by public support for active travel
+interventions: while only 3\% of trips in Ireland are made by cycling
+according to the latest Census data from 2016 ``83\% of respondents
+supported enhanced walking and cycling facilities''
+\citep{sustaina2019}.
 
@@
 
-{\centering
-
-\includegraphics{images/paste-1.png}
-
-\includegraphics{images/paste-3.png}
+\centering{
+
+\includegraphics{images/paste-8.png}
+\includegraphics{images/paste-9.png}
 
@@
 generated by the CRUSE Tool for Dublin City and surroundings (top) and
-rural County Mayo on the West coast of Ireland. Source: CRUSE Tool,
-available at \url{https://cruse.bike/}.}
-
-\end{figure}
+rural County Mayo on the West coast of Ireland (botton). Source: CRUSE
+Tool, available at \url{https://cruse.bike/}.}
+
+\end{figure}%
 
@@
 
-\hypertarget{sec-methods}{%
-\section{Methods and data}\label{sec-methods}}
+\section{Methods and data}\label{sec-methods}
 
@@
 
-\hypertarget{sec-disaggregation}{%
 \subsection{Disaggregation of origin-destination
-data}\label{sec-disaggregation}}
+data}\label{sec-disaggregation}
 
@@
 
-\hypertarget{sec-trip-purposes}{%
-\subsection{Additional trip purposes}\label{sec-trip-purposes}}
+\subsection{Additional trip purposes}\label{sec-trip-purposes}
 
@@
 
-{\centering
+\centering{
 
@@
 
-\end{figure}
-
-\hypertarget{routing}{%
-\subsection{Routing}\label{routing}}
+\end{figure}%
+
+\subsection{Routing}\label{routing}
 
@@
 
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics[width=2.60417in,height=\textheight]{images/naas_quietest_godutch.png}
-
-}
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics[width=2.60417in,height=\textheight]{images/naas_fastest_godutch.png}
-
-}
-
-}
-
-\end{minipage}%
+\begin{minipage}{0.50\linewidth}
+\includegraphics[width=2.60417in,height=\textheight]{images/naas_quietest_godutch.png}\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+\includegraphics[width=2.60417in,height=\textheight]{images/naas_fastest_godutch.png}\end{minipage}%
 \newline
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_quietest_godutch.png}
-
-}
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_fastest_godutch.png}
-
-}
-
-}
-
-\end{minipage}%
+\begin{minipage}{0.50\linewidth}
+\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_quietest_godutch.png}\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_fastest_godutch.png}\end{minipage}%
 
@@
 
-\end{figure}
-
-\hypertarget{sec-scenarios}{%
-\subsection{Scenarios of cycling potential}\label{sec-scenarios}}
+\end{figure}%
+
+\subsection{Scenarios of cycling potential}\label{sec-scenarios}
 
@@
 
-\end{figure}
-
-\hypertarget{baseline}{%
-\subsubsection{Baseline}\label{baseline}}
+\end{figure}%
+
+\subsubsection{Baseline}\label{baseline}
 
@@
 
-\hypertarget{near-market}{%
-\subsubsection{Near Market}\label{near-market}}
+\subsubsection{Near Market}\label{near-market}
 
@@
 
-\hypertarget{climate-action-plan}{%
-\subsubsection{Climate Action Plan}\label{climate-action-plan}}
+\subsubsection{Climate Action Plan}\label{climate-action-plan}
 
@@
 
-\hypertarget{go-dutch}{%
-\subsubsection{Go Dutch}\label{go-dutch}}
+\subsubsection{Go Dutch}\label{go-dutch}
 
@@
 
-\hypertarget{ebike}{%
-\subsubsection{Ebike}\label{ebike}}
+\subsubsection{Ebike}\label{ebike}
 
@@
 
-\hypertarget{sec-ui}{%
-\subsection{User interface}\label{sec-ui}}
+\subsection{User interface}\label{sec-ui}
 
@@
 
-\end{figure}
-
-\hypertarget{sec-results}{%
-\section{Results}\label{sec-results}}
+\end{figure}%
+
+\section{Results}\label{sec-results}
 
@@
 
-\end{figure}
+\end{figure}%
 
@@
 
-\hypertarget{sec-discussion}{%
-\section{Discussion}\label{sec-discussion}}
+\section{Discussion}\label{sec-discussion}
 
@@
 
-\hypertarget{sec-conclusions}{%
-\section{Conclusions}\label{sec-conclusions}}
+\section{Conclusions}\label{sec-conclusions}
 
@@
 
-\hypertarget{list-of-abbreviations}{%
-\section{List of abbreviations}\label{list-of-abbreviations}}
+\section{List of abbreviations}\label{list-of-abbreviations}
 
@@
 
-\hypertarget{declarations}{%
-\section{Declarations}\label{declarations}}
-
-\hypertarget{availability-of-data-and-material}{%
+\section{Declarations}\label{declarations}
+
 \subsection*{Availability of data and
-material}\label{availability-of-data-and-material}}
+material}\label{availability-of-data-and-material}
 \addcontentsline{toc}{subsection}{Availability of data and material}
@@
 
-\hypertarget{funding}{%
-\subsection*{Funding}\label{funding}}
+\subsection*{Funding}\label{funding}
 \addcontentsline{toc}{subsection}{Funding}
@@
 
-\hypertarget{acknowledgements}{%
-\subsection*{Acknowledgements}\label{acknowledgements}}
+\subsection*{Acknowledgements}\label{acknowledgements}
 \addcontentsline{toc}{subsection}{Acknowledgements}

### 2024-03-22 paper.qmd
Updates to paper

@@
 system("pdftk _site/cv.pdf _site/visions.pdf cat output _site/cv_and_visions.pdf")
+# latexdiffr::latexdiff("paper-v1.qmd", "paper.qmd")
+system("latexdiff paper-v1.tex paper.tex > paper-diff.tex")
+system("pdflatex paper-diff.tex")
 ```

### 2024-03-22 paper.qmd
Refactor intro

@@
 
-High energy transport systems are a major contributor to climate change, a leading cause of premature death and injury due to road traffic collisions, and a cause of disease due to airborne and noise pollutants.
-Conversely, evidence-based and effective transport *policies* have great potential to decarbonize economies, improve public health, and save lives.
-
+Transport systems dominated by heavy and powerful private cars are inefficient, dangerous, and unhealthy.
+Cars are a major and growing source of greenhouse gas emissions causing climate change [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018] , and a cause of disease due to physical inactivity and air, noise and (from tyres) microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024].
 Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport.
@@
 
-Recognizing the growing evidence of such impacts of poorly designed and performing transport systems, governments in many countries have set targets and taken actions.
+Growing evidence of the negative impacts of car-dependent transport systems has led governments in many countries to set targets and taken actions.
 In the context of climate, road safety and physicial inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
@@
 
-Another motivation for cycle network planning is the Road Infrastructure Safety Management (RISM) directive (2008/96/EC), which requires member states to implement a road safety management system (RSMS) for all public roads.
+Another motivation for cycle network planning at the European level is the Road Infrastructure Safety Management (RISM) directive (2008/96/EC), which requires member states to implement a road safety management system (RSMS) for all public roads.
 Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out" [@directiv2008].
@@
 For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
-Better data to inform road safety policies and interventions is a motivation for the CRUSE Tool outlined in this paper, and modeling active travel more broadly.
+Better data to inform road safety policies and interventions is a motivation for better estimates of 'baseline' levels of physical activity at high levels of geographic resolution [@tait2023].
 
+National governments are increasingly acting on the evidence.
 In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030 [@national2021].
-Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
-Cycling in Ireland represents only 3% of total modal share as of the latest Census (2016), but accounts for 20% of serious injuries and 7% of all fatalities.
-A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
+Doing so while simultaneously enabling rapid uptake of active modes will require key travel corridors to be identified 'cycle proofed'.
+Cycling in Ireland represents only 3% of total modal share as of the 2016 Census, but accounts for 20% of serious injuries and 7% of all fatalities, driving the development of cycling policy and infrastructure plans in Ireland.
 The main frameworks underpinning these efforts are the Climate Action Plan [@climate2022a], the National Development Plan, and the National Roads 2040 strategy [@national2023].
@@
 
-At the regional level, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
+At the regional level within Ireland, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings of high support for and capability for cycling: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
 Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
@@
 
-Further evidence of the importance of cycling in Ireland *already* is provided by the National Strategic Objective (NSO) from the National Development Plan, with €8.6 billion allocated to sustainable transport infrastructure including public transport and active travel interventions.
+Further evidence of the importance of cycling in Ireland is provided by the National Strategic Objective (NSO) from the National Development Plan, which allocates €8.6 billion to sustainable transport infrastructure including public transport and active travel interventions.
 Cycle infrastructure will be developed in synchrony with the BusConnects project, an entire redesign of the bus network in Dublin and Cork.
+It was in this context that Transport Infrastructure Ireland (TII) commissioned the research reported in this paper, which led to the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
+Building on previous work, including the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE Tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
 
-It was in this context that Transport Infrastructure Ireland (TII) commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
-Based on the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE Tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
-
-A key consideration for TII is to ensure both urban and rural Ireland were targeted for cycling infrastructure, as shown in @fig-dublin.
-To this end, TII extensively reviewed activity on Irish Roads and how the road network was being used by commuters, cyclists, heavy goods vehicles and other road users across the Irish transport network.
-The aim was for the tool to provide strong, national, systematic but locally-specific evidence to monitor cycling friendliness and safety in Ireland, to ensure strategic alignment with national, regional and local policies.
-The open source and publicly available nature of the tool ensures more inclusive and evidence-based conversations around cycle network planning between all stakeholders in the transport planning process.
-
-TII also requires evidence to support monitoring of growth in cycling, upgrades of existing cycling infrastructure, business case development and appraisal, and exposure information to estimate crash rates.
-Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
-These factors mean that the tool can be seen as an open access 'leverage point', providing key information for and supporting ambitious plans in many aspects the planning system, from network design to post-build monitoring [@lovelace2020].
+Stakeholder consultation by TII emphasized the need for a strong, national, systematic but locally-specific evidence base on cycle networks in both urban and rural Ireland.
+The evidence base had to scale nationally to support strategic alignment with national, regional and local policies, but also be useful for local network planning.
+TII also requires
 
@@
 <!-- ![](images/paste-1.png) -->
-![](images/paste-8.png)
-<!-- ![](images/paste-3.png) -->
+
+![](images/paste-8.png) <!-- ![](images/paste-3.png) -->
 ![](images/paste-9.png)
@@
 As the benefits of strategic planning for cycling become more apparent [@scappini2022] we expect the demand for tools such as CRUSE to grow.
+However, the approach is not without limitations and does not meet all requirements in Ireland or indeed any country where the methods are applied.
+
+A limitation of the approach as implemented in the CRUSE tool is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
+Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
+These factors mean that the tool can be seen as an open access 'leverage point', providing key information for and supporting ambitious plans in many aspects the planning system, from network design to post-build monitoring [@lovelace2020].
 

### 2024-03-22 paper.tex
Refactor intro

@@
 
-High energy transport systems are a major contributor to climate change,
-a leading cause of premature death and injury due to road traffic
-collisions, and a cause of disease due to airborne and noise pollutants.
-Conversely, evidence-based and effective transport \emph{policies} have
-great potential to decarbonize economies, improve public health, and
-save lives.
+Transport systems dominated by heavy and powerful private cars are
+inefficient, dangerous, and unhealthy. Cars are a major and growing
+source of greenhouse gas emissions causing climate change
+\citep{winkler2023}, a leading cause of premature death and injury due
+to road traffic collisions \citep{globals2018} , and a cause of disease
+due to physical inactivity and air, noise and (from tyres) microplastic
+pollution \citep{mattsson2023} \citep{welch2023} \citep{cavallaro2024}.
 

### 2024-03-22 paper.qmd
Updates

@@
 system("pdflatex paper-diff.tex")
+browseURL("paper-diff.pdf")
 ```

### 2024-03-22 paper.tex
Updates

@@
 pollution \citep{mattsson2023} \citep{welch2023} \citep{cavallaro2024}.
-
 Transport is responsible for 23\% of global greenhouse gas (GHG)
@@
 
-Recognizing the growing evidence of such impacts of poorly designed and
-performing transport systems, governments in many countries have set
-targets and taken actions. In the context of climate, road safety and
-physicial inactivity crises, policies to improve transport systems can
-be classified according to the `Avoid-Shift-Improve' (ASI) framework
+Growing evidence of the negative impacts of car-dependent transport
+systems has led governments in many countries to set targets and taken
+actions. In the context of climate, road safety and physicial inactivity
+crises, policies to improve transport systems can be classified
+according to the `Avoid-Shift-Improve' (ASI) framework
 \citep{jaramillo2022}. The ASI framework highlights the importance of
@@
 
-Another motivation for cycle network planning is the Road Infrastructure
-Safety Management (RISM) directive (2008/96/EC), which requires member
-states to implement a road safety management system (RSMS) for all
-public roads. Specifically, ``Member States shall ensure that the
-ranking of high accident concentration sections and the network safety
-ranking are carried out'' \citep{directiv2008}. Given that `safety' in
-this context is usefully quantified as the number of people killed and
-seriously injured (KSI) per distance traveled, the directive requires
-estimation of distance traveled by mode, down to the road link level.
-For active modes, about which there is a paucity of data compared with
-motorized modes, this is a major challenge. Better data to inform road
-safety policies and interventions is a motivation for the CRUSE Tool
-outlined in this paper, and modeling active travel more broadly.
+Another motivation for cycle network planning at the European level is
+the Road Infrastructure Safety Management (RISM) directive (2008/96/EC),
+which requires member states to implement a road safety management
+system (RSMS) for all public roads. Specifically, ``Member States shall
+ensure that the ranking of high accident concentration sections and the
+network safety ranking are carried out'' \citep{directiv2008}. Given
+that `safety' in this context is usefully quantified as the number of
+people killed and seriously injured (KSI) per distance traveled, the
+directive requires estimation of distance traveled by mode, down to the
+road link level. For active modes, about which there is a paucity of
+data compared with motorized modes, this is a major challenge. Better
+data to inform road safety policies and interventions is a motivation
+for better estimates of `baseline' levels of physical activity at high
+levels of geographic resolution \citep{tait2023}.
 
-In Ireland, the Road Safety Authority (RSA) has set the target of
-halving the number of road traffic deaths and serious injuries by 2030
+National governments are increasingly acting on the evidence. In
+Ireland, the Road Safety Authority (RSA) has set the target of halving
+the number of road traffic deaths and serious injuries by 2030
 \citep{national2021}. Doing so while simultaneously enabling rapid
-uptake of active modes will require key active travel routes to be
-identified and improved. Cycling in Ireland represents only 3\% of total
-modal share as of the latest Census (2016), but accounts for 20\% of
-serious injuries and 7\% of all fatalities. A proactive approach to the
-development of cycling policy and infrastructure is taking place in
-Ireland. The main frameworks underpinning these efforts are the Climate
-Action Plan \citep{climate2022a}, the National Development Plan, and the
-National Roads 2040 strategy \citep{national2023}. These policy
-frameworks are backed-up by public support for active travel
-interventions: while only 3\% of trips in Ireland are made by cycling
-according to the latest Census data from 2016 ``83\% of respondents
-supported enhanced walking and cycling facilities''
-\citep{sustaina2019}.
+uptake of active modes will require key travel corridors to be
+identified `cycle proofed'. Cycling in Ireland represents only 3\% of
+total modal share as of the 2016 Census, but accounts for 20\% of
+serious injuries and 7\% of all fatalities, driving the development of
+cycling policy and infrastructure plans in Ireland. The main frameworks
+underpinning these efforts are the Climate Action Plan
+\citep{climate2022a}, the National Development Plan, and the National
+Roads 2040 strategy \citep{national2023}. These policy frameworks are
+backed-up by public support for active travel interventions: while only
+3\% of trips in Ireland are made by cycling according to the latest
+Census data from 2016 ``83\% of respondents supported enhanced walking
+and cycling facilities'' \citep{sustaina2019}.
 
-At the regional level, the recently published Greater Dublin Area
-Transport Strategy \citep{greater} reinforces these findings: ``nearly a
-quarter of adults cycle at least once a week in the Dublin Metropolitan
-Area'' with cycling in the Dublin area taking up to 60,000 cars off the
-road today. Extrapolating this on a per population basis across Ireland,
-with around 40\% of the population living in Dublin, suggests that
-around 150,000 cars could be removed nationwide just by achieving Dublin
-levels of cycling in all counties (notwithstanding existing cycling
-trips and differences in trip distances). 7 in 10 trips in Ireland are
-by car \citep{national} . Cycling has the potential to replace a large
+At the regional level within Ireland, the recently published Greater
+Dublin Area Transport Strategy \citep{greater} reinforces these findings
+of high support for and capability for cycling: ``nearly a quarter of
+adults cycle at least once a week in the Dublin Metropolitan Area'' with
+cycling in the Dublin area taking up to 60,000 cars off the road today.
+Extrapolating this on a per population basis across Ireland, with around
+40\% of the population living in Dublin, suggests that around 150,000
+cars could be removed nationwide just by achieving Dublin levels of
+cycling in all counties (notwithstanding existing cycling trips and
+differences in trip distances). 7 in 10 trips in Ireland are by car
+\citep{national} . Cycling has the potential to replace a large
 proportion of these trips: ``A high priority must also be given to
@@
 
-Further evidence of the importance of cycling in Ireland \emph{already}
-is provided by the National Strategic Objective (NSO) from the National
-Development Plan, with €8.6 billion allocated to sustainable transport
+Further evidence of the importance of cycling in Ireland is provided by
+the National Strategic Objective (NSO) from the National Development
+Plan, which allocates €8.6 billion to sustainable transport
 infrastructure including public transport and active travel
@@
 the BusConnects project, an entire redesign of the bus network in Dublin
-and Cork.
+and Cork. It was in this context that Transport Infrastructure Ireland
+(TII) commissioned the research reported in this paper, which led to the
+Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
+Building on previous work, including the Propensity to Cycle Tool (PCT)
+for England and Wales, the CRUSE Tool was developed to provide evidence
+on current cycling levels and future cycling potential nationwide across
+Ireland.
 
-It was in this context that Transport Infrastructure Ireland (TII)
-commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool
-for Ireland. Based on the Propensity to Cycle Tool (PCT) for England and
-Wales, the CRUSE Tool was developed to provide evidence on current
-cycling levels and future cycling potential nationwide across Ireland.
-
-A key consideration for TII is to ensure both urban and rural Ireland
-were targeted for cycling infrastructure, as shown in
-Figure~\ref{fig-dublin}. To this end, TII extensively reviewed activity
-on Irish Roads and how the road network was being used by commuters,
-cyclists, heavy goods vehicles and other road users across the Irish
-transport network. The aim was for the tool to provide strong, national,
-systematic but locally-specific evidence to monitor cycling friendliness
-and safety in Ireland, to ensure strategic alignment with national,
-regional and local policies. The open source and publicly available
-nature of the tool ensures more inclusive and evidence-based
-conversations around cycle network planning between all stakeholders in
-the transport planning process.
-
-TII also requires evidence to support monitoring of growth in cycling,
-upgrades of existing cycling infrastructure, business case development
-and appraisal, and exposure information to estimate crash rates.
-Exposure information is a key part of the the European RISM Directive,
-for reporting collision rates of vulnerable road users by 2024. These
-factors mean that the tool can be seen as an open access `leverage
-point', providing key information for and supporting ambitious plans in
-many aspects the planning system, from network design to post-build
-monitoring \citep{lovelace2020}.
+Stakeholder consultation by TII emphasized the need for a strong,
+national, systematic but locally-specific evidence base on cycle
+networks in both urban and rural Ireland. The evidence base had to scale
+nationally to support strategic alignment with national, regional and
+local policies, but also be useful for local network planning. TII also
+requires
 
@@
 \citep{scappini2022} we expect the demand for tools such as CRUSE to
-grow.
+grow. However, the approach is not without limitations and does not meet
+all requirements in Ireland or indeed any country where the methods are
+applied.
+
+A limitation of the approach as implemented in the CRUSE tool is that
+the results are `static', reducing its ability to support monitoring of
+growth in cycling, upgrades of existing cycling infrastructure and
+dynamic (year-by-year) exposure information to estimate crash rates.
+Exposure information is a key part of the the European RISM Directive,
+for reporting collision rates of vulnerable road users by 2024. These
+factors mean that the tool can be seen as an open access `leverage
+point', providing key information for and supporting ambitious plans in
+many aspects the planning system, from network design to post-build
+monitoring \citep{lovelace2020}.
 

### 2024-03-23 paper.qmd
Add content from @eugenividal

@@
 
+# Tools for estimating cycling potential
+
+The tools for estimating geographical distribution of cycling potential have advanced substantially in recent years.
+Developed by both researchers and professionals, they have evolved from simple area-based static models tailored to specific regions to more complex generalisable open-source online maps at several levels of analysis open to the public.
+Table 1 summarises the most relevant.
+
+Parkin et al.
+(2007) presented a regression model to estimate the percentage of cycling trips to work in England and Wales based on socioeconomic, transportation, and physical factors at the small area (wards) level.
+They found that areas with a higher percentage of females, non-whites, car ownership, lower socioeconomic classes, income, distance to work, population density, poor highway conditions, hills, rainfall, and fewer off-road bicycle routes tended to have a lower proportion of cycling to work.
+
+Transport for London (2010) created the Analysis of Cycling Potential (ACP) tool to identify current trips that could be cycled instead of using mechanised modes.
+The tool is based on a mode choice model that assesses the likelihood of trips being cycled and uses data from the TfL London Travel Demand Survey as its primary input dataset.
+The tool generates tables, graphs, and heatmaps showing the potential for cycling for different trips purposes in London.
+This information has been used to guide local cycling initiatives, such as determining locations for new hire stations.
+The tool, however, it does not use origin-destination data, which makes it less useful for route-based interventions.
+An updated analysis was conducted in 2017 (Transport for London, 2016).
+
+Larsen et al.
+(2013) designed a tool to prioritize the placement of new cycling facilities in Montreal, Canada.
+The tool used survey, vehicle collision, and origin-destination data.
+To create the index, they first identified a set of relevant indicators for prioritising locations for infrastructure investment.
+Then, they imposed a grid and aggregated the indicators spatially, associating them with grid’s cells.
+Finally, they combined the aggregated grid cell indicators into the prioritization index.
+The tool generated heatmaps recommending cycling infrastructure upgrades in specific locations.
+
+Zhang et al.
+(2014) used a regression model to examine the effect of socioeconomic factors on willingness to cycle in Belo Horizonte City, Brazil, while considering individual heterogeneity.
+To develop the model, they employed a bicycle use survey conducted in Belo Horizonte.
+The results indicated a positive association between willingness to cycle and middle income, people renting apartments, and higher education, and a negatively association with distance to work and car use.
+Subsequently, using the calibrated model along with census and OD data, they estimated the willingness to commute cycling for the entire city.
+
+Phillips and Range (2017) created the Cycling Potential Tool (CPT) for Scotland to provide evidence-based insights for decision-making regarding local cycling infrastructure in Scotland.
+The CPT is a GIS toolbox consisting of two main modules: the base environmental module and the quality of service module.
+The base environmental module is based on eight weighted factors (population density, hilliness, physical barriers, access to services, existing cycling mode share, distance to work and school, and road speed) and produces heatmaps for each factor as well as a combined score.
+The quality of service module evaluates the network within each area of interest based on eight factors (surface condition, adjacent cyclists, comfort factor, conflict, distance between junctions, slope, access to services, and origin/destination).
+
+The Propensity to Cycle Tool (2017) was initially based on current and future potential uptake of cycling for single stage travel to work at desire line, zone, route, and route network levels (Lovelace et al.
+2017).
+Extensions of the PCT approach have included estimation of benefits at the individual level (Woodcock et al., 2018), addition of travel to school network (Goodman et al., 2019), and improved modeling of impacts on health, environmental and distributional outcomes (Woodcock et al., 2021).
+Initially developed just for England, the PCT was extended to cover all of Wales (for commuter data only) in 2018.
+The PCT approach has been applied in other countries, including Ireland (the topic of this paper), Scotland, and Portugal.
+In Portugal the ‘biclaR’ project, based on methods underlying the PCT, has been developed and deployed for the Lisbon metro region.
+The resulting evidence is available in an interactive web application hosted at biclar.tmlmobilidade.pt (Félix et al., 2022).
+biclaR includes estimates of impacts, using the World Health Organisation (WHO) ‘HEAT for Cycling’ tool and an ‘intermodality’ scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
+An important feature of the PCT is that it is open source and publicly available (at www.pct.bike), allowing its use by all stakeholders in the transport planning process.
+
+Silva et al.
+(2021)  designed the Gross Potential for Cycling (GPC) tool to identify key areas for the implementation of cycling infrastructure and other cycling measures.
+The GPC uses two sets of indicators: population-based indicators (age, potential demand density, employment density and motorization rate) and area-based indicators (accessibility, education, public transport, connectivity, land use mix and relative performance).
+These indicators are calculated considering topography, road hierarchy, and average congestion and presented on a scale from 5 to 10.
+They are then combined into an overall score, weighted by their impact on cycling according to the literature.
+The GPC is publicly available (at https://boost.up.pt/en/ferramentas/gpc) and is specially designed for local council-level practitioners.
+Its practical value has been assessed through a series of experiential workshops involving planning practitioners from various Portuguese contexts (Silva et al., 2022).
+
+The methods used to generate the evidence present in the CRUSE Tool, described in the next section, build on the PCT and seek to overcome three of the PCT methodological limitations: 1) low resolution of data, with routes starting and ending in administrative zone centroids, 2) limited coverage of trip purposes beyond travel to work and school, and 3) a web interface that was not user-friendly or intuitive.
+
 # Methods and data {#sec-methods}
-
+<!-- 
 The methods used to generate the evidence presented in the CRUSE Tool build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
@@
 
-Methods were developed to overcome each of these, as outlined in @sec-trip-purposes to @sec-ui.
+Methods were developed to overcome each of these, as outlined in @sec-trip-purposes to @sec-ui. -->
 

### 2024-03-23 paper.tex
Add content from @eugenividal

@@
 
+\section{Tools for estimating cycling
+potential}\label{tools-for-estimating-cycling-potential}
+
+The tools for estimating geographical distribution of cycling potential
+have advanced substantially in recent years. Developed by both
+researchers and professionals, they have evolved from simple area-based
+static models tailored to specific regions to more complex generalisable
+open-source online maps at several levels of analysis open to the
+public. Table 1 summarises the most relevant.
+
+Parkin et al.~(2007) presented a regression model to estimate the
+percentage of cycling trips to work in England and Wales based on
+socioeconomic, transportation, and physical factors at the small area
+(wards) level. They found that areas with a higher percentage of
+females, non-whites, car ownership, lower socioeconomic classes, income,
+distance to work, population density, poor highway conditions, hills,
+rainfall, and fewer off-road bicycle routes tended to have a lower
+proportion of cycling to work.
+
+Transport for London (2010) created the Analysis of Cycling Potential
+(ACP) tool to identify current trips that could be cycled instead of
+using mechanised modes. The tool is based on a mode choice model that
+assesses the likelihood of trips being cycled and uses data from the TfL
+London Travel Demand Survey as its primary input dataset. The tool
+generates tables, graphs, and heatmaps showing the potential for cycling
+for different trips purposes in London. This information has been used
+to guide local cycling initiatives, such as determining locations for
+new hire stations. The tool, however, it does not use origin-destination
+data, which makes it less useful for route-based interventions. An
+updated analysis was conducted in 2017 (Transport for London, 2016).
+
+Larsen et al.~(2013) designed a tool to prioritize the placement of new
+cycling facilities in Montreal, Canada. The tool used survey, vehicle
+collision, and origin-destination data. To create the index, they first
+identified a set of relevant indicators for prioritising locations for
+infrastructure investment. Then, they imposed a grid and aggregated the
+indicators spatially, associating them with grid's cells. Finally, they
+combined the aggregated grid cell indicators into the prioritization
+index. The tool generated heatmaps recommending cycling infrastructure
+upgrades in specific locations.
+
+Zhang et al.~(2014) used a regression model to examine the effect of
+socioeconomic factors on willingness to cycle in Belo Horizonte City,
+Brazil, while considering individual heterogeneity. To develop the
+model, they employed a bicycle use survey conducted in Belo Horizonte.
+The results indicated a positive association between willingness to
+cycle and middle income, people renting apartments, and higher
+education, and a negatively association with distance to work and car
+use. Subsequently, using the calibrated model along with census and OD
+data, they estimated the willingness to commute cycling for the entire
+city.
+
+Phillips and Range (2017) created the Cycling Potential Tool (CPT) for
+Scotland to provide evidence-based insights for decision-making
+regarding local cycling infrastructure in Scotland. The CPT is a GIS
+toolbox consisting of two main modules: the base environmental module
+and the quality of service module. The base environmental module is
+based on eight weighted factors (population density, hilliness, physical
+barriers, access to services, existing cycling mode share, distance to
+work and school, and road speed) and produces heatmaps for each factor
+as well as a combined score. The quality of service module evaluates the
+network within each area of interest based on eight factors (surface
+condition, adjacent cyclists, comfort factor, conflict, distance between
+junctions, slope, access to services, and origin/destination).
+
+The Propensity to Cycle Tool (2017) was initially based on current and
+future potential uptake of cycling for single stage travel to work at
+desire line, zone, route, and route network levels (Lovelace et
+al.~2017). Extensions of the PCT approach have included estimation of
+benefits at the individual level (Woodcock et al., 2018), addition of
+travel to school network (Goodman et al., 2019), and improved modeling
+of impacts on health, environmental and distributional outcomes
+(Woodcock et al., 2021). Initially developed just for England, the PCT
+was extended to cover all of Wales (for commuter data only) in 2018. The
+PCT approach has been applied in other countries, including Ireland (the
+topic of this paper), Scotland, and Portugal. In Portugal the `biclaR'
+project, based on methods underlying the PCT, has been developed and
+deployed for the Lisbon metro region. The resulting evidence is
+available in an interactive web application hosted at
+biclar.tmlmobilidade.pt (Félix et al., 2022). biclaR includes estimates
+of impacts, using the World Health Organisation (WHO) `HEAT for Cycling'
+tool and an `intermodality' scenario that combines cycling with
+currently available public transit options based on General Transit Feed
+Specification (GTFS) data. An important feature of the PCT is that it is
+open source and publicly available (at www.pct.bike), allowing its use
+by all stakeholders in the transport planning process.
+
+Silva et al.~(2021) designed the Gross Potential for Cycling (GPC) tool
+to identify key areas for the implementation of cycling infrastructure
+and other cycling measures. The GPC uses two sets of indicators:
+population-based indicators (age, potential demand density, employment
+density and motorization rate) and area-based indicators (accessibility,
+education, public transport, connectivity, land use mix and relative
+performance). These indicators are calculated considering topography,
+road hierarchy, and average congestion and presented on a scale from 5
+to 10. They are then combined into an overall score, weighted by their
+impact on cycling according to the literature. The GPC is publicly
+available (at https://boost.up.pt/en/ferramentas/gpc) and is specially
+designed for local council-level practitioners. Its practical value has
+been assessed through a series of experiential workshops involving
+planning practitioners from various Portuguese contexts (Silva et al.,
+2022).
+
+The methods used to generate the evidence present in the CRUSE Tool,
+described in the next section, build on the PCT and seek to overcome
+three of the PCT methodological limitations: 1) low resolution of data,
+with routes starting and ending in administrative zone centroids, 2)
+limited coverage of trip purposes beyond travel to work and school, and
+3) a web interface that was not user-friendly or intuitive.
+
 \section{Methods and data}\label{sec-methods}
-
-The methods used to generate the evidence presented in the CRUSE Tool
-build on the Propensity to Cycle Tool (PCT), which was originally funded
-by the UK's Department for Transport and developed by a multi-university
-team. An important feature of the PCT is that it is open source and
-publicly available (at \href{https://www.pct.bike/}{www.pct.bike}),
-allowing its use by all stakeholders in the transport planning process
-\citep{lovelace2017}. The PCT approach has had major policy impacts, as
-outlined in Research Excellence Framework (REF) impact case studies,
-which demonstrate that the tool ``revolutionized strategic cycle
-planning in England and Wales'' by overcoming the barriers to cycling
-investment imposed by lack of evidence on cycling
-potential\citep{lovelace2023}.
-
-The first version of the PCT was based on current and future potential
-uptake of cycling for single stage \emph{travel to work} at desire line,
-zone, route, and route network levels \citep{lovelace2016}. It was
-launched in April 2017 as the government-endorsed tool for strategic
-cycle network planning and part of the Cycling and Walking Investment
-Strategy \citep{cycling2017}. Extensions of the PCT approach have
-included estimation of benefits at the individual level
-\citep{woodcock2018}, addition of travel to school network
-\citep{goodman2019}, and improved modeling of impacts on health,
-environmental and distributional outcomes \citep{woodcock2021}.
-Initially developed just for England, the PCT was extended to cover all
-of Wales (for commuter data only) in 2018.
-
-The PCT approach has been applied in other countries, including Ireland
-(the topic of this paper), Scotland, and Portugal. In Portugal the
-`biclaR' project, based on methods underlying the PCT, has been
-developed and deployed for the Lisbon metro region. The resulting
-evidence is available in an interactive web application hosted at
-\href{https://biclar.tmlmobilidade.pt}{biclar.tmlmobilidade.pt}
-\citep{felix2023}. biclaR includes estimates of impacts, using the World
-Health Organisation (WHO) `HEAT for Cycling' tool and an `intermodality'
-scenario that combines cycling with currently available public transit
-options based on General Transit Feed Specification (GTFS) data.
-
-The CRUSE Tool seeks to overcome the following methodological
-limitations of the original PCT:
-
-\begin{itemize}
-\tightlist
-\item
-  Low resolution of data, with routes starting and ending in
-  administrative zone centroids
-\item
-  Limited coverage of trip purposes beyond travel to work and school
-\item
-  A web interface that was not user-friendly or intuitive
-\end{itemize}
-
-Methods were developed to overcome each of these, as outlined in
-Section~\ref{sec-trip-purposes} to Section~\ref{sec-ui}.
 

### 2024-03-23 paper.qmd
Add references to new section

--- 2024-03-22 paper.qmd
@@
 subtitle: A case study of Ireland
-bibliography: bibliography.bib
+bibliography:
+  # cruse.bib
+  - bibliography.bib
+  - cruse.bib
 csl: apa.csl
@@
 
+# Tools for estimating cycling potential
+
+Tools for estimating geographical distribution of cycling potential have advanced substantially in recent years.
+Developed by researchers and developers in academic, public and private sectors, they have evolved from simple area-based static models tailored to specific regions to more complex and (in some cases) more generalizable tools.
+Outputs include maps with results available at several levels of analysis with levels of availability ranging from being only available to researchers and in static maps to open access mapping systems available to the public.
+Some of the most relevant existing tools are summarised below and presented in Table 1.
+
+Area-based tools were developed first.
+Parkin and colleagues presented a regression model to estimate the percentage of cycling trips to work in England and Wales based on socioeconomic, transportation, and physical factors at the small area (wards) level in 2007 [@parkin_estimation_2007].
+They found that areas with a higher percentage of females, non-whites, car ownership, lower socioeconomic classes, income, distance to work, population density, poor highway conditions, hills, rainfall, and fewer off-road bicycle routes tended to have a lower proportion of cycling to work.
+
+Three years later, Transport for London created the Analysis of Cycling Potential (ACP) tool based on a mode choice model that assesses the likelihood of trips being cycled, with data from the TfL London Travel Demand Survey as its primary input.
+The tool generates tables, graphs, and heatmaps showing the potential for cycling for different trips purposes in London to guide local cycling initiatives, such as where new hire 'docking stations' should go.
+The tool does not use origin-destination data and cannot estimate cycling potential on specific routes [@transport_for_london_analysis_2010].
+
+Other tools that generate areal data to prioritize investment include a study that generated a 'prioritization index' for Montreal, Canada [@larsen_build_2013], a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil, and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
+The CPT is composed of the base environmental module — based on eight weighted factors (population density, hilliness, physical barriers, access to services, existing cycling mode share, distance to work and school, and road speed) — and the quality of service module, which evaluates each area of interest based on eight factors (surface condition, adjacent cyclists, comfort factor, conflict, distance between junctions, slope, access to services, and origin/destination).
+
+A more recent areal-based approach is the designed the Gross Potential for Cycling (GPC) tool to prioritize areas for cycling infrastructure and other cycling measures [@silva_gross_2021].
+The GPC uses two sets of indicators: population-based indicators (age, potential demand density, employment density and motorization rate) and area-based indicators (accessibility, education, public transport, connectivity, land use mix and relative performance).
+These indicators are calculated considering topography, road hierarchy, and average congestion and presented on a scale from 5 to 10.
+They are then combined into an overall score, weighted by their impact on cycling according to the literature.
+The ranking of areas by the GPC are [publicly available](https://boost.up.pt/en/ferramentas/gpc) and its practical value has been assessed through workshops [@silva_revealing_2022].
+
+Route-based tools emerged with increasing availability of origin-destination data, routing engines that can assign trips to networks, and improvements in computer hardware and software needed to generate route networks.
+A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace_mapping_2016].
+The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 @lovelace2017.
+Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modeling of impacts on health, environmental and distributional outcomes [@woodcock2021].
+Initially developed just for England, the PCT was extended to cover all of Wales (for commuter data only) in 2018.
+
+The PCT approach has been applied in other countries, including Ireland (the topic of this paper), Scotland, and Portugal.
+In Portugal the 'biclaR' project, based on methods underlying the PCT, has been developed and deployed for the Lisbon metro region.
+The resulting evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
+biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
+
+The approach presented in this paper seeks to overcome three limitations of previous tools to estimate cycling potential: 1) low resolution of data, with routes starting and ending in administrative zone centroids, 2) limited coverage of trip purposes beyond travel to work and school, and 3) a web interface that was not user-friendly or intuitive.
+
 # Methods and data {#sec-methods}
 
+```{=html}
+<!-- 
 The methods used to generate the evidence presented in the CRUSE Tool build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
@@
 
-Methods were developed to overcome each of these, as outlined in @sec-trip-purposes to @sec-ui.
-
+Methods were developed to overcome each of these, as outlined in @sec-trip-purposes to @sec-ui. -->
+```
 <!-- Todo: add section on input datasets -->

### 2024-03-23 paper.tex
Add references to new section

--- 2024-03-22 paper.tex
@@
 
-\section{Methods and data}\label{sec-methods}
+\section{Tools for estimating cycling
+potential}\label{tools-for-estimating-cycling-potential}
 
-The methods used to generate the evidence presented in the CRUSE Tool
-build on the Propensity to Cycle Tool (PCT), which was originally funded
-by the UK's Department for Transport and developed by a multi-university
-team. An important feature of the PCT is that it is open source and
-publicly available (at \href{https://www.pct.bike/}{www.pct.bike}),
-allowing its use by all stakeholders in the transport planning process
-\citep{lovelace2017}. The PCT approach has had major policy impacts, as
-outlined in Research Excellence Framework (REF) impact case studies,
-which demonstrate that the tool ``revolutionized strategic cycle
-planning in England and Wales'' by overcoming the barriers to cycling
-investment imposed by lack of evidence on cycling
-potential\citep{lovelace2023}.
+Tools for estimating geographical distribution of cycling potential have
+advanced substantially in recent years. Developed by researchers and
+developers in academic, public and private sectors, they have evolved
+from simple area-based static models tailored to specific regions to
+more complex and (in some cases) more generalizable tools. Outputs
+include maps with results available at several levels of analysis with
+levels of availability ranging from being only available to researchers
+and in static maps to open access mapping systems available to the
+public. Some of the most relevant existing tools are summarised below
+and presented in Table 1.
 
-The first version of the PCT was based on current and future potential
-uptake of cycling for single stage \emph{travel to work} at desire line,
-zone, route, and route network levels \citep{lovelace2016}. It was
-launched in April 2017 as the government-endorsed tool for strategic
-cycle network planning and part of the Cycling and Walking Investment
-Strategy \citep{cycling2017}. Extensions of the PCT approach have
-included estimation of benefits at the individual level
+Area-based tools were developed first. Parkin and colleagues presented a
+regression model to estimate the percentage of cycling trips to work in
+England and Wales based on socioeconomic, transportation, and physical
+factors at the small area (wards) level in 2007
+\citep{parkin_estimation_2007}. They found that areas with a higher
+percentage of females, non-whites, car ownership, lower socioeconomic
+classes, income, distance to work, population density, poor highway
+conditions, hills, rainfall, and fewer off-road bicycle routes tended to
+have a lower proportion of cycling to work.
+
+Three years later, Transport for London created the Analysis of Cycling
+Potential (ACP) tool based on a mode choice model that assesses the
+likelihood of trips being cycled, with data from the TfL London Travel
+Demand Survey as its primary input. The tool generates tables, graphs,
+and heatmaps showing the potential for cycling for different trips
+purposes in London to guide local cycling initiatives, such as where new
+hire `docking stations' should go. The tool does not use
+origin-destination data and cannot estimate cycling potential on
+specific routes \citep{transport_for_london_analysis_2010}.
+
+Other tools that generate areal data to prioritize investment include a
+study that generated a `prioritization index' for Montreal, Canada
+\citep{larsen_build_2013}, a `usage intensity index' based on stated
+preferences in Belo Horizonte, Brazil, and a Cycling Potential Tool
+(CPT) for Scotland \citep{phillips_development_2017}. The CPT is
+composed of the base environmental module --- based on eight weighted
+factors (population density, hilliness, physical barriers, access to
+services, existing cycling mode share, distance to work and school, and
+road speed) --- and the quality of service module, which evaluates each
+area of interest based on eight factors (surface condition, adjacent
+cyclists, comfort factor, conflict, distance between junctions, slope,
+access to services, and origin/destination).
+
+A more recent areal-based approach is the designed the Gross Potential
+for Cycling (GPC) tool to prioritize areas for cycling infrastructure
+and other cycling measures \citep{silva_gross_2021}. The GPC uses two
+sets of indicators: population-based indicators (age, potential demand
+density, employment density and motorization rate) and area-based
+indicators (accessibility, education, public transport, connectivity,
+land use mix and relative performance). These indicators are calculated
+considering topography, road hierarchy, and average congestion and
+presented on a scale from 5 to 10. They are then combined into an
+overall score, weighted by their impact on cycling according to the
+literature. The ranking of areas by the GPC are
+\href{https://boost.up.pt/en/ferramentas/gpc}{publicly available} and
+its practical value has been assessed through workshops
+\citep{silva_revealing_2022}.
+
+Route-based tools emerged with increasing availability of
+origin-destination data, routing engines that can assign trips to
+networks, and improvements in computer hardware and software needed to
+generate route networks. A prominent example is the the Propensity to
+Cycle Tool (PCT), first developed to estimate current and future levels
+of cycling at desire line, zone, route, and route network levels for
+case study cities \citep{lovelace_mapping_2016}. The approach was
+scaled-up to estimate estimate the potential benefits of uptake at zone
+and desire line levels nationally, and launched as a publicly available
+web application in 2017 \citet{lovelace2017}. Extensions of the PCT
+approach have included estimation of benefits at the individual level
 \citep{woodcock2018}, addition of travel to school network
@@
 
-The CRUSE Tool seeks to overcome the following methodological
-limitations of the original PCT:
+The approach presented in this paper seeks to overcome three limitations
+of previous tools to estimate cycling potential: 1) low resolution of
+data, with routes starting and ending in administrative zone centroids,
+2) limited coverage of trip purposes beyond travel to work and school,
+and 3) a web interface that was not user-friendly or intuitive.
 
-\begin{itemize}
-\tightlist
-\item
-  Low resolution of data, with routes starting and ending in
-  administrative zone centroids
-\item
-  Limited coverage of trip purposes beyond travel to work and school
-\item
-  A web interface that was not user-friendly or intuitive
-\end{itemize}
-
-Methods were developed to overcome each of these, as outlined in
-Section~\ref{sec-trip-purposes} to Section~\ref{sec-ui}.
+\section{Methods and data}\label{sec-methods}
 
@@
 
-  \bibliography{bibliography.bib}
+  \bibliography{bibliography.bib,cruse.bib}
 

### 2024-03-23 paper.qmd
Update paper with tools table

@@
     wrap: sentence
+execute: 
+  echo: false
+  message: false
 number-sections: true
@@
 #| include: false
+library(tidyverse)
+```
+
+```{r}
+#| include: false
 #| eval: false
-library(tidyverse)
 remotes::install_dev("tmap")
@@
 Some of the most relevant existing tools are summarised below and presented in Table 1.
+
+```{r}
+#| include: false
+tools = read_csv("tools.csv")
+tools = tools |>
+  transmute(
+    `Approach` = paste0(Tool, ", ", Reference),
+    `Coverage and input data` = paste0(Coverage, ", ", `Input data`),
+    `Outputs and availability` = Output,
+    License = Licence
+  )
+```
+
+```{r}
+knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
+  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
+  kableExtra::column_spec(1, width = "13em") |>
+  kableExtra::column_spec(2, width = "15em") |>
+  kableExtra::column_spec(3, width = "12em") |>
+  kableExtra::column_spec(4, width = "5em")
+```
 

### 2024-03-23 paper.tex
Update paper with tools table

@@
 
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \makeatletter
@@
 
+\begin{longtable}[t]{>{\raggedright\arraybackslash}p{13em}>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{12em}>{\raggedright\arraybackslash}p{5em}}
+\caption{\label{tab:unnamed-chunk-4}Summary of existing tools for estimating cycling potential.}\\
+\toprule
+Approach & Coverage and input data & Outputs and availability & License\\
+\midrule
+\cellcolor{gray!10}{Bicycle share model, Parkin et al.
+(2007)} & \cellcolor{gray!10}{England and Wales, Journeys to work OD census data, at the small-area (wards) level} & \cellcolor{gray!10}{Static tables in academic paper} & \cellcolor{gray!10}{Proprietary}\\
+Analysis of  Cycling Potential (ACP), Transport for London (2010, 2016) & London, London Travel Demand Survey & Static tables, graphs, and heatmaps, not publicly available & Proprietary\\
+\cellcolor{gray!10}{Prioritization index, Larsen et al.
+(2013)} & \cellcolor{gray!10}{Montreal, Survey, Road safety, and OD data} & \cellcolor{gray!10}{Map-based heatmap, not publicly available} & \cellcolor{gray!10}{Proprietary}\\
+Usage intensity index, Zhang et al. (2014) & Belo Horizonte, Survey, census, and OD data & Static tables and maps, not publicly available & Proprietary\\
+\cellcolor{gray!10}{The Cycling Potential Tool (CPT), Phillips and Range (2017)} & \cellcolor{gray!10}{Scotland, Environmental and socioeconomic data, at the small area (output areas) level} & \cellcolor{gray!10}{Maps showing cycling potential in each area} & \cellcolor{gray!10}{Proprietary}\\
+\addlinespace
+Propensity to Cycle Tool (PCT), Lovelace et al.
+(2017) & England and Wales, Journeys to work and school OD census data, at the small area (LSOA) level & Online maps, graphs, tables, publicly available at www.pct.bike & Open source\\
+\cellcolor{gray!10}{The    Gross Potential for Cycling tool (CPC), Silva et al.
+(2021 and 2022)} & \cellcolor{gray!10}{21 Portuguese cities, Land use and socio-demographic data, at the small area (census tract) level} & \cellcolor{gray!10}{Static maps showing cycling potential in different areas} & \cellcolor{gray!10}{NA}\\
+\bottomrule
+\end{longtable}
+
 Area-based tools were developed first. Parkin and colleagues presented a

### 2024-03-23 paper.qmd
Add table caption ++

@@
 Outputs include maps with results available at several levels of analysis with levels of availability ranging from being only available to researchers and in static maps to open access mapping systems available to the public.
-Some of the most relevant existing tools are summarised below and presented in Table 1.
+Some of the most relevant existing tools are summarised below and presented in @tbl-tools.
 
@@
 ```{r}
+#| label: tbl-tools
 knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>

### 2024-03-23 paper.tex
Add table caption ++

@@
 public. Some of the most relevant existing tools are summarised below
-and presented in Table 1.
+and presented in Table~\ref{tbl-tools}.
 
 \begin{longtable}[t]{>{\raggedright\arraybackslash}p{13em}>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{12em}>{\raggedright\arraybackslash}p{5em}}
-\caption{\label{tab:unnamed-chunk-4}Summary of existing tools for estimating cycling potential.}\\
+
+\caption{\label{tbl-tools}Summary of existing tools for estimating
+cycling potential.}
+
+\tabularnewline
+
+\\
 \toprule
@@
 \bottomrule
+
 \end{longtable}

### 2024-03-23 paper.qmd
some edits

@@
         address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
-  - name: "Gary O'Tool"
+  - name: "Gary O'Toole"
     affiliations:
@@
 CRUSE goes beyond previous work by: modelling networks at high levels of geographic resolution; simulating multiple trip purposes (social, shopping and personal utility trips plus recreational and cycle touring trips), in addition to using official data on trips to work and education locations; provision of estimates of 'quietness' (a proxy for cyclability) at the route segment level; and its provision of three network types: 'Fastest', 'Balanced' and 'Quietest' to help plan both arterial and residential cycle networks.
-Workshops and unstructure interview is stakeholders show that the tool is already being used in practice, to inform urban, inter-urban and rural network development.
+Workshops and unstructured interviews of stakeholders show that the tool is already being used in practice, to inform urban, inter-urban and rural network development.
 The approach and methods for strategic cycle network development tools outlined in this paper can be applied and built-on in other places to support the generation of evidence needed to support ambitious cycling policies and interventions internationally.
@@
 # latexdiffr::latexdiff("paper-v1.qmd", "paper.qmd")
-system("latexdiff paper-v1.tex paper.tex > paper-diff.tex")
-system("pdflatex paper-diff.tex")
-browseURL("paper-diff.pdf")
+system("latexdiff paper-v1.tex paper.tex > tracked-changes.tex")
+system("pdflatex tracked-changes.tex")
+browseURL("tracked-changes.pdf")
 ```
@@
 
-Growing evidence of the negative impacts of car-dependent transport systems has led governments in many countries to set targets and taken actions.
+Growing evidence of the negative impacts of car-dependent transport systems has led governments in many countries to set targets and take actions.
 In the context of climate, road safety and physicial inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
@@
 Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out" [@directiv2008].
-Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance traveled, the directive requires estimation of distance traveled by mode, down to the road link level.
+Given that 'safety' in this context is usefully quantified as the number of people killed or seriously injured (KSI) per distance traveled, the directive requires estimation of distance traveled by mode, down to the road link level.
 For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
@@
 The evidence base had to scale nationally to support strategic alignment with national, regional and local policies, but also be useful for local network planning.
-TII also requires
 
-::: {#fig-dublin}
-<!-- ![](images/paste-1.png) -->
+::: {#fig-landing}
+![](images/paste-10.png)
 
-![](images/paste-8.png) <!-- ![](images/paste-3.png) -->
-![](images/paste-9.png)
-
-Illustration of the density of route networks generated by the CRUSE Tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland (botton).
-Source: CRUSE Tool, available at <https://cruse.bike/>.
+Illustration of the landing page of the CRUSE Tool for Ireland.
 :::
 
-This paper aims to describe the design, features and potential use of the CRUSE Tool.
+This describes describe the design, features and potential use of an approach to strategic cycle network planning that is open, scalable, and evidence-based.
+We label the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and present a case study of its application in Ireland, the results of which are publicly available at [cruse.bike](https://cruse.bike/) in the CRUSE Tool for Ireland, the landing page of which is illustrated in @fig-landing.
 In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool.
 In @sec-results, we present the results of the CRUSE Tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
-In @sec-discussion, we discuss the implications of the results for policy and practice.
-In @sec-conclusions, we conclude the paper.
+In @sec-discussion, we discuss limitations and possible future improvements to the approach, and the implications of the results for Ireland before making concluding remarks in @sec-conclusions.
 
@@
 As shown in @fig-dublin, the resulting route networks are dense, even in rural areas.
+
+::: {#fig-dublin}
+<!-- ![](images/paste-1.png) -->
+
+![](images/paste-8.png) <!-- ![](images/paste-3.png) -->
+![](images/paste-9.png)
+
+Illustration of the density of route networks generated by the CRUSE Tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland (botton).
+Source: CRUSE Tool, publicly available at [cruse.bike](https://cruse.bike/).
+:::
 
@@
 
-The aim of the paper was to describe the design, features and potential use of the CRUSE Tool.
+The aim of the paper was to describe the design, features and potential use of the 'CRUSE' approach and resulting tool.
 As outlined in the previous section, the CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
 This provision of multiple scenarios of behavior change *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.
+
+## Uses of the tool in practice
 
@@
 It will also support reporting for the RISM Directive on road casualties as part of the road safety management system.
+
+## Comparison with other tools and contribution to the field
 

### 2024-03-23 paper.qmd
Tweak abstract

@@
         address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
-  - name: "Gary O'Toole"
+  - name: "Gary O'Tool"
     affiliations:
@@
 abstract: |
-  This paper describes an approach to the development of strategic cycle network planning tools.
+  This paper describes an approach for developing strategic cycle network planning tools.
 Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the methods used to generate the evidence presented in the tool, including disaggregation of origin-destination data, additional trip purposes, routing, scenarios of cycling potential, and the user interface.
@@
 CRUSE goes beyond previous work by: modelling networks at high levels of geographic resolution; simulating multiple trip purposes (social, shopping and personal utility trips plus recreational and cycle touring trips), in addition to using official data on trips to work and education locations; provision of estimates of 'quietness' (a proxy for cyclability) at the route segment level; and its provision of three network types: 'Fastest', 'Balanced' and 'Quietest' to help plan both arterial and residential cycle networks.
-Workshops and unstructured interviews of stakeholders show that the tool is already being used in practice, to inform urban, inter-urban and rural network development.
+Workshops and unstructure interview is stakeholders show that the tool is already being used in practice, to inform urban, inter-urban and rural network development.
 The approach and methods for strategic cycle network development tools outlined in this paper can be applied and built-on in other places to support the generation of evidence needed to support ambitious cycling policies and interventions internationally.
@@
 
-Growing evidence of the negative impacts of car-dependent transport systems has led governments in many countries to set targets and take actions.
+Growing evidence of the negative impacts of car-dependent transport systems has led governments in many countries to set targets and taken actions.
 In the context of climate, road safety and physicial inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
@@
 Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out" [@directiv2008].
-Given that 'safety' in this context is usefully quantified as the number of people killed or seriously injured (KSI) per distance traveled, the directive requires estimation of distance traveled by mode, down to the road link level.
+Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance traveled, the directive requires estimation of distance traveled by mode, down to the road link level.
 For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.

### 2024-03-24 paper.qmd
Update title

--- 2024-03-23 paper.qmd
@@
 ---
-title: "Cycle Route Uptake and Scenario Estimation (CRUSE): an open and nationally scalable approach for strategic cycle network planning tool development"
-subtitle: A case study of Ireland
+title: "Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools"
+# subtitle: A case study of Ireland
 bibliography:

### 2024-03-24 paper.tex
Update title

@@
 \hypersetup{
-  pdftitle={Cycle Route Uptake and Scenario Estimation (CRUSE): an open and nationally scalable approach for strategic cycle network planning tool development},
+  pdftitle={Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools},
   pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Tool; Dan Brennan; Suzanne Meade},
@@
 \begin{frontmatter}
-\title{Cycle Route Uptake and Scenario Estimation (CRUSE): an open and
-nationally scalable approach for strategic cycle network planning tool
-development \\\large{A case study of Ireland} }
+\title{Cycle Route Uptake and Scenario Estimation (CRUSE): an approach
+for developing strategic cycle network planning tools}
 \author[1]{Robin Lovelace%
@@
 \begin{abstract}
-This paper describes an approach to the development of strategic cycle
-network planning tools. Based on our experience developing and deploying
-the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland,
-we outline the methods used to generate the evidence presented in the
-tool, including disaggregation of origin-destination data, additional
-trip purposes, routing, scenarios of cycling potential, and the user
+This paper describes an approach for developing strategic cycle network
+planning tools. Based on our experience developing and deploying the
+Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we
+outline the methods used to generate the evidence presented in the tool,
+including disaggregation of origin-destination data, additional trip
+purposes, routing, scenarios of cycling potential, and the user
 interface. The CRUSE Tool provides estimates of cycling currently and
@@
 nationally to support strategic alignment with national, regional and
-local policies, but also be useful for local network planning. TII also
-requires
+local policies, but also be useful for local network planning.
 
@@
 
-\includegraphics{images/paste-8.png}
-\includegraphics{images/paste-9.png}
-
-}
-
-\caption{\label{fig-dublin}Illustration of the density of route networks
-generated by the CRUSE Tool for Dublin City and surroundings (top) and
-rural County Mayo on the West coast of Ireland (botton). Source: CRUSE
-Tool, available at \url{https://cruse.bike/}.}
+\includegraphics{images/paste-10.png}
+
+}
+
+\caption{\label{fig-landing}Illustration of the landing page of the
+CRUSE Tool for Ireland.}
 
@@
 
-This paper aims to describe the design, features and potential use of
-the CRUSE Tool. In Section~\ref{sec-methods}, we outline the methods
-used to generate the evidence presented in the CRUSE Tool. In
-Section~\ref{sec-results}, we present the results of the CRUSE Tool,
-including estimates of current cycling levels and future cycling
-potential at the national, regional and local levels. In
-Section~\ref{sec-discussion}, we discuss the implications of the results
-for policy and practice. In Section~\ref{sec-conclusions}, we conclude
-the paper.
+This describes describe the design, features and potential use of an
+approach to strategic cycle network planning that is open, scalable, and
+evidence-based. We label the approach Cycle Route Uptake and Scenario
+Estimation (CRUSE) and present a case study of its application in
+Ireland, the results of which are publicly available at
+\href{https://cruse.bike/}{cruse.bike} in the CRUSE Tool for Ireland,
+the landing page of which is illustrated in Figure~\ref{fig-landing}. In
+Section~\ref{sec-methods}, we outline the methods used to generate the
+evidence presented in the CRUSE Tool. In Section~\ref{sec-results}, we
+present the results of the CRUSE Tool, including estimates of current
+cycling levels and future cycling potential at the national, regional
+and local levels. In Section~\ref{sec-discussion}, we discuss
+limitations and possible future improvements to the approach, and the
+implications of the results for Ireland before making concluding remarks
+in Section~\ref{sec-conclusions}.
 
@@
 
+\begin{figure}
+
+\centering{
+
+\includegraphics{images/paste-8.png}
+\includegraphics{images/paste-9.png}
+
+}
+
+\caption{\label{fig-dublin}Illustration of the density of route networks
+generated by the CRUSE Tool for Dublin City and surroundings (top) and
+rural County Mayo on the West coast of Ireland (botton). Source: CRUSE
+Tool, publicly available at \href{https://cruse.bike/}{cruse.bike}.}
+
+\end{figure}%
+
 \subsection{Additional trip purposes}\label{sec-trip-purposes}
@@
 The aim of the paper was to describe the design, features and potential
-use of the CRUSE Tool. As outlined in the previous section, the CRUSE
-Tool provides evidence on current and potential future cycling levels
-across Ireland down to the street level, with potential assigned to
-`fastest', `balanced' and `quietest' route networks. This provision of
-multiple scenarios of behavior change \emph{and} multiple scenarios of
-investment, for example in cycling infrastructure next to major roads vs
-quiet residential streets, is a key feature of the CRUSE Tool.
+use of the `CRUSE' approach and resulting tool. As outlined in the
+previous section, the CRUSE Tool provides evidence on current and
+potential future cycling levels across Ireland down to the street level,
+with potential assigned to `fastest', `balanced' and `quietest' route
+networks. This provision of multiple scenarios of behavior change
+\emph{and} multiple scenarios of investment, for example in cycling
+infrastructure next to major roads vs quiet residential streets, is a
+key feature of the CRUSE Tool.
+
+\subsection{Uses of the tool in
+practice}\label{uses-of-the-tool-in-practice}
 
@@
 system.
+
+\subsection{Comparison with other tools and contribution to the
+field}\label{comparison-with-other-tools-and-contribution-to-the-field}
 

### 2024-03-24 paper.qmd
Shorten abstract

@@
 Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the methods used to generate the evidence presented in the tool, including disaggregation of origin-destination data, additional trip purposes, routing, scenarios of cycling potential, and the user interface.
-The CRUSE Tool provides estimates of cycling currently and under four scenarios of cycling potential: Near Market, Climate Action Plan, Go Dutch, and Ebike.
-The tool was commissioned by Transport Infrastructure Ireland and is publicly available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions.
-CRUSE goes beyond previous work by: modelling networks at high levels of geographic resolution; simulating multiple trip purposes (social, shopping and personal utility trips plus recreational and cycle touring trips), in addition to using official data on trips to work and education locations; provision of estimates of 'quietness' (a proxy for cyclability) at the route segment level; and its provision of three network types: 'Fastest', 'Balanced' and 'Quietest' to help plan both arterial and residential cycle networks.
+Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of cycling currently and under 'snapshot' scenarios of 'cycling futures' to inform investment decisions.
+The results are publicly available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions.
+CRUSE goes beyond previous work by: modelling networks at high levels of geographic resolution; simulating multiple trip purposes (social, shopping and personal utility trips plus recreational and cycle touring trips), in addition to using official data on trips to work and education locations; and by providing estimates of 'quietness' (a proxy for cyclability) at the route segment level.
+Three network types --- 'Fastest', 'Balanced' and 'Quietest' --- help plan both arterial and residential cycle networks.
 Workshops and unstructured interviews of stakeholders show that the tool is already being used in practice, to inform urban, inter-urban and rural network development.
-The approach and methods for strategic cycle network development tools outlined in this paper can be applied and built-on in other places to support the generation of evidence needed to support ambitious cycling policies and interventions internationally.
+The approach is flexible and open, allowing the underlying ideas and code to be used in other places to generate the evidence needed for effective cycling policies and interventions internationally.
 keywords: 

### 2024-03-24 paper.tex
Shorten abstract

@@
   pdftitle={Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools},
-  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Tool; Dan Brennan; Suzanne Meade},
+  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Toole; Dan Brennan; Suzanne Meade},
   pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
@@
 
-\author[2]{Gary O'Tool%
+\author[2]{Gary O'Toole%
 %
@@
 purposes, routing, scenarios of cycling potential, and the user
-interface. The CRUSE Tool provides estimates of cycling currently and
-under four scenarios of cycling potential: Near Market, Climate Action
-Plan, Go Dutch, and Ebike. The tool was commissioned by Transport
-Infrastructure Ireland and is publicly available at
-\url{https://cruse.bike/}, enabling planners, engineers, and other
-stakeholders to make more evidence-based decisions. CRUSE goes beyond
-previous work by: modelling networks at high levels of geographic
-resolution; simulating multiple trip purposes (social, shopping and
-personal utility trips plus recreational and cycle touring trips), in
-addition to using official data on trips to work and education
-locations; provision of estimates of `quietness' (a proxy for
-cyclability) at the route segment level; and its provision of three
-network types: `Fastest', `Balanced' and `Quietest' to help plan both
-arterial and residential cycle networks. Workshops and unstructure
-interview is stakeholders show that the tool is already being used in
-practice, to inform urban, inter-urban and rural network development.
-The approach and methods for strategic cycle network development tools
-outlined in this paper can be applied and built-on in other places to
-support the generation of evidence needed to support ambitious cycling
+interface. Commissioned and used by Transport Infrastructure Ireland,
+CRUSE provides estimates of cycling currently and under `snapshot'
+scenarios of `cycling futures' to inform investment decisions. The
+results are publicly available at \url{https://cruse.bike/}, enabling
+planners, engineers, and other stakeholders to make more evidence-based
+decisions. CRUSE goes beyond previous work by: modelling networks at
+high levels of geographic resolution; simulating multiple trip purposes
+(social, shopping and personal utility trips plus recreational and cycle
+touring trips), in addition to using official data on trips to work and
+education locations; and by providing estimates of `quietness' (a proxy
+for cyclability) at the route segment level. Three network types ---
+`Fastest', `Balanced' and `Quietest' --- help plan both arterial and
+residential cycle networks. Workshops and unstructured interviews of
+stakeholders show that the tool is already being used in practice, to
+inform urban, inter-urban and rural network development. The approach is
+flexible and open, allowing the underlying ideas and code to be used in
+other places to generate the evidence needed for effective cycling
 policies and interventions internationally.
@@
 Growing evidence of the negative impacts of car-dependent transport
-systems has led governments in many countries to set targets and taken
+systems has led governments in many countries to set targets and take
 actions. In the context of climate, road safety and physicial inactivity
@@
 that `safety' in this context is usefully quantified as the number of
-people killed and seriously injured (KSI) per distance traveled, the
+people killed or seriously injured (KSI) per distance traveled, the
 directive requires estimation of distance traveled by mode, down to the

### 2024-03-24 paper.qmd
Tighten up abstract

@@
   This paper describes an approach for developing strategic cycle network planning tools.
-Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the methods used to generate the evidence presented in the tool, including disaggregation of origin-destination data, additional trip purposes, routing, scenarios of cycling potential, and the user interface.
-Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of cycling currently and under 'snapshot' scenarios of 'cycling futures' to inform investment decisions.
-The results are publicly available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions.
-CRUSE goes beyond previous work by: modelling networks at high levels of geographic resolution; simulating multiple trip purposes (social, shopping and personal utility trips plus recreational and cycle touring trips), in addition to using official data on trips to work and education locations; and by providing estimates of 'quietness' (a proxy for cyclability) at the route segment level.
-Three network types --- 'Fastest', 'Balanced' and 'Quietest' --- help plan both arterial and residential cycle networks.
-Workshops and unstructured interviews of stakeholders show that the tool is already being used in practice, to inform urban, inter-urban and rural network development.
-The approach is flexible and open, allowing the underlying ideas and code to be used in other places to generate the evidence needed for effective cycling policies and interventions internationally.
+Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenarios generation, and development of an intuitive user interface.
+Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions.
+The publicly available results at https://cruse.bike/ enable planners, engineers, and other stakeholders to make more evidence-based decisions.
+CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring) in addition to using official origin-destination data; and providing estimates of 'quietness' (a proxy for cyclist comfort) at the route segment level.
+Three network types --- 'Fastest', 'Balanced', and 'Quietest' --- help plan both arterial and residential cycle networks.
+Workshops and unstructured interviews with stakeholders show that the tool is already being used in practice to inform urban, inter-urban, and rural network plans.
+The approach is flexible and open-source, allowing the underlying ideas and code to be adapted, supporting effective cycling policies and interventions internationally.
 keywords: 

### 2024-03-24 paper.qmd
Final tweaks to abstract

@@
   This paper describes an approach for developing strategic cycle network planning tools.
-Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenarios generation, and development of an intuitive user interface.
+Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenario generation, and development of an intuitive user interface.
 Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions.
 The publicly available results at https://cruse.bike/ enable planners, engineers, and other stakeholders to make more evidence-based decisions.
-CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring) in addition to using official origin-destination data; and providing estimates of 'quietness' (a proxy for cyclist comfort) at the route segment level.
+CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring) in addition to using official origin-destination data; and providing estimates of 'quietness' (a proxy for cyclist comfort and route preference) at the route segment level.
 Three network types --- 'Fastest', 'Balanced', and 'Quietest' --- help plan both arterial and residential cycle networks.
-Workshops and unstructured interviews with stakeholders show that the tool is already being used in practice to inform urban, inter-urban, and rural network plans.
+Workshops and unstructured interviews with stakeholders show that the tool is already being used in practice to inform urban, inter-urban, and rural network planning.
 The approach is flexible and open-source, allowing the underlying ideas and code to be adapted, supporting effective cycling policies and interventions internationally.

### 2024-03-24 paper.tex
Final tweaks to abstract

@@
 Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we
-outline the methods used to generate the evidence presented in the tool,
-including disaggregation of origin-destination data, additional trip
-purposes, routing, scenarios of cycling potential, and the user
+outline the underlying methods, including disaggregation of
+origin-destination data, incorporation of additional trip purposes,
+routing, scenarios generation, and development of an intuitive user
 interface. Commissioned and used by Transport Infrastructure Ireland,
-CRUSE provides estimates of cycling currently and under `snapshot'
-scenarios of `cycling futures' to inform investment decisions. The
-results are publicly available at \url{https://cruse.bike/}, enabling
-planners, engineers, and other stakeholders to make more evidence-based
-decisions. CRUSE goes beyond previous work by: modelling networks at
-high levels of geographic resolution; simulating multiple trip purposes
-(social, shopping and personal utility trips plus recreational and cycle
-touring trips), in addition to using official data on trips to work and
-education locations; and by providing estimates of `quietness' (a proxy
-for cyclability) at the route segment level. Three network types ---
-`Fastest', `Balanced' and `Quietest' --- help plan both arterial and
-residential cycle networks. Workshops and unstructured interviews of
-stakeholders show that the tool is already being used in practice, to
-inform urban, inter-urban and rural network development. The approach is
-flexible and open, allowing the underlying ideas and code to be used in
-other places to generate the evidence needed for effective cycling
-policies and interventions internationally.
+CRUSE provides estimates of current and potential future cycling levels
+under `snapshot' scenarios to inform investment decisions. The publicly
+available results at https://cruse.bike/ enable planners, engineers, and
+other stakeholders to make more evidence-based decisions. CRUSE goes
+beyond previous work by: modeling networks at high spatial resolution;
+simulating multiple trip purposes (social, shopping, personal utility,
+recreational, and cycle touring) in addition to using official
+origin-destination data; and providing estimates of `quietness' (a proxy
+for cyclist comfort) at the route segment level. Three network types ---
+`Fastest', `Balanced', and `Quietest' --- help plan both arterial and
+residential cycle networks. Workshops and unstructured interviews with
+stakeholders show that the tool is already being used in practice to
+inform urban, inter-urban, and rural network plans. The approach is
+flexible and open-source, allowing the underlying ideas and code to be
+adapted, supporting effective cycling policies and interventions
+internationally.
 \end{abstract}

### 2024-03-24 paper.qmd
Update cyclestreets sentence

@@
 Cars are a major and growing source of greenhouse gas emissions causing climate change [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018] , and a cause of disease due to physical inactivity and air, noise and (from tyres) microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024].
-Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport.
-Nearly half of these emissions (around 10% of global emissions) are from passenger cars [@jaramillo2022].
+Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport, with passenger cars accounting for nearly half of transport emissions (around 10% of global emissions) [@jaramillo2022].
 The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
@@
 
-CycleStreets' services can be used by anyone for free, but we commissioned a bespoke routing service for CRUSE, to enable:
+While CycleStreets offers a free routing API, we commissioned a custom routing service to enable:
 

### 2024-03-24 paper.tex
Update cyclestreets sentence

@@
 origin-destination data, incorporation of additional trip purposes,
-routing, scenarios generation, and development of an intuitive user
+routing, scenario generation, and development of an intuitive user
 interface. Commissioned and used by Transport Infrastructure Ireland,
@@
 origin-destination data; and providing estimates of `quietness' (a proxy
-for cyclist comfort) at the route segment level. Three network types ---
-`Fastest', `Balanced', and `Quietest' --- help plan both arterial and
-residential cycle networks. Workshops and unstructured interviews with
-stakeholders show that the tool is already being used in practice to
-inform urban, inter-urban, and rural network plans. The approach is
-flexible and open-source, allowing the underlying ideas and code to be
-adapted, supporting effective cycling policies and interventions
-internationally.
+for cyclist comfort and route preference) at the route segment level.
+Three network types --- `Fastest', `Balanced', and `Quietest' --- help
+plan both arterial and residential cycle networks. Workshops and
+unstructured interviews with stakeholders show that the tool is already
+being used in practice to inform urban, inter-urban, and rural network
+planning. The approach is flexible and open-source, allowing the
+underlying ideas and code to be adapted, supporting effective cycling
+policies and interventions internationally.
 \end{abstract}

### 2024-03-24 paper.qmd
Tweak sentence

@@
 
-The Climate Action Plan scenario is loosely based on the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which contains policies for action to achieve a 51% reduction in overall GHG emissions by 2030, on a path towards net-zero by 2050.
+The Climate Action Plan scenario models the transport emissions reductions targeted in the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which aims for a 51% cut in overall greenhouse gas emissions by 2030 as part of the pathway to net-zero emissions by 2050.
 For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.

### 2024-03-25 paper.qmd
Typo fix
More on Parkin approach

@@
   - cruse.bib
+  - references.bib
 csl: apa.csl
@@
 
-This describes describe the design, features and potential use of an approach to strategic cycle network planning that is open, scalable, and evidence-based.
+This paper describes an approach to strategic cycle network planning tool development that is open, scalable, and evidence-based.
 We label the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and present a case study of its application in Ireland, the results of which are publicly available at [cruse.bike](https://cruse.bike/) in the CRUSE Tool for Ireland, the landing page of which is illustrated in @fig-landing.
@@
 Developed by researchers and developers in academic, public and private sectors, they have evolved from simple area-based static models tailored to specific regions to more complex and (in some cases) more generalizable tools.
+These tools can be considered as a specific application of the concept of planning support systems (PSS) , which are designed to help planners and decision-makers to make more informed decisions [@geertman2009].
 Outputs include maps with results available at several levels of analysis with levels of availability ranging from being only available to researchers and in static maps to open access mapping systems available to the public.
+
+Previous studies have applied the concept of PSS to specific modes, including walking [@bencekri2024], public transport [@barmentlo2012] and cycling [@bencekri2023].
+While such prior work provides insight into the potential for approaches to transport planning that are both data-driven and participatory, the focus of this paper is on tools that focus attention on the *geographic distribution* of cycling potential and which answer the question "where to build".
+Not all papers reviewed in this section classify themselves as, or even mention, PSS, but all of them can be considered as PSS.
+This paper is informed by the thinking underlying PSS, including the idea that the resulting tools are more effective if they are modular and can inter-link with other tools and processes in the planning process: PSS can be conceived as a "toolbox with separate instruments that can work and talk to each other effectively (as in the open systems concept) and that will be selectively applied in varying configurations to support a particular phase in the planning process" [@geertman2002].
+We share this conception of PSS, emphasizing the benefits of a focus on a particular mode (cycling) and a particular phase in the planning process (strategic network development).
+
 Some of the most relevant existing tools are summarised below and presented in @tbl-tools.
@@
 
-Area-based tools were developed first.
-Parkin and colleagues presented a regression model to estimate the percentage of cycling trips to work in England and Wales based on socioeconomic, transportation, and physical factors at the small area (wards) level in 2007 [@parkin_estimation_2007].
+Area-based approaches were developed before network-based tools, such as a 2007 regression-based model to estimate the percentage of cycling trips to work in England and Wales based on socioeconomic, transportation, and physical factors at the small area (wards) level in 2007 [@parkin_estimation_2007].
 They found that areas with a higher percentage of females, non-whites, car ownership, lower socioeconomic classes, income, distance to work, population density, poor highway conditions, hills, rainfall, and fewer off-road bicycle routes tended to have a lower proportion of cycling to work.
+The approach also generated "forecasts for potential levels of bicycle use" with with increases in off-road routes, improvements of highway and, conversely, increases in car ownership and distance to work leading to decreases in cycling to work.
 
@@
 
+CRUSE is a "great tool for getting started with project outline documents".
+It's an "ideal tool for generating a strategic rationale for estimating demand and building the business case in the absence of cycle count data" (Declan Keenan, Senior Transport Planner working on active travel, personal communication).
+"It's a good representation of what's going on"
+
+"I like to know that the data on the county maps is from POWSCAR data which is gold standard".
+
 ## Comparison with other tools and contribution to the field

### 2024-03-25 paper.tex
More on Parkin approach

@@
 Transport is responsible for 23\% of global greenhouse gas (GHG)
-emissions, 70\% of which are from road transport. Nearly half of these
-emissions (around 10\% of global emissions) are from passenger cars
-\citep{jaramillo2022}. The transport system encourages, enables and in
-some cases enforces unsustainable lifestyles. Services that are only
-accessible by car lock-in car dependency
+emissions, 70\% of which are from road transport, with passenger cars
+accounting for nearly half of transport emissions (around 10\% of global
+emissions) \citep{jaramillo2022}. The transport system encourages,
+enables and in some cases enforces unsustainable lifestyles. Services
+that are only accessible by car lock-in car dependency
 \citep{gray2001, shergold2012, motte-baumvol2010}.
@@
 
-This describes describe the design, features and potential use of an
-approach to strategic cycle network planning that is open, scalable, and
-evidence-based. We label the approach Cycle Route Uptake and Scenario
-Estimation (CRUSE) and present a case study of its application in
-Ireland, the results of which are publicly available at
-\href{https://cruse.bike/}{cruse.bike} in the CRUSE Tool for Ireland,
-the landing page of which is illustrated in Figure~\ref{fig-landing}. In
-Section~\ref{sec-methods}, we outline the methods used to generate the
-evidence presented in the CRUSE Tool. In Section~\ref{sec-results}, we
-present the results of the CRUSE Tool, including estimates of current
-cycling levels and future cycling potential at the national, regional
-and local levels. In Section~\ref{sec-discussion}, we discuss
-limitations and possible future improvements to the approach, and the
-implications of the results for Ireland before making concluding remarks
-in Section~\ref{sec-conclusions}.
+This paper describes an approach to strategic cycle network planning
+tool development that is open, scalable, and evidence-based. We label
+the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and
+present a case study of its application in Ireland, the results of which
+are publicly available at \href{https://cruse.bike/}{cruse.bike} in the
+CRUSE Tool for Ireland, the landing page of which is illustrated in
+Figure~\ref{fig-landing}. In Section~\ref{sec-methods}, we outline the
+methods used to generate the evidence presented in the CRUSE Tool. In
+Section~\ref{sec-results}, we present the results of the CRUSE Tool,
+including estimates of current cycling levels and future cycling
+potential at the national, regional and local levels. In
+Section~\ref{sec-discussion}, we discuss limitations and possible future
+improvements to the approach, and the implications of the results for
+Ireland before making concluding remarks in
+Section~\ref{sec-conclusions}.
 
@@
 from simple area-based static models tailored to specific regions to
-more complex and (in some cases) more generalizable tools. Outputs
-include maps with results available at several levels of analysis with
-levels of availability ranging from being only available to researchers
-and in static maps to open access mapping systems available to the
-public. Some of the most relevant existing tools are summarised below
-and presented in Table~\ref{tbl-tools}.
+more complex and (in some cases) more generalizable tools. These tools
+can be considered as a specific application of the concept of planning
+support systems (PSS) , which are designed to help planners and
+decision-makers to make more informed decisions \citep{geertman2009}.
+Outputs include maps with results available at several levels of
+analysis with levels of availability ranging from being only available
+to researchers and in static maps to open access mapping systems
+available to the public.
+
+Previous studies have applied the concept of PSS to specific modes,
+including walking \citep{bencekri2024}, public transport
+\citep{barmentlo2012} and cycling \citep{bencekri2023}. While such prior
+work provides insight into the potential for approaches to transport
+planning that are both data-driven and participatory, the focus of this
+paper is on tools that focus attention on the \emph{geographic
+distribution} of cycling potential and which answer the question ``where
+to build''. Not all papers reviewed in this section classify themselves
+as, or even mention, PSS, but all of them can be considered as PSS. This
+paper is informed by the thinking underlying PSS, including the idea
+that the resulting tools are more effective if they are modular and can
+inter-link with other tools and processes in the planning process: PSS
+can be conceived as a ``toolbox with separate instruments that can work
+and talk to each other effectively (as in the open systems concept) and
+that will be selectively applied in varying configurations to support a
+particular phase in the planning process'' \citep{geertman2002}. We
+share this conception of PSS, emphasizing the benefits of a focus on a
+particular mode (cycling) and a particular phase in the planning process
+(strategic network development).
+
+Some of the most relevant existing tools are summarised below and
+presented in Table~\ref{tbl-tools}.
 
@@
 
-CycleStreets' services can be used by anyone for free, but we
-commissioned a bespoke routing service for CRUSE, to enable:
+While CycleStreets offers a free routing API, we commissioned a custom
+routing service to enable:
 
@@
 
-The Climate Action Plan scenario is loosely based on the Irish
-Government's
+The Climate Action Plan scenario models the transport emissions
+reductions targeted in the Irish Government's
 \href{https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/}{Climate
-Action Plan 2021}, which contains policies for action to achieve a 51\%
-reduction in overall GHG emissions by 2030, on a path towards net-zero
-by 2050. For transport, this includes 500,000 extra walking, cycling and
-public transport trips per day by 2030. In terms of car travel, the
-target is to ``Increas{[}e{]} the proportion of kilometers driven by
-passenger electric cars to between 40 and 45\% by 2030, in addition to a
-reduction of 10\% in kilometers driven by the remaining internal
-combustion engine cars.'' This equates to a 5.5 to 6\% reduction in
-total car km driven.
+Action Plan 2021}, which aims for a 51\% cut in overall greenhouse gas
+emissions by 2030 as part of the pathway to net-zero emissions by 2050.
+For transport, this includes 500,000 extra walking, cycling and public
+transport trips per day by 2030. In terms of car travel, the target is
+to ``Increas{[}e{]} the proportion of kilometers driven by passenger
+electric cars to between 40 and 45\% by 2030, in addition to a reduction
+of 10\% in kilometers driven by the remaining internal combustion engine
+cars.'' This equates to a 5.5 to 6\% reduction in total car km driven.
 
@@
 system.
+
+CRUSE is a ``great tool for getting started with project outline
+documents''. It's an ``ideal tool for generating a strategic rationale
+for estimating demand and building the business case in the absence of
+cycle count data'' (Declan Keenan, Senior Transport Planner working on
+active travel, personal communication). ``It's a good representation of
+what's going on''
+
+``I like to know that the data on the county maps is from POWSCAR data
+which is gold standard''.
 
@@
 
-  \bibliography{bibliography.bib,cruse.bib}
+  \bibliography{bibliography.bib,cruse.bib,references.bib}
 

### 2024-03-25 paper.qmd
Add new qualitative section to results

@@
 The main result of the work presented in this paper is an open access web application and datasets on current and potential future cycling levels in Ireland, with evidence provided at county and route segment levels.
-Like the PCT on which CRUSE builds, a key feature of the results is that they are systematic and available for all counties, and most routes on which cycling is permitted, in Ireland.
-The full set of results are therefore too extensive to present in their entirety in this paper.
+Like other national tools presented in @tbl-tools, a key feature of the results is that it provides a consistent and objective baseline for comparing cycling potentials in different areas.
+
+The approach generates estimates of cycling potential on every road in the country and are therefore too extensive to present in their entirety in this paper.
 Through the interactive web application, users can explore the data to generate the results that are most relevant to their needs, whether that is finding the cycling potential on a particular road or finding 'weak links' or barriers in the cycle network associated with a particular school, work place or other destination.
 
-Instead of trying to show such use cases, of which there are many hundreds, we present a selection of results to illustrate the main features of the CRUSE Tool.
+Instead of trying to show such use cases, of which there are many hundreds, we present a selection of results to illustrate the main features of the resulting evidence, in @sec-quantiative.
+We also present some qualitative results based on workshops and interviews with stakeholders in Ireland, in @sec-qualitative.
 
@@
 <!-- ## Local results -->
+
+## Selected quantitative results {#sec-quantiative}
 

### 2024-03-25 paper.qmd
edits

--- 2024-03-24 paper.qmd
@@
 For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
-Better data to inform road safety policies and interventions is a motivation for better estimates of 'baseline' levels of physical activity at high levels of geographic resolution [@tait2023].
+Better data to inform road safety policies and interventions is a motivation for better estimates of 'baseline' levels of physical activity at high geographic resolutions [@tait2023].
 
@@
 In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030 [@national2021].
-Doing so while simultaneously enabling rapid uptake of active modes will require key travel corridors to be identified 'cycle proofed'.
-Cycling in Ireland represents only 3% of total modal share as of the 2016 Census, but accounts for 20% of serious injuries and 7% of all fatalities, driving the development of cycling policy and infrastructure plans in Ireland.
+Doing so while simultaneously enabling rapid uptake of active modes will require key travel corridors to be identified and 'cycle proofed'.
+Cycling in Ireland represents only 3% of total modal share as of the 2016 Census, but accounts for 20% of serious injuries and 7% of all fatalities.
+Poor perception of safety has been found to represent the most important barrier to increased cycling in Ireland[@brick2018], and the need to improve road safety drives the development of national cycling policy and infrastructure plans.
 The main frameworks underpinning these efforts are the Climate Action Plan [@climate2022a], the National Development Plan, and the National Roads 2040 strategy [@national2023].
-<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ --> These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016 "83% of respondents supported enhanced walking and cycling facilities" [@sustaina2019].
+<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ -->
 
-At the regional level within Ireland, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings of high support for and capability for cycling: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
+At the regional level within Ireland, the recently published Greater Dublin Area Transport Strategy [@greater] reveals the high support for and capability for cycling: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
 Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
@@
 > This is followed by infrastructure type, the number of junctions along the route, traffic speed and cyclist volumes.
-> In terms if infrastructure, regardless of the level of cycling confidence, routes which have 'no facilities' or 'bus/cycle lanes' are the least favoured cycle route types.
+> In terms of infrastructure, regardless of the level of cycling confidence, routes which have 'no facilities' or 'bus/cycle lanes' are the least favoured cycle route types.
 
@@
 -   The results are fully reproducible (code to be released pending sign-off by TII's IT team), preventing 'cloud lock in' to a potentially monopolistic consultancy, and encouraging input from the wider open source community [@lovelace2021; @dhir2017].
--   By covering a wide range of trip purposes --- not just travel to work [@lovelace2017; @heinen2010] and travel school [@goodman2019] as covered in previously published research on open access national cycle network planning tools --- the results capture a higher proportion of cycling potential including key rural trips which are often under-represented in models of active travel.
+-   By covering a wide range of trip purposes --- not just travel to work [@lovelace2017; @heinen2010] and travel to school [@goodman2019] as covered in previously published research on open access national cycle network planning tools --- the results capture a higher proportion of cycling potential including key rural trips which are often under-represented in models of active travel.
 

### 2024-03-25 paper.tex
edits

@@
 \usepackage{amsmath,amssymb}
+\usepackage{lmodern}
 \usepackage{setspace}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
 \fi
@@
 \makeatletter
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 \makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
 \makeatother
 \makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
 \makeatother
@@
 \bibliographystyle{elsarticle-num}
-\usepackage{bookmark}
-
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \end{frontmatter}
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\section{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 
-\centering{
+{\centering
 
@@
 
-\end{figure}%
+\end{figure}
 
@@
 
+\hypertarget{tools-for-estimating-cycling-potential}{%
 \section{Tools for estimating cycling
-potential}\label{tools-for-estimating-cycling-potential}
+potential}\label{tools-for-estimating-cycling-potential}}
 
@@
 from simple area-based static models tailored to specific regions to
-more complex and (in some cases) more generalizable tools. These tools
-can be considered as a specific application of the concept of planning
-support systems (PSS) , which are designed to help planners and
-decision-makers to make more informed decisions \citep{geertman2009}.
-Outputs include maps with results available at several levels of
-analysis with levels of availability ranging from being only available
-to researchers and in static maps to open access mapping systems
-available to the public.
-
-Previous studies have applied the concept of PSS to specific modes,
-including walking \citep{bencekri2024}, public transport
-\citep{barmentlo2012} and cycling \citep{bencekri2023}. While such prior
-work provides insight into the potential for approaches to transport
-planning that are both data-driven and participatory, the focus of this
-paper is on tools that focus attention on the \emph{geographic
-distribution} of cycling potential and which answer the question ``where
-to build''. Not all papers reviewed in this section classify themselves
-as, or even mention, PSS, but all of them can be considered as PSS. This
-paper is informed by the thinking underlying PSS, including the idea
-that the resulting tools are more effective if they are modular and can
-inter-link with other tools and processes in the planning process: PSS
-can be conceived as a ``toolbox with separate instruments that can work
-and talk to each other effectively (as in the open systems concept) and
-that will be selectively applied in varying configurations to support a
-particular phase in the planning process'' \citep{geertman2002}. We
-share this conception of PSS, emphasizing the benefits of a focus on a
-particular mode (cycling) and a particular phase in the planning process
-(strategic network development).
-
-Some of the most relevant existing tools are summarised below and
-presented in Table~\ref{tbl-tools}.
-
-\begin{longtable}[t]{>{\raggedright\arraybackslash}p{13em}>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{12em}>{\raggedright\arraybackslash}p{5em}}
-
-\caption{\label{tbl-tools}Summary of existing tools for estimating
-cycling potential.}
-
-\tabularnewline
-
-\\
+more complex and (in some cases) more generalizable tools. Outputs
+include maps with results available at several levels of analysis with
+levels of availability ranging from being only available to researchers
+and in static maps to open access mapping systems available to the
+public. Some of the most relevant existing tools are summarised below
+and presented in Table~\ref{tbl-tools}.
+
+\begin{table}
+
+\caption{\label{tbl-tools}\textbf{?(caption)}}\begin{minipage}[t]{\linewidth}
+\subcaption{\label{tbl-tools-1}Summary of existing tools for estimating cycling potential.}
+
+{\centering
+
+\centering
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{13em}>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{12em}>{\raggedright\arraybackslash}p{5em}}
 \toprule
@@
 \midrule
-\cellcolor{gray!10}{Bicycle share model, Parkin et al.
-(2007)} & \cellcolor{gray!10}{England and Wales, Journeys to work OD census data, at the small-area (wards) level} & \cellcolor{gray!10}{Static tables in academic paper} & \cellcolor{gray!10}{Proprietary}\\
+\cellcolor{gray!6}{Bicycle share model, Parkin et al.
+(2007)} & \cellcolor{gray!6}{England and Wales, Journeys to work OD census data, at the small-area (wards) level} & \cellcolor{gray!6}{Static tables in academic paper} & \cellcolor{gray!6}{Proprietary}\\
 Analysis of  Cycling Potential (ACP), Transport for London (2010, 2016) & London, London Travel Demand Survey & Static tables, graphs, and heatmaps, not publicly available & Proprietary\\
-\cellcolor{gray!10}{Prioritization index, Larsen et al.
-(2013)} & \cellcolor{gray!10}{Montreal, Survey, Road safety, and OD data} & \cellcolor{gray!10}{Map-based heatmap, not publicly available} & \cellcolor{gray!10}{Proprietary}\\
+\cellcolor{gray!6}{Prioritization index, Larsen et al.
+(2013)} & \cellcolor{gray!6}{Montreal, Survey, Road safety, and OD data} & \cellcolor{gray!6}{Map-based heatmap, not publicly available} & \cellcolor{gray!6}{Proprietary}\\
 Usage intensity index, Zhang et al. (2014) & Belo Horizonte, Survey, census, and OD data & Static tables and maps, not publicly available & Proprietary\\
-\cellcolor{gray!10}{The Cycling Potential Tool (CPT), Phillips and Range (2017)} & \cellcolor{gray!10}{Scotland, Environmental and socioeconomic data, at the small area (output areas) level} & \cellcolor{gray!10}{Maps showing cycling potential in each area} & \cellcolor{gray!10}{Proprietary}\\
+\cellcolor{gray!6}{The Cycling Potential Tool (CPT), Phillips and Range (2017)} & \cellcolor{gray!6}{Scotland, Environmental and socioeconomic data, at the small area (output areas) level} & \cellcolor{gray!6}{Maps showing cycling potential in each area} & \cellcolor{gray!6}{Proprietary}\\
 \addlinespace
@@
 (2017) & England and Wales, Journeys to work and school OD census data, at the small area (LSOA) level & Online maps, graphs, tables, publicly available at www.pct.bike & Open source\\
-\cellcolor{gray!10}{The    Gross Potential for Cycling tool (CPC), Silva et al.
-(2021 and 2022)} & \cellcolor{gray!10}{21 Portuguese cities, Land use and socio-demographic data, at the small area (census tract) level} & \cellcolor{gray!10}{Static maps showing cycling potential in different areas} & \cellcolor{gray!10}{NA}\\
+\cellcolor{gray!6}{The    Gross Potential for Cycling tool (CPC), Silva et al.
+(2021 and 2022)} & \cellcolor{gray!6}{21 Portuguese cities, Land use and socio-demographic data, at the small area (census tract) level} & \cellcolor{gray!6}{Static maps showing cycling potential in different areas} & \cellcolor{gray!6}{NA}\\
 \bottomrule
-
-\end{longtable}
+\end{tabular}
+
+}
+
+\end{minipage}%
+
+\end{table}
 
@@
 
-\section{Methods and data}\label{sec-methods}
-
+\hypertarget{sec-methods}{%
+\section{Methods and data}\label{sec-methods}}
+
+\hypertarget{sec-disaggregation}{%
 \subsection{Disaggregation of origin-destination
-data}\label{sec-disaggregation}
+data}\label{sec-disaggregation}}
 
@@
 
-\end{figure}%
-
-\subsection{Additional trip purposes}\label{sec-trip-purposes}
+\end{figure}
+
+\hypertarget{sec-trip-purposes}{%
+\subsection{Additional trip purposes}\label{sec-trip-purposes}}
 
@@
 
-\end{figure}%
-
-\subsection{Routing}\label{routing}
+\end{figure}
+
+\hypertarget{routing}{%
+\subsection{Routing}\label{routing}}
 
@@
 
-\begin{minipage}{0.50\linewidth}
-\includegraphics[width=2.60417in,height=\textheight]{images/naas_quietest_godutch.png}\end{minipage}%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics[width=2.60417in,height=\textheight]{images/naas_quietest_godutch.png}
+
+}
+
+}
+
+\end{minipage}%
 %
-\begin{minipage}{0.50\linewidth}
-\includegraphics[width=2.60417in,height=\textheight]{images/naas_fastest_godutch.png}\end{minipage}%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics[width=2.60417in,height=\textheight]{images/naas_fastest_godutch.png}
+
+}
+
+}
+
+\end{minipage}%
 \newline
-\begin{minipage}{0.50\linewidth}
-\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_quietest_godutch.png}\end{minipage}%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_quietest_godutch.png}
+
+}
+
+}
+
+\end{minipage}%
 %
-\begin{minipage}{0.50\linewidth}
-\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_fastest_godutch.png}\end{minipage}%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_fastest_godutch.png}
+
+}
+
+}
+
+\end{minipage}%
 
@@
 
-\end{figure}%
-
-\subsection{Scenarios of cycling potential}\label{sec-scenarios}
+\end{figure}
+
+\hypertarget{sec-scenarios}{%
+\subsection{Scenarios of cycling potential}\label{sec-scenarios}}
 
@@
 
-\end{figure}%
-
-\subsubsection{Baseline}\label{baseline}
+\end{figure}
+
+\hypertarget{baseline}{%
+\subsubsection{Baseline}\label{baseline}}
 
@@
 
-\subsubsection{Near Market}\label{near-market}
+\hypertarget{near-market}{%
+\subsubsection{Near Market}\label{near-market}}
 
@@
 
-\subsubsection{Climate Action Plan}\label{climate-action-plan}
+\hypertarget{climate-action-plan}{%
+\subsubsection{Climate Action Plan}\label{climate-action-plan}}
 
@@
 
-\subsubsection{Go Dutch}\label{go-dutch}
+\hypertarget{go-dutch}{%
+\subsubsection{Go Dutch}\label{go-dutch}}
 
@@
 
-\subsubsection{Ebike}\label{ebike}
+\hypertarget{ebike}{%
+\subsubsection{Ebike}\label{ebike}}
 
@@
 
-\subsection{User interface}\label{sec-ui}
+\hypertarget{sec-ui}{%
+\subsection{User interface}\label{sec-ui}}
 
@@
 
-\end{figure}%
-
-\section{Results}\label{sec-results}
+\end{figure}
+
+\hypertarget{sec-results}{%
+\section{Results}\label{sec-results}}
 
@@
 
-\section{Discussion}\label{sec-discussion}
+\hypertarget{sec-discussion}{%
+\section{Discussion}\label{sec-discussion}}
 
@@
 
+\hypertarget{uses-of-the-tool-in-practice}{%
 \subsection{Uses of the tool in
-practice}\label{uses-of-the-tool-in-practice}
+practice}\label{uses-of-the-tool-in-practice}}
 
@@
 
-CRUSE is a ``great tool for getting started with project outline
-documents''. It's an ``ideal tool for generating a strategic rationale
-for estimating demand and building the business case in the absence of
-cycle count data'' (Declan Keenan, Senior Transport Planner working on
-active travel, personal communication). ``It's a good representation of
-what's going on''
-
-``I like to know that the data on the county maps is from POWSCAR data
-which is gold standard''.
-
+\hypertarget{comparison-with-other-tools-and-contribution-to-the-field}{%
 \subsection{Comparison with other tools and contribution to the
-field}\label{comparison-with-other-tools-and-contribution-to-the-field}
+field}\label{comparison-with-other-tools-and-contribution-to-the-field}}
 
@@
 
-\section{Conclusions}\label{sec-conclusions}
+\hypertarget{sec-conclusions}{%
+\section{Conclusions}\label{sec-conclusions}}
 
@@
 
-\section{List of abbreviations}\label{list-of-abbreviations}
+\hypertarget{list-of-abbreviations}{%
+\section{List of abbreviations}\label{list-of-abbreviations}}
 
@@
 
-\section{Declarations}\label{declarations}
-
+\hypertarget{declarations}{%
+\section{Declarations}\label{declarations}}
+
+\hypertarget{availability-of-data-and-material}{%
 \subsection*{Availability of data and
-material}\label{availability-of-data-and-material}
+material}\label{availability-of-data-and-material}}
 \addcontentsline{toc}{subsection}{Availability of data and material}
@@
 
-\subsection*{Funding}\label{funding}
+\hypertarget{funding}{%
+\subsection*{Funding}\label{funding}}
 \addcontentsline{toc}{subsection}{Funding}
@@
 
-\subsection*{Acknowledgements}\label{acknowledgements}
+\hypertarget{acknowledgements}{%
+\subsection*{Acknowledgements}\label{acknowledgements}}
 \addcontentsline{toc}{subsection}{Acknowledgements}
@@
 
-  \bibliography{bibliography.bib,cruse.bib,references.bib}
+  \bibliography{bibliography.bib,cruse.bib}
 

### 2024-03-25 paper.qmd
Add content from Declan

--- 2024-03-25 paper.qmd
@@
 <!-- The results for the other -->
+## Selected qualitative results {#sec-qualitative}
+
+Thoughout the project, we undertook a series of workshops and interviews with stakeholders in Ireland to gather feedback on the approach and resulting tool.
+The feedback was generally positive, with stakeholders commenting on the need for the tool, lack of data on cycling potential proving a barrier to decision-making and investment, and the potential for the tool.
+At an end of project workshop in December 2023, stakeholders were asked to provide feedback on the tool and its potential use in practice.
+Quotes and suggestions from these workshops, and a subsequent interview with a Senior Transport Planning using the evidence in their everyday work in Transport Infrastructure Ireland, and presented in this section (with quotes being attributed to named individuals where permission has been granted).
+
+Early in the project, we presented the approach to practitioners in Kildare and Limmerick.
+The feedback from the workshop in Kildare highlighted the need for the tool for practitioners at the county level, as highlighted in the following quote:
+
+> "It's a missing piece of evidence that will help new projects get built" (Senior Engineer, Sustainable Transport & Traffic Management, Kildare County Council)
+
+Practitioners in Kildare County Council that the evidence generated *was already valuable* the development of new cycle routes, especially when making the case during planning applications.
+They emphasized that cycle networks are often fragmented, and the tool allows them to demonstrate how individual local schemes can contribute to a larger, interconnected network.
+Council officers expressed a preference for visualizations that directly compare the fast and quiet routes, as they recognize the importance of both types of routes (users can switch between the on the 'Route types' page for each county although this functionality is not available in the landing page map due to other feedback requesting a simpler interface).
+
+In comparison with another tool (the National Transport Authority's cycling propensity tool), the CRUSE tool was seen by practitioners in Kildare as more user-friendly and intuitive, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
+Overall, the workshop participants concluded that the approach filled gap in evidence and will greatly assist in the development of new cycling projects.
+
+During the workshop with practitioners in Limerick, the estimates of Baseline and Go Dutch potential at the network levels were found to match well with local knowledge.
+The tool effectively highlighted the same sections on the transport network that experienced planners identified as priorities for investment in cycling, such as the South Circular Road.
+Limerick County Council expressed their interest in utilizing each network layer, particularly the quiet network, to support new cyclists who lack confidence to share space with motor traffic.
+Specific suggestions from the workshop included the suggestion to align scenarios with national strategy (resulting in the Climate Action Plan scenario) and to provide data downloads provided as 'Shapefiles' for further analysis (now implemented although the download format is GeoPackage, an open standard for spatial data).
+
+In a subsequent interview with a practitioner using the evidence generated by the CRUSE approach in their everyday work, we received the following feedback.
+The outputs were a "great tool for getting started with project outline documents" and "ideal for generating a strategic rationale for estimating demand and building the business case in the absence of cycle count data" (Declan Keenan, Senior Transport Planner working on active travel, personal communication).
+Declan also noted the alignment between local knowledge, cycle count data, and the outputs of the approach: "It's a good representation of what's going on" and the networks allign well in terms of network "shape" and the proportion of trips of trips on different parts of the network.
+There were two suggestions from Declan which have not yet been implemented: the provision of data showing only "gold standard" POWSCAR data, and the need for training on the tool to ensure that all staff can use it effectively.
+These, and other potential improvements, are discussed in the next section.
 
@@
 It will also support reporting for the RISM Directive on road casualties as part of the road safety management system.
-
-CRUSE is a "great tool for getting started with project outline documents".
-It's an "ideal tool for generating a strategic rationale for estimating demand and building the business case in the absence of cycle count data" (Declan Keenan, Senior Transport Planner working on active travel, personal communication).
-"It's a good representation of what's going on"
-
-"I like to know that the data on the county maps is from POWSCAR data which is gold standard".
 

### 2024-03-25 paper.tex
Add content from Declan

--- 2024-03-25 paper.tex
@@
 
-Area-based tools were developed first. Parkin and colleagues presented a
-regression model to estimate the percentage of cycling trips to work in
-England and Wales based on socioeconomic, transportation, and physical
-factors at the small area (wards) level in 2007
-\citep{parkin_estimation_2007}. They found that areas with a higher
+Area-based approaches were developed before network-based tools, such as
+a 2007 regression-based model to estimate the percentage of cycling
+trips to work in England and Wales based on socioeconomic,
+transportation, and physical factors at the small area (wards) level in
+2007 \citep{parkin_estimation_2007}. They found that areas with a higher
 percentage of females, non-whites, car ownership, lower socioeconomic
@@
 conditions, hills, rainfall, and fewer off-road bicycle routes tended to
-have a lower proportion of cycling to work.
+have a lower proportion of cycling to work. The approach also generated
+``forecasts for potential levels of bicycle use'' with with increases in
+off-road routes, improvements of highway and, conversely, increases in
+car ownership and distance to work leading to decreases in cycling to
+work.
 
@@
 levels in Ireland, with evidence provided at county and route segment
-levels. Like the PCT on which CRUSE builds, a key feature of the results
-is that they are systematic and available for all counties, and most
-routes on which cycling is permitted, in Ireland. The full set of
-results are therefore too extensive to present in their entirety in this
-paper. Through the interactive web application, users can explore the
-data to generate the results that are most relevant to their needs,
-whether that is finding the cycling potential on a particular road or
-finding `weak links' or barriers in the cycle network associated with a
-particular school, work place or other destination.
+levels. Like other national tools presented in Table~\ref{tbl-tools}, a
+key feature of the results is that it provides a consistent and
+objective baseline for comparing cycling potentials in different areas.
+
+The approach generates estimates of cycling potential on every road in
+the country and are therefore too extensive to present in their entirety
+in this paper. Through the interactive web application, users can
+explore the data to generate the results that are most relevant to their
+needs, whether that is finding the cycling potential on a particular
+road or finding `weak links' or barriers in the cycle network associated
+with a particular school, work place or other destination.
 
@@
 hundreds, we present a selection of results to illustrate the main
-features of the CRUSE Tool.
+features of the resulting evidence, in Section~\ref{sec-quantiative}. We
+also present some qualitative results based on workshops and interviews
+with stakeholders in Ireland, in Section~\ref{sec-qualitative}.
+
+\subsection{Selected quantitative results}\label{sec-quantiative}
 
@@
 
+\subsection{Selected qualitative results}\label{sec-qualitative}
+
+Thoughout the project, we undertook a series of workshops and interviews
+with stakeholders in Ireland to gather feedback on the approach and
+resulting tool. The feedback was generally positive, with stakeholders
+commenting on the need for the tool, lack of data on cycling potential
+proving a barrier to decision-making and investment, and the potential
+for the tool. At an end of project workshop in December 2023,
+stakeholders were asked to provide feedback on the tool and its
+potential use in practice. Quotes and suggestions from these workshops,
+and a subsequent interview with a Senior Transport Planning using the
+evidence in their everyday work in Transport Infrastructure Ireland, and
+presented in this section (with quotes being attributed to named
+individuals where permission has been granted).
+
+Early in the project, we presented the approach to practitioners in
+Kildare and Limmerick. The feedback from the workshop in Kildare
+highlighted the need for the tool for practitioners at the county level,
+as highlighted in the following quote:
+
+\begin{quote}
+``It's a missing piece of evidence that will help new projects get
+built'' (Senior Engineer, Sustainable Transport \& Traffic Management,
+Kildare County Council)
+\end{quote}
+
+Practitioners in Kildare County Council that the evidence generated
+\emph{was already valuable} the development of new cycle routes,
+especially when making the case during planning applications. They
+emphasized that cycle networks are often fragmented, and the tool allows
+them to demonstrate how individual local schemes can contribute to a
+larger, interconnected network. Council officers expressed a preference
+for visualizations that directly compare the fast and quiet routes, as
+they recognize the importance of both types of routes (users can switch
+between the on the `Route types' page for each county although this
+functionality is not available in the landing page map due to other
+feedback requesting a simpler interface).
+
+In comparison with another tool (the National Transport Authority's
+cycling propensity tool), the CRUSE tool was seen by practitioners in
+Kildare as more user-friendly and intuitive, with the ability to
+generate evidence at the route level, which is crucial for planning new
+cycling infrastructure. Overall, the workshop participants concluded
+that the approach filled gap in evidence and will greatly assist in the
+development of new cycling projects.
+
+During the workshop with practitioners in Limerick, the estimates of
+Baseline and Go Dutch potential at the network levels were found to
+match well with local knowledge. The tool effectively highlighted the
+same sections on the transport network that experienced planners
+identified as priorities for investment in cycling, such as the South
+Circular Road. Limerick County Council expressed their interest in
+utilizing each network layer, particularly the quiet network, to support
+new cyclists who lack confidence to share space with motor traffic.
+Specific suggestions from the workshop included the suggestion to align
+scenarios with national strategy (resulting in the Climate Action Plan
+scenario) and to provide data downloads provided as `Shapefiles' for
+further analysis (now implemented although the download format is
+GeoPackage, an open standard for spatial data).
+
+In a subsequent interview with a practitioner using the evidence
+generated by the CRUSE approach in their everyday work, we received the
+following feedback. The outputs were a ``great tool for getting started
+with project outline documents'' and ``ideal for generating a strategic
+rationale for estimating demand and building the business case in the
+absence of cycle count data'' (Declan Keenan, Senior Transport Planner
+working on active travel, personal communication). Declan also noted the
+alignment between local knowledge, cycle count data, and the outputs of
+the approach: ``It's a good representation of what's going on'' and the
+networks allign well in terms of network ``shape'' and the proportion of
+trips of trips on different parts of the network. There were two
+suggestions from Declan which have not yet been implemented: the
+provision of data showing only ``gold standard'' POWSCAR data, and the
+need for training on the tool to ensure that all staff can use it
+effectively. These, and other potential improvements, are discussed in
+the next section.
+
 \section{Discussion}\label{sec-discussion}
@@
 system.
-
-CRUSE is a ``great tool for getting started with project outline
-documents''. It's an ``ideal tool for generating a strategic rationale
-for estimating demand and building the business case in the absence of
-cycle count data'' (Declan Keenan, Senior Transport Planner working on
-active travel, personal communication). ``It's a good representation of
-what's going on''
-
-``I like to know that the data on the county maps is from POWSCAR data
-which is gold standard''.
 

### 2024-03-25 paper.qmd
docx out

@@
   # html: default
-  # docx: default
+  docx: default
 # # # # Uncomment for formatted version
-  elsevier-pdf:
-    linestretch: 2
-    keep-tex: true
-    journal:
-      name: European Transport Research Review
-      formatting: preprint
-      model: 3p
-      cite-style: super
+#   elsevier-pdf:
+#     linestretch: 2
+#     keep-tex: true
+#     journal:
+#       name: European Transport Research Review
+#       formatting: preprint
+#       model: 3p
+#       cite-style: super
 # # # #   # pdf: default
@@
 #| label: tbl-tools
-knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
-  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
-  kableExtra::column_spec(1, width = "13em") |>
-  kableExtra::column_spec(2, width = "15em") |>
-  kableExtra::column_spec(3, width = "12em") |>
-  kableExtra::column_spec(4, width = "5em")
+# For PDF:
+# knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
+#   kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
+#   kableExtra::column_spec(1, width = "13em") |>
+#   kableExtra::column_spec(2, width = "15em") |>
+#   kableExtra::column_spec(3, width = "12em") |>
+#   kableExtra::column_spec(4, width = "5em")
+# For HTML:
+knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.")
 ```

### 2024-03-25 paper.qmd
Update latex

@@
   # html: default
-  docx: default
-# # # # Uncomment for formatted version
-#   elsevier-pdf:
-#     linestretch: 2
-#     keep-tex: true
-#     journal:
-#       name: European Transport Research Review
-#       formatting: preprint
-#       model: 3p
-#       cite-style: super
+  # docx: default
+# # # Uncomment for formatted version
+  elsevier-pdf:
+    linestretch: 2
+    keep-tex: true
+    journal:
+      name: European Transport Research Review
+      formatting: preprint
+      model: 3p
+      cite-style: super
 # # # #   # pdf: default
@@
 # For PDF:
-# knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
-#   kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
-#   kableExtra::column_spec(1, width = "13em") |>
-#   kableExtra::column_spec(2, width = "15em") |>
-#   kableExtra::column_spec(3, width = "12em") |>
-#   kableExtra::column_spec(4, width = "5em")
+knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
+  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
+  kableExtra::column_spec(1, width = "13em") |>
+  kableExtra::column_spec(2, width = "15em") |>
+  kableExtra::column_spec(3, width = "12em") |>
+  kableExtra::column_spec(4, width = "5em")
 # For HTML:
-knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.")
+# knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.")
 ```

### 2024-03-25 paper.tex
Update latex

@@
 for better estimates of `baseline' levels of physical activity at high
-levels of geographic resolution \citep{tait2023}.
+geographic resolutions \citep{tait2023}.
 
@@
 uptake of active modes will require key travel corridors to be
-identified `cycle proofed'. Cycling in Ireland represents only 3\% of
-total modal share as of the 2016 Census, but accounts for 20\% of
-serious injuries and 7\% of all fatalities, driving the development of
-cycling policy and infrastructure plans in Ireland. The main frameworks
-underpinning these efforts are the Climate Action Plan
-\citep{climate2022a}, the National Development Plan, and the National
-Roads 2040 strategy \citep{national2023}. These policy frameworks are
-backed-up by public support for active travel interventions: while only
-3\% of trips in Ireland are made by cycling according to the latest
-Census data from 2016 ``83\% of respondents supported enhanced walking
-and cycling facilities'' \citep{sustaina2019}.
+identified and `cycle proofed'. Cycling in Ireland represents only 3\%
+of total modal share as of the 2016 Census, but accounts for 20\% of
+serious injuries and 7\% of all fatalities. Poor perception of safety
+has been found to represent the most important barrier to increased
+cycling in Ireland\citep{brick2018}, and the need to improve road safety
+drives the development of national cycling policy and infrastructure
+plans. The main frameworks underpinning these efforts are the Climate
+Action Plan \citep{climate2022a}, the National Development Plan, and the
+National Roads 2040 strategy \citep{national2023}.
 
 At the regional level within Ireland, the recently published Greater
-Dublin Area Transport Strategy \citep{greater} reinforces these findings
-of high support for and capability for cycling: ``nearly a quarter of
-adults cycle at least once a week in the Dublin Metropolitan Area'' with
-cycling in the Dublin area taking up to 60,000 cars off the road today.
-Extrapolating this on a per population basis across Ireland, with around
-40\% of the population living in Dublin, suggests that around 150,000
-cars could be removed nationwide just by achieving Dublin levels of
-cycling in all counties (notwithstanding existing cycling trips and
-differences in trip distances). 7 in 10 trips in Ireland are by car
-\citep{national} . Cycling has the potential to replace a large
-proportion of these trips: ``A high priority must also be given to
-cyclists, because trips by this mode have the potential to replace trips
-by private car, most specifically for short to medium distance trips,
-but increasingly for longer trips as ebikes extend the range of this
-mode'' \citep{greater}.
+Dublin Area Transport Strategy \citep{greater} reveals the high support
+for and capability for cycling: ``nearly a quarter of adults cycle at
+least once a week in the Dublin Metropolitan Area'' with cycling in the
+Dublin area taking up to 60,000 cars off the road today. Extrapolating
+this on a per population basis across Ireland, with around 40\% of the
+population living in Dublin, suggests that around 150,000 cars could be
+removed nationwide just by achieving Dublin levels of cycling in all
+counties (notwithstanding existing cycling trips and differences in trip
+distances). 7 in 10 trips in Ireland are by car \citep{national} .
+Cycling has the potential to replace a large proportion of these trips:
+``A high priority must also be given to cyclists, because trips by this
+mode have the potential to replace trips by private car, most
+specifically for short to medium distance trips, but increasingly for
+longer trips as ebikes extend the range of this mode'' \citep{greater}.
 
@@
 number of junctions along the route, traffic speed and cyclist volumes.
-In terms if infrastructure, regardless of the level of cycling
+In terms of infrastructure, regardless of the level of cycling
 confidence, routes which have `no facilities' or `bus/cycle lanes' are
@@
   By covering a wide range of trip purposes --- not just travel to work
-  \citep{lovelace2017, heinen2010} and travel school \citep{goodman2019}
-  as covered in previously published research on open access national
-  cycle network planning tools --- the results capture a higher
-  proportion of cycling potential including key rural trips which are
-  often under-represented in models of active travel.
+  \citep{lovelace2017, heinen2010} and travel to school
+  \citep{goodman2019} as covered in previously published research on
+  open access national cycle network planning tools --- the results
+  capture a higher proportion of cycling potential including key rural
+  trips which are often under-represented in models of active travel.
 \end{itemize}

### 2024-03-25 paper.qmd
Revision EVT

--- 2024-03-25 paper.qmd
@@
   - cruse.bib
-  - references.bib
 csl: apa.csl
@@
   # html: default
-  docx: default
+  # docx: default
 # # # # Uncomment for formatted version
-#   elsevier-pdf:
-#     linestretch: 2
-#     keep-tex: true
-#     journal:
-#       name: European Transport Research Review
-#       formatting: preprint
-#       model: 3p
-#       cite-style: super
+  elsevier-pdf:
+    linestretch: 2
+    keep-tex: true
+    journal:
+      name: European Transport Research Review
+      formatting: preprint
+      model: 3p
+      cite-style: super
 # # # #   # pdf: default
@@
   - Active Travel
+  - Transport Planning
+  - Collaborative Planning
 date: last-modified
@@
 <!-- Mobility kills. -->
+## Background
 
 Transport systems dominated by heavy and powerful private cars are inefficient, dangerous, and unhealthy.
-Cars are a major and growing source of greenhouse gas emissions causing climate change [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018] , and a cause of disease due to physical inactivity and air, noise and (from tyres) microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024].
-Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport, with passenger cars accounting for nearly half of transport emissions (around 10% of global emissions) [@jaramillo2022].
+Cars are a major and growing source of greenhouse gas (GHG) emissions [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018], and a cause of disease due to physical inactivity and air, noise and microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024].
+<!-- Shouldn't these three sources share the brackets?
+--> Transport is responsible for 23% of GHG emissions, 70% of which are from road transport, with passenger cars accounting for nearly half of transport emissions (around 10% of global emissions) [@jaramillo2022].
 The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
@@
 Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
-Furthermore, highly efficient ebikes --- which are seeing rapid uptake in Ireland and other countries --- outperform electric cars, which are too heavy and expensive.
+Furthermore, highly efficient ebikes --- which are seeing rapid uptake --- outperform electric cars, which are too heavy and expensive, for the majority of trips.
 Over-reliance on electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
@@
 Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
-7 in 10 trips in Ireland are by car [@national] .
+Seven in ten trips in Ireland are by car [@national].
 Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as ebikes extend the range of this mode" [@greater].
@@
 
-::: {#fig-landing}
-![](images/paste-10.png)
-
-Illustration of the landing page of the CRUSE Tool for Ireland.
-:::
+## Aim and content
 
 This paper describes an approach to strategic cycle network planning tool development that is open, scalable, and evidence-based.
-We label the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and present a case study of its application in Ireland, the results of which are publicly available at [cruse.bike](https://cruse.bike/) in the CRUSE Tool for Ireland, the landing page of which is illustrated in @fig-landing.
+We label the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and present a case study of its application in Ireland, the results of which are publicly available at [cruse.bike](https://cruse.bike/).
+
+In @sec-tools, we review existing current tools for estimating cycling potential.
 In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool.
@@
 
-# Tools for estimating cycling potential
+# Tools for estimating cycling potential {#sec-tools}
 
@@
 Developed by researchers and developers in academic, public and private sectors, they have evolved from simple area-based static models tailored to specific regions to more complex and (in some cases) more generalizable tools.
-These tools can be considered as a specific application of the concept of planning support systems (PSS) , which are designed to help planners and decision-makers to make more informed decisions [@geertman2009].
 Outputs include maps with results available at several levels of analysis with levels of availability ranging from being only available to researchers and in static maps to open access mapping systems available to the public.
-
-Previous studies have applied the concept of PSS to specific modes, including walking [@bencekri2024], public transport [@barmentlo2012] and cycling [@bencekri2023].
-While such prior work provides insight into the potential for approaches to transport planning that are both data-driven and participatory, the focus of this paper is on tools that focus attention on the *geographic distribution* of cycling potential and which answer the question "where to build".
-Not all papers reviewed in this section classify themselves as, or even mention, PSS, but all of them can be considered as PSS.
-This paper is informed by the thinking underlying PSS, including the idea that the resulting tools are more effective if they are modular and can inter-link with other tools and processes in the planning process: PSS can be conceived as a "toolbox with separate instruments that can work and talk to each other effectively (as in the open systems concept) and that will be selectively applied in varying configurations to support a particular phase in the planning process" [@geertman2002].
-We share this conception of PSS, emphasizing the benefits of a focus on a particular mode (cycling) and a particular phase in the planning process (strategic network development).
-
 Some of the most relevant existing tools are summarised below and presented in @tbl-tools.
@@
 #| label: tbl-tools
-# For PDF:
-# knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
-#   kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
-#   kableExtra::column_spec(1, width = "13em") |>
-#   kableExtra::column_spec(2, width = "15em") |>
-#   kableExtra::column_spec(3, width = "12em") |>
-#   kableExtra::column_spec(4, width = "5em")
-# For HTML:
-knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.")
+knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
+  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
+  kableExtra::column_spec(1, width = "13em") |>
+  kableExtra::column_spec(2, width = "15em") |>
+  kableExtra::column_spec(3, width = "12em") |>
+  kableExtra::column_spec(4, width = "5em")
 ```
 
-Area-based approaches were developed before network-based tools, such as a 2007 regression-based model to estimate the percentage of cycling trips to work in England and Wales based on socioeconomic, transportation, and physical factors at the small area (wards) level in 2007 [@parkin_estimation_2007].
+Area-based tools were developed first.
+Parkin and colleagues presented a regression model to estimate the percentage of cycling trips to work in England and Wales based on socioeconomic, transportation, and physical factors at the small area (wards) level in 2007 [@parkin_estimation_2007].
 They found that areas with a higher percentage of females, non-whites, car ownership, lower socioeconomic classes, income, distance to work, population density, poor highway conditions, hills, rainfall, and fewer off-road bicycle routes tended to have a lower proportion of cycling to work.
-The approach also generated "forecasts for potential levels of bicycle use" with with increases in off-road routes, improvements of highway and, conversely, increases in car ownership and distance to work leading to decreases in cycling to work.
 
@@
 A feature of active travel interventions is that they require dense networks of routes to be effective [@parkin2018].
-This means that high levels of geographic resolution are needed in the data used to estimate cycling potential.
+This means that data with high levels of geographic resolution are necessary to estimate cyclical potential.
 However, datasets on travel patterns are often only available at the level of administrative zones.
@@
 This method works fine when the OD data represents movement between small areas, but was not appropriate for generating route networks from the POWSCAR data because zone centroids are so far apart that the resulting route networks would be sparse and unrealistic.
-To tackle this issue we developed a new method for OD data disaggregation called 'jittering' [@lovelace2022].
-The method works by first disaggregating the OD data based on a 'disaggregation threshold' and then assigning each disaggregated 'sub-OD' pair to 'subpoints' within each zone.
+To tackle this issue we developed a new method for OD data disaggregation called 'jittering'.
+<!--This reference should be removed to anonymise the paper.
+They will guess that Lovelace is involved, but saying we developed and then citing is as if we did not remove authors names from the paper :) --> The method works by first disaggregating the OD data based on a 'disaggregation threshold' and then assigning each disaggregated 'sub-OD' pair to 'subpoints' within each zone.
 As shown in @fig-dublin, the resulting route networks are dense, even in rural areas.
@@
 
-However, travel to school and work only constitute around only 30% of all trips in Ireland.
+However, travel to school and work only constitute around only 30% of all trips in Ireland. <!-- Reference needed -->
 To overcome this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
-The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
-An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes.
+The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips.
+<!-- This is already implemented, right?-->
+An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes. <!--This should be Table 2--->
 
@@
 We calculated three network types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest'.
-As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the 'Fastest' routes minimize journey time, accounting for traffic lights and surface type.
+As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the fastest routes minimize journey time, accounting for traffic lights and surface type.
 The 'Quietest' route type minimizes busy sections of road, allowing high 'diversion factors' from the fastest route on quieter but less direct ways (often avoiding roads and interactions with motor vehicles altogether where possible).
@@
 These statistics can be revealing: in Kildare it suggests that around half of all cycling activity occurs on network segments that are hostile (with a high inferred level of traffic stress), even when efforts are taken to avoid busy segments.
-For the Fastest network, which may be more realistic for utility cycling, the proportion of cycling on hostile segments is even higher at 75%.
+For the fastest network, which may be more realistic for utility cycling, the proportion of cycling on hostile segments is even higher at 75%.
 
@@
 
-Illustration of the Quietest (left) and Fastest (right) route networks for Naas (top) and Newbridge (bottom).
+Illustration of the quietest (left) and fastest (right) route networks for Naas (top) and Newbridge (bottom) in County Kildare.
 :::
@@
 
-The Climate Action Plan scenario models the transport emissions reductions targeted in the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which aims for a 51% cut in overall greenhouse gas emissions by 2030 as part of the pathway to net-zero emissions by 2050.
+The Climate Action Plan scenario models the transport emissions reductions targeted in the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which aims for a 51% cut in overall GHG emissions by 2030 as part of the pathway to net-zero emissions by 2050.
 For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.
@@
 
-To model this decrease in car driver km, cycling uptake increases in line with the Go Dutch scenario.
+To model this decrease in car km driven, cycling uptake increases in line with the Go Dutch scenario.
 However, we only model shift from driving to cycling.
@@
 Under the Go Dutch scenario cycling reaches levels equivalent to those found in the Netherlands, taking account of the effects of route hilliness (measured as mean gradient) and route distance.
-This scenario uses the same model that was used PCT, allowing trips to shift from any other mode to cycling [@lovelace2017].
+This scenario uses the same model as the PCT, allowing trips to shift from any other mode to cycling [@lovelace2017].
 
@@
 The CRUSE web application is statically hosted, meaning that it can be accessed without the need for a server running software such as the R package 'shiny' or Python packages such as 'streamlit' or 'flask' in the background [@wickham2021].
+Its landing page is illustrated in @fig-landing).
+
+::: {#fig-landing}
+![](images/paste-10.png)
+
+Illustration of the landing page of the CRUSE Tool for Ireland.
+:::
+
 Typical intended user stories are illustrated in @fig-user-stories, which shows that the tool is designed to be used by both professional and non-professional users.
@@
 The main result of the work presented in this paper is an open access web application and datasets on current and potential future cycling levels in Ireland, with evidence provided at county and route segment levels.
-Like other national tools presented in @tbl-tools, a key feature of the results is that it provides a consistent and objective baseline for comparing cycling potentials in different areas.
-
-The approach generates estimates of cycling potential on every road in the country and are therefore too extensive to present in their entirety in this paper.
+Like the PCT on which CRUSE builds, a key feature of the results is that they are systematic and available for all counties, and most routes on which cycling is permitted, in Ireland.
+The full set of results are therefore too extensive to present in their entirety in this paper.
 Through the interactive web application, users can explore the data to generate the results that are most relevant to their needs, whether that is finding the cycling potential on a particular road or finding 'weak links' or barriers in the cycle network associated with a particular school, work place or other destination.
 
-Instead of trying to show such use cases, of which there are many hundreds, we present a selection of results to illustrate the main features of the resulting evidence, in @sec-quantiative.
-We also present some qualitative results based on workshops and interviews with stakeholders in Ireland, in @sec-qualitative.
+Instead of trying to show such use cases, of which there are many hundreds, we present a selection of results to illustrate the main features of the CRUSE Tool.
 
@@
 <!-- ## Local results -->
-
-## Selected quantitative results {#sec-quantiative}
 
@@
 Aside from the widely varying sizes and route networks of each city, with Dublin having by far the densest network, it is clear from the visualisation of cycling friendliness that there are many gaps in the networks.
-Dublin has had received most investment in cycling infrastructure and this is apparent from the high proportion of the route network that is green, even in the 'Fastest' route network representing routes that prioritize speed and directness over comfort and safety.
+Dublin has had received most investment in cycling infrastructure and this is apparent from the high proportion of the route network that is green, even in the fastest route network representing routes that prioritize speed and directness over comfort and safety.
 Dublin also has the highest mode share of cycling of the four cities, adding to the evidence base showing that investment in cycling infrastructure is effective in increasing cycling uptake.
 
-Although Dublin City has made progress, there are still many parts of the Fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
+Although Dublin City has made progress, there are still many parts of the fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
 According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians" [@walking2021].
@@
 <!-- The results for the other -->
-## Selected qualitative results {#sec-qualitative}
-
-Thoughout the project, we undertook a series of workshops and interviews with stakeholders in Ireland to gather feedback on the approach and resulting tool.
-The feedback was generally positive, with stakeholders commenting on the need for the tool, lack of data on cycling potential proving a barrier to decision-making and investment, and the potential for the tool.
-At an end of project workshop in December 2023, stakeholders were asked to provide feedback on the tool and its potential use in practice.
-Quotes and suggestions from these workshops, and a subsequent interview with a Senior Transport Planning using the evidence in their everyday work in Transport Infrastructure Ireland, and presented in this section (with quotes being attributed to named individuals where permission has been granted).
-
-Early in the project, we presented the approach to practitioners in Kildare and Limmerick.
-The feedback from the workshop in Kildare highlighted the need for the tool for practitioners at the county level, as highlighted in the following quote:
-
-> "It's a missing piece of evidence that will help new projects get built" (Senior Engineer, Sustainable Transport & Traffic Management, Kildare County Council)
-
-Practitioners in Kildare County Council that the evidence generated *was already valuable* the development of new cycle routes, especially when making the case during planning applications.
-They emphasized that cycle networks are often fragmented, and the tool allows them to demonstrate how individual local schemes can contribute to a larger, interconnected network.
-Council officers expressed a preference for visualizations that directly compare the fast and quiet routes, as they recognize the importance of both types of routes (users can switch between the on the 'Route types' page for each county although this functionality is not available in the landing page map due to other feedback requesting a simpler interface).
-
-In comparison with another tool (the National Transport Authority's cycling propensity tool), the CRUSE tool was seen by practitioners in Kildare as more user-friendly and intuitive, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
-Overall, the workshop participants concluded that the approach filled gap in evidence and will greatly assist in the development of new cycling projects.
-
-During the workshop with practitioners in Limerick, the estimates of Baseline and Go Dutch potential at the network levels were found to match well with local knowledge.
-The tool effectively highlighted the same sections on the transport network that experienced planners identified as priorities for investment in cycling, such as the South Circular Road.
-Limerick County Council expressed their interest in utilizing each network layer, particularly the quiet network, to support new cyclists who lack confidence to share space with motor traffic.
-Specific suggestions from the workshop included the suggestion to align scenarios with national strategy (resulting in the Climate Action Plan scenario) and to provide data downloads provided as 'Shapefiles' for further analysis (now implemented although the download format is GeoPackage, an open standard for spatial data).
-
-In a subsequent interview with a practitioner using the evidence generated by the CRUSE approach in their everyday work, we received the following feedback.
-The outputs were a "great tool for getting started with project outline documents" and "ideal for generating a strategic rationale for estimating demand and building the business case in the absence of cycle count data" (Declan Keenan, Senior Transport Planner working on active travel, personal communication).
-Declan also noted the alignment between local knowledge, cycle count data, and the outputs of the approach: "It's a good representation of what's going on" and the networks allign well in terms of network "shape" and the proportion of trips of trips on different parts of the network.
-There were two suggestions from Declan which have not yet been implemented: the provision of data showing only "gold standard" POWSCAR data, and the need for training on the tool to ensure that all staff can use it effectively.
-These, and other potential improvements, are discussed in the next section.
 
@@
 The aim of the paper was to describe the design, features and potential use of the 'CRUSE' approach and resulting tool.
-As outlined in the previous section, the CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
+As outlined in the previous section, the CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to fastest, balanced and quietest route networks.
 This provision of multiple scenarios of behavior change *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.
@@
 
-As the benefits of strategic planning for cycling become more apparent [@scappini2022] we expect the demand for tools such as CRUSE to grow.
+As the benefits of strategic planning for cycling become more apparent [@scappini2022], we expect the demand for tools such as CRUSE to grow.
 However, the approach is not without limitations and does not meet all requirements in Ireland or indeed any country where the methods are applied.
 
-A limitation of the approach as implemented in the CRUSE tool is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
+A limitation of the approach as implemented in the CRUSE Tool is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
 Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.

### 2024-03-25 paper-v1.qmd

--- 2024-03-25 paper.qmd
@@
 ---
-title: "Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools"
-# subtitle: A case study of Ireland
-bibliography:
-  # cruse.bib
-  - bibliography.bib
-  - cruse.bib
+title: "CRUSE to Safe Cycling in Ireland"
+subtitle: An Open Source Tool to Support Active Travel
+bibliography: bibliography.bib
 csl: apa.csl
@@
     wrap: sentence
-execute: 
-  echo: false
-  message: false
 number-sections: true
@@
 abstract: |
-  This paper describes an approach for developing strategic cycle network planning tools.
-Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenario generation, and development of an intuitive user interface.
-Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions.
-The publicly available results at https://cruse.bike/ enable planners, engineers, and other stakeholders to make more evidence-based decisions.
-CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring) in addition to using official origin-destination data; and providing estimates of 'quietness' (a proxy for cyclist comfort and route preference) at the route segment level.
-Three network types --- 'Fastest', 'Balanced', and 'Quietest' --- help plan both arterial and residential cycle networks.
-Workshops and unstructured interviews with stakeholders show that the tool is already being used in practice to inform urban, inter-urban, and rural network planning.
-The approach is flexible and open-source, allowing the underlying ideas and code to be adapted, supporting effective cycling policies and interventions internationally.
+  Legislation mandates improved road safety and decarbonisation of predominantly road-based transport network in Ireland and other countries.
+Increasing cycling transport, making it safer and more attractive, is essential to this end.
+Cycling in Ireland represents only 3% of total modal share as of the latest Census (2016), but accounts for 20% of serious injuries and 7% of all fatalities.
+A planning tool based on high resolution datasets on baseline and potential cycling levels could greatly assist in achieving this goal.
+
+  This paper describes the design, features and potential use of the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland was commissioned to support cycle network planning, safety interventions, business cases and project evaluation.
+The resulting web application is available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions.
+  CRUSE goes beyond previous work in several important ways.
+  It models networks at high levels of geographic resolution, simulates a wide range of trips including recreational, social, shopping and personal trips, in addition to trips to work and primary, secondary and tertiary education locations.
+  The tool provides key 'denominator' information to enable reporting of collision rates per billion km cycled.
+
+  Another key feature of CRUSE is its use of frequently-updated OpenStreetMap data for route calculation and "cycle friendliness" estimates of all links on the network.
+  CRUSE provides three network types: the 'Fastest' network highlights routes for directness; 'Quietest' prioritises cycle friendliness, and 'Balanced' represents a balance between speed and comfort.
+  Trips are generated based on origin and destination data from the 2016 Census in combination with modeled demand data to estimate current and potential future cycling levels at area, route, and network levels for each county in Ireland.
+
+  For investment in cycling to deliver on the huge potential benefits, cycling must become safer.
+  With growth in the ebike market, CRUSE will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
+  Provision of open access data in a publicly available web tool will address the challenges faced by counties across Ireland in meeting their sustainability commitments.
+  The experience of developing a deploying CRUSE are relevant to countries with sustainable transport policies worldwide.
 keywords: 
@@
   - Active Travel
-  - Transport Planning
-  - Collaborative Planning
 date: last-modified
@@
 #| include: false
+#| eval: false
 library(tidyverse)
-```
-
-```{r}
-#| include: false
-#| eval: false
 remotes::install_dev("tmap")
@@
 system("pdftk _site/cv.pdf _site/visions.pdf cat output _site/cv_and_visions.pdf")
-# latexdiffr::latexdiff("paper-v1.qmd", "paper.qmd")
-system("latexdiff paper-v1.tex paper.tex > tracked-changes.tex")
-system("pdflatex tracked-changes.tex")
-browseURL("tracked-changes.pdf")
 ```
@@
 <!-- Mobility kills. -->
-## Background
 
-Transport systems dominated by heavy and powerful private cars are inefficient, dangerous, and unhealthy.
-Cars are a major and growing source of greenhouse gas (GHG) emissions [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018], and a cause of disease due to physical inactivity and air, noise and microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024].
-<!-- Shouldn't these three sources share the brackets?
---> Transport is responsible for 23% of GHG emissions, 70% of which are from road transport, with passenger cars accounting for nearly half of transport emissions (around 10% of global emissions) [@jaramillo2022].
+High energy transport systems are a major contributor to climate change, a leading cause of premature death and injury due to road traffic collisions, and a cause of disease due to airborne and noise pollutants.
+Conversely, evidence-based and effective transport *policies* have great potential to decarbonize economies, improve public health, and save lives.
+
+Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport.
+Nearly half of these emissions (around 10% of global emissions) are from passenger cars [@jaramillo2022].
 The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
@@
 
-Growing evidence of the negative impacts of car-dependent transport systems has led governments in many countries to set targets and take actions.
+Recognizing the growing evidence of such impacts of poorly designed and performing transport systems, governments in many countries have set targets and taken actions.
 In the context of climate, road safety and physicial inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
@@
 Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
-Furthermore, highly efficient ebikes --- which are seeing rapid uptake --- outperform electric cars, which are too heavy and expensive, for the majority of trips.
+Furthermore, highly efficient ebikes --- which are seeing rapid uptake in Ireland and other countries --- outperform electric cars, which are too heavy and expensive.
 Over-reliance on electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
@@
 
-Another motivation for cycle network planning at the European level is the Road Infrastructure Safety Management (RISM) directive (2008/96/EC), which requires member states to implement a road safety management system (RSMS) for all public roads.
+Another motivation for cycle network planning is the Road Infrastructure Safety Management (RISM) directive (2008/96/EC), which requires member states to implement a road safety management system (RSMS) for all public roads.
 Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out" [@directiv2008].
-Given that 'safety' in this context is usefully quantified as the number of people killed or seriously injured (KSI) per distance traveled, the directive requires estimation of distance traveled by mode, down to the road link level.
+Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance traveled, the directive requires estimation of distance traveled by mode, down to the road link level.
 For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
-Better data to inform road safety policies and interventions is a motivation for better estimates of 'baseline' levels of physical activity at high levels of geographic resolution [@tait2023].
+Better data to inform road safety policies and interventions is a motivation for the CRUSE Tool outlined in this paper, and modeling active travel more broadly.
 
-National governments are increasingly acting on the evidence.
 In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030 [@national2021].
-Doing so while simultaneously enabling rapid uptake of active modes will require key travel corridors to be identified 'cycle proofed'.
-Cycling in Ireland represents only 3% of total modal share as of the 2016 Census, but accounts for 20% of serious injuries and 7% of all fatalities, driving the development of cycling policy and infrastructure plans in Ireland.
+Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
+A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
 The main frameworks underpinning these efforts are the Climate Action Plan [@climate2022a], the National Development Plan, and the National Roads 2040 strategy [@national2023].
@@
 
-At the regional level within Ireland, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings of high support for and capability for cycling: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
+At the regional level, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
 Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
-Seven in ten trips in Ireland are by car [@national].
+7 in 10 trips in Ireland are by car [@national] .
 Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as ebikes extend the range of this mode" [@greater].
 
-Further evidence of the importance of cycling in Ireland is provided by the National Strategic Objective (NSO) from the National Development Plan, which allocates €8.6 billion to sustainable transport infrastructure including public transport and active travel interventions.
+Further evidence of the importance of cycling in Ireland *already* is provided by the National Strategic Objective (NSO) from the National Development Plan, with €8.6 billion allocated to sustainable transport infrastructure including public transport and active travel interventions.
 Cycle infrastructure will be developed in synchrony with the BusConnects project, an entire redesign of the bus network in Dublin and Cork.
-It was in this context that Transport Infrastructure Ireland (TII) commissioned the research reported in this paper, which led to the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
-Building on previous work, including the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE Tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
 
-Stakeholder consultation by TII emphasized the need for a strong, national, systematic but locally-specific evidence base on cycle networks in both urban and rural Ireland.
-The evidence base had to scale nationally to support strategic alignment with national, regional and local policies, but also be useful for local network planning.
+It was in this context that Transport Infrastructure Ireland (TII) commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
+Based on the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE Tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
 
-## Aim and content
+A key consideration for TII is to ensure both urban and rural Ireland were targeted for cycling infrastructure, as shown in @fig-dublin.
+To this end, TII extensively reviewed activity on Irish Roads and how the road network was being used by commuters, cyclists, heavy goods vehicles and other road users across the Irish transport network.
+The aim was for the tool to provide strong, national, systematic but locally-specific evidence to monitor cycling friendliness and safety in Ireland, to ensure strategic alignment with national, regional and local policies.
+The open source and publicly available nature of the tool ensures more inclusive and evidence-based conversations around cycle network planning between all stakeholders in the transport planning process.
 
-This paper describes an approach to strategic cycle network planning tool development that is open, scalable, and evidence-based.
-We label the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and present a case study of its application in Ireland, the results of which are publicly available at [cruse.bike](https://cruse.bike/).
+TII also requires evidence to support monitoring of growth in cycling, upgrades of existing cycling infrastructure, business case development and appraisal, and exposure information to estimate crash rates.
+Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
+These factors mean that the tool can be seen as an open access 'leverage point', providing key information for and supporting ambitious plans in many aspects the planning system, from network design to post-build monitoring [@lovelace2020].
 
-In @sec-tools, we review existing current tools for estimating cycling potential.
+::: {#fig-dublin}
+![](images/paste-1.png)
+
+![](images/paste-3.png)
+
+Illustration of the density of route networks generated by the CRUSE Tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland.
+Source: CRUSE Tool, available at <https://cruse.bike/>.
+:::
+
+This paper aims to describe the design, features and potential use of the CRUSE Tool.
 In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool.
 In @sec-results, we present the results of the CRUSE Tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
-In @sec-discussion, we discuss limitations and possible future improvements to the approach, and the implications of the results for Ireland before making concluding remarks in @sec-conclusions.
-
-# Tools for estimating cycling potential {#sec-tools}
-
-Tools for estimating geographical distribution of cycling potential have advanced substantially in recent years.
-Developed by researchers and developers in academic, public and private sectors, they have evolved from simple area-based static models tailored to specific regions to more complex and (in some cases) more generalizable tools.
-Outputs include maps with results available at several levels of analysis with levels of availability ranging from being only available to researchers and in static maps to open access mapping systems available to the public.
-Some of the most relevant existing tools are summarised below and presented in @tbl-tools.
-
-```{r}
-#| include: false
-tools = read_csv("tools.csv")
-tools = tools |>
-  transmute(
-    `Approach` = paste0(Tool, ", ", Reference),
-    `Coverage and input data` = paste0(Coverage, ", ", `Input data`),
-    `Outputs and availability` = Output,
-    License = Licence
-  )
-```
-
-```{r}
-#| label: tbl-tools
-knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
-  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
-  kableExtra::column_spec(1, width = "13em") |>
-  kableExtra::column_spec(2, width = "15em") |>
-  kableExtra::column_spec(3, width = "12em") |>
-  kableExtra::column_spec(4, width = "5em")
-```
-
-Area-based tools were developed first.
-Parkin and colleagues presented a regression model to estimate the percentage of cycling trips to work in England and Wales based on socioeconomic, transportation, and physical factors at the small area (wards) level in 2007 [@parkin_estimation_2007].
-They found that areas with a higher percentage of females, non-whites, car ownership, lower socioeconomic classes, income, distance to work, population density, poor highway conditions, hills, rainfall, and fewer off-road bicycle routes tended to have a lower proportion of cycling to work.
-
-Three years later, Transport for London created the Analysis of Cycling Potential (ACP) tool based on a mode choice model that assesses the likelihood of trips being cycled, with data from the TfL London Travel Demand Survey as its primary input.
-The tool generates tables, graphs, and heatmaps showing the potential for cycling for different trips purposes in London to guide local cycling initiatives, such as where new hire 'docking stations' should go.
-The tool does not use origin-destination data and cannot estimate cycling potential on specific routes [@transport_for_london_analysis_2010].
-
-Other tools that generate areal data to prioritize investment include a study that generated a 'prioritization index' for Montreal, Canada [@larsen_build_2013], a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil, and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
-The CPT is composed of the base environmental module — based on eight weighted factors (population density, hilliness, physical barriers, access to services, existing cycling mode share, distance to work and school, and road speed) — and the quality of service module, which evaluates each area of interest based on eight factors (surface condition, adjacent cyclists, comfort factor, conflict, distance between junctions, slope, access to services, and origin/destination).
-
-A more recent areal-based approach is the designed the Gross Potential for Cycling (GPC) tool to prioritize areas for cycling infrastructure and other cycling measures [@silva_gross_2021].
-The GPC uses two sets of indicators: population-based indicators (age, potential demand density, employment density and motorization rate) and area-based indicators (accessibility, education, public transport, connectivity, land use mix and relative performance).
-These indicators are calculated considering topography, road hierarchy, and average congestion and presented on a scale from 5 to 10.
-They are then combined into an overall score, weighted by their impact on cycling according to the literature.
-The ranking of areas by the GPC are [publicly available](https://boost.up.pt/en/ferramentas/gpc) and its practical value has been assessed through workshops [@silva_revealing_2022].
-
-Route-based tools emerged with increasing availability of origin-destination data, routing engines that can assign trips to networks, and improvements in computer hardware and software needed to generate route networks.
-A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace_mapping_2016].
-The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 @lovelace2017.
-Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modeling of impacts on health, environmental and distributional outcomes [@woodcock2021].
-Initially developed just for England, the PCT was extended to cover all of Wales (for commuter data only) in 2018.
-
-The PCT approach has been applied in other countries, including Ireland (the topic of this paper), Scotland, and Portugal.
-In Portugal the 'biclaR' project, based on methods underlying the PCT, has been developed and deployed for the Lisbon metro region.
-The resulting evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
-biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
-
-The approach presented in this paper seeks to overcome three limitations of previous tools to estimate cycling potential: 1) low resolution of data, with routes starting and ending in administrative zone centroids, 2) limited coverage of trip purposes beyond travel to work and school, and 3) a web interface that was not user-friendly or intuitive.
+In @sec-discussion, we discuss the implications of the results for policy and practice.
+In @sec-conclusions, we conclude the paper.
 
@@
 
-```{=html}
-<!-- 
 The methods used to generate the evidence presented in the CRUSE Tool build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
@@
 
-Methods were developed to overcome each of these, as outlined in @sec-trip-purposes to @sec-ui. -->
-```
+Methods were developed to overcome each of these, as outlined in @sec-trip-purposes to @sec-ui.
+
 <!-- Todo: add section on input datasets -->
@@
 A feature of active travel interventions is that they require dense networks of routes to be effective [@parkin2018].
-This means that data with high levels of geographic resolution are necessary to estimate cyclical potential.
+This means that high levels of geographic resolution are needed in the data used to estimate cycling potential.
 However, datasets on travel patterns are often only available at the level of administrative zones.
@@
 This method works fine when the OD data represents movement between small areas, but was not appropriate for generating route networks from the POWSCAR data because zone centroids are so far apart that the resulting route networks would be sparse and unrealistic.
-To tackle this issue we developed a new method for OD data disaggregation called 'jittering'.
-<!--This reference should be removed to anonymise the paper.
-They will guess that Lovelace is involved, but saying we developed and then citing is as if we did not remove authors names from the paper :) --> The method works by first disaggregating the OD data based on a 'disaggregation threshold' and then assigning each disaggregated 'sub-OD' pair to 'subpoints' within each zone.
+To tackle this issue we developed a new method for OD data disaggregation called 'jittering' [@lovelace2022].
+The method works by first disaggregating the OD data based on a 'disaggregation threshold' and then assigning each disaggregated 'sub-OD' pair to 'subpoints' within each zone.
 As shown in @fig-dublin, the resulting route networks are dense, even in rural areas.
-
-::: {#fig-dublin}
-<!-- ![](images/paste-1.png) -->
-
-![](images/paste-8.png) <!-- ![](images/paste-3.png) -->
-![](images/paste-9.png)
-
-Illustration of the density of route networks generated by the CRUSE Tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland (botton).
-Source: CRUSE Tool, publicly available at [cruse.bike](https://cruse.bike/).
-:::
 
@@
 
-However, travel to school and work only constitute around only 30% of all trips in Ireland. <!-- Reference needed -->
+However, travel to school and work only constitute around only 30% of all trips in Ireland.
 To overcome this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
-The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips.
-<!-- This is already implemented, right?-->
-An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes. <!--This should be Table 2--->
+The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
+An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes.
 
@@
 
-While CycleStreets offers a free routing API, we commissioned a custom routing service to enable:
+CycleStreets' services can be used by anyone for free, but we commissioned a bespoke routing service for CRUSE, to enable:
 
@@
 We calculated three network types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest'.
-As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the fastest routes minimize journey time, accounting for traffic lights and surface type.
+As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the 'Fastest' routes minimize journey time, accounting for traffic lights and surface type.
 The 'Quietest' route type minimizes busy sections of road, allowing high 'diversion factors' from the fastest route on quieter but less direct ways (often avoiding roads and interactions with motor vehicles altogether where possible).
@@
 These statistics can be revealing: in Kildare it suggests that around half of all cycling activity occurs on network segments that are hostile (with a high inferred level of traffic stress), even when efforts are taken to avoid busy segments.
-For the fastest network, which may be more realistic for utility cycling, the proportion of cycling on hostile segments is even higher at 75%.
+For the Fastest network, which may be more realistic for utility cycling, the proportion of cycling on hostile segments is even higher at 75%.
 
@@
 
-Illustration of the quietest (left) and fastest (right) route networks for Naas (top) and Newbridge (bottom) in County Kildare.
+Illustration of the Quietest (left) and Fastest (right) route networks for Naas (top) and Newbridge (bottom).
 :::
@@
 
-The Climate Action Plan scenario models the transport emissions reductions targeted in the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which aims for a 51% cut in overall GHG emissions by 2030 as part of the pathway to net-zero emissions by 2050.
+The Climate Action Plan scenario is loosely based on the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which contains policies for action to achieve a 51% reduction in overall GHG emissions by 2030, on a path towards net-zero by 2050.
 For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.
@@
 
-To model this decrease in car km driven, cycling uptake increases in line with the Go Dutch scenario.
+To model this decrease in car driver km, cycling uptake increases in line with the Go Dutch scenario.
 However, we only model shift from driving to cycling.
@@
 Under the Go Dutch scenario cycling reaches levels equivalent to those found in the Netherlands, taking account of the effects of route hilliness (measured as mean gradient) and route distance.
-This scenario uses the same model as the PCT, allowing trips to shift from any other mode to cycling [@lovelace2017].
+This scenario uses the same model that was used PCT, allowing trips to shift from any other mode to cycling [@lovelace2017].
 
@@
 The CRUSE web application is statically hosted, meaning that it can be accessed without the need for a server running software such as the R package 'shiny' or Python packages such as 'streamlit' or 'flask' in the background [@wickham2021].
-Its landing page is illustrated in @fig-landing).
-
-::: {#fig-landing}
-![](images/paste-10.png)
-
-Illustration of the landing page of the CRUSE Tool for Ireland.
-:::
-
 Typical intended user stories are illustrated in @fig-user-stories, which shows that the tool is designed to be used by both professional and non-professional users.
@@
 ::: {#fig-user-stories}
+
 ![](images/mermaid-figure-1.png)
@@
 Hypothetical user stories for the CRUSE Tool illustrating that it can be used in different ways by different user groups.
+
 :::
@@
 Aside from the widely varying sizes and route networks of each city, with Dublin having by far the densest network, it is clear from the visualisation of cycling friendliness that there are many gaps in the networks.
-Dublin has had received most investment in cycling infrastructure and this is apparent from the high proportion of the route network that is green, even in the fastest route network representing routes that prioritize speed and directness over comfort and safety.
+Dublin has had received most investment in cycling infrastructure and this is apparent from the high proportion of the route network that is green, even in the 'Fastest' route network representing routes that prioritize speed and directness over comfort and safety.
 Dublin also has the highest mode share of cycling of the four cities, adding to the evidence base showing that investment in cycling infrastructure is effective in increasing cycling uptake.
 
-Although Dublin City has made progress, there are still many parts of the fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
+Although Dublin City has made progress, there are still many parts of the Fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
 According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians" [@walking2021].
@@
 
-The aim of the paper was to describe the design, features and potential use of the 'CRUSE' approach and resulting tool.
-As outlined in the previous section, the CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to fastest, balanced and quietest route networks.
+The aim of the paper was to describe the design, features and potential use of the CRUSE Tool.
+As outlined in the previous section, the CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
 This provision of multiple scenarios of behavior change *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.
-
-## Uses of the tool in practice
 
@@
 
-## Comparison with other tools and contribution to the field
-
 While the CRUSE Tool is not the first national and publicly available tool for cycling planning, it has some key features that make it relevant for other countries, regions and road authorities tasked with making their transport systems safe and sustainable:
@@
 
-As the benefits of strategic planning for cycling become more apparent [@scappini2022], we expect the demand for tools such as CRUSE to grow.
-However, the approach is not without limitations and does not meet all requirements in Ireland or indeed any country where the methods are applied.
-
-A limitation of the approach as implemented in the CRUSE Tool is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
-Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
-These factors mean that the tool can be seen as an open access 'leverage point', providing key information for and supporting ambitious plans in many aspects the planning system, from network design to post-build monitoring [@lovelace2020].
+As the benefits of strategic planning for cycling become more apparent [@scappini2022] we expect the demand for tools such as CRUSE to grow.
 

### 2024-03-25 paper-v1.tex

--- 2024-03-25 paper.tex
@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{setspace}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
 \fi
@@
 
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
 \makeatletter
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
 \@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
 \makeatother
@@
 \hypersetup{
-  pdftitle={Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools},
-  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Toole; Dan Brennan; Suzanne Meade},
+  pdftitle={CRUSE to Safe Cycling in Ireland},
+  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Tool; Dan Brennan; Suzanne Meade},
   pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
@@
 \begin{frontmatter}
-\title{Cycle Route Uptake and Scenario Estimation (CRUSE): an approach
-for developing strategic cycle network planning tools}
+\title{CRUSE to Safe Cycling in Ireland \\\large{An Open Source Tool to
+Support Active Travel} }
 \author[1]{Robin Lovelace%
@@
 
-\author[2]{Gary O'Toole%
+\author[2]{Gary O'Tool%
 %
@@
 \begin{abstract}
-This paper describes an approach for developing strategic cycle network
-planning tools. Based on our experience developing and deploying the
-Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we
-outline the underlying methods, including disaggregation of
-origin-destination data, incorporation of additional trip purposes,
-routing, scenario generation, and development of an intuitive user
-interface. Commissioned and used by Transport Infrastructure Ireland,
-CRUSE provides estimates of current and potential future cycling levels
-under `snapshot' scenarios to inform investment decisions. The publicly
-available results at https://cruse.bike/ enable planners, engineers, and
-other stakeholders to make more evidence-based decisions. CRUSE goes
-beyond previous work by: modeling networks at high spatial resolution;
-simulating multiple trip purposes (social, shopping, personal utility,
-recreational, and cycle touring) in addition to using official
-origin-destination data; and providing estimates of `quietness' (a proxy
-for cyclist comfort and route preference) at the route segment level.
-Three network types --- `Fastest', `Balanced', and `Quietest' --- help
-plan both arterial and residential cycle networks. Workshops and
-unstructured interviews with stakeholders show that the tool is already
-being used in practice to inform urban, inter-urban, and rural network
-planning. The approach is flexible and open-source, allowing the
-underlying ideas and code to be adapted, supporting effective cycling
-policies and interventions internationally.
+Legislation mandates improved road safety and decarbonisation of
+predominantly road-based transport network in Ireland and other
+countries. Increasing cycling transport, making it safer and more
+attractive, is essential to this end. Cycling in Ireland represents only
+3\% of total modal share as of the latest Census (2016), but accounts
+for 20\% of serious injuries and 7\% of all fatalities. A planning tool
+based on high resolution datasets on baseline and potential cycling
+levels could greatly assist in achieving this goal.
+
+This paper describes the design, features and potential use of the Cycle
+Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland was
+commissioned to support cycle network planning, safety interventions,
+business cases and project evaluation. The resulting web application is
+available at \url{https://cruse.bike/}, enabling planners, engineers,
+and other stakeholders to make more evidence-based decisions. CRUSE goes
+beyond previous work in several important ways. It models networks at
+high levels of geographic resolution, simulates a wide range of trips
+including recreational, social, shopping and personal trips, in addition
+to trips to work and primary, secondary and tertiary education
+locations. The tool provides key `denominator' information to enable
+reporting of collision rates per billion km cycled.
+
+Another key feature of CRUSE is its use of frequently-updated
+OpenStreetMap data for route calculation and ``cycle friendliness''
+estimates of all links on the network. CRUSE provides three network
+types: the `Fastest' network highlights routes for directness;
+`Quietest' prioritises cycle friendliness, and `Balanced' represents a
+balance between speed and comfort. Trips are generated based on origin
+and destination data from the 2016 Census in combination with modeled
+demand data to estimate current and potential future cycling levels at
+area, route, and network levels for each county in Ireland.
+
+For investment in cycling to deliver on the huge potential benefits,
+cycling must become safer. With growth in the ebike market, CRUSE will
+help inform inter-urban and rural networks to support the transfer of
+trips to sustainable modes for longer journeys. Provision of open access
+data in a publicly available web tool will address the challenges faced
+by counties across Ireland in meeting their sustainability commitments.
+The experience of developing a deploying CRUSE are relevant to countries
+with sustainable transport policies worldwide.
 \end{abstract}
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-Transport systems dominated by heavy and powerful private cars are
-inefficient, dangerous, and unhealthy. Cars are a major and growing
-source of greenhouse gas emissions causing climate change
-\citep{winkler2023}, a leading cause of premature death and injury due
-to road traffic collisions \citep{globals2018} , and a cause of disease
-due to physical inactivity and air, noise and (from tyres) microplastic
-pollution \citep{mattsson2023} \citep{welch2023} \citep{cavallaro2024}.
+\section{Introduction}\label{introduction}
+
+High energy transport systems are a major contributor to climate change,
+a leading cause of premature death and injury due to road traffic
+collisions, and a cause of disease due to airborne and noise pollutants.
+Conversely, evidence-based and effective transport \emph{policies} have
+great potential to decarbonize economies, improve public health, and
+save lives.
+
 Transport is responsible for 23\% of global greenhouse gas (GHG)
-emissions, 70\% of which are from road transport, with passenger cars
-accounting for nearly half of transport emissions (around 10\% of global
-emissions) \citep{jaramillo2022}. The transport system encourages,
-enables and in some cases enforces unsustainable lifestyles. Services
-that are only accessible by car lock-in car dependency
+emissions, 70\% of which are from road transport. Nearly half of these
+emissions (around 10\% of global emissions) are from passenger cars
+\citep{jaramillo2022}. The transport system encourages, enables and in
+some cases enforces unsustainable lifestyles. Services that are only
+accessible by car lock-in car dependency
 \citep{gray2001, shergold2012, motte-baumvol2010}.
 
-Growing evidence of the negative impacts of car-dependent transport
-systems has led governments in many countries to set targets and take
-actions. In the context of climate, road safety and physicial inactivity
-crises, policies to improve transport systems can be classified
-according to the `Avoid-Shift-Improve' (ASI) framework
+Recognizing the growing evidence of such impacts of poorly designed and
+performing transport systems, governments in many countries have set
+targets and taken actions. In the context of climate, road safety and
+physicial inactivity crises, policies to improve transport systems can
+be classified according to the `Avoid-Shift-Improve' (ASI) framework
 \citep{jaramillo2022}. The ASI framework highlights the importance of
@@
 
-Another motivation for cycle network planning at the European level is
-the Road Infrastructure Safety Management (RISM) directive (2008/96/EC),
-which requires member states to implement a road safety management
-system (RSMS) for all public roads. Specifically, ``Member States shall
-ensure that the ranking of high accident concentration sections and the
-network safety ranking are carried out'' \citep{directiv2008}. Given
-that `safety' in this context is usefully quantified as the number of
-people killed or seriously injured (KSI) per distance traveled, the
-directive requires estimation of distance traveled by mode, down to the
-road link level. For active modes, about which there is a paucity of
-data compared with motorized modes, this is a major challenge. Better
-data to inform road safety policies and interventions is a motivation
-for better estimates of `baseline' levels of physical activity at high
-levels of geographic resolution \citep{tait2023}.
-
-National governments are increasingly acting on the evidence. In
-Ireland, the Road Safety Authority (RSA) has set the target of halving
-the number of road traffic deaths and serious injuries by 2030
+Another motivation for cycle network planning is the Road Infrastructure
+Safety Management (RISM) directive (2008/96/EC), which requires member
+states to implement a road safety management system (RSMS) for all
+public roads. Specifically, ``Member States shall ensure that the
+ranking of high accident concentration sections and the network safety
+ranking are carried out'' \citep{directiv2008}. Given that `safety' in
+this context is usefully quantified as the number of people killed and
+seriously injured (KSI) per distance traveled, the directive requires
+estimation of distance traveled by mode, down to the road link level.
+For active modes, about which there is a paucity of data compared with
+motorized modes, this is a major challenge. Better data to inform road
+safety policies and interventions is a motivation for the CRUSE Tool
+outlined in this paper, and modeling active travel more broadly.
+
+In Ireland, the Road Safety Authority (RSA) has set the target of
+halving the number of road traffic deaths and serious injuries by 2030
 \citep{national2021}. Doing so while simultaneously enabling rapid
-uptake of active modes will require key travel corridors to be
-identified `cycle proofed'. Cycling in Ireland represents only 3\% of
-total modal share as of the 2016 Census, but accounts for 20\% of
-serious injuries and 7\% of all fatalities, driving the development of
-cycling policy and infrastructure plans in Ireland. The main frameworks
-underpinning these efforts are the Climate Action Plan
+uptake of active modes will require key active travel routes to be
+identified and improved. A proactive approach to the development of
+cycling policy and infrastructure is taking place in Ireland. The main
+frameworks underpinning these efforts are the Climate Action Plan
 \citep{climate2022a}, the National Development Plan, and the National
@@
 
-At the regional level within Ireland, the recently published Greater
-Dublin Area Transport Strategy \citep{greater} reinforces these findings
-of high support for and capability for cycling: ``nearly a quarter of
-adults cycle at least once a week in the Dublin Metropolitan Area'' with
-cycling in the Dublin area taking up to 60,000 cars off the road today.
-Extrapolating this on a per population basis across Ireland, with around
-40\% of the population living in Dublin, suggests that around 150,000
-cars could be removed nationwide just by achieving Dublin levels of
-cycling in all counties (notwithstanding existing cycling trips and
-differences in trip distances). 7 in 10 trips in Ireland are by car
-\citep{national} . Cycling has the potential to replace a large
+At the regional level, the recently published Greater Dublin Area
+Transport Strategy \citep{greater} reinforces these findings: ``nearly a
+quarter of adults cycle at least once a week in the Dublin Metropolitan
+Area'' with cycling in the Dublin area taking up to 60,000 cars off the
+road today. Extrapolating this on a per population basis across Ireland,
+with around 40\% of the population living in Dublin, suggests that
+around 150,000 cars could be removed nationwide just by achieving Dublin
+levels of cycling in all counties (notwithstanding existing cycling
+trips and differences in trip distances). 7 in 10 trips in Ireland are
+by car \citep{national} . Cycling has the potential to replace a large
 proportion of these trips: ``A high priority must also be given to
@@
 
-Further evidence of the importance of cycling in Ireland is provided by
-the National Strategic Objective (NSO) from the National Development
-Plan, which allocates €8.6 billion to sustainable transport
+Further evidence of the importance of cycling in Ireland \emph{already}
+is provided by the National Strategic Objective (NSO) from the National
+Development Plan, with €8.6 billion allocated to sustainable transport
 infrastructure including public transport and active travel
@@
 the BusConnects project, an entire redesign of the bus network in Dublin
-and Cork. It was in this context that Transport Infrastructure Ireland
-(TII) commissioned the research reported in this paper, which led to the
-Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
-Building on previous work, including the Propensity to Cycle Tool (PCT)
-for England and Wales, the CRUSE Tool was developed to provide evidence
-on current cycling levels and future cycling potential nationwide across
-Ireland.
-
-Stakeholder consultation by TII emphasized the need for a strong,
-national, systematic but locally-specific evidence base on cycle
-networks in both urban and rural Ireland. The evidence base had to scale
-nationally to support strategic alignment with national, regional and
-local policies, but also be useful for local network planning.
+and Cork.
+
+It was in this context that Transport Infrastructure Ireland (TII)
+commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool
+for Ireland. Based on the Propensity to Cycle Tool (PCT) for England and
+Wales, the CRUSE Tool was developed to provide evidence on current
+cycling levels and future cycling potential nationwide across Ireland.
+
+A key consideration for TII is to ensure both urban and rural Ireland
+were targeted for cycling infrastructure, as shown in
+Figure~\ref{fig-dublin}. To this end, TII extensively reviewed activity
+on Irish Roads and how the road network was being used by commuters,
+cyclists, heavy goods vehicles and other road users across the Irish
+transport network. The aim was for the tool to provide strong, national,
+systematic but locally-specific evidence to monitor cycling friendliness
+and safety in Ireland, to ensure strategic alignment with national,
+regional and local policies. The open source and publicly available
+nature of the tool ensures more inclusive and evidence-based
+conversations around cycle network planning between all stakeholders in
+the transport planning process.
+
+TII also requires evidence to support monitoring of growth in cycling,
+upgrades of existing cycling infrastructure, business case development
+and appraisal, and exposure information to estimate crash rates.
+Exposure information is a key part of the the European RISM Directive,
+for reporting collision rates of vulnerable road users by 2024. These
+factors mean that the tool can be seen as an open access `leverage
+point', providing key information for and supporting ambitious plans in
+many aspects the planning system, from network design to post-build
+monitoring \citep{lovelace2020}.
 
@@
 
-{\centering
-
-\includegraphics{images/paste-10.png}
-
-}
-
-\caption{\label{fig-landing}Illustration of the landing page of the
-CRUSE Tool for Ireland.}
-
-\end{figure}
-
-This paper describes an approach to strategic cycle network planning
-tool development that is open, scalable, and evidence-based. We label
-the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and
-present a case study of its application in Ireland, the results of which
-are publicly available at \href{https://cruse.bike/}{cruse.bike} in the
-CRUSE Tool for Ireland, the landing page of which is illustrated in
-Figure~\ref{fig-landing}. In Section~\ref{sec-methods}, we outline the
-methods used to generate the evidence presented in the CRUSE Tool. In
+\centering{
+
+\includegraphics{images/paste-1.png}
+
+\includegraphics{images/paste-3.png}
+
+}
+
+\caption{\label{fig-dublin}Illustration of the density of route networks
+generated by the CRUSE Tool for Dublin City and surroundings (top) and
+rural County Mayo on the West coast of Ireland. Source: CRUSE Tool,
+available at \url{https://cruse.bike/}.}
+
+\end{figure}%
+
+This paper aims to describe the design, features and potential use of
+the CRUSE Tool. In Section~\ref{sec-methods}, we outline the methods
+used to generate the evidence presented in the CRUSE Tool. In
 Section~\ref{sec-results}, we present the results of the CRUSE Tool,
@@
 potential at the national, regional and local levels. In
-Section~\ref{sec-discussion}, we discuss limitations and possible future
-improvements to the approach, and the implications of the results for
-Ireland before making concluding remarks in
-Section~\ref{sec-conclusions}.
-
-\hypertarget{tools-for-estimating-cycling-potential}{%
-\section{Tools for estimating cycling
-potential}\label{tools-for-estimating-cycling-potential}}
-
-Tools for estimating geographical distribution of cycling potential have
-advanced substantially in recent years. Developed by researchers and
-developers in academic, public and private sectors, they have evolved
-from simple area-based static models tailored to specific regions to
-more complex and (in some cases) more generalizable tools. Outputs
-include maps with results available at several levels of analysis with
-levels of availability ranging from being only available to researchers
-and in static maps to open access mapping systems available to the
-public. Some of the most relevant existing tools are summarised below
-and presented in Table~\ref{tbl-tools}.
-
-\begin{table}
-
-\caption{\label{tbl-tools}\textbf{?(caption)}}\begin{minipage}[t]{\linewidth}
-\subcaption{\label{tbl-tools-1}Summary of existing tools for estimating cycling potential.}
-
-{\centering
-
-\centering
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{13em}>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{12em}>{\raggedright\arraybackslash}p{5em}}
-\toprule
-Approach & Coverage and input data & Outputs and availability & License\\
-\midrule
-\cellcolor{gray!6}{Bicycle share model, Parkin et al.
-(2007)} & \cellcolor{gray!6}{England and Wales, Journeys to work OD census data, at the small-area (wards) level} & \cellcolor{gray!6}{Static tables in academic paper} & \cellcolor{gray!6}{Proprietary}\\
-Analysis of  Cycling Potential (ACP), Transport for London (2010, 2016) & London, London Travel Demand Survey & Static tables, graphs, and heatmaps, not publicly available & Proprietary\\
-\cellcolor{gray!6}{Prioritization index, Larsen et al.
-(2013)} & \cellcolor{gray!6}{Montreal, Survey, Road safety, and OD data} & \cellcolor{gray!6}{Map-based heatmap, not publicly available} & \cellcolor{gray!6}{Proprietary}\\
-Usage intensity index, Zhang et al. (2014) & Belo Horizonte, Survey, census, and OD data & Static tables and maps, not publicly available & Proprietary\\
-\cellcolor{gray!6}{The Cycling Potential Tool (CPT), Phillips and Range (2017)} & \cellcolor{gray!6}{Scotland, Environmental and socioeconomic data, at the small area (output areas) level} & \cellcolor{gray!6}{Maps showing cycling potential in each area} & \cellcolor{gray!6}{Proprietary}\\
-\addlinespace
-Propensity to Cycle Tool (PCT), Lovelace et al.
-(2017) & England and Wales, Journeys to work and school OD census data, at the small area (LSOA) level & Online maps, graphs, tables, publicly available at www.pct.bike & Open source\\
-\cellcolor{gray!6}{The    Gross Potential for Cycling tool (CPC), Silva et al.
-(2021 and 2022)} & \cellcolor{gray!6}{21 Portuguese cities, Land use and socio-demographic data, at the small area (census tract) level} & \cellcolor{gray!6}{Static maps showing cycling potential in different areas} & \cellcolor{gray!6}{NA}\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{minipage}%
-
-\end{table}
-
-Area-based tools were developed first. Parkin and colleagues presented a
-regression model to estimate the percentage of cycling trips to work in
-England and Wales based on socioeconomic, transportation, and physical
-factors at the small area (wards) level in 2007
-\citep{parkin_estimation_2007}. They found that areas with a higher
-percentage of females, non-whites, car ownership, lower socioeconomic
-classes, income, distance to work, population density, poor highway
-conditions, hills, rainfall, and fewer off-road bicycle routes tended to
-have a lower proportion of cycling to work.
-
-Three years later, Transport for London created the Analysis of Cycling
-Potential (ACP) tool based on a mode choice model that assesses the
-likelihood of trips being cycled, with data from the TfL London Travel
-Demand Survey as its primary input. The tool generates tables, graphs,
-and heatmaps showing the potential for cycling for different trips
-purposes in London to guide local cycling initiatives, such as where new
-hire `docking stations' should go. The tool does not use
-origin-destination data and cannot estimate cycling potential on
-specific routes \citep{transport_for_london_analysis_2010}.
-
-Other tools that generate areal data to prioritize investment include a
-study that generated a `prioritization index' for Montreal, Canada
-\citep{larsen_build_2013}, a `usage intensity index' based on stated
-preferences in Belo Horizonte, Brazil, and a Cycling Potential Tool
-(CPT) for Scotland \citep{phillips_development_2017}. The CPT is
-composed of the base environmental module --- based on eight weighted
-factors (population density, hilliness, physical barriers, access to
-services, existing cycling mode share, distance to work and school, and
-road speed) --- and the quality of service module, which evaluates each
-area of interest based on eight factors (surface condition, adjacent
-cyclists, comfort factor, conflict, distance between junctions, slope,
-access to services, and origin/destination).
-
-A more recent areal-based approach is the designed the Gross Potential
-for Cycling (GPC) tool to prioritize areas for cycling infrastructure
-and other cycling measures \citep{silva_gross_2021}. The GPC uses two
-sets of indicators: population-based indicators (age, potential demand
-density, employment density and motorization rate) and area-based
-indicators (accessibility, education, public transport, connectivity,
-land use mix and relative performance). These indicators are calculated
-considering topography, road hierarchy, and average congestion and
-presented on a scale from 5 to 10. They are then combined into an
-overall score, weighted by their impact on cycling according to the
-literature. The ranking of areas by the GPC are
-\href{https://boost.up.pt/en/ferramentas/gpc}{publicly available} and
-its practical value has been assessed through workshops
-\citep{silva_revealing_2022}.
-
-Route-based tools emerged with increasing availability of
-origin-destination data, routing engines that can assign trips to
-networks, and improvements in computer hardware and software needed to
-generate route networks. A prominent example is the the Propensity to
-Cycle Tool (PCT), first developed to estimate current and future levels
-of cycling at desire line, zone, route, and route network levels for
-case study cities \citep{lovelace_mapping_2016}. The approach was
-scaled-up to estimate estimate the potential benefits of uptake at zone
-and desire line levels nationally, and launched as a publicly available
-web application in 2017 \citet{lovelace2017}. Extensions of the PCT
-approach have included estimation of benefits at the individual level
+Section~\ref{sec-discussion}, we discuss the implications of the results
+for policy and practice. In Section~\ref{sec-conclusions}, we conclude
+the paper.
+
+\section{Methods and data}\label{sec-methods}
+
+The methods used to generate the evidence presented in the CRUSE Tool
+build on the Propensity to Cycle Tool (PCT), which was originally funded
+by the UK's Department for Transport and developed by a multi-university
+team. An important feature of the PCT is that it is open source and
+publicly available (at \href{https://www.pct.bike/}{www.pct.bike}),
+allowing its use by all stakeholders in the transport planning process
+\citep{lovelace2017}. The PCT approach has had major policy impacts, as
+outlined in Research Excellence Framework (REF) impact case studies,
+which demonstrate that the tool ``revolutionized strategic cycle
+planning in England and Wales'' by overcoming the barriers to cycling
+investment imposed by lack of evidence on cycling
+potential\citep{lovelace2023}.
+
+The first version of the PCT was based on current and future potential
+uptake of cycling for single stage \emph{travel to work} at desire line,
+zone, route, and route network levels \citep{lovelace2016}. It was
+launched in April 2017 as the government-endorsed tool for strategic
+cycle network planning and part of the Cycling and Walking Investment
+Strategy \citep{cycling2017}. Extensions of the PCT approach have
+included estimation of benefits at the individual level
 \citep{woodcock2018}, addition of travel to school network
@@
 
-The approach presented in this paper seeks to overcome three limitations
-of previous tools to estimate cycling potential: 1) low resolution of
-data, with routes starting and ending in administrative zone centroids,
-2) limited coverage of trip purposes beyond travel to work and school,
-and 3) a web interface that was not user-friendly or intuitive.
-
-\hypertarget{sec-methods}{%
-\section{Methods and data}\label{sec-methods}}
-
-\hypertarget{sec-disaggregation}{%
+The CRUSE Tool seeks to overcome the following methodological
+limitations of the original PCT:
+
+\begin{itemize}
+\tightlist
+\item
+  Low resolution of data, with routes starting and ending in
+  administrative zone centroids
+\item
+  Limited coverage of trip purposes beyond travel to work and school
+\item
+  A web interface that was not user-friendly or intuitive
+\end{itemize}
+
+Methods were developed to overcome each of these, as outlined in
+Section~\ref{sec-trip-purposes} to Section~\ref{sec-ui}.
+
 \subsection{Disaggregation of origin-destination
-data}\label{sec-disaggregation}}
+data}\label{sec-disaggregation}
 
@@
 
-\begin{figure}
-
-{\centering
-
-\includegraphics{images/paste-8.png}
-\includegraphics{images/paste-9.png}
-
-}
-
-\caption{\label{fig-dublin}Illustration of the density of route networks
-generated by the CRUSE Tool for Dublin City and surroundings (top) and
-rural County Mayo on the West coast of Ireland (botton). Source: CRUSE
-Tool, publicly available at \href{https://cruse.bike/}{cruse.bike}.}
-
-\end{figure}
-
-\hypertarget{sec-trip-purposes}{%
-\subsection{Additional trip purposes}\label{sec-trip-purposes}}
+\subsection{Additional trip purposes}\label{sec-trip-purposes}
 
@@
 
-While CycleStreets offers a free routing API, we commissioned a custom
-routing service to enable:
+CycleStreets' services can be used by anyone for free, but we
+commissioned a bespoke routing service for CRUSE, to enable:
 
@@
 
-\hypertarget{climate-action-plan}{%
-\subsubsection{Climate Action Plan}\label{climate-action-plan}}
-
-The Climate Action Plan scenario models the transport emissions
-reductions targeted in the Irish Government's
+\subsubsection{Climate Action Plan}\label{climate-action-plan}
+
+The Climate Action Plan scenario is loosely based on the Irish
+Government's
 \href{https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/}{Climate
-Action Plan 2021}, which aims for a 51\% cut in overall greenhouse gas
-emissions by 2030 as part of the pathway to net-zero emissions by 2050.
-For transport, this includes 500,000 extra walking, cycling and public
-transport trips per day by 2030. In terms of car travel, the target is
-to ``Increas{[}e{]} the proportion of kilometers driven by passenger
-electric cars to between 40 and 45\% by 2030, in addition to a reduction
-of 10\% in kilometers driven by the remaining internal combustion engine
-cars.'' This equates to a 5.5 to 6\% reduction in total car km driven.
+Action Plan 2021}, which contains policies for action to achieve a 51\%
+reduction in overall GHG emissions by 2030, on a path towards net-zero
+by 2050. For transport, this includes 500,000 extra walking, cycling and
+public transport trips per day by 2030. In terms of car travel, the
+target is to ``Increas{[}e{]} the proportion of kilometers driven by
+passenger electric cars to between 40 and 45\% by 2030, in addition to a
+reduction of 10\% in kilometers driven by the remaining internal
+combustion engine cars.'' This equates to a 5.5 to 6\% reduction in
+total car km driven.
 
@@
 
-\hypertarget{sec-discussion}{%
-\section{Discussion}\label{sec-discussion}}
+\section{Discussion}\label{sec-discussion}
 
 The aim of the paper was to describe the design, features and potential
-use of the `CRUSE' approach and resulting tool. As outlined in the
-previous section, the CRUSE Tool provides evidence on current and
-potential future cycling levels across Ireland down to the street level,
-with potential assigned to `fastest', `balanced' and `quietest' route
-networks. This provision of multiple scenarios of behavior change
-\emph{and} multiple scenarios of investment, for example in cycling
-infrastructure next to major roads vs quiet residential streets, is a
-key feature of the CRUSE Tool.
-
-\hypertarget{uses-of-the-tool-in-practice}{%
-\subsection{Uses of the tool in
-practice}\label{uses-of-the-tool-in-practice}}
+use of the CRUSE Tool. As outlined in the previous section, the CRUSE
+Tool provides evidence on current and potential future cycling levels
+across Ireland down to the street level, with potential assigned to
+`fastest', `balanced' and `quietest' route networks. This provision of
+multiple scenarios of behavior change \emph{and} multiple scenarios of
+investment, for example in cycling infrastructure next to major roads vs
+quiet residential streets, is a key feature of the CRUSE Tool.
 
@@
 
-\hypertarget{comparison-with-other-tools-and-contribution-to-the-field}{%
-\subsection{Comparison with other tools and contribution to the
-field}\label{comparison-with-other-tools-and-contribution-to-the-field}}
-
 While the CRUSE Tool is not the first national and publicly available
@@
 \citep{scappini2022} we expect the demand for tools such as CRUSE to
-grow. However, the approach is not without limitations and does not meet
-all requirements in Ireland or indeed any country where the methods are
-applied.
-
-A limitation of the approach as implemented in the CRUSE tool is that
-the results are `static', reducing its ability to support monitoring of
-growth in cycling, upgrades of existing cycling infrastructure and
-dynamic (year-by-year) exposure information to estimate crash rates.
-Exposure information is a key part of the the European RISM Directive,
-for reporting collision rates of vulnerable road users by 2024. These
-factors mean that the tool can be seen as an open access `leverage
-point', providing key information for and supporting ambitious plans in
-many aspects the planning system, from network design to post-build
-monitoring \citep{lovelace2020}.
-
-\hypertarget{sec-conclusions}{%
-\section{Conclusions}\label{sec-conclusions}}
+grow.
+
+\section{Conclusions}\label{sec-conclusions}
 
@@
 
-  \bibliography{bibliography.bib,cruse.bib}
+  \bibliography{bibliography.bib}
 

### 2024-03-25 paper.tex
Update tex

@@
   pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Toole; Dan Brennan; Suzanne Meade},
-  pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
+  pdfkeywords={Cycling, Open-source, Road Safety, Active
+Travel, Transport Planning, Collaborative Planning},
   colorlinks=true,
@@
 \begin{keyword}
-    Cycling \sep Open-source \sep Road Safety \sep 
-    Active Travel
+    Cycling \sep Open-source \sep Road Safety \sep Active
+Travel \sep Transport Planning \sep 
+    Collaborative Planning
 \end{keyword}
@@
 
+\subsection{Background}\label{background}
+
 Transport systems dominated by heavy and powerful private cars are
 inefficient, dangerous, and unhealthy. Cars are a major and growing
-source of greenhouse gas emissions causing climate change
-\citep{winkler2023}, a leading cause of premature death and injury due
-to road traffic collisions \citep{globals2018} , and a cause of disease
-due to physical inactivity and air, noise and (from tyres) microplastic
-pollution \citep{mattsson2023} \citep{welch2023} \citep{cavallaro2024}.
-Transport is responsible for 23\% of global greenhouse gas (GHG)
-emissions, 70\% of which are from road transport, with passenger cars
-accounting for nearly half of transport emissions (around 10\% of global
-emissions) \citep{jaramillo2022}. The transport system encourages,
-enables and in some cases enforces unsustainable lifestyles. Services
-that are only accessible by car lock-in car dependency
+source of greenhouse gas (GHG) emissions \citep{winkler2023}, a leading
+cause of premature death and injury due to road traffic collisions
+\citep{globals2018}, and a cause of disease due to physical inactivity
+and air, noise and microplastic pollution \citep{mattsson2023}
+\citep{welch2023} \citep{cavallaro2024}. Transport is responsible for
+23\% of GHG emissions, 70\% of which are from road transport, with
+passenger cars accounting for nearly half of transport emissions (around
+10\% of global emissions) \citep{jaramillo2022}. The transport system
+encourages, enables and in some cases enforces unsustainable lifestyles.
+Services that are only accessible by car lock-in car dependency
 \citep{gray2001, shergold2012, motte-baumvol2010}.
@@
 Furthermore, highly efficient ebikes --- which are seeing rapid uptake
-in Ireland and other countries --- outperform electric cars, which are
-too heavy and expensive. Over-reliance on electric cars could slow the
+--- outperform electric cars, which are too heavy and expensive, for the
+majority of trips. Over-reliance on electric cars could slow the
 transition away from car dependency and inadvertently enable ``high
@@
 counties (notwithstanding existing cycling trips and differences in trip
-distances). 7 in 10 trips in Ireland are by car \citep{national} .
+distances). Seven in ten trips in Ireland are by car \citep{national}.
 Cycling has the potential to replace a large proportion of these trips:
@@
 
-\begin{figure}
-
-\centering{
-
-\includegraphics{images/paste-10.png}
-
-}
-
-\caption{\label{fig-landing}Illustration of the landing page of the
-CRUSE Tool for Ireland.}
-
-\end{figure}%
+\subsection{Aim and content}\label{aim-and-content}
 
@@
 present a case study of its application in Ireland, the results of which
-are publicly available at \href{https://cruse.bike/}{cruse.bike} in the
-CRUSE Tool for Ireland, the landing page of which is illustrated in
-Figure~\ref{fig-landing}. In Section~\ref{sec-methods}, we outline the
-methods used to generate the evidence presented in the CRUSE Tool. In
-Section~\ref{sec-results}, we present the results of the CRUSE Tool,
+are publicly available at \href{https://cruse.bike/}{cruse.bike}.
+
+In Section~\ref{sec-tools}, we review existing current tools for
+estimating cycling potential. In Section~\ref{sec-methods}, we outline
+the methods used to generate the evidence presented in the CRUSE Tool.
+In Section~\ref{sec-results}, we present the results of the CRUSE Tool,
 including estimates of current cycling levels and future cycling
@@
 
-\section{Tools for estimating cycling
-potential}\label{tools-for-estimating-cycling-potential}
+\section{Tools for estimating cycling potential}\label{sec-tools}
 
@@
 networks of routes to be effective \citep{parkin2018}. This means that
-high levels of geographic resolution are needed in the data used to
-estimate cycling potential. However, datasets on travel patterns are
-often only available at the level of administrative zones. The Central
-Statistics Office (CSO) in Ireland provides Place of Work, School or
-College Census of Anonymized Records
+data with high levels of geographic resolution are necessary to estimate
+cyclical potential. However, datasets on travel patterns are often only
+available at the level of administrative zones. The Central Statistics
+Office (CSO) in Ireland provides Place of Work, School or College Census
+of Anonymized Records
 (\href{https://www.cso.ie/en/census/census2016reports/powscar/}{POWSCAR})
@@
 sparse and unrealistic. To tackle this issue we developed a new method
-for OD data disaggregation called `jittering' \citep{lovelace2022}. The
-method works by first disaggregating the OD data based on a
-`disaggregation threshold' and then assigning each disaggregated
-`sub-OD' pair to `subpoints' within each zone. As shown in
-Figure~\ref{fig-dublin}, the resulting route networks are dense, even in
-rural areas.
+for OD data disaggregation called `jittering'. The method works by first
+disaggregating the OD data based on a `disaggregation threshold' and
+then assigning each disaggregated `sub-OD' pair to `subpoints' within
+each zone. As shown in Figure~\ref{fig-dublin}, the resulting route
+networks are dense, even in rural areas.
 
@@
 but with the addition of categories based on the comprehensive POWSCAR
-data, and the need to include recreational trips and multi-stage trips
-(not yet implemented). An overview of the trip purposes used in CRUSE is
-presented in Figure~\ref{fig-trip-purposes}.
+data, and the need to include recreational trips and multi-stage trips.
+An overview of the trip purposes used in CRUSE is presented in
+Figure~\ref{fig-trip-purposes}.
 
@@
 \href{https://www.cyclestreets.net/help/journey/howitworks/}{website},
-the `Fastest' routes minimize journey time, accounting for traffic
-lights and surface type. The `Quietest' route type minimizes busy
-sections of road, allowing high `diversion factors' from the fastest
-route on quieter but less direct ways (often avoiding roads and
-interactions with motor vehicles altogether where possible). The
-`Balanced' route type is a compromize between the two, minimising
-journey time while avoiding the busiest roads. Allowing users to switch
-between these three route types enables them to consider the trade-offs
-between directness and `cycle friendliness' when planning new
-infrastructure, as shown in Figure~\ref{fig-route-types}.
+the fastest routes minimize journey time, accounting for traffic lights
+and surface type. The `Quietest' route type minimizes busy sections of
+road, allowing high `diversion factors' from the fastest route on
+quieter but less direct ways (often avoiding roads and interactions with
+motor vehicles altogether where possible). The `Balanced' route type is
+a compromize between the two, minimising journey time while avoiding the
+busiest roads. Allowing users to switch between these three route types
+enables them to consider the trade-offs between directness and `cycle
+friendliness' when planning new infrastructure, as shown in
+Figure~\ref{fig-route-types}.
 
@@
 (with a high inferred level of traffic stress), even when efforts are
-taken to avoid busy segments. For the Fastest network, which may be more
+taken to avoid busy segments. For the fastest network, which may be more
 realistic for utility cycling, the proportion of cycling on hostile
@@
 
-\caption{\label{fig-route-types}Illustration of the Quietest (left) and
-Fastest (right) route networks for Naas (top) and Newbridge (bottom).}
+\caption{\label{fig-route-types}Illustration of the quietest (left) and
+fastest (right) route networks for Naas (top) and Newbridge (bottom) in
+County Kildare.}
 
@@
 \href{https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/}{Climate
-Action Plan 2021}, which aims for a 51\% cut in overall greenhouse gas
-emissions by 2030 as part of the pathway to net-zero emissions by 2050.
-For transport, this includes 500,000 extra walking, cycling and public
+Action Plan 2021}, which aims for a 51\% cut in overall GHG emissions by
+2030 as part of the pathway to net-zero emissions by 2050. For
+transport, this includes 500,000 extra walking, cycling and public
 transport trips per day by 2030. In terms of car travel, the target is
@@
 
-To model this decrease in car driver km, cycling uptake increases in
+To model this decrease in car km driven, cycling uptake increases in
 line with the Go Dutch scenario. However, we only model shift from
@@
 hilliness (measured as mean gradient) and route distance. This scenario
-uses the same model that was used PCT, allowing trips to shift from any
-other mode to cycling \citep{lovelace2017}.
+uses the same model as the PCT, allowing trips to shift from any other
+mode to cycling \citep{lovelace2017}.
 
@@
 package `shiny' or Python packages such as `streamlit' or `flask' in the
-background \citep{wickham2021}. Typical intended user stories are
-illustrated in Figure~\ref{fig-user-stories}, which shows that the tool
-is designed to be used by both professional and non-professional users.
-For the main target audience, professional transport planners working at
-the county level, the tool provides a range of outputs, including
-estimates of cycling potential at the county and network level. The
-provision of balanced (the default), quietest and fastest route networks
-enables planners to consider the trade-offs between directness and
-`cycle friendliness' when planning new infrastructure. The tool also
-provides data downloads, enabling estimates of cycling potential on the
-network to be visualised and analysed with other tools such as QGIS,
-Python or R.
+background \citep{wickham2021}. Its landing page is illustrated in
+Figure~\ref{fig-landing}).
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{images/paste-10.png}
+
+}
+
+\caption{\label{fig-landing}Illustration of the landing page of the
+CRUSE Tool for Ireland.}
+
+\end{figure}%
+
+Typical intended user stories are illustrated in
+Figure~\ref{fig-user-stories}, which shows that the tool is designed to
+be used by both professional and non-professional users. For the main
+target audience, professional transport planners working at the county
+level, the tool provides a range of outputs, including estimates of
+cycling potential at the county and network level. The provision of
+balanced (the default), quietest and fastest route networks enables
+planners to consider the trade-offs between directness and `cycle
+friendliness' when planning new infrastructure. The tool also provides
+data downloads, enabling estimates of cycling potential on the network
+to be visualised and analysed with other tools such as QGIS, Python or
+R.
 
@@
 infrastructure and this is apparent from the high proportion of the
-route network that is green, even in the `Fastest' route network
+route network that is green, even in the fastest route network
 representing routes that prioritize speed and directness over comfort
@@
 Although Dublin City has made progress, there are still many parts of
-the Fastest route network, and even some parts of the Balanced and
+the fastest route network, and even some parts of the Balanced and
 Quietest route networks, that are not cycle friendly and which have high
@@
 potential future cycling levels across Ireland down to the street level,
-with potential assigned to `fastest', `balanced' and `quietest' route
+with potential assigned to fastest, balanced and quietest route
 networks. This provision of multiple scenarios of behavior change
@@
 As the benefits of strategic planning for cycling become more apparent
-\citep{scappini2022} we expect the demand for tools such as CRUSE to
+\citep{scappini2022}, we expect the demand for tools such as CRUSE to
 grow. However, the approach is not without limitations and does not meet
@@
 
-A limitation of the approach as implemented in the CRUSE tool is that
+A limitation of the approach as implemented in the CRUSE Tool is that
 the results are `static', reducing its ability to support monitoring of

### 2024-03-25 paper.qmd
Fix up references

@@
 # subtitle: A case study of Ireland
-bibliography:
-  # cruse.bib
-  - bibliography.bib
-  - cruse.bib
-  - references.bib
+bibliography: references.bib
 csl: apa.csl
@@
 A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace_mapping_2016].
-The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 @lovelace2017.
+The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 [@lovelace2017].
 Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modeling of impacts on health, environmental and distributional outcomes [@woodcock2021].
@@
 
-However, travel to school and work only constitute around only 30% of all trips in Ireland. <!-- Reference needed -->
+However, travel to school and work only constitute 39% of all trips in Ireland, excluding 'returning home' trips, according to the 2022 National Household Travel Survey (NHTS) [@national2022].
 To overcome this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
@@
 They emphasized that cycle networks are often fragmented, and the tool allows them to demonstrate how individual local schemes can contribute to a larger, interconnected network.
-Council officers expressed a preference for visualizations that directly compare the fast and quiet routes, as they recognize the importance of both types of routes (users can switch between the on the 'Route types' page for each county although this functionality is not available in the landing page map due to other feedback requesting a simpler interface).
-
+An unexpected piece of feedback from the workshop was the preference for visualizations that directly compare the fast and quiet routes, as they recognize the importance of both types of routes.
+In response to this we provided the 'Route types' page for each county, although this functionality is not available in the landing page map due to other feedback requesting a simpler interface.
 In comparison with another tool (the National Transport Authority's cycling propensity tool), the CRUSE tool was seen by practitioners in Kildare as more user-friendly and intuitive, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
@@
 Declan also noted the alignment between local knowledge, cycle count data, and the outputs of the approach: "It's a good representation of what's going on" and the networks allign well in terms of network "shape" and the proportion of trips of trips on different parts of the network.
-There were two suggestions from Declan which have not yet been implemented: the provision of data showing only "gold standard" POWSCAR data, and the need for training on the tool to ensure that all staff can use it effectively.
-These, and other potential improvements, are discussed in the next section.
+
+Findings from the qualitative results also suggest changes which have not been implemented due to time constraints, including the provision of route netework data showing only "gold standard" POWSCAR data (we did make this data available in the data downloads page), the need for further validation of results (especially in rural areas where counter datasets are scarce) and the need for training on the tool to ensure that it is well-understood and used effectively.
+These and other potential areas for further research and development are discussed in the next section.
 

### 2024-03-25 paper.tex
Fix up references

@@
 and desire line levels nationally, and launched as a publicly available
-web application in 2017 \citet{lovelace2017}. Extensions of the PCT
+web application in 2017 \citep{lovelace2017}. Extensions of the PCT
 approach have included estimation of benefits at the individual level
@@
 
-However, travel to school and work only constitute around only 30\% of
-all trips in Ireland. To overcome this issue, we developed a spatial
-interaction modeling methodology to estimate the number of trips between
-each OD pair for additional trip purposes. The classification of trip
-purposes used in the CRUSE Tool was guided mainly by the trip purpose
-classification found within the National Household Travel Survey (NHTS),
-but with the addition of categories based on the comprehensive POWSCAR
-data, and the need to include recreational trips and multi-stage trips.
-An overview of the trip purposes used in CRUSE is presented in
+However, travel to school and work only constitute 39\% of all trips in
+Ireland, excluding `returning home' trips, according to the 2022
+National Household Travel Survey (NHTS) \citep{national2022}. To
+overcome this issue, we developed a spatial interaction modeling
+methodology to estimate the number of trips between each OD pair for
+additional trip purposes. The classification of trip purposes used in
+the CRUSE Tool was guided mainly by the trip purpose classification
+found within the National Household Travel Survey (NHTS), but with the
+addition of categories based on the comprehensive POWSCAR data, and the
+need to include recreational trips and multi-stage trips. An overview of
+the trip purposes used in CRUSE is presented in
 Figure~\ref{fig-trip-purposes}.
@@
 them to demonstrate how individual local schemes can contribute to a
-larger, interconnected network. Council officers expressed a preference
-for visualizations that directly compare the fast and quiet routes, as
-they recognize the importance of both types of routes (users can switch
-between the on the `Route types' page for each county although this
-functionality is not available in the landing page map due to other
-feedback requesting a simpler interface).
-
-In comparison with another tool (the National Transport Authority's
-cycling propensity tool), the CRUSE tool was seen by practitioners in
-Kildare as more user-friendly and intuitive, with the ability to
-generate evidence at the route level, which is crucial for planning new
-cycling infrastructure. Overall, the workshop participants concluded
-that the approach filled gap in evidence and will greatly assist in the
+larger, interconnected network. An unexpected piece of feedback from the
+workshop was the preference for visualizations that directly compare the
+fast and quiet routes, as they recognize the importance of both types of
+routes. In response to this we provided the `Route types' page for each
+county, although this functionality is not available in the landing page
+map due to other feedback requesting a simpler interface. In comparison
+with another tool (the National Transport Authority's cycling propensity
+tool), the CRUSE tool was seen by practitioners in Kildare as more
+user-friendly and intuitive, with the ability to generate evidence at
+the route level, which is crucial for planning new cycling
+infrastructure. Overall, the workshop participants concluded that the
+approach filled gap in evidence and will greatly assist in the
 development of new cycling projects.
@@
 networks allign well in terms of network ``shape'' and the proportion of
-trips of trips on different parts of the network. There were two
-suggestions from Declan which have not yet been implemented: the
-provision of data showing only ``gold standard'' POWSCAR data, and the
-need for training on the tool to ensure that all staff can use it
-effectively. These, and other potential improvements, are discussed in
+trips of trips on different parts of the network.
+
+Findings from the qualitative results also suggest changes which have
+not been implemented due to time constraints, including the provision of
+route netework data showing only ``gold standard'' POWSCAR data (we did
+make this data available in the data downloads page), the need for
+further validation of results (especially in rural areas where counter
+datasets are scarce) and the need for training on the tool to ensure
+that it is well-understood and used effectively. These and other
+potential areas for further research and development are discussed in
 the next section.
@@
 
-  \bibliography{bibliography.bib,cruse.bib,references.bib}
+  \bibliography{references.bib}
 

### 2024-03-25 paper.qmd
Mention cycle tourism layer

@@
 <!-- Mobility kills. -->
+
 ## Background
@@
 
-However, travel to school and work only constitute 39% of all trips in Ireland, excluding 'returning home' trips, according to the 2022 National Household Travel Survey (NHTS) [@national2022].
-To overcome this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
+However, travel to school and work account for less than half (39%) of all trips in Ireland, excluding 'returning home' trips, according to the 2022 National Household Travel Survey (NHTS) [@national2022].
+This represents a substantial decline in the proportion of trips covered by POWSCAR data compared with pre-COVID survey results, with 51% of non-return trips being for work or education according to the 2017 release of the same report [@national2017].
+To tackle this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
 The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips.
-<!-- This is already implemented, right?-->
-An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes. <!--This should be Table 2--->
+<!-- This is already implemented, right?--> An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes.
+<!--This should be Table 2--->
 
@@
 
-Trip purposes used in the CRUSE Tool.
+Everyday trip purposes used in the CRUSE Tool.
 :::
@@
 ```
+Following feedback from stakeholders, we added another, 'non-everyday' trip purpose in an extension phase: cycle tourism.
+As outlined in the CRUSE Extension report [reference and link to be added on publication], this involved developing an inter-county spatial interaction model to estimate the number of trips between each county for cycle tourism, with trip attractors including campsites and major international transport hubs.
+
 ## Routing
@@
 <!-- The results for the other -->
+
 ## Selected qualitative results {#sec-qualitative}

### 2024-03-25 paper.tex
Mention cycle tourism layer

@@
 
-However, travel to school and work only constitute 39\% of all trips in
-Ireland, excluding `returning home' trips, according to the 2022
-National Household Travel Survey (NHTS) \citep{national2022}. To
-overcome this issue, we developed a spatial interaction modeling
-methodology to estimate the number of trips between each OD pair for
-additional trip purposes. The classification of trip purposes used in
-the CRUSE Tool was guided mainly by the trip purpose classification
-found within the National Household Travel Survey (NHTS), but with the
-addition of categories based on the comprehensive POWSCAR data, and the
-need to include recreational trips and multi-stage trips. An overview of
-the trip purposes used in CRUSE is presented in
-Figure~\ref{fig-trip-purposes}.
+However, travel to school and work account for less than half (39\%) of
+all trips in Ireland, excluding `returning home' trips, according to the
+2022 National Household Travel Survey (NHTS) \citep{national2022}. This
+represents a substantial decline in the proportion of trips covered by
+POWSCAR data compared with pre-COVID survey results, with 51\% of
+non-return trips being for work or education according to the 2017
+release of the same report \citep{national2017}. To tackle this issue,
+we developed a spatial interaction modeling methodology to estimate the
+number of trips between each OD pair for additional trip purposes. The
+classification of trip purposes used in the CRUSE Tool was guided mainly
+by the trip purpose classification found within the National Household
+Travel Survey (NHTS), but with the addition of categories based on the
+comprehensive POWSCAR data, and the need to include recreational trips
+and multi-stage trips. An overview of the trip purposes used in CRUSE is
+presented in Figure~\ref{fig-trip-purposes}.
 

### 2024-03-26 paper.qmd
Add limitations section

@@
 
-## Selected qualitative results {#sec-qualitative}
+## Stakeholder engagement and feedback {#sec-qualitative}
 
@@
 
-A limitation of the approach as implemented in the CRUSE Tool is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
+## Limitations and future research
+
+The approach has a number of limitations that should be understood by practitioners, researchers, advocates, policy makers using the tool.
+As highlighted in @sec-qualitative, we were only able to compare the results with a small number of cycle counters and based on feedback from workshops.
+As outlined in the CRUSE extension report [reference to be added on publication], we also undertook validation against Strava data.
+However, more data is needed to better understand the accuracy of the baseline results, especially for the 'non-POWSCAR' purposes.
+We hope this becomes possible as Ireland's cycle counter network expands, allowing refinements to the approach (espcially the spatial interaction model and parameters) to be made.
+Changes that could help mitigate this limitation on the user interface side include providing results with confidence intervals or as categories of cycling potential, rather than as central estimates, and the visualistion of the 'POWSCAR' network (which is based on Census data) in the web application.
+
+A limitation associated with the approach is its reliance on OpenStreetMap (OSM) data.
+The 'cycle friendliness' of a route is based on the OSM tags and the routes themselves are generated by the CycleStreets routing engine, which is based on OSM data.
+During a conference in Sligo where we presented prelimary results, a stakeholder with local knowledge pointed out that a quiet route along the canal was not being followed.
+In response, manual edits were made to the OSM network, and the results currently on the website reflect these changes.
+However, there are many other potential errors in the OSM data that have not been corrected.
+Future work could mitigate this by integrating the approach more closely with OSM (for example by providing links to enable one-click edits of OSM when users click on a route segment) or by setting up a workshop to 'crowdsource' corrections to the OSM data.
+
+A limitation of the approach is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
 Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.

### 2024-03-26 paper.qmd
More limitations

@@
 Future work could mitigate this by integrating the approach more closely with OSM (for example by providing links to enable one-click edits of OSM when users click on a route segment) or by setting up a workshop to 'crowdsource' corrections to the OSM data.
+Furthermore, because there are multiple possible users of the tool with different starging points, workshops and training events could be valuable if not essential to ensure that the tool is used effectively.
+Such events could bring together stakeholders from advocacy groups, local authorities, and national governments.
+The siloed nature of transport planning means that such diverse stakeholders are rarely the same room, meaning that a 'wide boundary' outcome of the tool could be to bring these groups together (if workshops based on the tool are successful).
 
-A limitation of the approach is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
+Another limitation of the approach is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
 Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.

### 2024-03-26 paper.qmd
Update with more on landing page

@@
   This paper describes an approach for developing strategic cycle network planning tools.
-Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenario generation, and development of an intuitive user interface.
+Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenario generation, and development of an intuitive user interface that has been tested and used by practitioners.
 Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions.
@@
 Following feedback from stakeholders, we added another, 'non-everyday' trip purpose in an extension phase: cycle tourism.
-As outlined in the CRUSE Extension report [reference and link to be added on publication], this involved developing an inter-county spatial interaction model to estimate the number of trips between each county for cycle tourism, with trip attractors including campsites and major international transport hubs.
+As outlined in the CRUSE Extension report \[reference and link to be added on publication\], this involved developing an inter-county spatial interaction model to estimate the number of trips between each county for cycle tourism, with trip attractors including campsites and major international transport hubs.
 
@@
 
-The CRUSE web application is statically hosted, meaning that it can be accessed without the need for a server running software such as the R package 'shiny' or Python packages such as 'streamlit' or 'flask' in the background [@wickham2021].
-Its landing page is illustrated in @fig-landing).
+The CRUSE web application is statically hosted, meaning it does not require a server running software such as the R package 'shiny' or Python packages such as 'streamlit' or 'flask' in the background [@wickham2021].
+This reduces hosting costs and the maintenance burden of the tool compared with tools such as the PCT, and makes the map-based interface more responsive, using the recently developed 'MapLibre' JavaScript package and 'PMTiles' "cloud native" vector tile format [@gonçalves2023].
+The default view in landing page is of a map of Ireland with county level results for the % trips that could be made by cycling under the 'Go Dutch' scenario, highlighting the intended use case for strategic cycle network planning and level of ambition.
+The key results, route network results, are available within an instant by zooming in on the map illustrated in @fig-landing: beyond a certain zoom level the network view appears.
+This differs from the user interface in tools such as the PCT and BiclaR which force users to click on a county before the key network results become visible: to get to the network view in the PCT takes around 10 seconds and multiple clicks, compared with simply zooming in on the landing page for CRUSE.
+
+Other improvements to the user interface in the national map on the landing page include a search bar in the top left and clearer legend for each network visualisation option (users can select any scenario or quietness or gradient depending on their needs).
+Following feedback from user testing, a 'geolocate' and 'full screen' button were added in the top right, enabling users to 'zoom' to their current location and to focus on the detailed geographic results, as shown in @fig-landing.
 
@@
 graph LR
-    A(Professional user) --> B[CRUSE landing page]
-    B --> C[CRUSE national map]
-    C --> D[CRUSE county map]
+    A(Professional user) --> B[Landing page]
+    B --> C[National map]
+    C --> D[County map]
     D --> F[Full county map]
@@
     DD --> EX{External tools}
-    X(Non-professional user) --> Y[CRUSE landing page]
-    Y --> Z[CRUSE national map]
-    Z --> R[CRUSE county level results]
+    X(Non-professional user) --> Y[Landing page]
+    Y --> Z[National map]
+    Z --> R[County level results]
     Z --> E(Exploration)
@@
 In response to this we provided the 'Route types' page for each county, although this functionality is not available in the landing page map due to other feedback requesting a simpler interface.
-In comparison with another tool (the National Transport Authority's cycling propensity tool), the CRUSE tool was seen by practitioners in Kildare as more user-friendly and intuitive, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
+In comparison with another tool (the National Transport Authority's cycling propensity tool), the CRUSE tool was seen by practitioners in Kildare as more user-friendly, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
 Overall, the workshop participants concluded that the approach filled gap in evidence and will greatly assist in the development of new cycling projects.
@@
 As highlighted in @sec-qualitative, we were only able to compare the results with a small number of cycle counters and based on feedback from workshops.
-As outlined in the CRUSE extension report [reference to be added on publication], we also undertook validation against Strava data.
+As outlined in the CRUSE extension report \[reference to be added on publication\], we also undertook validation against Strava data.
 However, more data is needed to better understand the accuracy of the baseline results, especially for the 'non-POWSCAR' purposes.

### 2024-03-26 paper.tex
Update with more on landing page

@@
 routing, scenario generation, and development of an intuitive user
-interface. Commissioned and used by Transport Infrastructure Ireland,
-CRUSE provides estimates of current and potential future cycling levels
-under `snapshot' scenarios to inform investment decisions. The publicly
-available results at https://cruse.bike/ enable planners, engineers, and
-other stakeholders to make more evidence-based decisions. CRUSE goes
-beyond previous work by: modeling networks at high spatial resolution;
+interface that has been tested and used by practitioners. Commissioned
+and used by Transport Infrastructure Ireland, CRUSE provides estimates
+of current and potential future cycling levels under `snapshot'
+scenarios to inform investment decisions. The publicly available results
+at https://cruse.bike/ enable planners, engineers, and other
+stakeholders to make more evidence-based decisions. CRUSE goes beyond
+previous work by: modeling networks at high spatial resolution;
 simulating multiple trip purposes (social, shopping, personal utility,
@@
 
-\caption{\label{fig-trip-purposes}Trip purposes used in the CRUSE Tool.}
+\caption{\label{fig-trip-purposes}Everyday trip purposes used in the
+CRUSE Tool.}
 
 \end{figure}%
+
+Following feedback from stakeholders, we added another, `non-everyday'
+trip purpose in an extension phase: cycle tourism. As outlined in the
+CRUSE Extension report {[}reference and link to be added on
+publication{]}, this involved developing an inter-county spatial
+interaction model to estimate the number of trips between each county
+for cycle tourism, with trip attractors including campsites and major
+international transport hubs.
 
@@
 
-The CRUSE web application is statically hosted, meaning that it can be
-accessed without the need for a server running software such as the R
-package `shiny' or Python packages such as `streamlit' or `flask' in the
-background \citep{wickham2021}. Its landing page is illustrated in
-Figure~\ref{fig-landing}).
+The CRUSE web application is statically hosted, meaning it does not
+require a server running software such as the R package `shiny' or
+Python packages such as `streamlit' or `flask' in the background
+\citep{wickham2021}. This reduces hosting costs and the maintenance
+burden of the tool compared with tools such as the PCT, and makes the
+map-based interface more responsive, using the recently developed
+`MapLibre' JavaScript package and `PMTiles' ``cloud native'' vector tile
+format \citep{gonçalves2023}. The default view in landing page is of a
+map of Ireland with county level results for the \% trips that could be
+made by cycling under the `Go Dutch' scenario, highlighting the intended
+use case for strategic cycle network planning and level of ambition. The
+key results, route network results, are available within an instant by
+zooming in on the map illustrated in Figure~\ref{fig-landing}: beyond a
+certain zoom level the network view appears. This differs from the user
+interface in tools such as the PCT and BiclaR which force users to click
+on a county before the key network results become visible: to get to the
+network view in the PCT takes around 10 seconds and multiple clicks,
+compared with simply zooming in on the landing page for CRUSE.
+
+Other improvements to the user interface in the national map on the
+landing page include a search bar in the top left and clearer legend for
+each network visualisation option (users can select any scenario or
+quietness or gradient depending on their needs). Following feedback from
+user testing, a `geolocate' and `full screen' button were added in the
+top right, enabling users to `zoom' to their current location and to
+focus on the detailed geographic results, as shown in
+Figure~\ref{fig-landing}.
 
@@
 
-\subsection{Selected qualitative results}\label{sec-qualitative}
+\subsection{Stakeholder engagement and feedback}\label{sec-qualitative}
 
@@
 tool), the CRUSE tool was seen by practitioners in Kildare as more
-user-friendly and intuitive, with the ability to generate evidence at
-the route level, which is crucial for planning new cycling
-infrastructure. Overall, the workshop participants concluded that the
-approach filled gap in evidence and will greatly assist in the
-development of new cycling projects.
+user-friendly, with the ability to generate evidence at the route level,
+which is crucial for planning new cycling infrastructure. Overall, the
+workshop participants concluded that the approach filled gap in evidence
+and will greatly assist in the development of new cycling projects.
 
@@
 
-A limitation of the approach as implemented in the CRUSE Tool is that
-the results are `static', reducing its ability to support monitoring of
-growth in cycling, upgrades of existing cycling infrastructure and
-dynamic (year-by-year) exposure information to estimate crash rates.
-Exposure information is a key part of the the European RISM Directive,
-for reporting collision rates of vulnerable road users by 2024. These
-factors mean that the tool can be seen as an open access `leverage
-point', providing key information for and supporting ambitious plans in
-many aspects the planning system, from network design to post-build
-monitoring \citep{lovelace2020}.
+\subsection{Limitations and future
+research}\label{limitations-and-future-research}
+
+The approach has a number of limitations that should be understood by
+practitioners, researchers, advocates, policy makers using the tool. As
+highlighted in Section~\ref{sec-qualitative}, we were only able to
+compare the results with a small number of cycle counters and based on
+feedback from workshops. As outlined in the CRUSE extension report
+{[}reference to be added on publication{]}, we also undertook validation
+against Strava data. However, more data is needed to better understand
+the accuracy of the baseline results, especially for the `non-POWSCAR'
+purposes. We hope this becomes possible as Ireland's cycle counter
+network expands, allowing refinements to the approach (espcially the
+spatial interaction model and parameters) to be made. Changes that could
+help mitigate this limitation on the user interface side include
+providing results with confidence intervals or as categories of cycling
+potential, rather than as central estimates, and the visualistion of the
+`POWSCAR' network (which is based on Census data) in the web
+application.
+
+A limitation associated with the approach is its reliance on
+OpenStreetMap (OSM) data. The `cycle friendliness' of a route is based
+on the OSM tags and the routes themselves are generated by the
+CycleStreets routing engine, which is based on OSM data. During a
+conference in Sligo where we presented prelimary results, a stakeholder
+with local knowledge pointed out that a quiet route along the canal was
+not being followed. In response, manual edits were made to the OSM
+network, and the results currently on the website reflect these changes.
+However, there are many other potential errors in the OSM data that have
+not been corrected. Future work could mitigate this by integrating the
+approach more closely with OSM (for example by providing links to enable
+one-click edits of OSM when users click on a route segment) or by
+setting up a workshop to `crowdsource' corrections to the OSM data.
+Furthermore, because there are multiple possible users of the tool with
+different starging points, workshops and training events could be
+valuable if not essential to ensure that the tool is used effectively.
+Such events could bring together stakeholders from advocacy groups,
+local authorities, and national governments. The siloed nature of
+transport planning means that such diverse stakeholders are rarely the
+same room, meaning that a `wide boundary' outcome of the tool could be
+to bring these groups together (if workshops based on the tool are
+successful).
+
+Another limitation of the approach is that the results are `static',
+reducing its ability to support monitoring of growth in cycling,
+upgrades of existing cycling infrastructure and dynamic (year-by-year)
+exposure information to estimate crash rates. Exposure information is a
+key part of the the European RISM Directive, for reporting collision
+rates of vulnerable road users by 2024. These factors mean that the tool
+can be seen as an open access `leverage point', providing key
+information for and supporting ambitious plans in many aspects the
+planning system, from network design to post-build monitoring
+\citep{lovelace2020}.
 

### 2024-03-26 paper.qmd
Routing descriptions

@@
 
-Routes in CRUSE are generated by CycleStreets, a not-for profit transport consultancy and web development company who provide application programming interfaces (APIs) supplying a range of datasets for cycle planning and advocacy.
-Although based in Cambridge, UK, CycleStreets provide routing services 'by cyclists, for cyclists' internationally, including in Ireland.
+Routes in CRUSE are generated by CycleStreets, a not-for profit transport consultancy and web development company who provide application programming interfaces (APIs) supplying a range of datasets for cycle planning and advocacy internationally, including in Ireland.
 The CycleStreets routing engine is based on OpenStreetMap (OSM) data, which is continuously updated by a global community of volunteers.
@@
 
-We calculated three network types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest'.
+The CycleStreets routing engine accounts for hilliness and traffic signals, based on substantial experience generating cycle routes for utility and leisure cyclists worldwide.
+The resulting routes are designed to indicate the route choices of confident, moderately confident, and less confident cyclists with three route options provided for each OD pair as described below.
+It would be possible to use other routing engines and custom profiles and cost functions in the future, for example to simulate the impact of implementing traffic signals which are optimised for cycling throughput, a feature not currently available in CycleStreets that could be explored in future work.
+
+We computed three route types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest'.
 As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the fastest routes minimize journey time, accounting for traffic lights and surface type.
@@
 Allowing users to switch between these three route types enables them to consider the trade-offs between directness and 'cycle friendliness' when planning new infrastructure, as shown in @fig-route-types.
-
-These map are available on the 'Route types' page for each county (see [cruse.bike/kildare/route-types](https://cruse.bike/kildare/route-types)).
+These different network maps are available on the 'Route types' page for each county (see [cruse.bike/kildare/route-types](https://cruse.bike/kildare/route-types) for Kildare, for example).
 In addition to showing the estimated routes under each scenario, the page also presents summary statistics on the network:

### 2024-03-26 paper.qmd
add tbl-trip_purposes

--- 2024-03-26 paper.qmd
@@
 # subtitle: A case study of Ireland
-bibliography: references.bib
+bibliography:
+  # cruse.bib
+  - bibliography.bib
+  - cruse.bib
+  - references.bib
 csl: apa.csl
@@
 
-## Background
-
 Transport systems dominated by heavy and powerful private cars are inefficient, dangerous, and unhealthy.
-Cars are a major and growing source of greenhouse gas (GHG) emissions [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018], and a cause of disease due to physical inactivity and air, noise and microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024].
-<!-- Shouldn't these three sources share the brackets?
---> Transport is responsible for 23% of GHG emissions, 70% of which are from road transport, with passenger cars accounting for nearly half of transport emissions (around 10% of global emissions) [@jaramillo2022].
+Cars are a major and growing source of greenhouse gas emissions causing climate change [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018] , and a cause of disease due to physical inactivity and air, noise and (from tyres) microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024].
+Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport, with passenger cars accounting for nearly half of transport emissions (around 10% of global emissions) [@jaramillo2022].
 The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
@@
 Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
-Seven in ten trips in Ireland are by car [@national].
+7 in 10 trips in Ireland are by car [@national] .
 Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as ebikes extend the range of this mode" [@greater].
@@
 
-## Aim and content
+::: {#fig-landing}
+![](images/paste-10.png)
+
+Illustration of the landing page of the CRUSE Tool for Ireland.
+:::
 
 This paper describes an approach to strategic cycle network planning tool development that is open, scalable, and evidence-based.
-We label the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and present a case study of its application in Ireland, the results of which are publicly available at [cruse.bike](https://cruse.bike/).
-
-In @sec-tools, we review existing current tools for estimating cycling potential.
+We label the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and present a case study of its application in Ireland, the results of which are publicly available at [cruse.bike](https://cruse.bike/) in the CRUSE Tool for Ireland, the landing page of which is illustrated in @fig-landing.
 In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool.
@@
 
-# Tools for estimating cycling potential {#sec-tools}
+# Tools for estimating cycling potential
 
@@
 A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace_mapping_2016].
-The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 [@lovelace2017].
+The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 @lovelace2017.
 Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modeling of impacts on health, environmental and distributional outcomes [@woodcock2021].
@@
 This method works fine when the OD data represents movement between small areas, but was not appropriate for generating route networks from the POWSCAR data because zone centroids are so far apart that the resulting route networks would be sparse and unrealistic.
-To tackle this issue we developed a new method for OD data disaggregation called 'jittering'.
-<!--This reference should be removed to anonymise the paper.
-They will guess that Lovelace is involved, but saying we developed and then citing is as if we did not remove authors names from the paper :) --> The method works by first disaggregating the OD data based on a 'disaggregation threshold' and then assigning each disaggregated 'sub-OD' pair to 'subpoints' within each zone.
+To tackle this issue we developed a new method for OD data disaggregation called 'jittering' [@lovelace2022].
+The method works by first disaggregating the OD data based on a 'disaggregation threshold' and then assigning each disaggregated 'sub-OD' pair to 'subpoints' within each zone.
 As shown in @fig-dublin, the resulting route networks are dense, even in rural areas.
@@
 
-However, travel to school and work account for less than half (39%) of all trips in Ireland, excluding 'returning home' trips, according to the 2022 National Household Travel Survey (NHTS) [@national2022].
-This represents a substantial decline in the proportion of trips covered by POWSCAR data compared with pre-COVID survey results, with 51% of non-return trips being for work or education according to the 2017 release of the same report [@national2017].
-To tackle this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
-The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips.
-<!-- This is already implemented, right?--> An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes.
-<!--This should be Table 2--->
+However, travel to school and work only constitute around only 30% of all trips in Ireland.
+To overcome this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
+The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
+An overview of the trip purposes used in CRUSE is presented in @tbl-trip_purposes.
 
-::: {#fig-trip-purposes}
-![](images/paste-4.png)
+```{r}
+#| include: false
+trip_purposes = read_csv("trip_purposes.csv") %>% 
+  mutate_if(is.character, utf8::utf8_encode)
+```
 
-Everyday trip purposes used in the CRUSE Tool.
-:::
+```{r}
+#| label: tbl-trip_purposes
+# For PDF:
+knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.", booktabs = TRUE) |>
+  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
+  kableExtra::column_spec(1, width = "13em") |>
+  kableExtra::column_spec(2, width = "15em") |>
+  kableExtra::column_spec(3, width = "12em") |>
+  kableExtra::column_spec(4, width = "5em")
+# For HTML:
+# knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.")
+```
 
@@
 ```
-Following feedback from stakeholders, we added another, 'non-everyday' trip purpose in an extension phase: cycle tourism.
-As outlined in the CRUSE Extension report [reference and link to be added on publication], this involved developing an inter-county spatial interaction model to estimate the number of trips between each county for cycle tourism, with trip attractors including campsites and major international transport hubs.
-
 ## Routing
@@
 
-The Climate Action Plan scenario models the transport emissions reductions targeted in the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which aims for a 51% cut in overall GHG emissions by 2030 as part of the pathway to net-zero emissions by 2050.
+The Climate Action Plan scenario models the transport emissions reductions targeted in the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which aims for a 51% cut in overall greenhouse gas emissions by 2030 as part of the pathway to net-zero emissions by 2050.
 For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.
@@
 <!-- The results for the other -->
-
-## Stakeholder engagement and feedback {#sec-qualitative}
+## Selected qualitative results {#sec-qualitative}
 
@@
 They emphasized that cycle networks are often fragmented, and the tool allows them to demonstrate how individual local schemes can contribute to a larger, interconnected network.
-An unexpected piece of feedback from the workshop was the preference for visualizations that directly compare the fast and quiet routes, as they recognize the importance of both types of routes.
-In response to this we provided the 'Route types' page for each county, although this functionality is not available in the landing page map due to other feedback requesting a simpler interface.
+Council officers expressed a preference for visualizations that directly compare the fast and quiet routes, as they recognize the importance of both types of routes (users can switch between the on the 'Route types' page for each county although this functionality is not available in the landing page map due to other feedback requesting a simpler interface).
+
 In comparison with another tool (the National Transport Authority's cycling propensity tool), the CRUSE tool was seen by practitioners in Kildare as more user-friendly and intuitive, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
@@
 Declan also noted the alignment between local knowledge, cycle count data, and the outputs of the approach: "It's a good representation of what's going on" and the networks allign well in terms of network "shape" and the proportion of trips of trips on different parts of the network.
-
-Findings from the qualitative results also suggest changes which have not been implemented due to time constraints, including the provision of route netework data showing only "gold standard" POWSCAR data (we did make this data available in the data downloads page), the need for further validation of results (especially in rural areas where counter datasets are scarce) and the need for training on the tool to ensure that it is well-understood and used effectively.
-These and other potential areas for further research and development are discussed in the next section.
+There were two suggestions from Declan which have not yet been implemented: the provision of data showing only "gold standard" POWSCAR data, and the need for training on the tool to ensure that all staff can use it effectively.
+These, and other potential improvements, are discussed in the next section.
 
@@
 The aim of the paper was to describe the design, features and potential use of the 'CRUSE' approach and resulting tool.
-As outlined in the previous section, the CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to fastest, balanced and quietest route networks.
+As outlined in the previous section, the CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
 This provision of multiple scenarios of behavior change *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.
@@
 
-As the benefits of strategic planning for cycling become more apparent [@scappini2022], we expect the demand for tools such as CRUSE to grow.
+As the benefits of strategic planning for cycling become more apparent [@scappini2022] we expect the demand for tools such as CRUSE to grow.
 However, the approach is not without limitations and does not meet all requirements in Ireland or indeed any country where the methods are applied.
 
-## Limitations and future research
-
-The approach has a number of limitations that should be understood by practitioners, researchers, advocates, policy makers using the tool.
-As highlighted in @sec-qualitative, we were only able to compare the results with a small number of cycle counters and based on feedback from workshops.
-As outlined in the CRUSE extension report [reference to be added on publication], we also undertook validation against Strava data.
-However, more data is needed to better understand the accuracy of the baseline results, especially for the 'non-POWSCAR' purposes.
-We hope this becomes possible as Ireland's cycle counter network expands, allowing refinements to the approach (espcially the spatial interaction model and parameters) to be made.
-Changes that could help mitigate this limitation on the user interface side include providing results with confidence intervals or as categories of cycling potential, rather than as central estimates, and the visualistion of the 'POWSCAR' network (which is based on Census data) in the web application.
-
-A limitation associated with the approach is its reliance on OpenStreetMap (OSM) data.
-The 'cycle friendliness' of a route is based on the OSM tags and the routes themselves are generated by the CycleStreets routing engine, which is based on OSM data.
-During a conference in Sligo where we presented prelimary results, a stakeholder with local knowledge pointed out that a quiet route along the canal was not being followed.
-In response, manual edits were made to the OSM network, and the results currently on the website reflect these changes.
-However, there are many other potential errors in the OSM data that have not been corrected.
-Future work could mitigate this by integrating the approach more closely with OSM (for example by providing links to enable one-click edits of OSM when users click on a route segment) or by setting up a workshop to 'crowdsource' corrections to the OSM data.
-Furthermore, because there are multiple possible users of the tool with different starging points, workshops and training events could be valuable if not essential to ensure that the tool is used effectively.
-Such events could bring together stakeholders from advocacy groups, local authorities, and national governments.
-The siloed nature of transport planning means that such diverse stakeholders are rarely the same room, meaning that a 'wide boundary' outcome of the tool could be to bring these groups together (if workshops based on the tool are successful).
-
-Another limitation of the approach is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
+A limitation of the approach as implemented in the CRUSE tool is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
 Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.

### 2024-03-26 paper.qmd
improved fig 5

@@
 
+```{r alternative}
+# https://www.sesync.org/resources/creating-visualizations-diagrammer
+schematic = DiagrammeR::grViz("
+digraph {
+  # graph statement
+  #################
+  graph [rankdir = LR
+         color = white
+         fontsize = 36]
+
+  # nodes (rectangles)
+  #################
+   node [shape = rectangle            
+       margin = 0.4
+       fontsize = 36
+       fixedsize = false
+       width = 1]
+
+   subgraph cluster_2 {
+     graph[shape = rectangle]
+         C2b [label = 'Exploration']
+         C2a [label = 'CRUSE\ncounty level results']
+         C2 [label = 'CRUSE\nnational map']
+         B2 [label = 'CRUSE\nlanding page']
+         A2 [label = 'Non professional\nuser'] 
+}
+
+   subgraph cluster_1 {
+   graph[shape = rectangle]
+         Dd [label = 'Data dowloads']
+         Dc [label = 'Quiet and fast route\nnetworks']
+         Db [label = 'County statistics']
+         Da [label = 'Full county map']
+         D [label = 'CRUSE\ncounty map']
+         C [label = 'CRUSE\nnational map']
+         B [label = 'CRUSE\nlanding page']
+         A [label = 'Professional user'] 
+}
+
+  # node (triangle)
+  #################
+  node [shape = diamond]
+         Dd1 [label = 'External\ntools']
+         C2b1 [label = 'Communication']
+  # edges 
+  #######
+  A2 -> B2
+  A -> B
+  B -> C
+  B2 -> C2
+  C -> D
+  C2 -> C2a
+  C2 -> C2b
+  C2b -> C2b1
+  D -> Da 
+  D -> Db 
+  D -> Dc 
+  D -> Dd 
+  Dd -> Dd1
+
+}
+")
+# schematic
+# htmlwidgets::saveWidget(schematic, "images/fig5.html")
+# browseURL("images/fig5.html")
+# webshot2::webshot(url = "images/fig5.html", file = "images/fig5.png")
+knitr::include_graphics("images/fig5.png")
+```
+
 ::: {#fig-user-stories}
-![](images/mermaid-figure-1.png)
+![](images/fig5.png)
 

### 2024-03-26 paper.qmd
correct error detecteds in improved figure

@@
          B2 [label = 'CRUSE\nlanding page']
-         A2 [label = 'Non professional\nuser'] 
+         A2 [label = 'Non-professional\nuser'] 
 }
@@
   C2 -> C2b
+  C2a -> C2b
   C2b -> C2b1

### 2024-03-26 paper.qmd
add edits from Kildare

@@
 
-Thoughout the project, we undertook a series of workshops and interviews with stakeholders in Ireland to gather feedback on the approach and resulting tool.
+Throughout the project, we undertook a series of workshops and interviews with stakeholders in Ireland to gather feedback on the approach and resulting tool.
 The feedback was generally positive, with stakeholders commenting on the need for the tool, lack of data on cycling potential proving a barrier to decision-making and investment, and the potential for the tool.
@@
 
-Early in the project, we presented the approach to practitioners in Kildare and Limmerick.
+Early in the project, we presented the approach to practitioners in Kildare and Limerick.
 The feedback from the workshop in Kildare highlighted the need for the tool for practitioners at the county level, as highlighted in the following quote:
 
-> "It's a missing piece of evidence that will help new projects get built" (Senior Engineer, Sustainable Transport & Traffic Management, Kildare County Council)
+> "It's a missing piece of evidence that will help new projects get built" (Dónal Hodgins, Senior Engineer, Sustainable Transport & Traffic Management, Kildare County Council)
 
-Practitioners in Kildare County Council that the evidence generated *was already valuable* the development of new cycle routes, especially when making the case during planning applications.
+Practitioners in Kildare County Council found that the evidence generated *was already valuable* the development of new cycle routes, especially when making the case during planning applications.
 They emphasized that cycle networks are often fragmented, and the tool allows them to demonstrate how individual local schemes can contribute to a larger, interconnected network.
@@
 
-In comparison with another tool (the National Transport Authority's cycling propensity tool), the CRUSE tool was seen by practitioners in Kildare as more user-friendly and intuitive, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
-Overall, the workshop participants concluded that the approach filled gap in evidence and will greatly assist in the development of new cycling projects.
+In comparison with another cycling propensity tool, the CRUSE Tool was seen by practitioners in Kildare as more user-friendly and intuitive, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
+Overall, the workshop participants concluded that the approach filled a gap in evidence and will greatly assist in the development of new cycling projects.
 
@@
 The outputs were a "great tool for getting started with project outline documents" and "ideal for generating a strategic rationale for estimating demand and building the business case in the absence of cycle count data" (Declan Keenan, Senior Transport Planner working on active travel, personal communication).
-Declan also noted the alignment between local knowledge, cycle count data, and the outputs of the approach: "It's a good representation of what's going on" and the networks allign well in terms of network "shape" and the proportion of trips of trips on different parts of the network.
+Declan also noted the alignment between local knowledge, cycle count data, and the outputs of the approach: "It's a good representation of what's going on" and the networks align well in terms of network "shape" and the proportion of trips of trips on different parts of the network.
 There were two suggestions from Declan which have not yet been implemented: the provision of data showing only "gold standard" POWSCAR data, and the need for training on the tool to ensure that all staff can use it effectively.

### 2024-03-26 paper.tex

@@
 application programming interfaces (APIs) supplying a range of datasets
-for cycle planning and advocacy. Although based in Cambridge, UK,
-CycleStreets provide routing services `by cyclists, for cyclists'
-internationally, including in Ireland. The CycleStreets routing engine
-is based on OpenStreetMap (OSM) data, which is continuously updated by a
-global community of volunteers.
+for cycle planning and advocacy internationally, including in Ireland.
+The CycleStreets routing engine is based on OpenStreetMap (OSM) data,
+which is continuously updated by a global community of volunteers.
 
@@
 
-We calculated three network types for each disaggregated (`jittered') OD
+The CycleStreets routing engine accounts for hilliness and traffic
+signals, based on substantial experience generating cycle routes for
+utility and leisure cyclists worldwide. The resulting routes are
+designed to indicate the route choices of confident, moderately
+confident, and less confident cyclists with three route options provided
+for each OD pair as described below. It would be possible to use other
+routing engines and custom profiles and cost functions in the future,
+for example to simulate the impact of implementing traffic signals which
+are optimised for cycling throughput, a feature not currently available
+in CycleStreets that could be explored in future work.
+
+We computed three route types for each disaggregated (`jittered') OD
 pair: `Fastest', `Balanced' and `Quietest'. As outlined on the
@@
 friendliness' when planning new infrastructure, as shown in
-Figure~\ref{fig-route-types}.
-
-These map are available on the `Route types' page for each county (see
-\href{https://cruse.bike/kildare/route-types}{cruse.bike/kildare/route-types}).
-In addition to showing the estimated routes under each scenario, the
-page also presents summary statistics on the network:
+Figure~\ref{fig-route-types}. These different network maps are available
+on the `Route types' page for each county (see
+\href{https://cruse.bike/kildare/route-types}{cruse.bike/kildare/route-types}
+for Kildare, for example). In addition to showing the estimated routes
+under each scenario, the page also presents summary statistics on the
+network:
 

### 2024-03-26 paper.qmd
add tab2 (#37)
* add tab2

* add tbl-trip_purposes

* improved fig 5

* add new fig5 image

* correct error detecteds in improved figure

* add edits from Kildare

---------

Co-authored-by: Robin Lovelace <Robinlovelace@users.noreply.github.com>

--- 2024-03-26 paper.qmd
@@
 
+## Background
+
 Transport systems dominated by heavy and powerful private cars are inefficient, dangerous, and unhealthy.
-Cars are a major and growing source of greenhouse gas emissions causing climate change [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018] , and a cause of disease due to physical inactivity and air, noise and (from tyres) microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024].
-Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport, with passenger cars accounting for nearly half of transport emissions (around 10% of global emissions) [@jaramillo2022].
+Cars are a major and growing source of greenhouse gas (GHG) emissions [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018], and a cause of disease due to physical inactivity and air, noise and microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024].
+<!-- Shouldn't these three sources share the brackets?
+--> Transport is responsible for 23% of GHG emissions, 70% of which are from road transport, with passenger cars accounting for nearly half of transport emissions (around 10% of global emissions) [@jaramillo2022].
 The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
@@
 
-However, travel to school and work only constitute around only 30% of all trips in Ireland.
-To overcome this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
-The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
+However, travel to school and work account for less than half (39%) of all trips in Ireland, excluding 'returning home' trips, according to the 2022 National Household Travel Survey (NHTS) [@national2022].
+This represents a substantial decline in the proportion of trips covered by POWSCAR data compared with pre-COVID survey results, with 51% of non-return trips being for work or education according to the 2017 release of the same report [@national2017].
+To tackle this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
+The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips.
 An overview of the trip purposes used in CRUSE is presented in @tbl-trip_purposes.
@@
 ```
+Following feedback from stakeholders, we added another, 'non-everyday' trip purpose in an extension phase: cycle tourism.
+As outlined in the CRUSE Extension report \[reference and link to be added on publication\], this involved developing an inter-county spatial interaction model to estimate the number of trips between each county for cycle tourism, with trip attractors including campsites and major international transport hubs.
+
 ## Routing
 
-Routes in CRUSE are generated by CycleStreets, a not-for profit transport consultancy and web development company who provide application programming interfaces (APIs) supplying a range of datasets for cycle planning and advocacy.
-Although based in Cambridge, UK, CycleStreets provide routing services 'by cyclists, for cyclists' internationally, including in Ireland.
+Routes in CRUSE are generated by CycleStreets, a not-for profit transport consultancy and web development company who provide application programming interfaces (APIs) supplying a range of datasets for cycle planning and advocacy internationally, including in Ireland.
 The CycleStreets routing engine is based on OpenStreetMap (OSM) data, which is continuously updated by a global community of volunteers.
@@
 
-We calculated three network types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest'.
-As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the 'Fastest' routes minimize journey time, accounting for traffic lights and surface type.
+The CycleStreets routing engine accounts for hilliness and traffic signals, based on substantial experience generating cycle routes for utility and leisure cyclists worldwide.
+The resulting routes are designed to indicate the route choices of confident, moderately confident, and less confident cyclists with three route options provided for each OD pair as described below.
+It would be possible to use other routing engines and custom profiles and cost functions in the future, for example to simulate the impact of implementing traffic signals which are optimised for cycling throughput, a feature not currently available in CycleStreets that could be explored in future work.
+
+We computed three route types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest'.
+As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the fastest routes minimize journey time, accounting for traffic lights and surface type.
 The 'Quietest' route type minimizes busy sections of road, allowing high 'diversion factors' from the fastest route on quieter but less direct ways (often avoiding roads and interactions with motor vehicles altogether where possible).
@@
 
-The CRUSE web application is statically hosted, meaning that it can be accessed without the need for a server running software such as the R package 'shiny' or Python packages such as 'streamlit' or 'flask' in the background [@wickham2021].
+The CRUSE web application is statically hosted, meaning it does not require a server running software such as the R package 'shiny' or Python packages such as 'streamlit' or 'flask' in the background [@wickham2021].
+This reduces hosting costs and the maintenance burden of the tool compared with tools such as the PCT, and makes the map-based interface more responsive, using the recently developed 'MapLibre' JavaScript package and 'PMTiles' "cloud native" vector tile format [@gonçalves2023].
+The default view in landing page is of a map of Ireland with county level results for the % trips that could be made by cycling under the 'Go Dutch' scenario, highlighting the intended use case for strategic cycle network planning and level of ambition.
+The key results, route network results, are available within an instant by zooming in on the map illustrated in @fig-landing: beyond a certain zoom level the network view appears.
+This differs from the user interface in tools such as the PCT and BiclaR which force users to click on a county before the key network results become visible: to get to the network view in the PCT takes around 10 seconds and multiple clicks, compared with simply zooming in on the landing page for CRUSE.
+
+Other improvements to the user interface in the national map on the landing page include a search bar in the top left and clearer legend for each network visualisation option (users can select any scenario or quietness or gradient depending on their needs).
+Following feedback from user testing, a 'geolocate' and 'full screen' button were added in the top right, enabling users to 'zoom' to their current location and to focus on the detailed geographic results, as shown in @fig-landing.
+
+::: {#fig-landing}
+![](images/paste-10.png)
+
+Illustration of the landing page of the CRUSE Tool for Ireland.
+:::
+
 Typical intended user stories are illustrated in @fig-user-stories, which shows that the tool is designed to be used by both professional and non-professional users.
@@
 <!-- The results for the other -->
-## Selected qualitative results {#sec-qualitative}
 
-Thoughout the project, we undertook a series of workshops and interviews with stakeholders in Ireland to gather feedback on the approach and resulting tool.
+## Stakeholder engagement and feedback {#sec-qualitative}
+
+Throughout the project, we undertook a series of workshops and interviews with stakeholders in Ireland to gather feedback on the approach and resulting tool.
 The feedback was generally positive, with stakeholders commenting on the need for the tool, lack of data on cycling potential proving a barrier to decision-making and investment, and the potential for the tool.
@@
 
-Early in the project, we presented the approach to practitioners in Kildare and Limmerick.
+Early in the project, we presented the approach to practitioners in Kildare and Limerick.
 The feedback from the workshop in Kildare highlighted the need for the tool for practitioners at the county level, as highlighted in the following quote:
 
-> "It's a missing piece of evidence that will help new projects get built" (Senior Engineer, Sustainable Transport & Traffic Management, Kildare County Council)
+> "It's a missing piece of evidence that will help new projects get built" (Dónal Hodgins, Senior Engineer, Sustainable Transport & Traffic Management, Kildare County Council)
 
@@
 They emphasized that cycle networks are often fragmented, and the tool allows them to demonstrate how individual local schemes can contribute to a larger, interconnected network.
-Council officers expressed a preference for visualizations that directly compare the fast and quiet routes, as they recognize the importance of both types of routes (users can switch between the on the 'Route types' page for each county although this functionality is not available in the landing page map due to other feedback requesting a simpler interface).
-
-In comparison with another tool (the National Transport Authority's cycling propensity tool), the CRUSE tool was seen by practitioners in Kildare as more user-friendly and intuitive, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
+An unexpected piece of feedback from the workshop was the preference for visualizations that directly compare the fast and quiet routes, as they recognize the importance of both types of routes.
+In response to this we provided the 'Route types' page for each county, although this functionality is not available in the landing page map due to other feedback requesting a simpler interface.
+In comparison with another tool (the National Transport Authority's cycling propensity tool), the CRUSE tool was seen by practitioners in Kildare as more user-friendly, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
 Overall, the workshop participants concluded that the approach filled gap in evidence and will greatly assist in the development of new cycling projects.
@@
 
-As the benefits of strategic planning for cycling become more apparent [@scappini2022] we expect the demand for tools such as CRUSE to grow.
+As the benefits of strategic planning for cycling become more apparent [@scappini2022], we expect the demand for tools such as CRUSE to grow.
 However, the approach is not without limitations and does not meet all requirements in Ireland or indeed any country where the methods are applied.
 
-A limitation of the approach as implemented in the CRUSE tool is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
+## Limitations and future research
+
+The approach has a number of limitations that should be understood by practitioners, researchers, advocates, policy makers using the tool.
+As highlighted in @sec-qualitative, we were only able to compare the results with a small number of cycle counters and based on feedback from workshops.
+As outlined in the CRUSE extension report \[reference to be added on publication\], we also undertook validation against Strava data.
+However, more data is needed to better understand the accuracy of the baseline results, especially for the 'non-POWSCAR' purposes.
+We hope this becomes possible as Ireland's cycle counter network expands, allowing refinements to the approach (espcially the spatial interaction model and parameters) to be made.
+Changes that could help mitigate this limitation on the user interface side include providing results with confidence intervals or as categories of cycling potential, rather than as central estimates, and the visualistion of the 'POWSCAR' network (which is based on Census data) in the web application.
+
+A limitation associated with the approach is its reliance on OpenStreetMap (OSM) data.
+The 'cycle friendliness' of a route is based on the OSM tags and the routes themselves are generated by the CycleStreets routing engine, which is based on OSM data.
+During a conference in Sligo where we presented prelimary results, a stakeholder with local knowledge pointed out that a quiet route along the canal was not being followed.
+In response, manual edits were made to the OSM network, and the results currently on the website reflect these changes.
+However, there are many other potential errors in the OSM data that have not been corrected.
+Future work could mitigate this by integrating the approach more closely with OSM (for example by providing links to enable one-click edits of OSM when users click on a route segment) or by setting up a workshop to 'crowdsource' corrections to the OSM data.
+Furthermore, because there are multiple possible users of the tool with different starging points, workshops and training events could be valuable if not essential to ensure that the tool is used effectively.
+Such events could bring together stakeholders from advocacy groups, local authorities, and national governments.
+The siloed nature of transport planning means that such diverse stakeholders are rarely the same room, meaning that a 'wide boundary' outcome of the tool could be to bring these groups together (if workshops based on the tool are successful).
+
+Another limitation of the approach is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
 Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.

### 2024-03-26 paper.qmd
add Declan comments

@@
 
-Practitioners in Kildare County Council that the evidence generated *was already valuable* the development of new cycle routes, especially when making the case during planning applications.
+Practitioners in Kildare County Council found that the evidence generated *was already valuable* for the development of new cycle routes, especially when making the case during planning applications.
 They emphasized that cycle networks are often fragmented, and the tool allows them to demonstrate how individual local schemes can contribute to a larger, interconnected network.
@@
 In response to this we provided the 'Route types' page for each county, although this functionality is not available in the landing page map due to other feedback requesting a simpler interface.
-In comparison with another tool (the National Transport Authority's cycling propensity tool), the CRUSE tool was seen by practitioners in Kildare as more user-friendly, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
+In comparison with another tool cycling propensity tool, the CRUSE tool was seen by practitioners in Kildare as more user-friendly, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
 Overall, the workshop participants concluded that the approach filled gap in evidence and will greatly assist in the development of new cycling projects.
@@
 In a subsequent interview with a practitioner using the evidence generated by the CRUSE approach in their everyday work, we received the following feedback.
-The outputs were a "great tool for getting started with project outline documents" and "ideal for generating a strategic rationale for estimating demand and building the business case in the absence of cycle count data" (Declan Keenan, Senior Transport Planner working on active travel, personal communication).
-Declan also noted the alignment between local knowledge, cycle count data, and the outputs of the approach: "It's a good representation of what's going on" and the networks allign well in terms of network "shape" and the proportion of trips of trips on different parts of the network.
+The outputs were found to be a "useful tool for getting schemes started; and provides good detail to supplement the strategic rationale for intervention within Project Outline Documents, via the cycle friendliness maps and the baseline / near market cyclist estimates”.
+As a result, the CRUSE tool “is ideal for generating a strategic rationale for estimating demand and building the business case in the absence of cycle count data; it is also very useful at the detailed design stage and the CRUSE estimates can (and have been used) to supplement the economic appraisal of cycling schemes” (Declan Keenan, Senior Transport Planner TII working on active travel, personal communication).
+Declan also noted a strong alignment in the South Dublin area between cycle count data, and the outputs of the approach: “It’s forms a strong representation of what’s going on in the areas where TII have collated some cycling count data, e.g.
+in the vicinity of the M50 in the areas of Ballinteer and Sandyford.
+Some additional network checking against count data would also be useful, to ensure good representation across other settlements.” Where cycling data is available, the networks align well in terms of network “shape” and the proportion of trips on different parts of the network.
 
-Findings from the qualitative results also suggest changes which have not been implemented due to time constraints, including the provision of route netework data showing only "gold standard" POWSCAR data (we did make this data available in the data downloads page), the need for further validation of results (especially in rural areas where counter datasets are scarce) and the need for training on the tool to ensure that it is well-understood and used effectively.
-These and other potential areas for further research and development are discussed in the next section.
+Where cycling data is available, the networks align well in terms of network “shape” and the proportion of trips on different parts of the network.
+There were two suggestions from Declan which have not yet been implemented: the provision of data showing POWSCAR and recreational estimates/data separately, and the need for training on the tool to ensure that all staff can use it effectively.
+These, and other potential improvements, are discussed in the next section.
 

### 2024-03-26 paper.qmd
review previous comments by Donal were included

@@
 In response to this we provided the 'Route types' page for each county, although this functionality is not available in the landing page map due to other feedback requesting a simpler interface.
-In comparison with another tool cycling propensity tool, the CRUSE tool was seen by practitioners in Kildare as more user-friendly, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
+In comparison with another cycling propensity tool, the CRUSE tool was seen by practitioners in Kildare as more user-friendly, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
 Overall, the workshop participants concluded that the approach filled gap in evidence and will greatly assist in the development of new cycling projects.
@@
 
-Where cycling data is available, the networks align well in terms of network “shape” and the proportion of trips on different parts of the network.
 There were two suggestions from Declan which have not yet been implemented: the provision of data showing POWSCAR and recreational estimates/data separately, and the need for training on the tool to ensure that all staff can use it effectively.

### 2024-03-26 paper.qmd
Resub evt3 (#39)
* add Declan comments

* review previous comments by Donal were included

* Update paper.qmd

---------

Co-authored-by: eugenividal <eugenividal@hotmail.com>

@@
 In response to this we provided the 'Route types' page for each county, although this functionality is not available in the landing page map due to other feedback requesting a simpler interface.
-In comparison with another cycling propensity tool, the CRUSE tool was seen by practitioners in Kildare as more user-friendly, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
+In comparison with another tool for transport planning, CRUSE was seen by practitioners in Kildare as more user-friendly.
+They found the evidence at the route level (unavailable in other tools) crucial for planning new cycling infrastructure.
 Overall, the workshop participants concluded that the approach filled gap in evidence and will greatly assist in the development of new cycling projects.

### 2024-03-26 paper.qmd
Update table code, fix #40

@@
 The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips.
-An overview of the trip purposes used in CRUSE is presented in @tbl-trip_purposes.
+An overview of the trip purposes used in CRUSE is presented in @tbl-trip-purposes.
 
@@
 ```{r}
-#| label: tbl-trip_purposes
+#| label: tbl-trip-purposes
 # For PDF:

### 2024-03-26 paper.tex
Update table code, fix #40

@@
 and multi-stage trips. An overview of the trip purposes used in CRUSE is
-presented in Figure~\ref{fig-trip-purposes}.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{images/paste-4.png}
-
-}
-
-\caption{\label{fig-trip-purposes}Everyday trip purposes used in the
-CRUSE Tool.}
-
-\end{figure}%
+presented in Table~\ref{tbl-trip-purposes}.
+
+\begin{longtable}[t]{>{\raggedright\arraybackslash}p{13em}>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{12em}>{\raggedright\arraybackslash}p{5em}}
+
+\caption{\label{tbl-trip-purposes}Trip purposes used in the CRUSE Tool.}
+
+\tabularnewline
+
+\\
+\toprule
+Trip purpose & Description & Primary source(s) & Confidence\\
+\midrule
+\cellcolor{gray!10}{Work (commute)} & \cellcolor{gray!10}{Commuting to/from workplaces} & \cellcolor{gray!10}{Census 2016 POWSCAR Data} & \cellcolor{gray!10}{High confidence - data both origins and destinations for trip purposes}\\
+Primary Education & Primary school trips & Census 2016 POWSCAR Data & Dept. of Education Schools database,High confidence - data both origins and destinations for trip purposes\\
+\cellcolor{gray!10}{Secondary Education} & \cellcolor{gray!10}{Secondary school trips} & \cellcolor{gray!10}{Census 2016 POWSCAR Data} & \cellcolor{gray!10}{Dept.
+of Education Schools database,High confidence - data both origins and destinations for trip purposes}\\
+Tertiary Education & Tertiary education trips & Census 2016 POWSCAR Data - Geodirectory categories for 'Tertiary Education' and 'Other Education' & High confidence - data both origins and destinations for trip purposes\\
+\cellcolor{gray!10}{Social} & \cellcolor{gray!10}{NHTS classification, including trips to/from sports facilities, leisure, and cultural destinations (e.g.
+restaurants, cinemas, gyms etc.)} & \cellcolor{gray!10}{Trip rates from NHTS} & \cellcolor{gray!10}{Geodirectory businesses classified according to relevant NACE codes,"Medium Confidence - Method should identify clusters of activity for this trip purpose, although individual trips or destinations come with significant uncertainty"}\\
+\addlinespace
+Shopping & NHTS classification, including trips to/from supermarkets and shops & Trip rates from NHTS & Geodirectory businesses classified according to relevant NACE codes,"Medium Confidence - Method should identify clusters of activity for this trip purpose, although individual trips or destinations come with significant uncertainty"\\
+\cellcolor{gray!10}{Personal/other} & \cellcolor{gray!10}{Combination of two NHTS classifications, and includes trips to medical, personal services and others} & \cellcolor{gray!10}{Trip rates from NHTS} & \cellcolor{gray!10}{Geodirectory businesses classified according to relevant NACE codes,"Medium Confidence - Method should identify clusters of activity for this trip purpose, although individual trips or destinations come with significant uncertainty"}\\
+Tourism/recreational & Trips by non-resident visitors to/from visitor attractions & Geodirectory listings for accommodation (origins) and destinations (destinations) & Failte Ireland accommodation database, Failte Ireland attraction database,Low confidence - lack of data regarding tourist trip rates\\
+\cellcolor{gray!10}{Public Transport} & \cellcolor{gray!10}{Trips to/from public transport} & \cellcolor{gray!10}{NTA Heavy Rail Census} & \cellcolor{gray!10}{NA}\\
+\bottomrule
+
+\end{longtable}
 
@@
 
+\includegraphics[width=3.31in,height=\textheight]{images/fig5.png}
+
 \begin{figure}
@@
 
-\includegraphics{images/mermaid-figure-1.png}
+\includegraphics{images/fig5.png}
 
@@
 
-Thoughout the project, we undertook a series of workshops and interviews
-with stakeholders in Ireland to gather feedback on the approach and
-resulting tool. The feedback was generally positive, with stakeholders
-commenting on the need for the tool, lack of data on cycling potential
-proving a barrier to decision-making and investment, and the potential
-for the tool. At an end of project workshop in December 2023,
-stakeholders were asked to provide feedback on the tool and its
-potential use in practice. Quotes and suggestions from these workshops,
-and a subsequent interview with a Senior Transport Planning using the
-evidence in their everyday work in Transport Infrastructure Ireland, and
-presented in this section (with quotes being attributed to named
-individuals where permission has been granted).
+Throughout the project, we undertook a series of workshops and
+interviews with stakeholders in Ireland to gather feedback on the
+approach and resulting tool. The feedback was generally positive, with
+stakeholders commenting on the need for the tool, lack of data on
+cycling potential proving a barrier to decision-making and investment,
+and the potential for the tool. At an end of project workshop in
+December 2023, stakeholders were asked to provide feedback on the tool
+and its potential use in practice. Quotes and suggestions from these
+workshops, and a subsequent interview with a Senior Transport Planning
+using the evidence in their everyday work in Transport Infrastructure
+Ireland, and presented in this section (with quotes being attributed to
+named individuals where permission has been granted).
 
 Early in the project, we presented the approach to practitioners in
-Kildare and Limmerick. The feedback from the workshop in Kildare
+Kildare and Limerick. The feedback from the workshop in Kildare
 highlighted the need for the tool for practitioners at the county level,
@@
 ``It's a missing piece of evidence that will help new projects get
-built'' (Senior Engineer, Sustainable Transport \& Traffic Management,
-Kildare County Council)
+built'' (Dónal Hodgins, Senior Engineer, Sustainable Transport \&
+Traffic Management, Kildare County Council)
 \end{quote}
 
-Practitioners in Kildare County Council that the evidence generated
-\emph{was already valuable} the development of new cycle routes,
-especially when making the case during planning applications. They
-emphasized that cycle networks are often fragmented, and the tool allows
-them to demonstrate how individual local schemes can contribute to a
-larger, interconnected network. An unexpected piece of feedback from the
-workshop was the preference for visualizations that directly compare the
-fast and quiet routes, as they recognize the importance of both types of
-routes. In response to this we provided the `Route types' page for each
-county, although this functionality is not available in the landing page
-map due to other feedback requesting a simpler interface. In comparison
-with another tool (the National Transport Authority's cycling propensity
-tool), the CRUSE tool was seen by practitioners in Kildare as more
-user-friendly, with the ability to generate evidence at the route level,
-which is crucial for planning new cycling infrastructure. Overall, the
-workshop participants concluded that the approach filled gap in evidence
-and will greatly assist in the development of new cycling projects.
+Practitioners in Kildare County Council found that the evidence
+generated \emph{was already valuable} for the development of new cycle
+routes, especially when making the case during planning applications.
+They emphasized that cycle networks are often fragmented, and the tool
+allows them to demonstrate how individual local schemes can contribute
+to a larger, interconnected network. An unexpected piece of feedback
+from the workshop was the preference for visualizations that directly
+compare the fast and quiet routes, as they recognize the importance of
+both types of routes. In response to this we provided the `Route types'
+page for each county, although this functionality is not available in
+the landing page map due to other feedback requesting a simpler
+interface. In comparison with another tool for transport planning, CRUSE
+was seen by practitioners in Kildare as more user-friendly. They found
+the evidence at the route level (unavailable in other tools) crucial for
+planning new cycling infrastructure. Overall, the workshop participants
+concluded that the approach filled gap in evidence and will greatly
+assist in the development of new cycling projects.
 
@@
 generated by the CRUSE approach in their everyday work, we received the
-following feedback. The outputs were a ``great tool for getting started
-with project outline documents'' and ``ideal for generating a strategic
-rationale for estimating demand and building the business case in the
-absence of cycle count data'' (Declan Keenan, Senior Transport Planner
-working on active travel, personal communication). Declan also noted the
-alignment between local knowledge, cycle count data, and the outputs of
-the approach: ``It's a good representation of what's going on'' and the
-networks allign well in terms of network ``shape'' and the proportion of
-trips of trips on different parts of the network.
-
-Findings from the qualitative results also suggest changes which have
-not been implemented due to time constraints, including the provision of
-route netework data showing only ``gold standard'' POWSCAR data (we did
-make this data available in the data downloads page), the need for
-further validation of results (especially in rural areas where counter
-datasets are scarce) and the need for training on the tool to ensure
-that it is well-understood and used effectively. These and other
-potential areas for further research and development are discussed in
-the next section.
+following feedback. The outputs were found to be a ``useful tool for
+getting schemes started; and provides good detail to supplement the
+strategic rationale for intervention within Project Outline Documents,
+via the cycle friendliness maps and the baseline / near market cyclist
+estimates''. As a result, the CRUSE tool ``is ideal for generating a
+strategic rationale for estimating demand and building the business case
+in the absence of cycle count data; it is also very useful at the
+detailed design stage and the CRUSE estimates can (and have been used)
+to supplement the economic appraisal of cycling schemes'' (Declan
+Keenan, Senior Transport Planner TII working on active travel, personal
+communication). Declan also noted a strong alignment in the South Dublin
+area between cycle count data, and the outputs of the approach: ``It's
+forms a strong representation of what's going on in the areas where TII
+have collated some cycling count data, e.g.~in the vicinity of the M50
+in the areas of Ballinteer and Sandyford. Some additional network
+checking against count data would also be useful, to ensure good
+representation across other settlements.'' Where cycling data is
+available, the networks align well in terms of network ``shape'' and the
+proportion of trips on different parts of the network.
+
+There were two suggestions from Declan which have not yet been
+implemented: the provision of data showing POWSCAR and recreational
+estimates/data separately, and the need for training on the tool to
+ensure that all staff can use it effectively. These, and other potential
+improvements, are discussed in the next section.
 

### 2024-03-26 paper.qmd
Edit csv and table code, fix #41

@@
 knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.", booktabs = TRUE) |>
-  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
-  kableExtra::column_spec(1, width = "13em") |>
-  kableExtra::column_spec(2, width = "15em") |>
-  kableExtra::column_spec(3, width = "12em") |>
-  kableExtra::column_spec(4, width = "5em")
+  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE, font_size = 8) |>
+  kableExtra::column_spec(1, width = "5em") |>
+  kableExtra::column_spec(2, width = "12em") |>
+  kableExtra::column_spec(3, width = "15em") |>
+  kableExtra::column_spec(4, width = "15em")
 # For HTML:

### 2024-03-26 paper.tex
Edit csv and table code, fix #41

@@
 
-\begin{longtable}[t]{>{\raggedright\arraybackslash}p{13em}>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{12em}>{\raggedright\arraybackslash}p{5em}}
+\begingroup\fontsize{8}{10}\selectfont
+
+\begin{longtable}[t]{>{\raggedright\arraybackslash}p{5em}>{\raggedright\arraybackslash}p{12em}>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{15em}}
 
@@
 \midrule
-\cellcolor{gray!10}{Work (commute)} & \cellcolor{gray!10}{Commuting to/from workplaces} & \cellcolor{gray!10}{Census 2016 POWSCAR Data} & \cellcolor{gray!10}{High confidence - data both origins and destinations for trip purposes}\\
-Primary Education & Primary school trips & Census 2016 POWSCAR Data & Dept. of Education Schools database,High confidence - data both origins and destinations for trip purposes\\
-\cellcolor{gray!10}{Secondary Education} & \cellcolor{gray!10}{Secondary school trips} & \cellcolor{gray!10}{Census 2016 POWSCAR Data} & \cellcolor{gray!10}{Dept.
-of Education Schools database,High confidence - data both origins and destinations for trip purposes}\\
-Tertiary Education & Tertiary education trips & Census 2016 POWSCAR Data - Geodirectory categories for 'Tertiary Education' and 'Other Education' & High confidence - data both origins and destinations for trip purposes\\
-\cellcolor{gray!10}{Social} & \cellcolor{gray!10}{NHTS classification, including trips to/from sports facilities, leisure, and cultural destinations (e.g.
-restaurants, cinemas, gyms etc.)} & \cellcolor{gray!10}{Trip rates from NHTS} & \cellcolor{gray!10}{Geodirectory businesses classified according to relevant NACE codes,"Medium Confidence - Method should identify clusters of activity for this trip purpose, although individual trips or destinations come with significant uncertainty"}\\
+\cellcolor{gray!10}{Work (commute)} & \cellcolor{gray!10}{Commuting to / from workplaces} & \cellcolor{gray!10}{Census 2016 POWSCAR Data} & \cellcolor{gray!10}{High confidence - data both origins and destinations for trip purposes}\\
+Primary Education & Primary school trips & Census 2016 POWSCAR Data, Schools database & High confidence - data both origins and destinations for trip purposes\\
+\cellcolor{gray!10}{Secondary Education} & \cellcolor{gray!10}{Secondary school trips} & \cellcolor{gray!10}{Census 2016 POWSCAR Data, Schools database} & \cellcolor{gray!10}{High confidence - data both origins and destinations for trip purposes}\\
+Tertiary Education & Tertiary education trips & Census 2016 POWSCAR Data - Geodirectory categories & High confidence - data both origins and destinations for trip purposes\\
+\cellcolor{gray!10}{Social} & \cellcolor{gray!10}{NHTS classification (e.g.
+restaurants, cinemas, gyms etc.)} & \cellcolor{gray!10}{Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes} & \cellcolor{gray!10}{Medium Confidence}\\
 \addlinespace
-Shopping & NHTS classification, including trips to/from supermarkets and shops & Trip rates from NHTS & Geodirectory businesses classified according to relevant NACE codes,"Medium Confidence - Method should identify clusters of activity for this trip purpose, although individual trips or destinations come with significant uncertainty"\\
-\cellcolor{gray!10}{Personal/other} & \cellcolor{gray!10}{Combination of two NHTS classifications, and includes trips to medical, personal services and others} & \cellcolor{gray!10}{Trip rates from NHTS} & \cellcolor{gray!10}{Geodirectory businesses classified according to relevant NACE codes,"Medium Confidence - Method should identify clusters of activity for this trip purpose, although individual trips or destinations come with significant uncertainty"}\\
-Tourism/recreational & Trips by non-resident visitors to/from visitor attractions & Geodirectory listings for accommodation (origins) and destinations (destinations) & Failte Ireland accommodation database, Failte Ireland attraction database,Low confidence - lack of data regarding tourist trip rates\\
-\cellcolor{gray!10}{Public Transport} & \cellcolor{gray!10}{Trips to/from public transport} & \cellcolor{gray!10}{NTA Heavy Rail Census} & \cellcolor{gray!10}{NA}\\
+Shopping & NHTS classificationfrom supermarkets and shops & Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes & Medium Confidence\\
+\cellcolor{gray!10}{Personal / other} & \cellcolor{gray!10}{Combination of two NHTS classifications, and includes trips to medical, personal services and others} & \cellcolor{gray!10}{Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes} & \cellcolor{gray!10}{Medium Confidence}\\
+Tourism / recreational & Trips by non-resident visitors to / from visitor attractions & Geodirectory listings for accommodation (origins) and destinations (destinations) & Failte Ireland accommodation database, Failte Ireland attraction database\\
 \bottomrule
@@
 \end{longtable}
+
+\endgroup{}
 

### 2024-03-26 paper.qmd
Fix up table 1 location

--- 2024-03-26 paper.qmd
@@
 We share this conception of PSS, emphasizing the benefits of a focus on a particular mode (cycling) and a particular phase in the planning process (strategic network development).
+A selection of existing tools are summarised below and presented in @tbl-tools.
 
-Some of the most relevant existing tools are summarised below and presented in @tbl-tools.
+\newpage
 
@@
 #| label: tbl-trip-purposes
+#| tbl-cap: "Trip purposes used in the CRUSE Tool."
 # For PDF:
-knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.", booktabs = TRUE) |>
-  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
-  kableExtra::column_spec(1, width = "13em") |>
-  kableExtra::column_spec(2, width = "15em") |>
-  kableExtra::column_spec(3, width = "12em") |>
-  kableExtra::column_spec(4, width = "5em")
+knitr::kable(trip_purposes, booktabs = TRUE) |>
+  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE, font_size = 8) |>
+  kableExtra::column_spec(1, width = "5em") |>
+  kableExtra::column_spec(2, width = "12em") |>
+  kableExtra::column_spec(3, width = "15em") |>
+  kableExtra::column_spec(4, width = "15em")
 # For HTML:

### 2024-03-27 paper-v1.tex
Add v2 sub files

@@
   pdftitle={CRUSE to Safe Cycling in Ireland},
-  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Tool; Dan Brennan; Suzanne Meade},
+  pdfauthor={},
   pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
@@
 Support Active Travel} }
-\author[1]{Robin Lovelace%
-\corref{cor1}%
-}
- \ead{r.lovelace@leeds.ac.uk} 
-\author[1]{Joey Talbot%
-%
-}
-
-\author[1]{Eugeni Vidal-Tortosa%
-%
-}
-
-\author[1]{Hussein Mahfouz%
-%
-}
-
-\author[2]{Elaine Brick%
-%
-}
-
-\author[3]{Peter Wright%
-%
-}
-
-\author[2]{Gary O'Tool%
-%
-}
-
-\author[4]{Dan Brennan%
-%
-}
-
-\author[4]{Suzanne Meade%
-%
-}
-
-\affiliation[1]{organization={University of
-Leeds},addressline={Institute for Transport Studies, Leeds, LS2 9JT,
-United Kingdom},postcodesep={}}
-\affiliation[2]{organization={AECOM},addressline={Unit 6, Galway
-Technology Park, Parkmore, Galway, Ireland},postcodesep={}}
-\affiliation[3]{organization={AECOM},addressline={Winslade House,
-Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY,
-UK},postcodesep={}}
-\affiliation[4]{organization={Transport Infrastructure
-Ireland},addressline={Parkgate Business Centre, Parkgate Street, Dublin
-8, D08 DK10, Ireland},postcodesep={}}
-
-\cortext[cor1]{Corresponding author}
 

### 2024-03-27 paper.qmd
Add v2 sub files

@@
 pdf-engine: pdflatex
-knitr:
-  opts_chunk:
-    fig.path: images/
 format: 
-  # html: default
+  html: default
   # docx: default
 # # # Uncomment for formatted version
-  elsevier-pdf:
-    linestretch: 2
-    keep-tex: true
-    journal:
-      name: European Transport Research Review
-      formatting: preprint
-      model: 3p
-      cite-style: super
+  # elsevier-pdf:
+  #   linestretch: 2
+  #   keep-tex: true
+  #   journal:
+  #     name: European Transport Research Review
+  #     formatting: preprint
+  #     model: 3p
+  #     cite-style: super
 # # # #   # pdf: default
 
-# # Uncomment the following line to produce the cover page
-author:
-  - name: Robin Lovelace
-    email: r.lovelace@leeds.ac.uk
-    corresponding: true
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Joey Talbot
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Eugeni Vidal-Tortosa
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Hussein Mahfouz
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Elaine Brick
-    affiliations:
-      - name: AECOM
-        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-  - name: Peter Wright
-    affiliations:
-      - name: AECOM
-        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
-  - name: "Gary O'Toole"
-    affiliations:
-      - name: AECOM
-        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-  - name: Dan Brennan
-    affiliations:
-      - name: Transport Infrastructure Ireland
-        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
-  - name: Suzanne Meade
-    affiliations:
-      - name: Transport Infrastructure Ireland
-        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+# # # Uncomment the following line to produce the cover page
+# author:
+#   - name: Robin Lovelace
+#     email: r.lovelace@leeds.ac.uk
+#     corresponding: true
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Joey Talbot
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Eugeni Vidal-Tortosa
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Hussein Mahfouz
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Elaine Brick
+#     affiliations:
+#       - name: AECOM
+#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+#   - name: Peter Wright
+#     affiliations:
+#       - name: AECOM
+#         address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
+#   - name: "Gary O'Toole"
+#     affiliations:
+#       - name: AECOM
+#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+#   - name: Dan Brennan
+#     affiliations:
+#       - name: Transport Infrastructure Ireland
+#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+#   - name: Suzanne Meade
+#     affiliations:
+#       - name: Transport Infrastructure Ireland
+#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
 abstract: |
@@
 ::: {#fig-dublin}
-<!-- ![](images/paste-1.png) -->
+<!-- ![](paste-1.png) -->
 
-![](images/paste-8.png) <!-- ![](images/paste-3.png) -->
-![](images/paste-9.png)
+![](paste-8.png) <!-- ![](paste-3.png) -->
+![](paste-9.png)
 
@@
 An overview of the trip purposes used in CRUSE is presented in @tbl-trip-purposes.
+
+\clearpage
 
@@
 
-We computed three route types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest'.
+We computed three route types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest', based on the observation that different cyclists have different preferences for route types and the fastest or shortest path may not be the most heavily used [@desjardins2022].
 As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the fastest routes minimize journey time, accounting for traffic lights and surface type.
 The 'Quietest' route type minimizes busy sections of road, allowing high 'diversion factors' from the fastest route on quieter but less direct ways (often avoiding roads and interactions with motor vehicles altogether where possible).
-The 'Balanced' route type is a compromize between the two, minimising journey time while avoiding the busiest roads.
+The 'Balanced' route type is a compromise between the two, minimizing journey time while avoiding the busiest roads.
 Allowing users to switch between these three route types enables them to consider the trade-offs between directness and 'cycle friendliness' when planning new infrastructure, as shown in @fig-route-types.
@@
 
-<!-- ![](images/paste-5.png) -->
+<!-- ![](paste-5.png) -->
 
-<!-- ![](images/paste-6.png) -->
+<!-- ![](paste-6.png) -->
 
-<!-- Illustration of the Fastest and Quietest route networks for central and East Limmerick. -->
+<!-- Illustration of the Fastest and Quietest route networks for central and East limerick. -->
 
@@
 ::: {#fig-scenarios}
-<!-- ![](images/paste-7.png) -->
+<!-- ![](paste-7.png) -->
 
-![](images/scenarios_kildare.png)
+![](scenarios_kildare.png)
 
@@
 This reduces hosting costs and the maintenance burden of the tool compared with tools such as the PCT, and makes the map-based interface more responsive, using the recently developed 'MapLibre' JavaScript package and 'PMTiles' "cloud native" vector tile format [@gonçalves2023].
-The default view in landing page is of a map of Ireland with county level results for the % trips that could be made by cycling under the 'Go Dutch' scenario, highlighting the intended use case for strategic cycle network planning and level of ambition.
+These open source technologies were used to generate default view in the landing page: a map of Ireland with county level results for the % trips that could be made by cycling under the 'Go Dutch' scenario, highlighting the intended use case for strategic cycle network planning and level of ambition, as illustrated in @fig-landing.
+
+::: {#fig-landing}
+![](paste-10.png)
+
+![](paste-11.png)
+
+Illustration of the landing page of the CRUSE Tool for Ireland, showing the default 'zoomed out' view of the national map, with county level results showing the proportion of trips cycled under the 'Go Dutch' scenario (top).
+The default 'zoomed out' view enables users to click on a county for more detailed results.
+The 'zoomed in' view shows the national network view for the 'Go Dutch' scenario for the Limerick area (bottom).
+:::
+
 The key results, route network results, are available within an instant by zooming in on the map illustrated in @fig-landing: beyond a certain zoom level the network view appears.
@@
 Following feedback from user testing, a 'geolocate' and 'full screen' button were added in the top right, enabling users to 'zoom' to their current location and to focus on the detailed geographic results, as shown in @fig-landing.
-
-::: {#fig-landing}
-![](images/paste-10.png)
-
-Illustration of the landing page of the CRUSE Tool for Ireland.
-:::
 
@@
 ```{r alternative}
-# https://www.sesync.org/resources/creating-visualizations-diagrammer
+#| fig-cap: Hypothetical user stories for the CRUSE Tool illustrating that it can be used in different ways by different user groups.
+#| label: fig-user-stories
 schematic = DiagrammeR::grViz("
@@
 # schematic
-# htmlwidgets::saveWidget(schematic, "images/fig5.html")
-# browseURL("images/fig5.html")
-# webshot2::webshot(url = "images/fig5.html", file = "images/fig5.png")
-knitr::include_graphics("images/fig5.png")
+# htmlwidgets::saveWidget(schematic, "fig5.html")
+# browseURL("fig5.html")
+# webshot2::webshot(url = "fig5.html", file = "fig5.png")
+# knitr::include_graphics("fig5.png")
+schematic
 ```
 
-::: {#fig-user-stories}
-![](images/fig5.png)
+```{=html}
+<!-- ::: {#fig-user-stories}
+![](fig5.png)
 
-Hypothetical user stories for the CRUSE Tool illustrating that it can be used in different ways by different user groups.
-:::
-
+::: -->
+```
 # Results {#sec-results}
@@
 
-To illustrate the results in urban areas, we took the top 4 cities in Ireland by population: Dublin City, Cork, Limerick and Galway (with city populations of 0.6, 0.2, 0.1 and 0.1 million respectively).
-In @fig-city-results we present the cycle route networks within a 3km radius circle centred on each of these cities.
+To illustrate the results in urban areas, we explored route networks in the three largest cities in Ireland: Dublin, Cork, and Limerick.
+@fig-city-results shows the 'fastest' and 'quietest' cycle networks within a 6km wide square centred on each of these cities.
 
@@
 # Results for one city are shown below
-cities = c("dublin", "cork", "limerick", "galway")
-city_name = "dublin"
+# cities = c("dublin", "cork", "limerick", "galway")
+library(tmap)
+cities = c("dublin", "cork", "limerick")
+city_name = "limerick"
 city_buffer = zonebuilder::zb_zone(city_name)
@@
   tm_lines(col = "Quietness", palette = "RdYlGn", lwd = 5)
-tmap_save(m, "./images/rnet_fastest_city_top.png", width = 10, height = 10)
+tmap_save(m, "./rnet_fastest_city_top.png", width = 10, height = 10)
 # Generalisation to balanced and quiestest route types:
-route_types = c("fastest", "balanced", "quietest")
+route_types = c("fastest", "quietest")
 dublin_rnets = purrr::map(route_types, function(route_type) {
@@
     filter(circle_id <= 2)
+  city_buffer_5km = stplanr::geo_bb(city_buffer_5km)
   city_buffer_union = sf::st_union(city_buffer_5km)
-  route_types = c("fastest", "balanced", "quietest")
+  route_types = c("fastest", "quietest")
   rnet_types = purrr::map(route_types, function(route_type) {
@@
 cities_rnet_types = do.call(rbind, cities_rnet_types)
+# cities_rnet_types_1km = do.call(rbind, cities_rnet_types_1km)
 summary(cities_rnet_types$`Bicycle (Go Dutch)`)
@@
 m_types
-tmap_save(m_types, "./images/rnet_types.png", width = 12, height = 10)
-browseURL("./images/rnet_types.png")
+tmap_save(m_types, "./rnet_types.png", width = 12, height = 16)
+browseURL("./rnet_types.png")
 ```
@@
 ::: {#fig-city-results}
-![](images/rnet_types.png)
+![](rnet_types.png)
 
-Fastest, balanced and quietest cycle networks within 3km circular areas containing the four largest cities in Ireland.
+Fastest (left) and quietest (right) cycle networks in 6 km squares centered on central Dublin (top panels) Cork (middle) and Limerick (bottom).
 The networks are coloured by cycle friendliness, with greener segments representing more cycle friendly segments.
@@
 
-Aside from the widely varying sizes and route networks of each city, with Dublin having by far the densest network, it is clear from the visualisation of cycling friendliness that there are many gaps in the networks.
-Dublin has had received most investment in cycling infrastructure and this is apparent from the high proportion of the route network that is green, even in the fastest route network representing routes that prioritize speed and directness over comfort and safety.
-Dublin also has the highest mode share of cycling of the four cities, adding to the evidence base showing that investment in cycling infrastructure is effective in increasing cycling uptake.
+Aside from the varying sizes and route network shapes associated with each city, it is clear from the visualisation of cycling friendliness that none of the cities can be considered to have a 'complete' cycle network.
+Of the three, Dublin has clearly had most investment cycle infrastructure, as shown by the relatively high proportion of the route network that is green in both network types.
 
-Although Dublin City has made progress, there are still many parts of the fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
+@fig-city-results highlights the value of generating different network types: looking only at the fastest network it would be hard to differentiate between Cork and Limerick in terms of cycle network quality.
+However, Limerick clearly has a higher number of quiet routes that can be cycled on for people willing to take a longer route to avoid busy roads: its quiet network (bottom right in @fig-city-results) contains more green segments than the quietest network for Cork (middle right in @fig-city-results).
+The lack of continuous sections of quiet routes in Cork, even in the quietest network, means that people who are less confident cycling *cannot* avoid mixing with motor traffic for many common trips.
+In an ideal network, conversely, even the fastest route network would be predominantly cycle friendly, meaning that the most direct and fast routes are also quiet, with low levels of traffic stress.
+
+There are still many parts of the fastest route network, and even some parts of the quietest route networks, that are not cycle friendly and which have high cycling potential, even in Dublin.
 According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians" [@walking2021].
-The results for the Dublin area could help prioritise investment in cycling infrastructure in the city.
-
+Assuming that these findings apply in other cities, the results can help prioritise where such infrastructure should go: the results presented in @fig-city-results should bring attention to where there are wide red lines in fastest and quietest networks, with a reasonable strategy being to prioritise a continuous quiet network first and eventually to 'cycle proof' the fastest network also.
+However, a feature of the approach is that it is not prescriptive and it does not tell planners what to build where; as highlighted below practitioners can use the results in a variety of ways.
 <!-- The results for the other -->
@@
 
-While the CRUSE Tool is not the first national and publicly available tool for cycling planning, it has some key features that make it relevant for other countries, regions and road authorities tasked with making their transport systems safe and sustainable:
+While the CRUSE approach presented in this paper is not the first national and publicly available tool for cycling planning, it has some key features that make it relevant for other countries, regions and road authorities tasked with making their transport systems safe and sustainable:
 
+-   Compared with the Propensity to Cycle Tool (PCT), which provides either travel to work [@lovelace2017] or to school cycling networks [@goodman2019], the approach presented here captures a higher proportion of cycling potential including key rural trips which are often under-represented in models of active travel.
 -   The results are *open access* meaning that any stakeholder in the planning system can access the evidence.
@@
 -   The results are fully reproducible (code to be released pending sign-off by TII's IT team), preventing 'cloud lock in' to a potentially monopolistic consultancy, and encouraging input from the wider open source community [@lovelace2021; @dhir2017].
--   By covering a wide range of trip purposes --- not just travel to work [@lovelace2017; @heinen2010] and travel to school [@goodman2019] as covered in previously published research on open access national cycle network planning tools --- the results capture a higher proportion of cycling potential including key rural trips which are often under-represented in models of active travel.
+-   The tool has a user interface that allows people to see the key results quickly, without having to click on regional result pages, as is the case with the BiclaR and PCT projects.
 
-As the benefits of strategic planning for cycling become more apparent [@scappini2022], we expect the demand for tools such as CRUSE to grow.
+As the benefits of strategic planning for cycling become more apparent [@scappini2022], we expect the demand for the approach outlined in this paper, and resulting tools, to grow.
 However, the approach is not without limitations and does not meet all requirements in Ireland or indeed any country where the methods are applied.
@@
 Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
-These factors mean that the tool can be seen as an open access 'leverage point', providing key information for and supporting ambitious plans in many aspects the planning system, from network design to post-build monitoring [@lovelace2020].
+Regular updates to the model to generate yearly baseline cycle flow estimates could tackle this limitation, although would take substantial resources.
+
+More broadly, the tool's outputs are limited to just one mode of active travel (cycling), ignoring walking and wheeling, including wheelchair use and a range of wheeled devices such as scooters, bike trailers and e-cargo bikes that can be used to escort children to school.
+Sustainable transport policies should plan for walking, wheeling and cycling, and there are many co-benefits of broadly defined active travel interventions that benefit all active modes, such as measures to reduce heavy motor traffic speeds and volumes in areas and along corridors with high active travel potential.
+This raises the question of whether other active modes should be incorporated into the results using the OD-based approach outlined in this paper, or whether different modeling approaches are needed to properly capture the shorter distance trips typically made by walking [@cooper2018].
+Furthermore, the estimates of cycling presented in the CRUSE Tool omit multi-modal trips including public transport, and omit trip chaining, due to the need to capture a large portion of cycling potential within the resource constraints of the project.
 
@@
 
-The CRUSE Tool and the underlying research and methods are not without limitations, suggesting future areas of research, data collection needs, and policy application.
-The route network level results have not been validated to the extent we had planned at the outset of the project.
-We compared route network level results with 'ground truth' data from cycle counter datasets across Dublin to test different network types and to ensure that additional trip types added to work and educational trips increased the quality of fit under the baseline scenario.
-However, the size of the counter network was insufficient to provide an opportunity for robust evaluation of model performance, suggesting a combined program of new count data collection and analysis should be a priority for future work, with findings feeding directly into work to improve the outputs of the the CRUSE Tool.
-
-More broadly, the tool's outputs are limited to just one mode of active travel (cycling), ignoring walking and wheeling, including wheelchair use and a range of wheeled devices such as scooters, bike trailers and e-cargo bikes that can be used to escort children to school.
-Sustainable transport policies should plan for walking, wheeling and cycling, and there are many co-benefits of broadly defined active travel interventions that benefit all active modes, such as measures to reduce heavy motor traffic speeds and volumes in areas and along corridors with high active travel potential.
-This raises the question of whether other active modes should be incorporated into the results using the OD-based approach outlined in this paper, or whether different modeling approaches are needed to properly capture the shorter distance trips typically made by walking [@cooper2018].
-Furthermore, the estimates of cycling presented in the CRUSE Tool omit multi-modal trips including public transport, and omit trip chaining, due to the need to capture a large portion of cycling potential within the resource constraints of the project.
-
-The CRUSE Tool is already used in practice to support more ambitious and data-driven planning for safe cycling routes in multiple counties across Ireland.
-We hope that the underlying approach, and the publicly available evidence provided in the CRUSE Tool itself, provide a basis for future reproducible research, open source software development into strategic cycle network planning tools in Ireland and other countries.
-Active travel represents a win-win-win for physical activity, environment, and local economic opportunities.
-In combination with broader sustainable mobility measures and policies to reduce motor traffic speeds and volumes, tools such as CRUSE can support the fast and fair decarbonisation of transport systems worldwide.
+As outlined in @sec-qualitative, the tool is already used in practice to support more ambitious and data-driven planning for safe cycling routes in multiple counties across Ireland.
+We hope that the underlying approach provides a basis for future research and development.
+In combination with broader sustainable mobility measures and policies to reduce motor traffic speeds and volumes, evidence generated by the approach outlined in this paper can support the fast and fair decarbonisation of transport systems worldwide.
 

### 2024-03-27 paper.tex
Fix up table 1 location
Add v2 sub files

@@
   pdftitle={Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools},
-  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Toole; Dan Brennan; Suzanne Meade},
   pdfkeywords={Cycling, Open-source, Road Safety, Active
@@
 for developing strategic cycle network planning tools}
-\author[1]{Robin Lovelace%
-\corref{cor1}%
-}
- \ead{r.lovelace@leeds.ac.uk} 
-\author[1]{Joey Talbot%
-%
-}
-
-\author[1]{Eugeni Vidal-Tortosa%
-%
-}
-
-\author[1]{Hussein Mahfouz%
-%
-}
-
-\author[2]{Elaine Brick%
-%
-}
-
-\author[3]{Peter Wright%
-%
-}
-
-\author[2]{Gary O'Toole%
-%
-}
-
-\author[4]{Dan Brennan%
-%
-}
-
-\author[4]{Suzanne Meade%
-%
-}
-
-\affiliation[1]{organization={University of
-Leeds},addressline={Institute for Transport Studies, Leeds, LS2 9JT,
-United Kingdom},postcodesep={}}
-\affiliation[2]{organization={AECOM},addressline={Unit 6, Galway
-Technology Park, Parkmore, Galway, Ireland},postcodesep={}}
-\affiliation[3]{organization={AECOM},addressline={Winslade House,
-Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY,
-UK},postcodesep={}}
-\affiliation[4]{organization={Transport Infrastructure
-Ireland},addressline={Parkgate Business Centre, Parkgate Street, Dublin
-8, D08 DK10, Ireland},postcodesep={}}
 
@@
 particular mode (cycling) and a particular phase in the planning process
-(strategic network development).
-
-Some of the most relevant existing tools are summarised below and
-presented in Table~\ref{tbl-tools}.
+(strategic network development). A selection of existing tools are
+summarised below and presented in Table~\ref{tbl-tools}.
+
+\newpage
 
@@
 
-\includegraphics{images/paste-8.png}
-\includegraphics{images/paste-9.png}
+\includegraphics{paste-8.png} \includegraphics{paste-9.png}
 
@@
 
+\clearpage
+
 \begingroup\fontsize{8}{10}\selectfont
@@
 
-\\
 \toprule
@@
 \cellcolor{gray!10}{Social} & \cellcolor{gray!10}{NHTS classification (e.g.
-restaurants, cinemas, gyms etc.)} & \cellcolor{gray!10}{Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes} & \cellcolor{gray!10}{Medium Confidence}\\
+restaurants, cinemas, gyms etc.)} & \cellcolor{gray!10}{Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes} & \cellcolor{gray!10}{Medium confidence}\\
 \addlinespace
-Shopping & NHTS classificationfrom supermarkets and shops & Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes & Medium Confidence\\
-\cellcolor{gray!10}{Personal / other} & \cellcolor{gray!10}{Combination of two NHTS classifications, and includes trips to medical, personal services and others} & \cellcolor{gray!10}{Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes} & \cellcolor{gray!10}{Medium Confidence}\\
-Tourism / recreational & Trips by non-resident visitors to / from visitor attractions & Geodirectory listings for accommodation (origins) and destinations (destinations) & Failte Ireland accommodation database, Failte Ireland attraction database\\
+Shopping & NHTS classificationfrom supermarkets and shops & Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes & Medium confidence\\
+\cellcolor{gray!10}{Personal / other} & \cellcolor{gray!10}{Combination of two NHTS classifications, and includes trips to medical, personal services and others} & \cellcolor{gray!10}{Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes} & \cellcolor{gray!10}{Medium confidence}\\
+Tourism / recreational & Trips by non-resident visitors to / from visitor attractions & Geodirectory listings for accommodation (origins) and destinations (destinations), Failte Ireland accommodation database, Failte Ireland attraction database & Low confidence - lack of data regarding tourist trip rates\\
 \bottomrule
@@
 We computed three route types for each disaggregated (`jittered') OD
-pair: `Fastest', `Balanced' and `Quietest'. As outlined on the
-CycleStreets'
+pair: `Fastest', `Balanced' and `Quietest', based on the observation
+that different cyclists have different preferences for route types and
+the fastest or shortest path may not be the most heavily used
+\citep{desjardins2022}. As outlined on the CycleStreets'
 \href{https://www.cyclestreets.net/help/journey/howitworks/}{website},
@@
 motor vehicles altogether where possible). The `Balanced' route type is
-a compromize between the two, minimising journey time while avoiding the
+a compromise between the two, minimizing journey time while avoiding the
 busiest roads. Allowing users to switch between these three route types
@@
 \begin{minipage}{0.50\linewidth}
-\includegraphics[width=2.60417in,height=\textheight]{images/naas_quietest_godutch.png}\end{minipage}%
+\includegraphics[width=2.60417in,height=\textheight]{naas_quietest_godutch.png}\end{minipage}%
 %
 \begin{minipage}{0.50\linewidth}
-\includegraphics[width=2.60417in,height=\textheight]{images/naas_fastest_godutch.png}\end{minipage}%
+\includegraphics[width=2.60417in,height=\textheight]{naas_fastest_godutch.png}\end{minipage}%
 \newline
 \begin{minipage}{0.50\linewidth}
-\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_quietest_godutch.png}\end{minipage}%
+\includegraphics[width=2.60417in,height=\textheight]{newbridge_quietest_godutch.png}\end{minipage}%
 %
 \begin{minipage}{0.50\linewidth}
-\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_fastest_godutch.png}\end{minipage}%
+\includegraphics[width=2.60417in,height=\textheight]{newbridge_fastest_godutch.png}\end{minipage}%
 
@@
 `MapLibre' JavaScript package and `PMTiles' ``cloud native'' vector tile
-format \citep{gonçalves2023}. The default view in landing page is of a
-map of Ireland with county level results for the \% trips that could be
-made by cycling under the `Go Dutch' scenario, highlighting the intended
-use case for strategic cycle network planning and level of ambition. The
-key results, route network results, are available within an instant by
-zooming in on the map illustrated in Figure~\ref{fig-landing}: beyond a
-certain zoom level the network view appears. This differs from the user
-interface in tools such as the PCT and BiclaR which force users to click
-on a county before the key network results become visible: to get to the
-network view in the PCT takes around 10 seconds and multiple clicks,
-compared with simply zooming in on the landing page for CRUSE.
+format \citep{gonçalves2023}. These open source technologies were used
+to generate default view in the landing page: a map of Ireland with
+county level results for the \% trips that could be made by cycling
+under the `Go Dutch' scenario, highlighting the intended use case for
+strategic cycle network planning and level of ambition, as illustrated
+in Figure~\ref{fig-landing}.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{paste-10.png}
+
+\includegraphics{paste-11.png}
+
+}
+
+\caption{\label{fig-landing}Illustration of the landing page of the
+CRUSE Tool for Ireland, showing the default `zoomed out' view of the
+national map, with county level results showing the proportion of trips
+cycled under the `Go Dutch' scenario (top). The default `zoomed out'
+view enables users to click on a county for more detailed results. The
+`zoomed in' view shows the national network view for the `Go Dutch'
+scenario for the Limerick area (bottom).}
+
+\end{figure}%
+
+The key results, route network results, are available within an instant
+by zooming in on the map illustrated in Figure~\ref{fig-landing}: beyond
+a certain zoom level the network view appears. This differs from the
+user interface in tools such as the PCT and BiclaR which force users to
+click on a county before the key network results become visible: to get
+to the network view in the PCT takes around 10 seconds and multiple
+clicks, compared with simply zooming in on the landing page for CRUSE.
 
@@
 Figure~\ref{fig-landing}.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{images/paste-10.png}
-
-}
-
-\caption{\label{fig-landing}Illustration of the landing page of the
-CRUSE Tool for Ireland.}
-
-\end{figure}%
 
@@
 
-\includegraphics[width=3.31in,height=\textheight]{images/fig5.png}
-
 \begin{figure}
@@
 
-\includegraphics{images/fig5.png}
+\includegraphics{fig-user-stories-1.pdf}
 
@@
 
-To illustrate the results in urban areas, we took the top 4 cities in
-Ireland by population: Dublin City, Cork, Limerick and Galway (with city
-populations of 0.6, 0.2, 0.1 and 0.1 million respectively). In
-Figure~\ref{fig-city-results} we present the cycle route networks within
-a 3km radius circle centred on each of these cities.
+To illustrate the results in urban areas, we explored route networks in
+the three largest cities in Ireland: Dublin, Cork, and Limerick.
+Figure~\ref{fig-city-results} shows the `fastest' and `quietest' cycle
+networks within a 6km wide square centred on each of these cities.
 
@@
 
-\caption{\label{fig-city-results}Fastest, balanced and quietest cycle
-networks within 3km circular areas containing the four largest cities in
-Ireland. The networks are coloured by cycle friendliness, with greener
-segments representing more cycle friendly segments. Widths vary in
-proportion to the potential level of cycling under the `Go Dutch'
-scenario.}
+\caption{\label{fig-city-results}Fastest (left) and quietest (right)
+cycle networks in 6 km squares centered on central Dublin (top panels)
+Cork (middle) and Limerick (bottom). The networks are coloured by cycle
+friendliness, with greener segments representing more cycle friendly
+segments. Widths vary in proportion to the potential level of cycling
+under the `Go Dutch' scenario.}
 
@@
 
-Aside from the widely varying sizes and route networks of each city,
-with Dublin having by far the densest network, it is clear from the
-visualisation of cycling friendliness that there are many gaps in the
-networks. Dublin has had received most investment in cycling
-infrastructure and this is apparent from the high proportion of the
-route network that is green, even in the fastest route network
-representing routes that prioritize speed and directness over comfort
-and safety. Dublin also has the highest mode share of cycling of the
-four cities, adding to the evidence base showing that investment in
-cycling infrastructure is effective in increasing cycling uptake.
-
-Although Dublin City has made progress, there are still many parts of
-the fastest route network, and even some parts of the Balanced and
-Quietest route networks, that are not cycle friendly and which have high
-cycling potential. According to a recent report, 71\% of residents in
-the Dublin Metropolitan Area support ``more cycle tracks along roads,
-physically separated from traffic and pedestrians'' \citep{walking2021}.
-The results for the Dublin area could help prioritise investment in
-cycling infrastructure in the city.
+Aside from the varying sizes and route network shapes associated with
+each city, it is clear from the visualisation of cycling friendliness
+that none of the cities can be considered to have a `complete' cycle
+network. Of the three, Dublin has clearly had most investment cycle
+infrastructure, as shown by the relatively high proportion of the route
+network that is green in both network types.
+
+Figure~\ref{fig-city-results} highlights the value of generating
+different network types: looking only at the fastest network it would be
+hard to differentiate between Cork and Limerick in terms of cycle
+network quality. However, Limerick clearly has a higher number of quiet
+routes that can be cycled on for people willing to take a longer route
+to avoid busy roads: its quiet network (bottom right in
+Figure~\ref{fig-city-results}) contains more green segments than the
+quietest network for Cork (middle right in
+Figure~\ref{fig-city-results}). The lack of continuous sections of quiet
+routes in Cork, even in the quietest network, means that people who are
+less confident cycling \emph{cannot} avoid mixing with motor traffic for
+many common trips. In an ideal network, conversely, even the fastest
+route network would be predominantly cycle friendly, meaning that the
+most direct and fast routes are also quiet, with low levels of traffic
+stress.
+
+There are still many parts of the fastest route network, and even some
+parts of the quietest route networks, that are not cycle friendly and
+which have high cycling potential, even in Dublin. According to a recent
+report, 71\% of residents in the Dublin Metropolitan Area support ``more
+cycle tracks along roads, physically separated from traffic and
+pedestrians'' \citep{walking2021}. Assuming that these findings apply in
+other cities, the results can help prioritise where such infrastructure
+should go: the results presented in Figure~\ref{fig-city-results} should
+bring attention to where there are wide red lines in fastest and
+quietest networks, with a reasonable strategy being to prioritise a
+continuous quiet network first and eventually to `cycle proof' the
+fastest network also. However, a feature of the approach is that it is
+not prescriptive and it does not tell planners what to build where; as
+highlighted below practitioners can use the results in a variety of
+ways.
 
@@
 
-While the CRUSE Tool is not the first national and publicly available
-tool for cycling planning, it has some key features that make it
-relevant for other countries, regions and road authorities tasked with
-making their transport systems safe and sustainable:
+While the CRUSE approach presented in this paper is not the first
+national and publicly available tool for cycling planning, it has some
+key features that make it relevant for other countries, regions and road
+authorities tasked with making their transport systems safe and
+sustainable:
 
@@
 \tightlist
+\item
+  Compared with the Propensity to Cycle Tool (PCT), which provides
+  either travel to work \citep{lovelace2017} or to school cycling
+  networks \citep{goodman2019}, the approach presented here captures a
+  higher proportion of cycling potential including key rural trips which
+  are often under-represented in models of active travel.
 \item
@@
 \item
-  By covering a wide range of trip purposes --- not just travel to work
-  \citep{lovelace2017, heinen2010} and travel to school
-  \citep{goodman2019} as covered in previously published research on
-  open access national cycle network planning tools --- the results
-  capture a higher proportion of cycling potential including key rural
-  trips which are often under-represented in models of active travel.
+  The tool has a user interface that allows people to see the key
+  results quickly, without having to click on regional result pages, as
+  is the case with the BiclaR and PCT projects.
 \end{itemize}
@@
 As the benefits of strategic planning for cycling become more apparent
-\citep{scappini2022}, we expect the demand for tools such as CRUSE to
-grow. However, the approach is not without limitations and does not meet
-all requirements in Ireland or indeed any country where the methods are
-applied.
+\citep{scappini2022}, we expect the demand for the approach outlined in
+this paper, and resulting tools, to grow. However, the approach is not
+without limitations and does not meet all requirements in Ireland or
+indeed any country where the methods are applied.
 
@@
 key part of the the European RISM Directive, for reporting collision
-rates of vulnerable road users by 2024. These factors mean that the tool
-can be seen as an open access `leverage point', providing key
-information for and supporting ambitious plans in many aspects the
-planning system, from network design to post-build monitoring
-\citep{lovelace2020}.
-
-\section{Conclusions}\label{sec-conclusions}
-
-The CRUSE Tool is an open access web application for strategic cycle
-network planning and prioritization of road safety interventions across
-Ireland. Building on previous work, it provides a nationally consistent
-evidence base, providing valuable insights to planners and other
-stakeholders in the transport planning process, at national and local
-levels. Because the results are available at the route segment level,
-the tool can be used to identify `weak links' in the cycle network
-\citep{vybornova2022}, and to prioritize investment in cycling
-infrastructure to maximize health, equality, and other benefits
-\citep{mahfouz, woodcock2021}. Furthermore, the provision of the
-evidence in a free and publicly available website, hosted at
-\href{https://cruse.bike}{cruse.bike}, means that it can be used by
-anyone, encouraging wider participation and more evidence-based debate
-about transport planning.
-
-A key feature of the project methodologically is its calculation of
-current and future potential not only for travel to work and travel to
-school, but also for other trip purposes, including recreational trips.
-This required the development of spatial interaction models and
-estimation of the relative attractiveness of different destinations for
-different trip purposes, an area of active research where new
-developments could be incorporated into the tool in future
-\citep{hasova2022}.
-
-The CRUSE Tool and the underlying research and methods are not without
-limitations, suggesting future areas of research, data collection needs,
-and policy application. The route network level results have not been
-validated to the extent we had planned at the outset of the project. We
-compared route network level results with `ground truth' data from cycle
-counter datasets across Dublin to test different network types and to
-ensure that additional trip types added to work and educational trips
-increased the quality of fit under the baseline scenario. However, the
-size of the counter network was insufficient to provide an opportunity
-for robust evaluation of model performance, suggesting a combined
-program of new count data collection and analysis should be a priority
-for future work, with findings feeding directly into work to improve the
-outputs of the the CRUSE Tool.
+rates of vulnerable road users by 2024. Regular updates to the model to
+generate yearly baseline cycle flow estimates could tackle this
+limitation, although would take substantial resources.
 
@@
 
-The CRUSE Tool is already used in practice to support more ambitious and
-data-driven planning for safe cycling routes in multiple counties across
-Ireland. We hope that the underlying approach, and the publicly
-available evidence provided in the CRUSE Tool itself, provide a basis
-for future reproducible research, open source software development into
-strategic cycle network planning tools in Ireland and other countries.
-Active travel represents a win-win-win for physical activity,
-environment, and local economic opportunities. In combination with
-broader sustainable mobility measures and policies to reduce motor
-traffic speeds and volumes, tools such as CRUSE can support the fast and
-fair decarbonisation of transport systems worldwide.
+\section{Conclusions}\label{sec-conclusions}
+
+The CRUSE Tool is an open access web application for strategic cycle
+network planning and prioritization of road safety interventions across
+Ireland. Building on previous work, it provides a nationally consistent
+evidence base, providing valuable insights to planners and other
+stakeholders in the transport planning process, at national and local
+levels. Because the results are available at the route segment level,
+the tool can be used to identify `weak links' in the cycle network
+\citep{vybornova2022}, and to prioritize investment in cycling
+infrastructure to maximize health, equality, and other benefits
+\citep{mahfouz, woodcock2021}. Furthermore, the provision of the
+evidence in a free and publicly available website, hosted at
+\href{https://cruse.bike}{cruse.bike}, means that it can be used by
+anyone, encouraging wider participation and more evidence-based debate
+about transport planning.
+
+A key feature of the project methodologically is its calculation of
+current and future potential not only for travel to work and travel to
+school, but also for other trip purposes, including recreational trips.
+This required the development of spatial interaction models and
+estimation of the relative attractiveness of different destinations for
+different trip purposes, an area of active research where new
+developments could be incorporated into the tool in future
+\citep{hasova2022}.
+
+As outlined in Section~\ref{sec-qualitative}, the tool is already used
+in practice to support more ambitious and data-driven planning for safe
+cycling routes in multiple counties across Ireland. We hope that the
+underlying approach provides a basis for future research and
+development. In combination with broader sustainable mobility measures
+and policies to reduce motor traffic speeds and volumes, evidence
+generated by the approach outlined in this paper can support the fast
+and fair decarbonisation of transport systems worldwide.
 

### 2024-03-27 paper.qmd

@@
 pdf-engine: pdflatex
+knitr:
+  opts_chunk:
+    fig.path: images/
 format: 
-  html: default
+  # html: default
   # docx: default
 # # # Uncomment for formatted version
-  # elsevier-pdf:
-  #   linestretch: 2
-  #   keep-tex: true
-  #   journal:
-  #     name: European Transport Research Review
-  #     formatting: preprint
-  #     model: 3p
-  #     cite-style: super
+  elsevier-pdf:
+    linestretch: 2
+    keep-tex: true
+    journal:
+      name: European Transport Research Review
+      formatting: preprint
+      model: 3p
+      cite-style: super
 # # # #   # pdf: default
 
-# # # Uncomment the following line to produce the cover page
-# author:
-#   - name: Robin Lovelace
-#     email: r.lovelace@leeds.ac.uk
-#     corresponding: true
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Joey Talbot
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Eugeni Vidal-Tortosa
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Hussein Mahfouz
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Elaine Brick
-#     affiliations:
-#       - name: AECOM
-#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-#   - name: Peter Wright
-#     affiliations:
-#       - name: AECOM
-#         address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
-#   - name: "Gary O'Toole"
-#     affiliations:
-#       - name: AECOM
-#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-#   - name: Dan Brennan
-#     affiliations:
-#       - name: Transport Infrastructure Ireland
-#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
-#   - name: Suzanne Meade
-#     affiliations:
-#       - name: Transport Infrastructure Ireland
-#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+# # Uncomment the following line to produce the cover page
+author:
+  - name: Robin Lovelace
+    email: r.lovelace@leeds.ac.uk
+    corresponding: true
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Joey Talbot
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Eugeni Vidal-Tortosa
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Hussein Mahfouz
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Elaine Brick
+    affiliations:
+      - name: AECOM
+        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+  - name: Peter Wright
+    affiliations:
+      - name: AECOM
+        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
+  - name: "Gary O'Toole"
+    affiliations:
+      - name: AECOM
+        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+  - name: Dan Brennan
+    affiliations:
+      - name: Transport Infrastructure Ireland
+        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+  - name: Suzanne Meade
+    affiliations:
+      - name: Transport Infrastructure Ireland
+        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
 abstract: |
@@
 ::: {#fig-dublin}
-<!-- ![](paste-1.png) -->
+<!-- ![](images/paste-1.png) -->
 
-![](paste-8.png) <!-- ![](paste-3.png) -->
-![](paste-9.png)
+![](images/paste-8.png) <!-- ![](images/paste-3.png) -->
+![](images/paste-9.png)
 
@@
 
-<!-- ![](paste-5.png) -->
+<!-- ![](images/paste-5.png) -->
 
-<!-- ![](paste-6.png) -->
+<!-- ![](images/paste-6.png) -->
 
@@
 ::: {#fig-route-types layout-nrow="2"}
-![](naas_quietest_godutch.png){width="250"}
+![](images/naas_quietest_godutch.png){width="250"}
 
-![](naas_fastest_godutch.png){width="250"}
+![](images/naas_fastest_godutch.png){width="250"}
 
-![](newbridge_quietest_godutch.png){width="250"}
+![](images/newbridge_quietest_godutch.png){width="250"}
 
-![](newbridge_fastest_godutch.png){width="250"}
+![](images/newbridge_fastest_godutch.png){width="250"}
 
@@
 ::: {#fig-scenarios}
-<!-- ![](paste-7.png) -->
+<!-- ![](images/paste-7.png) -->
 
-![](scenarios_kildare.png)
+![](images/scenarios_kildare.png)
 
@@
 ::: {#fig-landing}
-![](paste-10.png)
+![](images/paste-10.png)
 
-![](paste-11.png)
+![](images/paste-11.png)
 
@@
 # schematic
-# htmlwidgets::saveWidget(schematic, "fig5.html")
-# browseURL("fig5.html")
-# webshot2::webshot(url = "fig5.html", file = "fig5.png")
-# knitr::include_graphics("fig5.png")
+# htmlwidgets::saveWidget(schematic, "images/fig5.html")
+# browseURL("images/fig5.html")
+# webshot2::webshot(url = "images/fig5.html", file = "images/fig5.png")
+# knitr::include_graphics("images/fig5.png")
 schematic
@@
 <!-- ::: {#fig-user-stories}
-![](fig5.png)
+![](images/fig5.png)
 
@@
   tm_lines(col = "Quietness", palette = "RdYlGn", lwd = 5)
-tmap_save(m, "./rnet_fastest_city_top.png", width = 10, height = 10)
+tmap_save(m, "./images/rnet_fastest_city_top.png", width = 10, height = 10)
 # Generalisation to balanced and quiestest route types:
@@
 m_types
-tmap_save(m_types, "./rnet_types.png", width = 12, height = 16)
-browseURL("./rnet_types.png")
+tmap_save(m_types, "./images/rnet_types.png", width = 12, height = 16)
+browseURL("./images/rnet_types.png")
 ```
@@
 ::: {#fig-city-results}
-![](rnet_types.png)
+![](images/rnet_types.png)
 

### 2024-03-27 paper.tex

@@
   pdftitle={Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools},
+  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Toole; Dan Brennan; Suzanne Meade},
   pdfkeywords={Cycling, Open-source, Road Safety, Active
@@
 for developing strategic cycle network planning tools}
+\author[1]{Robin Lovelace%
+\corref{cor1}%
+}
+ \ead{r.lovelace@leeds.ac.uk} 
+\author[1]{Joey Talbot%
+%
+}
+
+\author[1]{Eugeni Vidal-Tortosa%
+%
+}
+
+\author[1]{Hussein Mahfouz%
+%
+}
+
+\author[2]{Elaine Brick%
+%
+}
+
+\author[3]{Peter Wright%
+%
+}
+
+\author[2]{Gary O'Toole%
+%
+}
+
+\author[4]{Dan Brennan%
+%
+}
+
+\author[4]{Suzanne Meade%
+%
+}
+
+\affiliation[1]{organization={University of
+Leeds},addressline={Institute for Transport Studies, Leeds, LS2 9JT,
+United Kingdom},postcodesep={}}
+\affiliation[2]{organization={AECOM},addressline={Unit 6, Galway
+Technology Park, Parkmore, Galway, Ireland},postcodesep={}}
+\affiliation[3]{organization={AECOM},addressline={Winslade House,
+Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY,
+UK},postcodesep={}}
+\affiliation[4]{organization={Transport Infrastructure
+Ireland},addressline={Parkgate Business Centre, Parkgate Street, Dublin
+8, D08 DK10, Ireland},postcodesep={}}
 
@@
 
-\includegraphics{paste-8.png} \includegraphics{paste-9.png}
+\includegraphics{images/paste-8.png}
+\includegraphics{images/paste-9.png}
 
@@
 \begin{minipage}{0.50\linewidth}
-\includegraphics[width=2.60417in,height=\textheight]{naas_quietest_godutch.png}\end{minipage}%
+\includegraphics[width=2.60417in,height=\textheight]{images/naas_quietest_godutch.png}\end{minipage}%
 %
 \begin{minipage}{0.50\linewidth}
-\includegraphics[width=2.60417in,height=\textheight]{naas_fastest_godutch.png}\end{minipage}%
+\includegraphics[width=2.60417in,height=\textheight]{images/naas_fastest_godutch.png}\end{minipage}%
 \newline
 \begin{minipage}{0.50\linewidth}
-\includegraphics[width=2.60417in,height=\textheight]{newbridge_quietest_godutch.png}\end{minipage}%
+\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_quietest_godutch.png}\end{minipage}%
 %
 \begin{minipage}{0.50\linewidth}
-\includegraphics[width=2.60417in,height=\textheight]{newbridge_fastest_godutch.png}\end{minipage}%
+\includegraphics[width=2.60417in,height=\textheight]{images/newbridge_fastest_godutch.png}\end{minipage}%
 
@@
 
-\includegraphics{scenarios_kildare.png}
+\includegraphics{images/scenarios_kildare.png}
 
@@
 
-\includegraphics{paste-10.png}
-
-\includegraphics{paste-11.png}
+\includegraphics{images/paste-10.png}
+
+\includegraphics{images/paste-11.png}
 
@@
 
-\includegraphics{fig-user-stories-1.pdf}
+\includegraphics{images/fig-user-stories-1.pdf}
 
@@
 
-\includegraphics{rnet_types.png}
+\includegraphics{images/rnet_types.png}
 

### 2024-03-27 paper.qmd
Try html

@@
 format: 
-  # html: default
+  html: default
   # docx: default
 # # # Uncomment for formatted version
-  elsevier-pdf:
-    linestretch: 2
-    keep-tex: true
-    journal:
-      name: European Transport Research Review
-      formatting: preprint
-      model: 3p
-      cite-style: super
+  # elsevier-pdf:
+  #   linestretch: 2
+  #   keep-tex: true
+  #   journal:
+  #     name: European Transport Research Review
+  #     formatting: preprint
+  #     model: 3p
+  #     cite-style: super
 # # # #   # pdf: default

### 2024-04-17 paper.qmd
HTML
Update paper

@@
 ---
+
+Note: preprint now available here: https://osf.io/preprints/osf/a72yh
+
+Please cite as
+
+Lovelace, Robin, Joey Talbot, Eugeni Vidal-Tortosa, Hussein Mahfouz, Elaine Brick, Peter Wright, Gary O'Toole, et al.
+2024.
+“Cycle Route Uptake and Scenario Estimation (CRUSE): An Approach for Developing Strategic Cycle Network Planning Tools.” OSF Preprints.
+March 29.
+doi:10.31219/osf.io/a72yh.
 
@@
 #| label: tbl-tools
-# For PDF:
-knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
-  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
-  kableExtra::column_spec(1, width = "13em") |>
-  kableExtra::column_spec(2, width = "15em") |>
-  kableExtra::column_spec(3, width = "12em") |>
-  kableExtra::column_spec(4, width = "5em")
+# # For PDF:
+# knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
+#   kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
+#   kableExtra::column_spec(1, width = "13em") |>
+#   kableExtra::column_spec(2, width = "15em") |>
+#   kableExtra::column_spec(3, width = "12em") |>
+#   kableExtra::column_spec(4, width = "5em")
 # For HTML:
-# knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.")
+knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.")
 ```
@@
 # For PDF:
-knitr::kable(trip_purposes, booktabs = TRUE) |>
-  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE, font_size = 8) |>
-  kableExtra::column_spec(1, width = "5em") |>
-  kableExtra::column_spec(2, width = "12em") |>
-  kableExtra::column_spec(3, width = "15em") |>
-  kableExtra::column_spec(4, width = "15em")
+# knitr::kable(trip_purposes, booktabs = TRUE) |>
+#   kableExtra::kable_styling(latex_options = "striped", full_width = FALSE, font_size = 8) |>
+#   kableExtra::column_spec(1, width = "5em") |>
+#   kableExtra::column_spec(2, width = "12em") |>
+#   kableExtra::column_spec(3, width = "15em") |>
+#   kableExtra::column_spec(4, width = "15em")
 # For HTML:
-# knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.")
+knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.")
 ```

### 2024-04-17 paper.qmd
Tidyverse
Add more details
Minor updates

@@
 
+<details>
+
 Please cite as
@@
 
+This is an open and reproducible paper.
+See [github.com/cruse-bike](https://github.com/cruse-bike) for the code and data used to generate the results presented here, and the code underlying the web app hosted at [cruse.bike](https://cruse.bike/).
+
+</details>
+
 {{< pagebreak >}}
@@
 #| include: false
-library(tidyverse)
+library(readr)
+library(dplyr)
 ```
@@
 #| label: fig-user-stories
-schematic = DiagrammeR::grViz("
-digraph {
-  # graph statement
-  #################
-  graph [rankdir = LR
-         color = white
-         fontsize = 36]
+# schematic = DiagrammeR::grViz("
+# digraph {
+#   # graph statement
+#   #################
+#   graph [rankdir = LR
+#          color = white
+#          fontsize = 36]
 
-  # nodes (rectangles)
-  #################
-   node [shape = rectangle            
-       margin = 0.4
-       fontsize = 36
-       fixedsize = false
-       width = 1]
+#   # nodes (rectangles)
+#   #################
+#    node [shape = rectangle            
+#        margin = 0.4
+#        fontsize = 36
+#        fixedsize = false
+#        width = 1]
 
-   subgraph cluster_2 {
-     graph[shape = rectangle]
-         C2b [label = 'Exploration']
-         C2a [label = 'CRUSE\ncounty level results']
-         C2 [label = 'CRUSE\nnational map']
-         B2 [label = 'CRUSE\nlanding page']
-         A2 [label = 'Non-professional\nuser'] 
-}
+#    subgraph cluster_2 {
+#      graph[shape = rectangle]
+#          C2b [label = 'Exploration']
+#          C2a [label = 'CRUSE\ncounty level results']
+#          C2 [label = 'CRUSE\nnational map']
+#          B2 [label = 'CRUSE\nlanding page']
+#          A2 [label = 'Non-professional\nuser'] 
+# }
 
-   subgraph cluster_1 {
-   graph[shape = rectangle]
-         Dd [label = 'Data dowloads']
-         Dc [label = 'Quiet and fast route\nnetworks']
-         Db [label = 'County statistics']
-         Da [label = 'Full county map']
-         D [label = 'CRUSE\ncounty map']
-         C [label = 'CRUSE\nnational map']
-         B [label = 'CRUSE\nlanding page']
-         A [label = 'Professional user'] 
-}
+#    subgraph cluster_1 {
+#    graph[shape = rectangle]
+#          Dd [label = 'Data dowloads']
+#          Dc [label = 'Quiet and fast route\nnetworks']
+#          Db [label = 'County statistics']
+#          Da [label = 'Full county map']
+#          D [label = 'CRUSE\ncounty map']
+#          C [label = 'CRUSE\nnational map']
+#          B [label = 'CRUSE\nlanding page']
+#          A [label = 'Professional user'] 
+# }
 
-  # node (triangle)
-  #################
-  node [shape = diamond]
-         Dd1 [label = 'External\ntools']
-         C2b1 [label = 'Communication']
-  # edges 
-  #######
-  A2 -> B2
-  A -> B
-  B -> C
-  B2 -> C2
-  C -> D
-  C2 -> C2a
-  C2 -> C2b
-  C2a -> C2b
-  C2b -> C2b1
-  D -> Da 
-  D -> Db 
-  D -> Dc 
-  D -> Dd 
-  Dd -> Dd1
+#   # node (triangle)
+#   #################
+#   node [shape = diamond]
+#          Dd1 [label = 'External\ntools']
+#          C2b1 [label = 'Communication']
+#   # edges 
+#   #######
+#   A2 -> B2
+#   A -> B
+#   B -> C
+#   B2 -> C2
+#   C -> D
+#   C2 -> C2a
+#   C2 -> C2b
+#   C2a -> C2b
+#   C2b -> C2b1
+#   D -> Da 
+#   D -> Db 
+#   D -> Dc 
+#   D -> Dd 
+#   Dd -> Dd1
 
-}
-")
+# }
+# ")
 # schematic
@@
 # webshot2::webshot(url = "images/fig5.html", file = "images/fig5.png")
-# knitr::include_graphics("images/fig5.png")
-schematic
+knitr::include_graphics("images/fig5.png")
 ```

### 2024-06-15 paper.qmd

@@
 Developed by researchers and developers in academic, public and private sectors, they have evolved from simple area-based static models tailored to specific regions to more complex and (in some cases) more generalizable tools.
-These tools can be considered as a specific application of the concept of planning support systems (PSS) , which are designed to help planners and decision-makers to make more informed decisions [@geertman2009].
+These tools can be considered as a specific application of the concept of planning support systems (PSS), which are designed to help planners and decision-makers to make more informed decisions [@geertman2009].
 Outputs include maps with results available at several levels of analysis with levels of availability ranging from being only available to researchers and in static maps to open access mapping systems available to the public.
@@
 #| label: tbl-tools
+#| tbl-cap: "Summary of existing tools for estimating cycling potential."
 # # For PDF:
@@
 # For HTML:
-knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.")
+knitr::kable(tools)
 ```
@@
 
-Other tools that generate areal data to prioritize investment include a study that generated a 'prioritization index' for Montreal, Canada [@larsen_build_2013], a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil, and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
+Other tools that generate areal data to prioritize investment include a study that generated a<span style="color: blue;"> 'bikeability index' for the Metro Vancouver region [@winters_mapping_2013] and later called Bike Score and available for over 160 US and Canadian cities [@winters_bike_2016] -- its methodology is [publicly available](https://www.walkscore.com/bike-score-methodology.shtml); a </span>'prioritization index' for Montreal, Canada [@larsen_build_2013]; a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil [@zhang_prioritizing_2014]; and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
 The CPT is composed of the base environmental module — based on eight weighted factors (population density, hilliness, physical barriers, access to services, existing cycling mode share, distance to work and school, and road speed) — and the quality of service module, which evaluates each area of interest based on eight factors (surface condition, adjacent cyclists, comfort factor, conflict, distance between junctions, slope, access to services, and origin/destination).
 
-A more recent areal-based approach is the designed the Gross Potential for Cycling (GPC) tool to prioritize areas for cycling infrastructure and other cycling measures [@silva_gross_2021].
+A more recent areal-based approach is the designed Gross Potential for Cycling (GPC) tool to prioritize areas for cycling infrastructure and other cycling measures  [@silva_gross_2021].
 The GPC uses two sets of indicators: population-based indicators (age, potential demand density, employment density and motorization rate) and area-based indicators (accessibility, education, public transport, connectivity, land use mix and relative performance).
@@
 The ranking of areas by the GPC are [publicly available](https://boost.up.pt/en/ferramentas/gpc) and its practical value has been assessed through workshops [@silva_revealing_2022].
+<span style="color: blue;">The GPC tool has also been used to investigate the combination of factors contributing to high levels cycling potential [@lopes_impact_2021]. 
+Another recent areal-based approach is the Relative Equity Impact of Bicycle Planning (TIRE) [@cunha_assessing_2023]. 
+The TIRE assesses the impact of cycling network allocation on the accessibility levels of different socioeconomic groups.</span>
 
 Route-based tools emerged with increasing availability of origin-destination data, routing engines that can assign trips to networks, and improvements in computer hardware and software needed to generate route networks.
-A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace_mapping_2016].
+A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace2016].
 The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 [@lovelace2017].
@@
 biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
+
+<span style="color: blue;">Another route-based tool is The Cycling Routing Algorithm for Network Connectivity (CRANC) [@gehrke_cycling-focused_2020]. 
+The CRANC tool evaluates the benefits of new bicycle facilities for the accessibility of various populations and neighbourhoods. </span>
 
@@
 # For HTML:
-knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.")
+knitr::kable(trip_purposes)
 ```

### 2024-06-21 paper.qmd
Test commit

@@
 # # # #   # pdf: default
-
 # # Uncomment the following line to produce the cover page

### 2024-06-21 paper.qmd
Remove excess word

@@
 
-A more recent areal-based approach is the designed Gross Potential for Cycling (GPC) tool to prioritize areas for cycling infrastructure and other cycling measures  [@silva_gross_2021].
+A more recent areal-based approach is the Gross Potential for Cycling (GPC) tool to prioritize areas for cycling infrastructure and other cycling measures  [@silva_gross_2021].
 The GPC uses two sets of indicators: population-based indicators (age, potential demand density, employment density and motorization rate) and area-based indicators (accessibility, education, public transport, connectivity, land use mix and relative performance).

### 2024-06-21 paper.qmd
Add zoomed-in view to Figure 4

@@
 
-Lovelace, Robin, Joey Talbot, Eugeni Vidal-Tortosa, Hussein Mahfouz, Elaine Brick, Peter Wright, Gary O'Toole, et al.
-2024.
+Lovelace, Robin, Joey Talbot, Eugeni Vidal-Tortosa, Hussein Mahfouz, Elaine Brick, Peter Wright, Gary O'Toole, et al. 2024.
 “Cycle Route Uptake and Scenario Estimation (CRUSE): An Approach for Developing Strategic Cycle Network Planning Tools.” OSF Preprints.
@@
 
-Other tools that generate areal data to prioritize investment include a study that generated a<span style="color: blue;"> 'bikeability index' for the Metro Vancouver region [@winters_mapping_2013] and later called Bike Score and available for over 160 US and Canadian cities [@winters_bike_2016] -- its methodology is [publicly available](https://www.walkscore.com/bike-score-methodology.shtml); a </span>'prioritization index' for Montreal, Canada [@larsen_build_2013]; a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil [@zhang_prioritizing_2014]; and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
+Other tools that generate areal data to prioritize investment include a study that generated a[ 'bikeability index' for the Metro Vancouver region [@winters_mapping_2013] and later called Bike Score and available for over 160 US and Canadian cities [@winters_bike_2016] -- its methodology is [publicly available](https://www.walkscore.com/bike-score-methodology.shtml); a ]{style="color: blue;"}'prioritization index' for Montreal, Canada [@larsen_build_2013]; a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil [@zhang_prioritizing_2014]; and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
 The CPT is composed of the base environmental module — based on eight weighted factors (population density, hilliness, physical barriers, access to services, existing cycling mode share, distance to work and school, and road speed) — and the quality of service module, which evaluates each area of interest based on eight factors (surface condition, adjacent cyclists, comfort factor, conflict, distance between junctions, slope, access to services, and origin/destination).
@@
 The ranking of areas by the GPC are [publicly available](https://boost.up.pt/en/ferramentas/gpc) and its practical value has been assessed through workshops [@silva_revealing_2022].
-<span style="color: blue;">The GPC tool has also been used to investigate the combination of factors contributing to high levels cycling potential [@lopes_impact_2021]. 
+[The GPC tool has also been used to investigate the combination of factors contributing to high levels cycling potential [@lopes_impact_2021].
 Another recent areal-based approach is the Relative Equity Impact of Bicycle Planning (TIRE) [@cunha_assessing_2023].
-The TIRE assesses the impact of cycling network allocation on the accessibility levels of different socioeconomic groups.</span>
+The TIRE assesses the impact of cycling network allocation on the accessibility levels of different socioeconomic groups.]{style="color: blue;"}
 
@@
 
-<span style="color: blue;">Another route-based tool is The Cycling Routing Algorithm for Network Connectivity (CRANC) [@gehrke_cycling-focused_2020]. 
-The CRANC tool evaluates the benefits of new bicycle facilities for the accessibility of various populations and neighbourhoods. </span>
+[Another route-based tool is The Cycling Routing Algorithm for Network Connectivity (CRANC) [@gehrke_cycling-focused_2020].
+The CRANC tool evaluates the benefits of new bicycle facilities for the accessibility of various populations and neighbourhoods.
+]{style="color: blue;"}
 

### 2024-06-21 paper.qmd
Describe qualitative approach

@@
 
-Other tools that generate areal data to prioritize investment include a study that generated a[ 'bikeability index' for the Metro Vancouver region [@winters_mapping_2013] and later called Bike Score and available for over 160 US and Canadian cities [@winters_bike_2016] -- its methodology is [publicly available](https://www.walkscore.com/bike-score-methodology.shtml); a ]{style="color: blue;"}'prioritization index' for Montreal, Canada [@larsen_build_2013]; a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil [@zhang_prioritizing_2014]; and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
+Other tools that generate areal data to prioritize investment include a study that generated a['bikeability index' for the Metro Vancouver region [@winters_mapping_2013] and later called Bike Score and available for over 160 US and Canadian cities [@winters_bike_2016] -- its methodology is [publicly available](https://www.walkscore.com/bike-score-methodology.shtml); a]{style="color: blue;"}'prioritization index' for Montreal, Canada [@larsen_build_2013]; a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil [@zhang_prioritizing_2014]; and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
 The CPT is composed of the base environmental module — based on eight weighted factors (population density, hilliness, physical barriers, access to services, existing cycling mode share, distance to work and school, and road speed) — and the quality of service module, which evaluates each area of interest based on eight factors (surface condition, adjacent cyclists, comfort factor, conflict, distance between junctions, slope, access to services, and origin/destination).
@@
 [Another route-based tool is The Cycling Routing Algorithm for Network Connectivity (CRANC) [@gehrke_cycling-focused_2020].
-The CRANC tool evaluates the benefits of new bicycle facilities for the accessibility of various populations and neighbourhoods.
-]{style="color: blue;"}
+The CRANC tool evaluates the benefits of new bicycle facilities for the accessibility of various populations and neighbourhoods.]{style="color: blue;"}
 
@@
 ```
+## Stakeholder engagement {#sec-stakeholder-engagement}
+
+The CRUSE Tool was developed for, and in close collaboration with, practitioners in Ireland.
+Although the overall approach of the project was quantitative, stakeholder feedback and qualitative data was deemed essential to ensure the tool was fit for purpose, mitigating against the risk that the project would be 'technocentric' and a 'problem looking for a solution'.
+From the outset of the project, we aimed to follow @coppola2022's advocacy for a "human-centric approach in which personalization, purpose, ethics and inclusivity are key", rather than an 'off-the-shelf' product that failed to understand, let alone meet, local needs.
+As outlined in the overview of papers in the collection Inclusive and Collaborative Transport cited above, the 'digital transformation' taking place in transport planning (and many other sectors) has the potential to increase existing inequalities [@coppola2022].
+A wide range of methods and measure can be taken to ensure that digital tools are inclusive \[...\].
+
+<!-- Details from final workshop:
+
+Can you please join me and my colleagues from TII and ITS Leeds for 1 hour on Thursday the 7th December at 12:00 – 13:00 for a workshop and demonstration of the open source CRUSE tool) which is used primarily to highlight the cycle friendliness of roads in Ireland, avail of cycle scenario/trip data and highlight areas to target for active travel investment.
+This tool is free to use for transport planners or member of the general public and can be found here: CRUSE .
+We have now added recreational cycling trips to the CRUSE tool and would welcome the chance to present this work and ask for feedback.
+
+Please see below a draft agenda for the workshop till 1pm:
+
+-   Introduction (5 min)
+-   Changes to CRUSE: extension work (15 min)
+-   Testing new website (10 min)
+-   Feedback, Q&A + discussion (30 min)
+    - General feedback on new site
+    - Feedback on the recreational outputs
+    - Feedback on the tourism network -->
+
+<!-- – What are main issues you face in
+understanding infrastructure?, e.g.
+– What is happening at the moment?
+– Where to invest in the future?
+– What will change?
+– How will planning vary depending on trip purpose?
+– What sort of scenarios interested in testing?
+– How useful is having different networks for
+planning infrastructure and/or more detail,
+e.g. there could be an option for variation in
+access for professionals and public (the latter
+remaining simple) -->
+
+Due to time and skill constraints in the team developing the CRUSE tool (we lacked researchers with a background in qualitative methods), we opted for the well-tested and cost-effective approach of holding workshops with stakeholders in case study area, an approach that is documented with reference to participatory development of digital tools for supporting energy transitions [@hewitt2020].
+We held a total of three workshops over the course of the project and one open-ended live demo during a 2-day event in Sligo in which we collected feedback from dozens of people.
+The first two workshops took place in May 2022, in which key staff involved in cycle network planning in Kildare and Limerick counties were presented the tool.
+The county-specific approach allowed the workshop to focus on specific areas and key corridors.
+The local transport planners were asked to provide feedback in a semi-structured interview approach, with questions on the evidence they currently have access, how they may be able to use the tool, and requests for new features.
+
+The third and final workshop took place on the 7^th^ December 2023 and included descriptions of recent changes to the tool, with an aim being to get feedback on the new recreational and cycle touring evidence, with representatives from county-level transport planning bodies, Sport Ireland, and Waterways Ireland.
+The open-ended live demo was undertaken at a stall at TII’s National Roads and Greenways Conference in Sligo in September 2022, where much of the feedback was verbal and not collected.
+Much of this feedback from the workshops and live demo outlined above was in the form of ad-hoc comments and suggestions and not all of the qualitative data received was recorded in a structured way, a limitation that we refer back to in @sec-discussion.
+
 # Results {#sec-results}
@@
 
-## Stakeholder engagement and feedback {#sec-qualitative}
+## Stakeholder feedback {#sec-qualitative}
 
@@
 
+A final limitation of the approach relates to *how* it was developed.
+As described in @sec-stakeholder-engagement, the development of the tool was participatory, but only insofar as people who attended our workshops could provide feedback.
+The people who did provide feedback were deliberately chosen as the target audience.
+In future work we would like to collect feedback from a wider range of people, including campaigners, service providers (e.g.
+staff involved in travel planning for schools and new developments), and people who not involved in the planning process.
+
 # Conclusions {#sec-conclusions}

### 2024-06-23 paper.qmd
Workshops took place online
mark and minor changes to robin changes

@@
 
-Other tools that generate areal data to prioritize investment include a study that generated a['bikeability index' for the Metro Vancouver region [@winters_mapping_2013] and later called Bike Score and available for over 160 US and Canadian cities [@winters_bike_2016] -- its methodology is [publicly available](https://www.walkscore.com/bike-score-methodology.shtml); a]{style="color: blue;"}'prioritization index' for Montreal, Canada [@larsen_build_2013]; a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil [@zhang_prioritizing_2014]; and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
+Other tools that generate areal data to prioritize investment include a study that generated a ['bikeability index' for the Metro Vancouver region [@winters_mapping_2013], later renamed Bike Score and available for over 160 US and Canadian cities [@winters_bike_2016] -- its methodology is [openly accessible](https://www.walkscore.com/bike-score-methodology.shtml); a]{style="color: blue;"} 'prioritization index' for Montreal, Canada [@larsen_build_2013]; a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil [@zhang_prioritizing_2014]; and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
 The CPT is composed of the base environmental module — based on eight weighted factors (population density, hilliness, physical barriers, access to services, existing cycling mode share, distance to work and school, and road speed) — and the quality of service module, which evaluates each area of interest based on eight factors (surface condition, adjacent cyclists, comfort factor, conflict, distance between junctions, slope, access to services, and origin/destination).
@@
 [The GPC tool has also been used to investigate the combination of factors contributing to high levels cycling potential [@lopes_impact_2021].
-Another recent areal-based approach is the Relative Equity Impact of Bicycle Planning (TIRE) [@cunha_assessing_2023].
-The TIRE assesses the impact of cycling network allocation on the accessibility levels of different socioeconomic groups.]{style="color: blue;"}
+Another recent areal-based approach is the Relative Equity Impact of Bicycle Planning (TIRE) [@cunha_assessing_2023] tool.
+The TIRE tool assesses the impact of cycling network allocation on the accessibility levels of different socioeconomic groups.]{style="color: blue;"}
 
@@
 ```
-## Stakeholder engagement {#sec-stakeholder-engagement}
 
-The CRUSE Tool was developed for, and in close collaboration with, practitioners in Ireland.
+## [Stakeholder engagement]{style="color: blue;"} {#sec-stakeholder-engagement}
+
+[The CRUSE Tool was developed for, and in close collaboration with, practitioners in Ireland.
 Although the overall approach of the project was quantitative, stakeholder feedback and qualitative data was deemed essential to ensure the tool was fit for purpose, mitigating against the risk that the project would be 'technocentric' and a 'problem looking for a solution'.
 From the outset of the project, we aimed to follow @coppola2022's advocacy for a "human-centric approach in which personalization, purpose, ethics and inclusivity are key", rather than an 'off-the-shelf' product that failed to understand, let alone meet, local needs.
-As outlined in the overview of papers in the collection Inclusive and Collaborative Transport cited above, the 'digital transformation' taking place in transport planning (and many other sectors) has the potential to increase existing inequalities [@coppola2022].
-A wide range of methods and measure can be taken to ensure that digital tools are inclusive \[...\].
+As outlined by @coppola2022, the 'digital transformation' taking place in transport planning (and many other sectors) has the potential to increase existing inequalities [@coppola2022].
+A wide range of methods and measure can be taken to ensure that digital tools are accessible to their recipients.]{style="color: blue;"}
 
@@
 
-Due to time and skill constraints in the team developing the CRUSE tool (we lacked researchers with a background in qualitative methods), we opted for the well-tested and cost-effective approach of holding workshops with stakeholders in case study area, an approach that is documented with reference to participatory development of digital tools for supporting energy transitions [@hewitt2020].
+[The stakeholder engagement was carried out through workshops with stakeholders in case study area, an approach that is documented with reference to participatory development of digital tools for supporting energy transitions [@hewitt2020].
 We held a total of three workshops over the course of the project and one open-ended live demo during a 2-day event in Sligo in which we collected feedback from dozens of people.
-The first two workshops took place in May 2022, in which key staff involved in cycle network planning in Kildare and Limerick counties were presented the tool.
+Other than the live demo, all of the workshops took place online.]{style="color: blue;"}
+
+[The first two workshops took place in May 2022, during which the tool was introduced to key staff involved in cycle network planning in Kildare and Limerick counties.
 The county-specific approach allowed the workshop to focus on specific areas and key corridors.
 The local transport planners were asked to provide feedback in a semi-structured interview approach, with questions on the evidence they currently have access, how they may be able to use the tool, and requests for new features.
+The third and final workshop took place on the 7^th^ December 2023 and included descriptions of recent changes to the tool, with an aim being to get feedback on the new recreational and cycle touring evidence, with representatives from county-level transport planning bodies, Sport Ireland, and Waterways Ireland.]{style="color: blue;"}
 
-The third and final workshop took place on the 7^th^ December 2023 and included descriptions of recent changes to the tool, with an aim being to get feedback on the new recreational and cycle touring evidence, with representatives from county-level transport planning bodies, Sport Ireland, and Waterways Ireland.
-The open-ended live demo was undertaken at a stall at TII’s National Roads and Greenways Conference in Sligo in September 2022, where much of the feedback was verbal and not collected.
-Much of this feedback from the workshops and live demo outlined above was in the form of ad-hoc comments and suggestions and not all of the qualitative data received was recorded in a structured way, a limitation that we refer back to in @sec-discussion.
+[The open-ended live demo was undertaken at a stall at TII’s National Roads and Greenways Conference in Sligo in September 2022, where much of the feedback was verbal and not collected.
+Much of this feedback from the workshops and live demo outlined above was in the form of ad-hoc comments and suggestions and not all of the qualitative data received was recorded in a structured way, a limitation that we refer back to in @sec-discussion.]{style="color: blue;"}
 
@@
 
-A final limitation of the approach relates to *how* it was developed.
+[A final limitation of the approach relates to *how* it was developed.
 As described in @sec-stakeholder-engagement, the development of the tool was participatory, but only insofar as people who attended our workshops could provide feedback.
@@
 In future work we would like to collect feedback from a wider range of people, including campaigners, service providers (e.g.
-staff involved in travel planning for schools and new developments), and people who not involved in the planning process.
+staff involved in travel planning for schools and new developments), and people who not involved in the planning process.]{style="color: blue;"}
 

### 2024-06-23 paper.qmd
Prep PDF version for publication

@@
 format: 
-  html: default
-  # docx: default
-# # # Uncomment for formatted version
-  # elsevier-pdf:
-  #   linestretch: 2
-  #   keep-tex: true
-  #   journal:
-  #     name: European Transport Research Review
-  #     formatting: preprint
-  #     model: 3p
-  #     cite-style: super
+  # html: default
+  # Uncomment for formatted version
+  elsevier-pdf:
+    linestretch: 2
+    keep-tex: true
+    journal:
+      name: European Transport Research Review
+      formatting: preprint
+      model: 3p
+      cite-style: super
 # # # #   # pdf: default
 # # Uncomment the following line to produce the cover page
-author:
-  - name: Robin Lovelace
-    email: r.lovelace@leeds.ac.uk
-    corresponding: true
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Joey Talbot
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Eugeni Vidal-Tortosa
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Hussein Mahfouz
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Elaine Brick
-    affiliations:
-      - name: AECOM
-        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-  - name: Peter Wright
-    affiliations:
-      - name: AECOM
-        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
-  - name: "Gary O'Toole"
-    affiliations:
-      - name: AECOM
-        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-  - name: Dan Brennan
-    affiliations:
-      - name: Transport Infrastructure Ireland
-        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
-  - name: Suzanne Meade
-    affiliations:
-      - name: Transport Infrastructure Ireland
-        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+# author:
+#   - name: Robin Lovelace
+#     email: r.lovelace@leeds.ac.uk
+#     corresponding: true
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Joey Talbot
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Eugeni Vidal-Tortosa
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Hussein Mahfouz
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Elaine Brick
+#     affiliations:
+#       - name: AECOM
+#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+#   - name: Peter Wright
+#     affiliations:
+#       - name: AECOM
+#         address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
+#   - name: "Gary O'Toole"
+#     affiliations:
+#       - name: AECOM
+#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+#   - name: Dan Brennan
+#     affiliations:
+#       - name: Transport Infrastructure Ireland
+#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+#   - name: Suzanne Meade
+#     affiliations:
+#       - name: Transport Infrastructure Ireland
+#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
 abstract: |
@@
 # latexdiffr::latexdiff("paper-v1.qmd", "paper.qmd")
-system("latexdiff paper-v1.tex paper.tex > tracked-changes.tex")
-system("pdflatex tracked-changes.tex")
-browseURL("tracked-changes.pdf")
+system("latexdiff paper-r1.tex paper.tex > tracked-changes-r2.tex")
+system("pdflatex tracked-changes-r2.tex")
+browseURL("tracked-changes-r2.pdf")
 ```
@@
 # # For PDF:
-# knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
-#   kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
-#   kableExtra::column_spec(1, width = "13em") |>
-#   kableExtra::column_spec(2, width = "15em") |>
-#   kableExtra::column_spec(3, width = "12em") |>
-#   kableExtra::column_spec(4, width = "5em")
-# For HTML:
-knitr::kable(tools)
+knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
+  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
+  kableExtra::column_spec(1, width = "13em") |>
+  kableExtra::column_spec(2, width = "15em") |>
+  kableExtra::column_spec(3, width = "12em") |>
+  kableExtra::column_spec(4, width = "5em")
+# # For HTML:
+# knitr::kable(tools)
 ```

### 2024-06-23 paper-r1.qmd
R2 version

--- 2024-06-21 paper.qmd
@@
 format: 
-  html: default
-  # docx: default
-# # # Uncomment for formatted version
-  # elsevier-pdf:
-  #   linestretch: 2
-  #   keep-tex: true
-  #   journal:
-  #     name: European Transport Research Review
-  #     formatting: preprint
-  #     model: 3p
-  #     cite-style: super
+ elsevier-pdf:
+   linestretch: 2
+   keep-tex: true
+   journal:
+     name: European Transport Research Review
+     formatting: preprint
+     model: 3p
+     cite-style: super
 # # # #   # pdf: default
+
 # # Uncomment the following line to produce the cover page
-author:
-  - name: Robin Lovelace
-    email: r.lovelace@leeds.ac.uk
-    corresponding: true
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Joey Talbot
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Eugeni Vidal-Tortosa
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Hussein Mahfouz
-    affiliations:
-      - name: University of Leeds
-        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-  - name: Elaine Brick
-    affiliations:
-      - name: AECOM
-        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-  - name: Peter Wright
-    affiliations:
-      - name: AECOM
-        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
-  - name: "Gary O'Toole"
-    affiliations:
-      - name: AECOM
-        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-  - name: Dan Brennan
-    affiliations:
-      - name: Transport Infrastructure Ireland
-        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
-  - name: Suzanne Meade
-    affiliations:
-      - name: Transport Infrastructure Ireland
-        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+# author:
+#   - name: Robin Lovelace
+#     email: r.lovelace@leeds.ac.uk
+#     corresponding: true
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Joey Talbot
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Eugeni Vidal-Tortosa
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Hussein Mahfouz
+#     affiliations:
+#       - name: University of Leeds
+#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+#   - name: Elaine Brick
+#     affiliations:
+#       - name: AECOM
+#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+#   - name: Peter Wright
+#     affiliations:
+#       - name: AECOM
+#         address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
+#   - name: "Gary O'Toole"
+#     affiliations:
+#       - name: AECOM
+#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+#   - name: Dan Brennan
+#     affiliations:
+#       - name: Transport Infrastructure Ireland
+#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+#   - name: Suzanne Meade
+#     affiliations:
+#       - name: Transport Infrastructure Ireland
+#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
 abstract: |
@@
 
-Lovelace, Robin, Joey Talbot, Eugeni Vidal-Tortosa, Hussein Mahfouz, Elaine Brick, Peter Wright, Gary O'Toole, et al. 2024.
+Lovelace, Robin, Joey Talbot, Eugeni Vidal-Tortosa, Hussein Mahfouz, Elaine Brick, Peter Wright, Gary O'Toole, et al.
+2024.
 “Cycle Route Uptake and Scenario Estimation (CRUSE): An Approach for Developing Strategic Cycle Network Planning Tools.” OSF Preprints.
@@
 Developed by researchers and developers in academic, public and private sectors, they have evolved from simple area-based static models tailored to specific regions to more complex and (in some cases) more generalizable tools.
-These tools can be considered as a specific application of the concept of planning support systems (PSS), which are designed to help planners and decision-makers to make more informed decisions [@geertman2009].
+These tools can be considered as a specific application of the concept of planning support systems (PSS) , which are designed to help planners and decision-makers to make more informed decisions [@geertman2009].
 Outputs include maps with results available at several levels of analysis with levels of availability ranging from being only available to researchers and in static maps to open access mapping systems available to the public.
@@
 #| label: tbl-tools
-#| tbl-cap: "Summary of existing tools for estimating cycling potential."
 # # For PDF:
@@
 # For HTML:
-knitr::kable(tools)
+knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.")
 ```
@@
 
-Other tools that generate areal data to prioritize investment include a study that generated a['bikeability index' for the Metro Vancouver region [@winters_mapping_2013] and later called Bike Score and available for over 160 US and Canadian cities [@winters_bike_2016] -- its methodology is [publicly available](https://www.walkscore.com/bike-score-methodology.shtml); a]{style="color: blue;"}'prioritization index' for Montreal, Canada [@larsen_build_2013]; a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil [@zhang_prioritizing_2014]; and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
+Other tools that generate areal data to prioritize investment include a study that generated a 'prioritization index' for Montreal, Canada [@larsen_build_2013], a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil, and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
 The CPT is composed of the base environmental module — based on eight weighted factors (population density, hilliness, physical barriers, access to services, existing cycling mode share, distance to work and school, and road speed) — and the quality of service module, which evaluates each area of interest based on eight factors (surface condition, adjacent cyclists, comfort factor, conflict, distance between junctions, slope, access to services, and origin/destination).
 
-A more recent areal-based approach is the Gross Potential for Cycling (GPC) tool to prioritize areas for cycling infrastructure and other cycling measures [@silva_gross_2021].
+A more recent areal-based approach is the designed the Gross Potential for Cycling (GPC) tool to prioritize areas for cycling infrastructure and other cycling measures [@silva_gross_2021].
 The GPC uses two sets of indicators: population-based indicators (age, potential demand density, employment density and motorization rate) and area-based indicators (accessibility, education, public transport, connectivity, land use mix and relative performance).
@@
 The ranking of areas by the GPC are [publicly available](https://boost.up.pt/en/ferramentas/gpc) and its practical value has been assessed through workshops [@silva_revealing_2022].
-[The GPC tool has also been used to investigate the combination of factors contributing to high levels cycling potential [@lopes_impact_2021].
-Another recent areal-based approach is the Relative Equity Impact of Bicycle Planning (TIRE) [@cunha_assessing_2023].
-The TIRE assesses the impact of cycling network allocation on the accessibility levels of different socioeconomic groups.]{style="color: blue;"}
 
 Route-based tools emerged with increasing availability of origin-destination data, routing engines that can assign trips to networks, and improvements in computer hardware and software needed to generate route networks.
-A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace2016].
+A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace_mapping_2016].
 The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 [@lovelace2017].
@@
 biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.
-
-[Another route-based tool is The Cycling Routing Algorithm for Network Connectivity (CRANC) [@gehrke_cycling-focused_2020].
-The CRANC tool evaluates the benefits of new bicycle facilities for the accessibility of various populations and neighbourhoods.]{style="color: blue;"}
 
@@
 # For HTML:
-knitr::kable(trip_purposes)
+knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.")
 ```
@@
 ```
-## Stakeholder engagement {#sec-stakeholder-engagement}
-
-The CRUSE Tool was developed for, and in close collaboration with, practitioners in Ireland.
-Although the overall approach of the project was quantitative, stakeholder feedback and qualitative data was deemed essential to ensure the tool was fit for purpose, mitigating against the risk that the project would be 'technocentric' and a 'problem looking for a solution'.
-From the outset of the project, we aimed to follow @coppola2022's advocacy for a "human-centric approach in which personalization, purpose, ethics and inclusivity are key", rather than an 'off-the-shelf' product that failed to understand, let alone meet, local needs.
-As outlined in the overview of papers in the collection Inclusive and Collaborative Transport cited above, the 'digital transformation' taking place in transport planning (and many other sectors) has the potential to increase existing inequalities [@coppola2022].
-A wide range of methods and measure can be taken to ensure that digital tools are inclusive \[...\].
-
-<!-- Details from final workshop:
-
-Can you please join me and my colleagues from TII and ITS Leeds for 1 hour on Thursday the 7th December at 12:00 – 13:00 for a workshop and demonstration of the open source CRUSE tool) which is used primarily to highlight the cycle friendliness of roads in Ireland, avail of cycle scenario/trip data and highlight areas to target for active travel investment.
-This tool is free to use for transport planners or member of the general public and can be found here: CRUSE .
-We have now added recreational cycling trips to the CRUSE tool and would welcome the chance to present this work and ask for feedback.
-
-Please see below a draft agenda for the workshop till 1pm:
-
--   Introduction (5 min)
--   Changes to CRUSE: extension work (15 min)
--   Testing new website (10 min)
--   Feedback, Q&A + discussion (30 min)
-    - General feedback on new site
-    - Feedback on the recreational outputs
-    - Feedback on the tourism network -->
-
-<!-- – What are main issues you face in
-understanding infrastructure?, e.g.
-– What is happening at the moment?
-– Where to invest in the future?
-– What will change?
-– How will planning vary depending on trip purpose?
-– What sort of scenarios interested in testing?
-– How useful is having different networks for
-planning infrastructure and/or more detail,
-e.g. there could be an option for variation in
-access for professionals and public (the latter
-remaining simple) -->
-
-Due to time and skill constraints in the team developing the CRUSE tool (we lacked researchers with a background in qualitative methods), we opted for the well-tested and cost-effective approach of holding workshops with stakeholders in case study area, an approach that is documented with reference to participatory development of digital tools for supporting energy transitions [@hewitt2020].
-We held a total of three workshops over the course of the project and one open-ended live demo during a 2-day event in Sligo in which we collected feedback from dozens of people.
-The first two workshops took place in May 2022, in which key staff involved in cycle network planning in Kildare and Limerick counties were presented the tool.
-The county-specific approach allowed the workshop to focus on specific areas and key corridors.
-The local transport planners were asked to provide feedback in a semi-structured interview approach, with questions on the evidence they currently have access, how they may be able to use the tool, and requests for new features.
-
-The third and final workshop took place on the 7^th^ December 2023 and included descriptions of recent changes to the tool, with an aim being to get feedback on the new recreational and cycle touring evidence, with representatives from county-level transport planning bodies, Sport Ireland, and Waterways Ireland.
-The open-ended live demo was undertaken at a stall at TII’s National Roads and Greenways Conference in Sligo in September 2022, where much of the feedback was verbal and not collected.
-Much of this feedback from the workshops and live demo outlined above was in the form of ad-hoc comments and suggestions and not all of the qualitative data received was recorded in a structured way, a limitation that we refer back to in @sec-discussion.
-
 # Results {#sec-results}
@@
 
-## Stakeholder feedback {#sec-qualitative}
+## Stakeholder engagement and feedback {#sec-qualitative}
 
@@
 
-A final limitation of the approach relates to *how* it was developed.
-As described in @sec-stakeholder-engagement, the development of the tool was participatory, but only insofar as people who attended our workshops could provide feedback.
-The people who did provide feedback were deliberately chosen as the target audience.
-In future work we would like to collect feedback from a wider range of people, including campaigners, service providers (e.g.
-staff involved in travel planning for schools and new developments), and people who not involved in the planning process.
-
 # Conclusions {#sec-conclusions}

### 2024-06-23 paper.tex
R2 version

--- 2024-03-27 paper.tex
@@
 \setstretch{2}
+Note: preprint now available here: https://osf.io/preprints/osf/a72yh
+
+Please cite as
+
+Lovelace, Robin, Joey Talbot, Eugeni Vidal-Tortosa, Hussein Mahfouz,
+Elaine Brick, Peter Wright, Gary O'Toole, et al.~2024. ``Cycle Route
+Uptake and Scenario Estimation (CRUSE): An Approach for Developing
+Strategic Cycle Network Planning Tools.'' OSF Preprints. March 29.
+doi:10.31219/osf.io/a72yh.
+
+This is an open and reproducible paper. See
+\href{https://github.com/cruse-bike}{github.com/cruse-bike} for the code
+and data used to generate the results presented here, and the code
+underlying the web app hosted at \href{https://cruse.bike/}{cruse.bike}.
+
 \newpage{}
@@
 can be considered as a specific application of the concept of planning
-support systems (PSS) , which are designed to help planners and
+support systems (PSS), which are designed to help planners and
 decision-makers to make more informed decisions \citep{geertman2009}.
@@
 \midrule
-\cellcolor{gray!10}{Bicycle share model, Parkin et al.
+\cellcolor{gray!10}{Bicycle share model, Parkin, J.
+et al.
 (2007)} & \cellcolor{gray!10}{England and Wales, Journeys to work OD census data, at the small-area (wards) level} & \cellcolor{gray!10}{Static tables in academic paper} & \cellcolor{gray!10}{Proprietary}\\
-Analysis of  Cycling Potential (ACP), Transport for London (2010, 2016) & London, London Travel Demand Survey & Static tables, graphs, and heatmaps, not publicly available & Proprietary\\
-\cellcolor{gray!10}{Prioritization index, Larsen et al.
-(2013)} & \cellcolor{gray!10}{Montreal, Survey, Road safety, and OD data} & \cellcolor{gray!10}{Map-based heatmap, not publicly available} & \cellcolor{gray!10}{Proprietary}\\
-Usage intensity index, Zhang et al. (2014) & Belo Horizonte, Survey, census, and OD data & Static tables and maps, not publicly available & Proprietary\\
-\cellcolor{gray!10}{The Cycling Potential Tool (CPT), Phillips and Range (2017)} & \cellcolor{gray!10}{Scotland, Environmental and socioeconomic data, at the small area (output areas) level} & \cellcolor{gray!10}{Maps showing cycling potential in each area} & \cellcolor{gray!10}{Proprietary}\\
+Analysis of Cycling Potential (ACP), Transport for London (2010 and 2016) & London, London Travel Demand Survey & Static tables, graphs, and heatmaps, not publicly available & Proprietary\\
+\cellcolor{gray!10}{{}[Bikeability index/Bike Score]\{style="color: blue;"\}, [Winters, M.
+et al.
+(2013 and 2016)]\{style="color: blue;"\}} & \cellcolor{gray!10}{{}[Metro Vancouver region/160 US and Canadian cities]\{style="color: blue;"\}, [Cycling infrastructure, topography, destinations and road connectivity data]\{style="color: blue;"\}} & \cellcolor{gray!10}{{}[Map-based heatmap, not publicly available]\{style="color: blue;"\}} & \cellcolor{gray!10}{{}[Proprietary]\{style="color: blue;"\}}\\
+Prioritization index, Larsen, J. et al. (2013) & Montreal, Survey, Road safety, and OD data & Map-based heatmap, not publicly available & Proprietary\\
+\cellcolor{gray!10}{Usage intensity index, Zhang, D.
+et al.
+(2014)} & \cellcolor{gray!10}{Belo Horizonte, Survey, census, and OD data} & \cellcolor{gray!10}{Static tables and maps, not publicly available} & \cellcolor{gray!10}{Proprietary}\\
 \addlinespace
-Propensity to Cycle Tool (PCT), Lovelace et al.
-(2017) & England and Wales, Journeys to work and school OD census data, at the small area (LSOA) level & Online maps, graphs, tables, publicly available at www.pct.bike & Open source\\
-\cellcolor{gray!10}{The    Gross Potential for Cycling tool (CPC), Silva et al.
-(2021 and 2022)} & \cellcolor{gray!10}{21 Portuguese cities, Land use and socio-demographic data, at the small area (census tract) level} & \cellcolor{gray!10}{Static maps showing cycling potential in different areas} & \cellcolor{gray!10}{NA}\\
+The Cycling Potential Tool (CPT), Phillips, L., \& Range, A.
+(2017) & Scotland, Environmental and socioeconomic data, at the small area (output areas) level & Maps showing cycling potential in each area & Proprietary\\
+\cellcolor{gray!10}{Propensity to Cycle Tool (PCT), Lovelace, R.
+et al.
+(2017)} & \cellcolor{gray!10}{England and Wales, Journeys to work and school OD census data, at the small area (LSOA) level} & \cellcolor{gray!10}{Online maps, graphs, tables, publicly available at www.pct.bike} & \cellcolor{gray!10}{Open source}\\
+{}[Cyclist Routing Algorithm for Network Connectivity (CRANC)]\{style="color: blue;"\}, [Gehrke, S.
+R.
+et al.
+(2020)]\{style="color: blue;"\} & {}[Boston region]\{style="color: blue;"\}, [Road segment characteristics, elevation, and destination acessibility data, at the census blocks level]\{style="color: blue;"\} & {}[Graphs and heatmaps, not publicly available]\{style="color: blue;"\} & {}[NA]\{style="color: blue;"\}\\
+\cellcolor{gray!10}{The Gross Potential for Cycling tool (CPC), Silva, C.
+et al.
+(2021 and 2022) [and Lopez et al.
+(2021)]\{style="color: blue;"\}} & \cellcolor{gray!10}{21 Portuguese cities, Land use and socio-demographic data, at the small area (census tract) level} & \cellcolor{gray!10}{Static maps showing cycling potential in different areas} & \cellcolor{gray!10}{NA}\\
+{}[The relative equity impact of bicycle planning (TIRE)]\{style="color: blue;"\}, [Cunha, I., \& Silva, C.
+(2023)]\{style="color: blue;"\} & {}[Lisbon]\{style="color: blue;"\}, [Cycling infrastructure, topog­raphy, and location of points of interest data, at the hexagon cells level]\{style="color: blue;"\} & {}[Graphs and heatmaps, not publicly available]\{style="color: blue;"\} & {}[NA]\{style="color: blue;"\}\\
 \bottomrule
@@
 Other tools that generate areal data to prioritize investment include a
-study that generated a `prioritization index' for Montreal, Canada
-\citep{larsen_build_2013}, a `usage intensity index' based on stated
-preferences in Belo Horizonte, Brazil, and a Cycling Potential Tool
-(CPT) for Scotland \citep{phillips_development_2017}. The CPT is
-composed of the base environmental module --- based on eight weighted
-factors (population density, hilliness, physical barriers, access to
-services, existing cycling mode share, distance to work and school, and
-road speed) --- and the quality of service module, which evaluates each
-area of interest based on eight factors (surface condition, adjacent
-cyclists, comfort factor, conflict, distance between junctions, slope,
-access to services, and origin/destination).
-
-A more recent areal-based approach is the designed the Gross Potential
-for Cycling (GPC) tool to prioritize areas for cycling infrastructure
-and other cycling measures \citep{silva_gross_2021}. The GPC uses two
-sets of indicators: population-based indicators (age, potential demand
-density, employment density and motorization rate) and area-based
-indicators (accessibility, education, public transport, connectivity,
-land use mix and relative performance). These indicators are calculated
-considering topography, road hierarchy, and average congestion and
-presented on a scale from 5 to 10. They are then combined into an
-overall score, weighted by their impact on cycling according to the
-literature. The ranking of areas by the GPC are
+study that generated a {`bikeability index' for the Metro Vancouver
+region \citep{winters_mapping_2013}, later renamed Bike Score and
+available for over 160 US and Canadian cities \citep{winters_bike_2016}
+-- its methodology is
+\href{https://www.walkscore.com/bike-score-methodology.shtml}{openly
+accessible}; a} `prioritization index' for Montreal, Canada
+\citep{larsen_build_2013}; a `usage intensity index' based on stated
+preferences in Belo Horizonte, Brazil \citep{zhang_prioritizing_2014};
+and a Cycling Potential Tool (CPT) for Scotland
+\citep{phillips_development_2017}. The CPT is composed of the base
+environmental module --- based on eight weighted factors (population
+density, hilliness, physical barriers, access to services, existing
+cycling mode share, distance to work and school, and road speed) --- and
+the quality of service module, which evaluates each area of interest
+based on eight factors (surface condition, adjacent cyclists, comfort
+factor, conflict, distance between junctions, slope, access to services,
+and origin/destination).
+
+A more recent areal-based approach is the Gross Potential for Cycling
+(GPC) tool to prioritize areas for cycling infrastructure and other
+cycling measures \citep{silva_gross_2021}. The GPC uses two sets of
+indicators: population-based indicators (age, potential demand density,
+employment density and motorization rate) and area-based indicators
+(accessibility, education, public transport, connectivity, land use mix
+and relative performance). These indicators are calculated considering
+topography, road hierarchy, and average congestion and presented on a
+scale from 5 to 10. They are then combined into an overall score,
+weighted by their impact on cycling according to the literature. The
+ranking of areas by the GPC are
 \href{https://boost.up.pt/en/ferramentas/gpc}{publicly available} and
 its practical value has been assessed through workshops
-\citep{silva_revealing_2022}.
+\citep{silva_revealing_2022}. {The GPC tool has also been used to
+investigate the combination of factors contributing to high levels
+cycling potential \citep{lopes_impact_2021}. Another recent areal-based
+approach is the Relative Equity Impact of Bicycle Planning (TIRE)
+\citep{cunha_assessing_2023} tool. The TIRE tool assesses the impact of
+cycling network allocation on the accessibility levels of different
+socioeconomic groups.}
 
@@
 of cycling at desire line, zone, route, and route network levels for
-case study cities \citep{lovelace_mapping_2016}. The approach was
-scaled-up to estimate estimate the potential benefits of uptake at zone
-and desire line levels nationally, and launched as a publicly available
-web application in 2017 \citep{lovelace2017}. Extensions of the PCT
-approach have included estimation of benefits at the individual level
+case study cities \citep{lovelace2016}. The approach was scaled-up to
+estimate estimate the potential benefits of uptake at zone and desire
+line levels nationally, and launched as a publicly available web
+application in 2017 \citep{lovelace2017}. Extensions of the PCT approach
+have included estimation of benefits at the individual level
 \citep{woodcock2018}, addition of travel to school network
@@
 options based on General Transit Feed Specification (GTFS) data.
+
+{Another route-based tool is The Cycling Routing Algorithm for Network
+Connectivity (CRANC) \citep{gehrke_cycling-focused_2020}. The CRANC tool
+evaluates the benefits of new bicycle facilities for the accessibility
+of various populations and neighbourhoods.}
 
@@
 
-\begingroup\fontsize{8}{10}\selectfont
-
-\begin{longtable}[t]{>{\raggedright\arraybackslash}p{5em}>{\raggedright\arraybackslash}p{12em}>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{15em}}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0653}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2869}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.4460}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2017}}@{}}
 
@@
 
-\toprule
-Trip purpose & Description & Primary source(s) & Confidence\\
-\midrule
-\cellcolor{gray!10}{Work (commute)} & \cellcolor{gray!10}{Commuting to / from workplaces} & \cellcolor{gray!10}{Census 2016 POWSCAR Data} & \cellcolor{gray!10}{High confidence - data both origins and destinations for trip purposes}\\
-Primary Education & Primary school trips & Census 2016 POWSCAR Data, Schools database & High confidence - data both origins and destinations for trip purposes\\
-\cellcolor{gray!10}{Secondary Education} & \cellcolor{gray!10}{Secondary school trips} & \cellcolor{gray!10}{Census 2016 POWSCAR Data, Schools database} & \cellcolor{gray!10}{High confidence - data both origins and destinations for trip purposes}\\
-Tertiary Education & Tertiary education trips & Census 2016 POWSCAR Data - Geodirectory categories & High confidence - data both origins and destinations for trip purposes\\
-\cellcolor{gray!10}{Social} & \cellcolor{gray!10}{NHTS classification (e.g.
-restaurants, cinemas, gyms etc.)} & \cellcolor{gray!10}{Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes} & \cellcolor{gray!10}{Medium confidence}\\
-\addlinespace
-Shopping & NHTS classificationfrom supermarkets and shops & Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes & Medium confidence\\
-\cellcolor{gray!10}{Personal / other} & \cellcolor{gray!10}{Combination of two NHTS classifications, and includes trips to medical, personal services and others} & \cellcolor{gray!10}{Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes} & \cellcolor{gray!10}{Medium confidence}\\
-Tourism / recreational & Trips by non-resident visitors to / from visitor attractions & Geodirectory listings for accommodation (origins) and destinations (destinations), Failte Ireland accommodation database, Failte Ireland attraction database & Low confidence - lack of data regarding tourist trip rates\\
-\bottomrule
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Trip purpose
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Description
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Primary source(s)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Confidence
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Work (commute) & Commuting to / from workplaces & Census 2016 POWSCAR
+Data & High confidence - data both origins and destinations for trip
+purposes \\
+Primary Education & Primary school trips & Census 2016 POWSCAR Data,
+Schools database & High confidence - data both origins and destinations
+for trip purposes \\
+Secondary Education & Secondary school trips & Census 2016 POWSCAR Data,
+Schools database & High confidence - data both origins and destinations
+for trip purposes \\
+Tertiary Education & Tertiary education trips & Census 2016 POWSCAR Data
+- Geodirectory categories & High confidence - data both origins and
+destinations for trip purposes \\
+Social & NHTS classification (e.g.~restaurants, cinemas, gyms etc.) &
+Trip rates from NHTS, Geodirectory businesses classified according to
+relevant NACE codes & Medium confidence \\
+Shopping & NHTS classificationfrom supermarkets and shops & Trip rates
+from NHTS, Geodirectory businesses classified according to relevant NACE
+codes & Medium confidence \\
+Personal / other & Combination of two NHTS classifications, and includes
+trips to medical, personal services and others & Trip rates from NHTS,
+Geodirectory businesses classified according to relevant NACE codes &
+Medium confidence \\
+Tourism / recreational & Trips by non-resident visitors to / from
+visitor attractions & Geodirectory listings for accommodation (origins)
+and destinations (destinations), Failte Ireland accommodation database,
+Failte Ireland attraction database & Low confidence - lack of data
+regarding tourist trip rates \\
 
 \end{longtable}
-
-\endgroup{}
 
@@
 
-\includegraphics{fig-user-stories-1.pdf}
+\includegraphics[width=3.31in,height=\textheight]{images/fig5.png}
 
@@
 \end{figure}%
+
+\subsection{\texorpdfstring{{Stakeholder
+engagement}}{Stakeholder engagement}}\label{sec-stakeholder-engagement}
+
+{The CRUSE Tool was developed for, and in close collaboration with,
+practitioners in Ireland. Although the overall approach of the project
+was quantitative, stakeholder feedback and qualitative data was deemed
+essential to ensure the tool was fit for purpose, mitigating against the
+risk that the project would be `technocentric' and a `problem looking
+for a solution'. From the outset of the project, we aimed to follow
+\citet{coppola2022}`s advocacy for a ``human-centric approach in which
+personalization, purpose, ethics and inclusivity are key'', rather than
+an 'off-the-shelf' product that failed to understand, let alone meet,
+local needs. As outlined by \citet{coppola2022}, the `digital
+transformation' taking place in transport planning (and many other
+sectors) has the potential to increase existing inequalities
+\citep{coppola2022}. A wide range of methods and measure can be taken to
+ensure that digital tools are accessible to their recipients.}
+
+{The stakeholder engagement was carried out through workshops with
+stakeholders in case study area, an approach that is documented with
+reference to participatory development of digital tools for supporting
+energy transitions \citep{hewitt2020}. We held a total of three
+workshops over the course of the project and one open-ended live demo
+during a 2-day event in Sligo in which we collected feedback from dozens
+of people. Other than the live demo, all of the workshops took place
+online.}
+
+{The first two workshops took place in May 2022, during which the tool
+was introduced to key staff involved in cycle network planning in
+Kildare and Limerick counties. The county-specific approach allowed the
+workshop to focus on specific areas and key corridors. The local
+transport planners were asked to provide feedback in a semi-structured
+interview approach, with questions on the evidence they currently have
+access, how they may be able to use the tool, and requests for new
+features. The third and final workshop took place on the
+7\textsuperscript{th} December 2023 and included descriptions of recent
+changes to the tool, with an aim being to get feedback on the new
+recreational and cycle touring evidence, with representatives from
+county-level transport planning bodies, Sport Ireland, and Waterways
+Ireland.}
+
+{The open-ended live demo was undertaken at a stall at TII's National
+Roads and Greenways Conference in Sligo in September 2022, where much of
+the feedback was verbal and not collected. Much of this feedback from
+the workshops and live demo outlined above was in the form of ad-hoc
+comments and suggestions and not all of the qualitative data received
+was recorded in a structured way, a limitation that we refer back to in
+Section~\ref{sec-discussion}.}
 
@@
 
-\subsection{Stakeholder engagement and feedback}\label{sec-qualitative}
+\subsection{Stakeholder feedback}\label{sec-qualitative}
 
@@
 
+{A final limitation of the approach relates to \emph{how} it was
+developed. As described in Section~\ref{sec-stakeholder-engagement}, the
+development of the tool was participatory, but only insofar as people
+who attended our workshops could provide feedback. The people who did
+provide feedback were deliberately chosen as the target audience. In
+future work we would like to collect feedback from a wider range of
+people, including campaigners, service providers (e.g.~staff involved in
+travel planning for schools and new developments), and people who not
+involved in the planning process.}
+
 \section{Conclusions}\label{sec-conclusions}

### 2024-06-23 paper.qmd
Minor tweaks

@@
   This paper describes an approach for developing strategic cycle network planning tools.
-Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenario generation, and development of an intuitive user interface that has been tested and used by practitioners.
-Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions.
+Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data with the open source 'odjitter' software, incorporation of additional trip purposes, routing, scenario generation, and development of an intuitive user interface that is tested and used by practitioners.
+Commissioned by the national infrastructure agency Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions.
 The publicly available results at https://cruse.bike/ enable planners, engineers, and other stakeholders to make more evidence-based decisions.
-CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring) in addition to using official origin-destination data; and providing estimates of 'quietness' (a proxy for cyclist comfort and route preference) at the route segment level.
+CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring), supplementing official origin-destination datasets on travel for work and education; and providing estimates of 'quietness' (a proxy for cyclist comfort and route preference) at the route segment level.
 Three network types --- 'Fastest', 'Balanced', and 'Quietest' --- help plan both arterial and residential cycle networks.
-Workshops and unstructured interviews with stakeholders show that the tool is already being used in practice to inform urban, inter-urban, and rural network planning.
-The approach is flexible and open-source, allowing the underlying ideas and code to be adapted, supporting effective cycling policies and interventions internationally.
+Workshops and unstructured interviews with stakeholders were used to inform the development of the tool.
+Feedback shows that the tool has a wide range of uses and is already being used in practice to inform urban, inter-urban, and rural cycle network designs.
+The approach is flexible and open-source, allowing the underlying ideas and code to be adapted, supporting more evidence-based and effective cycling policies and interventions internationally.
 keywords: 
@@
 
-\clearpage
-
 ```{r}
@@
 # For PDF:
-# knitr::kable(trip_purposes, booktabs = TRUE) |>
-#   kableExtra::kable_styling(latex_options = "striped", full_width = FALSE, font_size = 8) |>
-#   kableExtra::column_spec(1, width = "5em") |>
-#   kableExtra::column_spec(2, width = "12em") |>
-#   kableExtra::column_spec(3, width = "15em") |>
-#   kableExtra::column_spec(4, width = "15em")
-# For HTML:
-knitr::kable(trip_purposes)
+knitr::kable(trip_purposes, booktabs = TRUE) |>
+  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE, font_size = 8) |>
+  kableExtra::column_spec(1, width = "5em") |>
+  kableExtra::column_spec(2, width = "12em") |>
+  kableExtra::column_spec(3, width = "15em") |>
+  kableExtra::column_spec(4, width = "15em")
+# # For HTML:
+# knitr::kable(trip_purposes)
 ```
@@
 
-The CRUSE web application is statically hosted, meaning it does not require a server running software such as the R package 'shiny' or Python packages such as 'streamlit' or 'flask' in the background [@wickham2021].
-This reduces hosting costs and the maintenance burden of the tool compared with tools such as the PCT, and makes the map-based interface more responsive, using the recently developed 'MapLibre' JavaScript package and 'PMTiles' "cloud native" vector tile format [@gonçalves2023].
-These open source technologies were used to generate default view in the landing page: a map of Ireland with county level results for the % trips that could be made by cycling under the 'Go Dutch' scenario, highlighting the intended use case for strategic cycle network planning and level of ambition, as illustrated in @fig-landing.
+The CRUSE web application is 'statically hosted' rather than 'dynamically hosted' [@wickham2021].
+This means the tool does not require a resource-intensive server, reducing hosting costs and the maintenance burden of the tool compared with tools such as the PCT.
+The use recently-developed web mapping technology makes the tool responsive [@gonçalves2023].
+Open source technologies were used throughout the tool's user interface, increasing the ease with which people in other countries can build on the open source code.
+The 'landing page' shows the proportion of trips that could be made by cycling under the 'Go Dutch' scenario at the county level, highlighting the intended use case for strategic cycle network planning (@fig-landing).
 
@@
 
-## [Stakeholder engagement]{style="color: blue;"} {#sec-stakeholder-engagement}
+## Stakeholder engagement {#sec-stakeholder-engagement}
 
-[The CRUSE Tool was developed for, and in close collaboration with, practitioners in Ireland.
+The CRUSE Tool was developed for, and in close collaboration with, practitioners in the Republic of Ireland.
 Although the overall approach of the project was quantitative, stakeholder feedback and qualitative data was deemed essential to ensure the tool was fit for purpose, mitigating against the risk that the project would be 'technocentric' and a 'problem looking for a solution'.
@@
 As outlined by @coppola2022, the 'digital transformation' taking place in transport planning (and many other sectors) has the potential to increase existing inequalities [@coppola2022].
-A wide range of methods and measure can be taken to ensure that digital tools are accessible to their recipients.]{style="color: blue;"}
+A wide range of methods and measure can be taken to ensure that digital tools are accessible.
 
@@
 
-[The stakeholder engagement was carried out through workshops with stakeholders in case study area, an approach that is documented with reference to participatory development of digital tools for supporting energy transitions [@hewitt2020].
+The stakeholder engagement was carried out through workshops with stakeholders in case study area, an approach that is documented with reference to participatory development of digital tools for supporting energy transitions [@hewitt2020].
 We held a total of three workshops over the course of the project and one open-ended live demo during a 2-day event in Sligo in which we collected feedback from dozens of people.
-Other than the live demo, all of the workshops took place online.]{style="color: blue;"}
+Other than the live demo, all of the workshops took place online.
 
-[The first two workshops took place in May 2022, during which the tool was introduced to key staff involved in cycle network planning in Kildare and Limerick counties.
+The first two workshops took place in May 2022, during which the tool was introduced to key staff involved in cycle network planning in Kildare and Limerick counties.
 The county-specific approach allowed the workshop to focus on specific areas and key corridors.
 The local transport planners were asked to provide feedback in a semi-structured interview approach, with questions on the evidence they currently have access, how they may be able to use the tool, and requests for new features.
-The third and final workshop took place on the 7^th^ December 2023 and included descriptions of recent changes to the tool, with an aim being to get feedback on the new recreational and cycle touring evidence, with representatives from county-level transport planning bodies, Sport Ireland, and Waterways Ireland.]{style="color: blue;"}
+The third and final workshop took place on the 7^th^ December 2023 and included descriptions of recent changes to the tool, with an aim being to get feedback on the new recreational and cycle touring evidence, with representatives from county-level transport planning bodies, Sport Ireland, and Waterways Ireland.
 

### 2024-06-23 paper.tex
Minor tweaks

@@
 outline the underlying methods, including disaggregation of
-origin-destination data, incorporation of additional trip purposes,
-routing, scenario generation, and development of an intuitive user
-interface that has been tested and used by practitioners. Commissioned
-and used by Transport Infrastructure Ireland, CRUSE provides estimates
-of current and potential future cycling levels under `snapshot'
-scenarios to inform investment decisions. The publicly available results
-at https://cruse.bike/ enable planners, engineers, and other
-stakeholders to make more evidence-based decisions. CRUSE goes beyond
-previous work by: modeling networks at high spatial resolution;
-simulating multiple trip purposes (social, shopping, personal utility,
-recreational, and cycle touring) in addition to using official
-origin-destination data; and providing estimates of `quietness' (a proxy
-for cyclist comfort and route preference) at the route segment level.
-Three network types --- `Fastest', `Balanced', and `Quietest' --- help
-plan both arterial and residential cycle networks. Workshops and
-unstructured interviews with stakeholders show that the tool is already
-being used in practice to inform urban, inter-urban, and rural network
-planning. The approach is flexible and open-source, allowing the
-underlying ideas and code to be adapted, supporting effective cycling
-policies and interventions internationally.
+origin-destination data with the open source `odjitter' software,
+incorporation of additional trip purposes, routing, scenario generation,
+and development of an intuitive user interface that is tested and used
+by practitioners. Commissioned by the national infrastructure agency
+Transport Infrastructure Ireland, CRUSE provides estimates of current
+and potential future cycling levels under `snapshot' scenarios to inform
+investment decisions. The publicly available results at
+https://cruse.bike/ enable planners, engineers, and other stakeholders
+to make more evidence-based decisions. CRUSE goes beyond previous work
+by: modeling networks at high spatial resolution; simulating multiple
+trip purposes (social, shopping, personal utility, recreational, and
+cycle touring), supplementing official origin-destination datasets on
+travel for work and education; and providing estimates of `quietness' (a
+proxy for cyclist comfort and route preference) at the route segment
+level. Three network types --- `Fastest', `Balanced', and `Quietest' ---
+help plan both arterial and residential cycle networks. Workshops and
+unstructured interviews with stakeholders were used to inform the
+development of the tool. Feedback shows that the tool has a wide range
+of uses and is already being used in practice to inform urban,
+inter-urban, and rural cycle network designs. The approach is flexible
+and open-source, allowing the underlying ideas and code to be adapted,
+supporting more evidence-based and effective cycling policies and
+interventions internationally.
 \end{abstract}
@@
 Analysis of Cycling Potential (ACP), Transport for London (2010 and 2016) & London, London Travel Demand Survey & Static tables, graphs, and heatmaps, not publicly available & Proprietary\\
-\cellcolor{gray!10}{{}[Bikeability index/Bike Score]\{style="color: blue;"\}, [Winters, M.
+\cellcolor{gray!10}{Bikeability index/Bike Score, Winters, M.
 et al.
-(2013 and 2016)]\{style="color: blue;"\}} & \cellcolor{gray!10}{{}[Metro Vancouver region/160 US and Canadian cities]\{style="color: blue;"\}, [Cycling infrastructure, topography, destinations and road connectivity data]\{style="color: blue;"\}} & \cellcolor{gray!10}{{}[Map-based heatmap, not publicly available]\{style="color: blue;"\}} & \cellcolor{gray!10}{{}[Proprietary]\{style="color: blue;"\}}\\
+(2013 and 2016)} & \cellcolor{gray!10}{Metro Vancouver region/160 US and Canadian cities, Cycling infrastructure, topography, destinations and road connectivity data} & \cellcolor{gray!10}{Map-based heatmap, not publicly available} & \cellcolor{gray!10}{Proprietary}\\
 Prioritization index, Larsen, J. et al. (2013) & Montreal, Survey, Road safety, and OD data & Map-based heatmap, not publicly available & Proprietary\\
@@
 (2017)} & \cellcolor{gray!10}{England and Wales, Journeys to work and school OD census data, at the small area (LSOA) level} & \cellcolor{gray!10}{Online maps, graphs, tables, publicly available at www.pct.bike} & \cellcolor{gray!10}{Open source}\\
-{}[Cyclist Routing Algorithm for Network Connectivity (CRANC)]\{style="color: blue;"\}, [Gehrke, S.
+Cyclist Routing Algorithm for Network Connectivity (CRANC), Gehrke, S.
 R.
 et al.
-(2020)]\{style="color: blue;"\} & {}[Boston region]\{style="color: blue;"\}, [Road segment characteristics, elevation, and destination acessibility data, at the census blocks level]\{style="color: blue;"\} & {}[Graphs and heatmaps, not publicly available]\{style="color: blue;"\} & {}[NA]\{style="color: blue;"\}\\
+(2020) & Boston region, Road segment characteristics, elevation, and destination acessibility data, at the census blocks level & Graphs and heatmaps, not publicly available & NA\\
 \cellcolor{gray!10}{The Gross Potential for Cycling tool (CPC), Silva, C.
 et al.
-(2021 and 2022) [and Lopez et al.
-(2021)]\{style="color: blue;"\}} & \cellcolor{gray!10}{21 Portuguese cities, Land use and socio-demographic data, at the small area (census tract) level} & \cellcolor{gray!10}{Static maps showing cycling potential in different areas} & \cellcolor{gray!10}{NA}\\
-{}[The relative equity impact of bicycle planning (TIRE)]\{style="color: blue;"\}, [Cunha, I., \& Silva, C.
-(2023)]\{style="color: blue;"\} & {}[Lisbon]\{style="color: blue;"\}, [Cycling infrastructure, topog­raphy, and location of points of interest data, at the hexagon cells level]\{style="color: blue;"\} & {}[Graphs and heatmaps, not publicly available]\{style="color: blue;"\} & {}[NA]\{style="color: blue;"\}\\
+(2021 and 2022) and Lopez et al.
+(2021)} & \cellcolor{gray!10}{21 Portuguese cities, Land use and socio-demographic data, at the small area (census tract) level} & \cellcolor{gray!10}{Static maps showing cycling potential in different areas} & \cellcolor{gray!10}{NA}\\
+The relative equity impact of bicycle planning (TIRE), Cunha, I., \& Silva, C.
+(2023) & Lisbon, Cycling infrastructure, topog­raphy, and location of points of interest data, at the hexagon cells level & Graphs and heatmaps, not publicly available & NA\\
 \bottomrule
@@
 
-\clearpage
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0653}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2869}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.4460}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2017}}@{}}
+\begingroup\fontsize{8}{10}\selectfont
+
+\begin{longtable}[t]{>{\raggedright\arraybackslash}p{5em}>{\raggedright\arraybackslash}p{12em}>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{15em}}
 
@@
 
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Trip purpose
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Description
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Primary source(s)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Confidence
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-Work (commute) & Commuting to / from workplaces & Census 2016 POWSCAR
-Data & High confidence - data both origins and destinations for trip
-purposes \\
-Primary Education & Primary school trips & Census 2016 POWSCAR Data,
-Schools database & High confidence - data both origins and destinations
-for trip purposes \\
-Secondary Education & Secondary school trips & Census 2016 POWSCAR Data,
-Schools database & High confidence - data both origins and destinations
-for trip purposes \\
-Tertiary Education & Tertiary education trips & Census 2016 POWSCAR Data
-- Geodirectory categories & High confidence - data both origins and
-destinations for trip purposes \\
-Social & NHTS classification (e.g.~restaurants, cinemas, gyms etc.) &
-Trip rates from NHTS, Geodirectory businesses classified according to
-relevant NACE codes & Medium confidence \\
-Shopping & NHTS classificationfrom supermarkets and shops & Trip rates
-from NHTS, Geodirectory businesses classified according to relevant NACE
-codes & Medium confidence \\
-Personal / other & Combination of two NHTS classifications, and includes
-trips to medical, personal services and others & Trip rates from NHTS,
-Geodirectory businesses classified according to relevant NACE codes &
-Medium confidence \\
-Tourism / recreational & Trips by non-resident visitors to / from
-visitor attractions & Geodirectory listings for accommodation (origins)
-and destinations (destinations), Failte Ireland accommodation database,
-Failte Ireland attraction database & Low confidence - lack of data
-regarding tourist trip rates \\
+\toprule
+Trip purpose & Description & Primary source(s) & Confidence\\
+\midrule
+\cellcolor{gray!10}{Work (commute)} & \cellcolor{gray!10}{Commuting to / from workplaces} & \cellcolor{gray!10}{Census 2016 POWSCAR Data} & \cellcolor{gray!10}{High confidence - data both origins and destinations for trip purposes}\\
+Primary Education & Primary school trips & Census 2016 POWSCAR Data, Schools database & High confidence - data both origins and destinations for trip purposes\\
+\cellcolor{gray!10}{Secondary Education} & \cellcolor{gray!10}{Secondary school trips} & \cellcolor{gray!10}{Census 2016 POWSCAR Data, Schools database} & \cellcolor{gray!10}{High confidence - data both origins and destinations for trip purposes}\\
+Tertiary Education & Tertiary education trips & Census 2016 POWSCAR Data - Geodirectory categories & High confidence - data both origins and destinations for trip purposes\\
+\cellcolor{gray!10}{Social} & \cellcolor{gray!10}{NHTS classification (e.g.
+restaurants, cinemas, gyms etc.)} & \cellcolor{gray!10}{Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes} & \cellcolor{gray!10}{Medium confidence}\\
+\addlinespace
+Shopping & NHTS classificationfrom supermarkets and shops & Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes & Medium confidence\\
+\cellcolor{gray!10}{Personal / other} & \cellcolor{gray!10}{Combination of two NHTS classifications, and includes trips to medical, personal services and others} & \cellcolor{gray!10}{Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes} & \cellcolor{gray!10}{Medium confidence}\\
+Tourism / recreational & Trips by non-resident visitors to / from visitor attractions & Geodirectory listings for accommodation (origins) and destinations (destinations), Failte Ireland accommodation database, Failte Ireland attraction database & Low confidence - lack of data regarding tourist trip rates\\
+\bottomrule
 
 \end{longtable}
+
+\endgroup{}
 
@@
 
-The CRUSE web application is statically hosted, meaning it does not
-require a server running software such as the R package `shiny' or
-Python packages such as `streamlit' or `flask' in the background
-\citep{wickham2021}. This reduces hosting costs and the maintenance
-burden of the tool compared with tools such as the PCT, and makes the
-map-based interface more responsive, using the recently developed
-`MapLibre' JavaScript package and `PMTiles' ``cloud native'' vector tile
-format \citep{gonçalves2023}. These open source technologies were used
-to generate default view in the landing page: a map of Ireland with
-county level results for the \% trips that could be made by cycling
-under the `Go Dutch' scenario, highlighting the intended use case for
-strategic cycle network planning and level of ambition, as illustrated
-in Figure~\ref{fig-landing}.
+The CRUSE web application is `statically hosted' rather than
+`dynamically hosted' \citep{wickham2021}. This means the tool does not
+require a resource-intensive server, reducing hosting costs and the
+maintenance burden of the tool compared with tools such as the PCT. The
+use recently-developed web mapping technology makes the tool responsive
+\citep{gonçalves2023}. Open source technologies were used throughout the
+tool's user interface, increasing the ease with which people in other
+countries can build on the open source code. The `landing page' shows
+the proportion of trips that could be made by cycling under the `Go
+Dutch' scenario at the county level, highlighting the intended use case
+for strategic cycle network planning (Figure~\ref{fig-landing}).
 
@@
 
-\subsection{\texorpdfstring{{Stakeholder
-engagement}}{Stakeholder engagement}}\label{sec-stakeholder-engagement}
-
-{The CRUSE Tool was developed for, and in close collaboration with,
-practitioners in Ireland. Although the overall approach of the project
-was quantitative, stakeholder feedback and qualitative data was deemed
-essential to ensure the tool was fit for purpose, mitigating against the
-risk that the project would be `technocentric' and a `problem looking
-for a solution'. From the outset of the project, we aimed to follow
-\citet{coppola2022}`s advocacy for a ``human-centric approach in which
-personalization, purpose, ethics and inclusivity are key'', rather than
-an 'off-the-shelf' product that failed to understand, let alone meet,
-local needs. As outlined by \citet{coppola2022}, the `digital
-transformation' taking place in transport planning (and many other
-sectors) has the potential to increase existing inequalities
+\subsection{Stakeholder engagement}\label{sec-stakeholder-engagement}
+
+The CRUSE Tool was developed for, and in close collaboration with,
+practitioners in the Republic of Ireland. Although the overall approach
+of the project was quantitative, stakeholder feedback and qualitative
+data was deemed essential to ensure the tool was fit for purpose,
+mitigating against the risk that the project would be `technocentric'
+and a `problem looking for a solution'. From the outset of the project,
+we aimed to follow \citet{coppola2022}`s advocacy for a ``human-centric
+approach in which personalization, purpose, ethics and inclusivity are
+key'', rather than an 'off-the-shelf' product that failed to understand,
+let alone meet, local needs. As outlined by \citet{coppola2022}, the
+`digital transformation' taking place in transport planning (and many
+other sectors) has the potential to increase existing inequalities
 \citep{coppola2022}. A wide range of methods and measure can be taken to
-ensure that digital tools are accessible to their recipients.}
-
-{The stakeholder engagement was carried out through workshops with
+ensure that digital tools are accessible.
+
+The stakeholder engagement was carried out through workshops with
 stakeholders in case study area, an approach that is documented with
@@
 of people. Other than the live demo, all of the workshops took place
-online.}
-
-{The first two workshops took place in May 2022, during which the tool
+online.
+
+The first two workshops took place in May 2022, during which the tool
 was introduced to key staff involved in cycle network planning in
@@
 county-level transport planning bodies, Sport Ireland, and Waterways
-Ireland.}
+Ireland.
 

### 2024-06-23 paper.qmd
Make paper camera-ready

@@
 
-Note: preprint now available here: https://osf.io/preprints/osf/a72yh
+<!-- Note: preprint now available here: https://osf.io/preprints/osf/a72yh
 
@@
 
-</details>
+</details> -->
 
@@
 #| label: fig-user-stories
+#| out-width: 100%
 # schematic = DiagrammeR::grViz("
@@
 Although the overall approach of the project was quantitative, stakeholder feedback and qualitative data was deemed essential to ensure the tool was fit for purpose, mitigating against the risk that the project would be 'technocentric' and a 'problem looking for a solution'.
-From the outset of the project, we aimed to follow @coppola2022's advocacy for a "human-centric approach in which personalization, purpose, ethics and inclusivity are key", rather than an 'off-the-shelf' product that failed to understand, let alone meet, local needs.
-As outlined by @coppola2022, the 'digital transformation' taking place in transport planning (and many other sectors) has the potential to increase existing inequalities [@coppola2022].
+From the outset of the project, we aimed to follow a "human-centric approach in which personalization, purpose, ethics and inclusivity are key", rather than an 'off-the-shelf' product that failed to understand, let alone meet, local needs.
+The 'digital transformation' taking place in transport planning (and many other sectors) has the potential to increase existing inequalities [@coppola2022].
 A wide range of methods and measure can be taken to ensure that digital tools are accessible.

### 2024-06-23 paper.tex
Make paper camera-ready

@@
 \setstretch{2}
-Note: preprint now available here: https://osf.io/preprints/osf/a72yh
-
-Please cite as
-
-Lovelace, Robin, Joey Talbot, Eugeni Vidal-Tortosa, Hussein Mahfouz,
-Elaine Brick, Peter Wright, Gary O'Toole, et al.~2024. ``Cycle Route
-Uptake and Scenario Estimation (CRUSE): An Approach for Developing
-Strategic Cycle Network Planning Tools.'' OSF Preprints. March 29.
-doi:10.31219/osf.io/a72yh.
-
-This is an open and reproducible paper. See
-\href{https://github.com/cruse-bike}{github.com/cruse-bike} for the code
-and data used to generate the results presented here, and the code
-underlying the web app hosted at \href{https://cruse.bike/}{cruse.bike}.
-
 \newpage{}
@@
 
-\includegraphics[width=3.31in,height=\textheight]{images/fig5.png}
+\includegraphics[width=1\textwidth,height=\textheight]{images/fig5.png}
 
@@
 and a `problem looking for a solution'. From the outset of the project,
-we aimed to follow \citet{coppola2022}`s advocacy for a ``human-centric
-approach in which personalization, purpose, ethics and inclusivity are
-key'', rather than an 'off-the-shelf' product that failed to understand,
-let alone meet, local needs. As outlined by \citet{coppola2022}, the
-`digital transformation' taking place in transport planning (and many
-other sectors) has the potential to increase existing inequalities
-\citep{coppola2022}. A wide range of methods and measure can be taken to
-ensure that digital tools are accessible.
+we aimed to follow a ``human-centric approach in which personalization,
+purpose, ethics and inclusivity are key'', rather than an
+`off-the-shelf' product that failed to understand, let alone meet, local
+needs. The `digital transformation' taking place in transport planning
+(and many other sectors) has the potential to increase existing
+inequalities \citep{coppola2022}. A wide range of methods and measure
+can be taken to ensure that digital tools are accessible.
 

### 2024-06-23 paper.qmd
Update for submission requirements

@@
 pdf-engine: pdflatex
-knitr:
-  opts_chunk:
-    fig.path: images/
 format: 
@@
 ::: {#fig-dublin}
-<!-- ![](images/paste-1.png) -->
 
-![](images/paste-8.png) <!-- ![](images/paste-3.png) -->
-![](images/paste-9.png)
+![](paste-8.png)
+![](paste-9.png)
 
@@
 
-<!-- ![](images/paste-5.png) -->
+<!-- ![](paste-5.png) -->
 
-<!-- ![](images/paste-6.png) -->
+<!-- ![](paste-6.png) -->
 
@@
 ::: {#fig-landing}
-![](images/paste-10.png)
+![](paste-10.png)
 
-![](images/paste-11.png)
+![](paste-11.png)
 
@@
 # schematic
-# htmlwidgets::saveWidget(schematic, "images/fig5.html")
-# browseURL("images/fig5.html")
-# webshot2::webshot(url = "images/fig5.html", file = "images/fig5.png")
-knitr::include_graphics("images/fig5.png")
+# htmlwidgets::saveWidget(schematic, "fig5.html")
+# browseURL("fig5.html")
+# webshot2::webshot(url = "fig5.html", file = "fig5.png")
+knitr::include_graphics("fig5.png")
 ```
@@
 <!-- ::: {#fig-user-stories}
-![](images/fig5.png)
+![](fig5.png)
 
@@
   tm_lines(col = "Quietness", palette = "RdYlGn", lwd = 5)
-tmap_save(m, "./images/rnet_fastest_city_top.png", width = 10, height = 10)
+tmap_save(m, "./rnet_fastest_city_top.png", width = 10, height = 10)
 # Generalisation to balanced and quiestest route types:
@@
 m_types
-tmap_save(m_types, "./images/rnet_types.png", width = 12, height = 16)
-browseURL("./images/rnet_types.png")
+tmap_save(m_types, "./rnet_types.png", width = 12, height = 16)
+browseURL("./rnet_types.png")
 ```

### 2024-06-23 paper.tex
Update for submission requirements

@@
 
-\includegraphics{images/paste-8.png}
-\includegraphics{images/paste-9.png}
+\includegraphics{paste-8.png}
 
@@
 
-\includegraphics{images/paste-10.png}
-
-\includegraphics{images/paste-11.png}
+\includegraphics{paste-10.png}
+
+\includegraphics{paste-11.png}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{images/fig5.png}
+\includegraphics[width=1\textwidth,height=\textheight]{fig5.png}
 

### 2024-06-24 paper-r1.qmd
Anonymise r1 version

@@
 ---
-
-Note: preprint now available here: https://osf.io/preprints/osf/a72yh
-
-<details>
-
-Please cite as
-
-Lovelace, Robin, Joey Talbot, Eugeni Vidal-Tortosa, Hussein Mahfouz, Elaine Brick, Peter Wright, Gary O'Toole, et al.
-2024.
-“Cycle Route Uptake and Scenario Estimation (CRUSE): An Approach for Developing Strategic Cycle Network Planning Tools.” OSF Preprints.
-March 29.
-doi:10.31219/osf.io/a72yh.
-
-This is an open and reproducible paper.
-See [github.com/cruse-bike](https://github.com/cruse-bike) for the code and data used to generate the results presented here, and the code underlying the web app hosted at [cruse.bike](https://cruse.bike/).
-
-</details>
 

### 2024-06-24 paper.qmd
Add code to generate citations in tracked changes, close #46

@@
 library(tmap)
-# Save title page (1st page of paper.pdf):
-system("pdftk _site/paper.pdf cat 1 output _site/title_page.pdf")
+# # Save title page (1st page of paper.pdf):
+# system("pdftk _site/paper.pdf cat 1 output _site/title_page.pdf")
 # Combine cv.pdf with visions.pdf:
-system("pdftk _site/cv.pdf _site/visions.pdf cat output _site/cv_and_visions.pdf")
+# system("pdftk _site/cv.pdf _site/visions.pdf cat output _site/cv_and_visions.pdf")
 # latexdiffr::latexdiff("paper-v1.qmd", "paper.qmd")
@@
 system("pdflatex tracked-changes-r2.tex")
+# To debug this search for: 
+# https://www.google.com/search?client=ubuntu-sn&channel=fs&q=latexdiff+references+question+marks
+# From that: https://tex.stackexchange.com/questions/63852/question-mark-or-bold-citation-key-instead-of-citation-number
+# Bibliography:
+system("bibtex tracked-changes-r2.aux")
+system("pdflatex tracked-changes-r2.tex")
+system("pdflatex tracked-changes-r2.tex")
+
 browseURL("tracked-changes-r2.pdf")

### 2024-06-25 paper.qmd
proofread eugeni

@@
 Developed by researchers and developers in academic, public and private sectors, they have evolved from simple area-based static models tailored to specific regions to more complex and (in some cases) more generalizable tools.
-These tools can be considered as a specific application of the concept of planning support systems (PSS), which are designed to help planners and decision-makers to make more informed decisions [@geertman2009].
+These tools can be considered as a specific application of the concept of planning support systems (PSS), which are designed to help planners and decision-makers in making better-informed decisions [@geertman2009].
 Outputs include maps with results available at several levels of analysis with levels of availability ranging from being only available to researchers and in static maps to open access mapping systems available to the public.
@@
 They found that areas with a higher percentage of females, non-whites, car ownership, lower socioeconomic classes, income, distance to work, population density, poor highway conditions, hills, rainfall, and fewer off-road bicycle routes tended to have a lower proportion of cycling to work.
-The approach also generated "forecasts for potential levels of bicycle use" with with increases in off-road routes, improvements of highway and, conversely, increases in car ownership and distance to work leading to decreases in cycling to work.
+The approach also generated "forecasts for potential levels of bicycle use" with increases in off-road routes, improvements of highway and, conversely, increases in car ownership and distance to work leading to decreases in cycling to work.
 
 Three years later, Transport for London created the Analysis of Cycling Potential (ACP) tool based on a mode choice model that assesses the likelihood of trips being cycled, with data from the TfL London Travel Demand Survey as its primary input.
-The tool generates tables, graphs, and heatmaps showing the potential for cycling for different trips purposes in London to guide local cycling initiatives, such as where new hire 'docking stations' should go.
-The tool does not use origin-destination data and cannot estimate cycling potential on specific routes [@transport_for_london_analysis_2010].
+The ACP tool generates tables, graphs, and heatmaps showing the potential for cycling for different trips purposes in London to guide local cycling initiatives, such as where new hire 'docking stations' should go.
+However, it does not use origin-destination data and cannot estimate cycling potential on specific routes [@transport_for_london_analysis_2010].
 
-Other tools that generate areal data to prioritize investment include a study that generated a ['bikeability index' for the Metro Vancouver region [@winters_mapping_2013], later renamed Bike Score and available for over 160 US and Canadian cities [@winters_bike_2016] -- its methodology is [openly accessible](https://www.walkscore.com/bike-score-methodology.shtml); a]{style="color: blue;"} 'prioritization index' for Montreal, Canada [@larsen_build_2013]; a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil [@zhang_prioritizing_2014]; and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
+Other tools that generate areal data to prioritize investment include a study that generated a ['bikeability index' for the Metro Vancouver region [@winters_mapping_2013], later renamed Bike Score and available for over 160 US and Canadian cities [@winters_bike_2016]; a]{style="color: blue;"} 'prioritization index' for Montreal, Canada [@larsen_build_2013]; a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil [@zhang_prioritizing_2014]; and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
 The CPT is composed of the base environmental module — based on eight weighted factors (population density, hilliness, physical barriers, access to services, existing cycling mode share, distance to work and school, and road speed) — and the quality of service module, which evaluates each area of interest based on eight factors (surface condition, adjacent cyclists, comfort factor, conflict, distance between junctions, slope, access to services, and origin/destination).
@@
 A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace2016].
-The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 [@lovelace2017].
+The approach was scaled-up to estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 [@lovelace2017].
 Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modeling of impacts on health, environmental and distributional outcomes [@woodcock2021].
@@
 
-[Another route-based tool is The Cycling Routing Algorithm for Network Connectivity (CRANC) [@gehrke_cycling-focused_2020].
-The CRANC tool evaluates the benefits of new bicycle facilities for the accessibility of various populations and neighbourhoods.]{style="color: blue;"}
+[Another route-based tool is The Cycling Routing Algorithm for Network Connectivity (CRANC) [@gehrke_cycling-focused_2020], which evaluates the benefits of new bicycle facilities for the accessibility of various populations and neighbourhoods.]{style="color: blue;"}
 
@@
 ```
-Following feedback from stakeholders, we added another, 'non-everyday' trip purpose in an extension phase: cycle tourism.
+Following feedback from stakeholders, we added another 'non-everyday' trip purpose in an extension phase: cycle tourism.
 As outlined in the CRUSE Extension report \[reference and link to be added on publication\], this involved developing an inter-county spatial interaction model to estimate the number of trips between each county for cycle tourism, with trip attractors including campsites and major international transport hubs.
@@
 The datasets outlined above were used to generate estimates of cycling currently (the 'Baseline' scenario) and under four scenarios of cycling potential: Near Market, Climate Action Plan, Go Dutch, and Ebike.
-Each of these scenarios is described below, as illustrated in @fig-scenarios.
+Each of these scenarios is described below. @fig-scenarios displays a graph showing the proportion of bicycle trips by distance for Kildare in each scenario.
 <!-- TODO: add table. -->
@@
 
-Throughout the project, we undertook a series of workshops and interviews with stakeholders in Ireland to gather feedback on the approach and resulting tool.
-The feedback was generally positive, with stakeholders commenting on the need for the tool, lack of data on cycling potential proving a barrier to decision-making and investment, and the potential for the tool.
+The feedback from the workshops and interviews was generally positive, with stakeholders commenting on the need for the tool, lack of data on cycling potential proving a barrier to decision-making and investment, and the potential for the tool.
 At an end of project workshop in December 2023, stakeholders were asked to provide feedback on the tool and its potential use in practice.

### 2024-06-25 paper.qmd
minor edits

@@
 
-Illustration of the density of route networks generated by the CRUSE Tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland (botton).
+Illustration of the density of route networks generated by the CRUSE Tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland (bottom).
 Source: CRUSE Tool, publicly available at [cruse.bike](https://cruse.bike/).
@@
 The default 'zoomed out' view enables users to click on a county for more detailed results.
-The 'zoomed in' view shows the national network view for the 'Go Dutch' scenario for the Limerick area (bottom).
+The 'zoomed in' view (bottom) shows the national network for the 'Go Dutch' scenario in the Limerick area.
 :::
@@
 
-The stakeholder engagement was carried out through workshops with stakeholders in case study area, an approach that is documented with reference to participatory development of digital tools for supporting energy transitions [@hewitt2020].
+The stakeholder engagement was carried out through workshops with stakeholders in case study areas, an approach that is documented with reference to participatory development of digital tools for supporting energy transitions [@hewitt2020].
 We held a total of three workshops over the course of the project and one open-ended live demo during a 2-day event in Sligo in which we collected feedback from dozens of people.
@@
 The county-specific approach allowed the workshop to focus on specific areas and key corridors.
-The local transport planners were asked to provide feedback in a semi-structured interview approach, with questions on the evidence they currently have access, how they may be able to use the tool, and requests for new features.
+The local transport planners were asked to provide feedback in a semi-structured interview approach, with questions on the evidence they currently have access to, how they may be able to use the tool, and requests for new features.
 The third and final workshop took place on the 7^th^ December 2023 and included descriptions of recent changes to the tool, with an aim being to get feedback on the new recreational and cycle touring evidence, with representatives from county-level transport planning bodies, Sport Ireland, and Waterways Ireland.

### 2024-06-26 paper.qmd
Keep only page 1 on title page

@@
 # # Save title page (1st page of paper.pdf):
-# system("pdftk _site/paper.pdf cat 1 output _site/title_page.pdf")
+system("pdftk title_page.pdf cat 1 output title_page1.pdf")
+browseURL("title_page1.pdf")
 # Combine cv.pdf with visions.pdf:

### 2024-06-26 paper.tex
Fix fig 6 size

@@
 support systems (PSS), which are designed to help planners and
-decision-makers to make more informed decisions \citep{geertman2009}.
-Outputs include maps with results available at several levels of
-analysis with levels of availability ranging from being only available
-to researchers and in static maps to open access mapping systems
-available to the public.
+decision-makers in making better-informed decisions
+\citep{geertman2009}. Outputs include maps with results available at
+several levels of analysis with levels of availability ranging from
+being only available to researchers and in static maps to open access
+mapping systems available to the public.
 
@@
 have a lower proportion of cycling to work. The approach also generated
-``forecasts for potential levels of bicycle use'' with with increases in
+``forecasts for potential levels of bicycle use'' with increases in
 off-road routes, improvements of highway and, conversely, increases in
@@
 likelihood of trips being cycled, with data from the TfL London Travel
-Demand Survey as its primary input. The tool generates tables, graphs,
-and heatmaps showing the potential for cycling for different trips
-purposes in London to guide local cycling initiatives, such as where new
-hire `docking stations' should go. The tool does not use
+Demand Survey as its primary input. The ACP tool generates tables,
+graphs, and heatmaps showing the potential for cycling for different
+trips purposes in London to guide local cycling initiatives, such as
+where new hire `docking stations' should go. However, it does not use
 origin-destination data and cannot estimate cycling potential on
@@
 region \citep{winters_mapping_2013}, later renamed Bike Score and
-available for over 160 US and Canadian cities \citep{winters_bike_2016}
--- its methodology is
-\href{https://www.walkscore.com/bike-score-methodology.shtml}{openly
-accessible}; a} `prioritization index' for Montreal, Canada
+available for over 160 US and Canadian cities \citep{winters_bike_2016};
+a} `prioritization index' for Montreal, Canada
 \citep{larsen_build_2013}; a `usage intensity index' based on stated
@@
 case study cities \citep{lovelace2016}. The approach was scaled-up to
-estimate estimate the potential benefits of uptake at zone and desire
-line levels nationally, and launched as a publicly available web
-application in 2017 \citep{lovelace2017}. Extensions of the PCT approach
-have included estimation of benefits at the individual level
-\citep{woodcock2018}, addition of travel to school network
-\citep{goodman2019}, and improved modeling of impacts on health,
-environmental and distributional outcomes \citep{woodcock2021}.
-Initially developed just for England, the PCT was extended to cover all
-of Wales (for commuter data only) in 2018.
+estimate the potential benefits of uptake at zone and desire line levels
+nationally, and launched as a publicly available web application in 2017
+\citep{lovelace2017}. Extensions of the PCT approach have included
+estimation of benefits at the individual level \citep{woodcock2018},
+addition of travel to school network \citep{goodman2019}, and improved
+modeling of impacts on health, environmental and distributional outcomes
+\citep{woodcock2021}. Initially developed just for England, the PCT was
+extended to cover all of Wales (for commuter data only) in 2018.
 
@@
 {Another route-based tool is The Cycling Routing Algorithm for Network
-Connectivity (CRANC) \citep{gehrke_cycling-focused_2020}. The CRANC tool
+Connectivity (CRANC) \citep{gehrke_cycling-focused_2020}, which
 evaluates the benefits of new bicycle facilities for the accessibility
@@
 generated by the CRUSE Tool for Dublin City and surroundings (top) and
-rural County Mayo on the West coast of Ireland (botton). Source: CRUSE
+rural County Mayo on the West coast of Ireland (bottom). Source: CRUSE
 Tool, publicly available at \href{https://cruse.bike/}{cruse.bike}.}
@@
 
-Following feedback from stakeholders, we added another, `non-everyday'
+Following feedback from stakeholders, we added another `non-everyday'
 trip purpose in an extension phase: cycle tourism. As outlined in the
@@
 potential: Near Market, Climate Action Plan, Go Dutch, and Ebike. Each
-of these scenarios is described below, as illustrated in
-Figure~\ref{fig-scenarios}.
+of these scenarios is described below. Figure~\ref{fig-scenarios}
+displays a graph showing the proportion of bicycle trips by distance for
+Kildare in each scenario.
 
@@
 view enables users to click on a county for more detailed results. The
-`zoomed in' view shows the national network view for the `Go Dutch'
-scenario for the Limerick area (bottom).}
+`zoomed in' view (bottom) shows the national network for the `Go Dutch'
+scenario in the Limerick area.}
 
@@
 The stakeholder engagement was carried out through workshops with
-stakeholders in case study area, an approach that is documented with
+stakeholders in case study areas, an approach that is documented with
 reference to participatory development of digital tools for supporting
@@
 interview approach, with questions on the evidence they currently have
-access, how they may be able to use the tool, and requests for new
+access to, how they may be able to use the tool, and requests for new
 features. The third and final workshop took place on the
@@
 
-\includegraphics{rnet_types.png}
+\includegraphics[width=0.9\textwidth,height=\textheight]{rnet_types.png}
 
@@
 
-Throughout the project, we undertook a series of workshops and
-interviews with stakeholders in Ireland to gather feedback on the
-approach and resulting tool. The feedback was generally positive, with
-stakeholders commenting on the need for the tool, lack of data on
+The feedback from the workshops and interviews was generally positive,
+with stakeholders commenting on the need for the tool, lack of data on
 cycling potential proving a barrier to decision-making and investment,

### 2024-06-26 paper.qmd
Fix fig 6 size
Re-word description of Parkin work

@@
 They found that areas with a higher percentage of females, non-whites, car ownership, lower socioeconomic classes, income, distance to work, population density, poor highway conditions, hills, rainfall, and fewer off-road bicycle routes tended to have a lower proportion of cycling to work.
-The approach also generated "forecasts for potential levels of bicycle use" with increases in off-road routes, improvements of highway and, conversely, increases in car ownership and distance to work leading to decreases in cycling to work.
+The approach also generated "forecasts for potential levels of bicycle use" associated with increasing in off-road routes and, conversely, increasing car ownership levels and distances travelled to work leading to decreases in cycling.
 
@@
 ::: {#fig-city-results}
-![](rnet_types.png)
+
+![](rnet_types.png){width="90%"}
 

### 2024-06-26 paper.tex
Add County Mayo image

@@
 have a lower proportion of cycling to work. The approach also generated
-``forecasts for potential levels of bicycle use'' with increases in
-off-road routes, improvements of highway and, conversely, increases in
-car ownership and distance to work leading to decreases in cycling to
-work.
+``forecasts for potential levels of bicycle use'' associated with
+increasing in off-road routes and, conversely, increasing car ownership
+levels and distances travelled to work leading to decreases in cycling.
 
@@
 \includegraphics{paste-8.png}
+
+\includegraphics{paste-9.png}
 

### 2024-06-26 paper.qmd
Add County Mayo image
Final typo fixes

@@
 Three network types --- 'Fastest', 'Balanced', and 'Quietest' --- help plan both arterial and residential cycle networks.
-Workshops and unstructured interviews with stakeholders were used to inform the development of the tool.
+Workshops with stakeholders were used to inform the development of the tool.
 Feedback shows that the tool has a wide range of uses and is already being used in practice to inform urban, inter-urban, and rural cycle network designs.
@@
 ![](paste-8.png)
+
 ![](paste-9.png)
@@
 Source: CRUSE Tool, publicly available at [cruse.bike](https://cruse.bike/).
+
 :::
@@
 The county-specific approach allowed the workshop to focus on specific areas and key corridors.
-The local transport planners were asked to provide feedback in a semi-structured interview approach, with questions on the evidence they currently have access to, how they may be able to use the tool, and requests for new features.
+During the workshops, after the tool was introduced and presented, participants were asked questions in a semi-structured interview format, with questions on the evidence they currently have access to, how they may be able to use the tool, and requests for new features.
 The third and final workshop took place on the 7^th^ December 2023 and included descriptions of recent changes to the tool, with an aim being to get feedback on the new recreational and cycle touring evidence, with representatives from county-level transport planning bodies, Sport Ireland, and Waterways Ireland.
@@
 Instead of trying to show such use cases, of which there are many hundreds, we present a selection of results to illustrate the main features of the resulting evidence, in @sec-quantiative.
-We also present some qualitative results based on workshops and interviews with stakeholders in Ireland, in @sec-qualitative.
+We also present some qualitative results based on workshops with stakeholders in Ireland, in @sec-qualitative.
 
@@
 
-The feedback from the workshops and interviews was generally positive, with stakeholders commenting on the need for the tool, lack of data on cycling potential proving a barrier to decision-making and investment, and the potential for the tool.
+The feedback from the workshops was generally positive, with stakeholders commenting on the need for the tool, lack of data on cycling potential proving a barrier to decision-making and investment, and the potential for the tool.
 At an end of project workshop in December 2023, stakeholders were asked to provide feedback on the tool and its potential use in practice.
-Quotes and suggestions from these workshops, and a subsequent interview with a Senior Transport Planning using the evidence in their everyday work in Transport Infrastructure Ireland, and presented in this section (with quotes being attributed to named individuals where permission has been granted).
+Quotes and suggestions from these workshops, and a subsequent interview with a transport planner using the evidence in their everyday work in Transport Infrastructure Ireland, are presented in this section (with quotes being attributed to named individuals where permission has been granted).
 
@@
 
-[A final limitation of the approach relates to *how* it was developed.
+A final limitation of the approach relates to *how* it was developed.
 As described in @sec-stakeholder-engagement, the development of the tool was participatory, but only insofar as people who attended our workshops could provide feedback.
@@
 In future work we would like to collect feedback from a wider range of people, including campaigners, service providers (e.g.
-staff involved in travel planning for schools and new developments), and people who not involved in the planning process.]{style="color: blue;"}
+staff involved in travel planning for schools and new developments), and groups who are under-represented in transport planning processes.
 

### 2024-06-26 paper.tex
Remove corresponding author
Add hyperlink to abstract

@@
 investment decisions. The publicly available results at
-https://cruse.bike/ enable planners, engineers, and other stakeholders
-to make more evidence-based decisions. CRUSE goes beyond previous work
-by: modeling networks at high spatial resolution; simulating multiple
-trip purposes (social, shopping, personal utility, recreational, and
-cycle touring), supplementing official origin-destination datasets on
-travel for work and education; and providing estimates of `quietness' (a
-proxy for cyclist comfort and route preference) at the route segment
-level. Three network types --- `Fastest', `Balanced', and `Quietest' ---
-help plan both arterial and residential cycle networks. Workshops and
-unstructured interviews with stakeholders were used to inform the
-development of the tool. Feedback shows that the tool has a wide range
-of uses and is already being used in practice to inform urban,
-inter-urban, and rural cycle network designs. The approach is flexible
-and open-source, allowing the underlying ideas and code to be adapted,
-supporting more evidence-based and effective cycling policies and
-interventions internationally.
+\url{https://cruse.bike/} enable planners, engineers, and other
+stakeholders to make more evidence-based decisions. CRUSE goes beyond
+previous work by: modeling networks at high spatial resolution;
+simulating multiple trip purposes (social, shopping, personal utility,
+recreational, and cycle touring), supplementing official
+origin-destination datasets on travel for work and education; and
+providing estimates of `quietness' (a proxy for cyclist comfort and
+route preference) at the route segment level. Three network types ---
+`Fastest', `Balanced', and `Quietest' --- help plan both arterial and
+residential cycle networks. Workshops with stakeholders were used to
+inform the development of the tool. Feedback shows that the tool has a
+wide range of uses and is already being used in practice to inform
+urban, inter-urban, and rural cycle network designs. The approach is
+flexible and open-source, allowing the underlying ideas and code to be
+adapted, supporting more evidence-based and effective cycling policies
+and interventions internationally.
 \end{abstract}
@@
 Kildare and Limerick counties. The county-specific approach allowed the
-workshop to focus on specific areas and key corridors. The local
-transport planners were asked to provide feedback in a semi-structured
-interview approach, with questions on the evidence they currently have
-access to, how they may be able to use the tool, and requests for new
-features. The third and final workshop took place on the
-7\textsuperscript{th} December 2023 and included descriptions of recent
-changes to the tool, with an aim being to get feedback on the new
-recreational and cycle touring evidence, with representatives from
-county-level transport planning bodies, Sport Ireland, and Waterways
-Ireland.
+workshop to focus on specific areas and key corridors. During the
+workshops, after the tool was introduced and presented, participants
+were asked questions in a semi-structured interview format, with
+questions on the evidence they currently have access to, how they may be
+able to use the tool, and requests for new features. The third and final
+workshop took place on the 7\textsuperscript{th} December 2023 and
+included descriptions of recent changes to the tool, with an aim being
+to get feedback on the new recreational and cycle touring evidence, with
+representatives from county-level transport planning bodies, Sport
+Ireland, and Waterways Ireland.
 
@@
 features of the resulting evidence, in Section~\ref{sec-quantiative}. We
-also present some qualitative results based on workshops and interviews
-with stakeholders in Ireland, in Section~\ref{sec-qualitative}.
+also present some qualitative results based on workshops with
+stakeholders in Ireland, in Section~\ref{sec-qualitative}.
 
@@
 
-The feedback from the workshops and interviews was generally positive,
-with stakeholders commenting on the need for the tool, lack of data on
+The feedback from the workshops was generally positive, with
+stakeholders commenting on the need for the tool, lack of data on
 cycling potential proving a barrier to decision-making and investment,
@@
 and its potential use in practice. Quotes and suggestions from these
-workshops, and a subsequent interview with a Senior Transport Planning
-using the evidence in their everyday work in Transport Infrastructure
-Ireland, and presented in this section (with quotes being attributed to
-named individuals where permission has been granted).
+workshops, and a subsequent interview with a transport planner using the
+evidence in their everyday work in Transport Infrastructure Ireland, are
+presented in this section (with quotes being attributed to named
+individuals where permission has been granted).
 
@@
 
-{A final limitation of the approach relates to \emph{how} it was
+A final limitation of the approach relates to \emph{how} it was
 developed. As described in Section~\ref{sec-stakeholder-engagement}, the
@@
 people, including campaigners, service providers (e.g.~staff involved in
-travel planning for schools and new developments), and people who not
-involved in the planning process.}
+travel planning for schools and new developments), and groups who are
+under-represented in transport planning processes.
 

### 2024-08-06 paper.qmd
Add hyperlink to abstract
Update with smaller Figure 2
KableExtra

@@
 number-sections: true
-latex-clean: false
-pdf-engine: pdflatex
-format: 
-  # html: default
-  # Uncomment for formatted version
-  elsevier-pdf:
-    linestretch: 2
-    keep-tex: true
-    journal:
-      name: European Transport Research Review
-      formatting: preprint
-      model: 3p
-      cite-style: super
 # # # #   # pdf: default
 # # Uncomment the following line to produce the cover page
-# author:
-#   - name: Robin Lovelace
-#     email: r.lovelace@leeds.ac.uk
-#     corresponding: true
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Joey Talbot
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Eugeni Vidal-Tortosa
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Hussein Mahfouz
-#     affiliations:
-#       - name: University of Leeds
-#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
-#   - name: Elaine Brick
-#     affiliations:
-#       - name: AECOM
-#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-#   - name: Peter Wright
-#     affiliations:
-#       - name: AECOM
-#         address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
-#   - name: "Gary O'Toole"
-#     affiliations:
-#       - name: AECOM
-#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
-#   - name: Dan Brennan
-#     affiliations:
-#       - name: Transport Infrastructure Ireland
-#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
-#   - name: Suzanne Meade
-#     affiliations:
-#       - name: Transport Infrastructure Ireland
-#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+author:
+  - name: Robin Lovelace
+    email: r.lovelace@leeds.ac.uk
+    corresponding: true
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Joey Talbot
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Eugeni Vidal-Tortosa
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Hussein Mahfouz
+    affiliations:
+      - name: University of Leeds
+        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
+  - name: Elaine Brick
+    affiliations:
+      - name: AECOM
+        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+  - name: Peter Wright
+    affiliations:
+      - name: AECOM
+        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
+  - name: "Gary O'Toole"
+    affiliations:
+      - name: AECOM
+        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
+  - name: Dan Brennan
+    affiliations:
+      - name: Transport Infrastructure Ireland
+        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
+  - name: Suzanne Meade
+    affiliations:
+      - name: Transport Infrastructure Ireland
+        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
 abstract: |
@@
 Commissioned by the national infrastructure agency Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions.
-The publicly available results at https://cruse.bike/ enable planners, engineers, and other stakeholders to make more evidence-based decisions.
+The publicly available results at <https://cruse.bike/> enable planners, engineers, and other stakeholders to make more evidence-based decisions.
 CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring), supplementing official origin-destination datasets on travel for work and education; and providing estimates of 'quietness' (a proxy for cyclist comfort and route preference) at the route segment level.
@@
 ---
+
+```{r}
+#| include: false
+#| eval: false
+latex-clean: false
+pdf-engine: pdflatex
+format: 
+  # html: default
+  # Uncomment for formatted version
+  elsevier-pdf:
+    linestretch: 2
+    keep-tex: true
+    journal:
+      name: European Transport Research Review
+      formatting: preprint
+      model: 3p
+      cite-style: super
+```
 
@@
 
-![](scenarios_kildare.png)
+![](scenarios_kildare.png){width="60%"}
 

### 2024-08-06 paper.tex
Update with smaller Figure 2
KableExtra

@@
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
+\makeatother
 
@@
 \journal{European Transport Research Review}
+
 \ifLuaTeX
@@
 
-\includegraphics{scenarios_kildare.png}
+\includegraphics[width=0.6\textwidth,height=\textheight]{scenarios_kildare.png}
 

### 2024-08-06 paper.qmd
Update paper with typo fixes thanks to Eugeni

@@
 Growing evidence of the negative impacts of car-dependent transport systems has led governments in many countries to set targets and take actions.
-In the context of climate, road safety and physicial inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
+In the context of climate, road safety and physical inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
 The ASI framework highlights the importance of demand reduction (*avoid*ing unnecessary trips), in addition to mode *shift* to sustainable modes and *improvement* of existing energy converters, in that order.
@@
 Stakeholder consultation by TII emphasized the need for a strong, national, systematic but locally-specific evidence base on cycle networks in both urban and rural Ireland.
-The evidence base had to scale nationally to support strategic alignment with national, regional and local policies, but also be useful for local network planning.
+To support evidence-based planning, the tool had to scale nationally to support strategic alignment with national, regional and local policies, but also be useful for local network planning.
 
@@
 We share this conception of PSS, emphasizing the benefits of a focus on a particular mode (cycling) and a particular phase in the planning process (strategic network development).
-A selection of existing tools are summarised below and presented in @tbl-tools.
+A selection of existing tools is summarised below and presented in @tbl-tools.
 
@@
 They are then combined into an overall score, weighted by their impact on cycling according to the literature.
-The ranking of areas by the GPC are [publicly available](https://boost.up.pt/en/ferramentas/gpc) and its practical value has been assessed through workshops [@silva_revealing_2022].
+The ranking of areas by the GPC is [publicly available](https://boost.up.pt/en/ferramentas/gpc) and its practical value has been assessed through workshops [@silva_revealing_2022].
 [The GPC tool has also been used to investigate the combination of factors contributing to high levels cycling potential [@lopes_impact_2021].
@@
 Route-based tools emerged with increasing availability of origin-destination data, routing engines that can assign trips to networks, and improvements in computer hardware and software needed to generate route networks.
-A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace2016].
+A prominent example is the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace2016].
 The approach was scaled-up to estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 [@lovelace2017].
@@
 The PCT approach has been applied in other countries, including Ireland (the topic of this paper), Scotland, and Portugal.
-In Portugal the 'biclaR' project, based on methods underlying the PCT, has been developed and deployed for the Lisbon metro region.
+In Portugal, the 'biclaR' project, based on methods underlying the PCT, has been developed and deployed for the Lisbon metro region.
 The resulting evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
@@
 
-Routes in CRUSE are generated by CycleStreets, a not-for profit transport consultancy and web development company who provide application programming interfaces (APIs) supplying a range of datasets for cycle planning and advocacy internationally, including in Ireland.
+Routes in CRUSE are generated by CycleStreets, a not-for profit transport consultancy and web development company that provides application programming interfaces (APIs) supplying a range of datasets for cycle planning and advocacy internationally, including in Ireland.
 The CycleStreets routing engine is based on OpenStreetMap (OSM) data, which is continuously updated by a global community of volunteers.
@@
 Other improvements to the user interface in the national map on the landing page include a search bar in the top left and clearer legend for each network visualisation option (users can select any scenario or quietness or gradient depending on their needs).
-Following feedback from user testing, a 'geolocate' and 'full screen' button were added in the top right, enabling users to 'zoom' to their current location and to focus on the detailed geographic results, as shown in @fig-landing.
+Following feedback from user testing, a 'geolocate' and 'full screen' buttons were added in the top right, enabling users to 'zoom' to their current location and to focus on the detailed geographic results, as shown in @fig-landing.
 
@@
 The CRUSE Tool was developed for, and in close collaboration with, practitioners in the Republic of Ireland.
-Although the overall approach of the project was quantitative, stakeholder feedback and qualitative data was deemed essential to ensure the tool was fit for purpose, mitigating against the risk that the project would be 'technocentric' and a 'problem looking for a solution'.
+Although the overall approach of the project was quantitative, stakeholder feedback and qualitative data were deemed essential to ensure the tool was fit for purpose, mitigating against the risk that the project would be 'technocentric' and a 'problem looking for a solution'.
 From the outset of the project, we aimed to follow a "human-centric approach in which personalization, purpose, ethics and inclusivity are key", rather than an 'off-the-shelf' product that failed to understand, let alone meet, local needs.
 The 'digital transformation' taking place in transport planning (and many other sectors) has the potential to increase existing inequalities [@coppola2022].
-A wide range of methods and measure can be taken to ensure that digital tools are accessible.
+A wide range of methods and measures can be taken to ensure that digital tools are accessible.
 
@@
 
-The approach generates estimates of cycling potential on every road in the country and are therefore too extensive to present in their entirety in this paper.
-Through the interactive web application, users can explore the data to generate the results that are most relevant to their needs, whether that is finding the cycling potential on a particular road or finding 'weak links' or barriers in the cycle network associated with a particular school, work place or other destination.
+The approach generates estimates of cycling potential on every road in the country.
+The full results are therefore too extensive to present in their entirety in this paper.
+Through the interactive web application, hosted at [cruse.bike](https://cruse.bike/), users can explore the data to generate the results that are most relevant to their needs.
+That could be finding the cycling potential on a particular road or finding 'weak links', barriers in the cycle network associated with a particular school, work place or other destination, or other features of the network that could be improved to increase cycling uptake.
 
@@
 An unexpected piece of feedback from the workshop was the preference for visualizations that directly compare the fast and quiet routes, as they recognize the importance of both types of routes.
-In response to this we provided the 'Route types' page for each county, although this functionality is not available in the landing page map due to other feedback requesting a simpler interface.
+In response to this, we provided the 'Route types' page for each county, although this functionality is not available in the landing page map due to other feedback requesting a simpler interface.
 In comparison with another tool for transport planning, CRUSE was seen by practitioners in Kildare as more user-friendly.
 They found the evidence at the route level (unavailable in other tools) crucial for planning new cycling infrastructure.
-Overall, the workshop participants concluded that the approach filled gap in evidence and will greatly assist in the development of new cycling projects.
+Overall, the workshop participants concluded that the approach filled a gap in the evidence and will greatly assist in the development of new cycling projects.
 
@@
 However, more data is needed to better understand the accuracy of the baseline results, especially for the 'non-POWSCAR' purposes.
-We hope this becomes possible as Ireland's cycle counter network expands, allowing refinements to the approach (espcially the spatial interaction model and parameters) to be made.
+We hope this becomes possible as Ireland's cycle counter network expands, allowing refinements to the approach (especially the spatial interaction model and parameters) to be made.
 Changes that could help mitigate this limitation on the user interface side include providing results with confidence intervals or as categories of cycling potential, rather than as central estimates, and the visualistion of the 'POWSCAR' network (which is based on Census data) in the web application.
@@
 The 'cycle friendliness' of a route is based on the OSM tags and the routes themselves are generated by the CycleStreets routing engine, which is based on OSM data.
-During a conference in Sligo where we presented prelimary results, a stakeholder with local knowledge pointed out that a quiet route along the canal was not being followed.
+During a conference in Sligo where we presented preliminary results, a stakeholder with local knowledge pointed out that a quiet route along the canal was not being followed.
 In response, manual edits were made to the OSM network, and the results currently on the website reflect these changes.
@@
 Future work could mitigate this by integrating the approach more closely with OSM (for example by providing links to enable one-click edits of OSM when users click on a route segment) or by setting up a workshop to 'crowdsource' corrections to the OSM data.
-Furthermore, because there are multiple possible users of the tool with different starging points, workshops and training events could be valuable if not essential to ensure that the tool is used effectively.
+Furthermore, because there are multiple possible users of the tool with different starting points, workshops and training events could be valuable if not essential to ensure that the tool is used effectively.
 Such events could bring together stakeholders from advocacy groups, local authorities, and national governments.
-The siloed nature of transport planning means that such diverse stakeholders are rarely the same room, meaning that a 'wide boundary' outcome of the tool could be to bring these groups together (if workshops based on the tool are successful).
+The siloed nature of transport planning means that such diverse stakeholders are rarely in the same room, meaning that a 'wide boundary' outcome of the tool could be to bring these groups together (if workshops based on the tool are successful).
 
 Another limitation of the approach is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
-Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
+Exposure information is a key part of the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
 Regular updates to the model to generate yearly baseline cycle flow estimates could tackle this limitation, although would take substantial resources.
@@
 In future work we would like to collect feedback from a wider range of people, including campaigners, service providers (e.g.
-staff involved in travel planning for schools and new developments), and groups who are under-represented in transport planning processes.
+staff involved in travel planning for schools and new developments), and groups that are under-represented in transport planning processes.
 

### 2024-08-06 paper.qmd
Add acknowledgements

--- 2024-08-06 paper.qmd
@@
 
-\[To be added post review.\]
+Thanks to Catherine Swift and Ciaran Maguire for their support in the development of the CRUSE Tool during the early stages.
+Thanks to everyone who attended the workshops and provided feedback on the tool, especially Dónal Hodgins, Kieran Cullinane, and Declan Keenan.
+Thanks to Yael Arbell for comments on an earlier version of the manuscript that informed discussion of stakeholder engagement.
+Finally, thanks and to the European Transport Research Review staff and to the anonymous reviewers for their constructive feedback.

### 2024-08-06 paper.qmd
Add link to source code
Update with note

@@
 
-<!-- Note: preprint now available here: https://osf.io/preprints/osf/a72yh
+Note: this paper has been accepted for publication in the European Transport Research Review.
+Preprint now available here: https://osf.io/preprints/osf/a72yh
 
@@
 “Cycle Route Uptake and Scenario Estimation (CRUSE): An Approach for Developing Strategic Cycle Network Planning Tools.” OSF Preprints.
-March 29.
-doi:10.31219/osf.io/a72yh.
 
@@
 
-</details> -->
+</details>
 
@@
 Data was obtained from the Central Statistics Office (CSO) and Transport Infrastructure Ireland (TII) under license and cannot be shared publicly.
-The code used to generate the results presented in this paper is fully reproducible and available at \[to be added post review.\]
+The code used to generate the results presented in this paper is reproducible and available at https://github.com/cruse-bike/
 

### 2024-08-06 paper.qmd
Final PDF?

@@
 number-sections: true
+latex-clean: false
+pdf-engine: pdflatex
+format: 
+  # html: default
+  # Uncomment for formatted version
+  elsevier-pdf:
+    linestretch: 2
+    keep-tex: true
+    journal:
+      name: European Transport Research Review
+      formatting: preprint
+      model: 3p
+      cite-style: super
 # # # #   # pdf: default
@@
 #| eval: false
-latex-clean: false
-pdf-engine: pdflatex
-format: 
-  # html: default
-  # Uncomment for formatted version
-  elsevier-pdf:
-    linestretch: 2
-    keep-tex: true
-    journal:
-      name: European Transport Research Review
-      formatting: preprint
-      model: 3p
-      cite-style: super
 ```

### 2024-08-06 paper.qmd
HTML again

--- 2024-08-06 paper.qmd
@@
 number-sections: true
-# # # #   # pdf: default
-# # Uncomment the following line to produce the cover page
 author:

### 2024-09-14 paper.qmd
Add contributions

@@
 Finally, thanks and to the European Transport Research Review staff and to the anonymous reviewers for their constructive feedback.
+
+## Author contributions {.unnumbered}
+
+<!-- Conceptualization
+Data curation
+Formal analysis
+Funding acquisition
+Investigation
+Methodology
+Project administration
+Software
+Resources
+Supervision
+Validation
+Visualization
+Writing – original draft
+Writing – review & editing -->
+
+RL, JT, EVT, HM: Conceptualization, Methodology, Formal analysis, Validation, Writing – original draft, Writing – review & editing
+
+EB, PW, GOT: Project administration, Validation
+
+DB, SM: Resources, Validation, Project administration, Funding acquisition
